---
title: "Transcript filters, GTFs & isoform quantification (stage 02-1)"
output:
  html_document:
    df_print: paged
---

Stage 02-1 (was `02-SUPPA2-analysis.Rmd`). Cohort-wide (not per-transcript_filter):
runs once and produces the shared inputs the per-filter stages consume —
`filtered_transcript_ids.rds` (the three transcript filters: `transcripts`,
`tsl_filtered`, `biotype_filtered`), the per-filter `gencode.v29.{tf}.gtf`, and
`isoform_quantifications_subset.tsv.gz` (RSEM isoform quants restricted to the
`rna_to_use` cohort = samples with both RNA-Seq and epigenomic data). These feed
`02-2-rnaseq-normalisation` (isoform quants + filter IDs), the external rMATS run
(GTFs), and `03-prepare-aggregation`.

**SUPPA2 is reference-only:** the `generateEvents`/`psiPerEvent` chunks below are
`eval=FALSE` and NOT part of the active pipeline — rMATS (via
`02-3-rmats-event-filtering`) is the PSI-quantification tool. The SUPPA2 output is
kept for cross-checking only.

Make Splicing Analysis folder:

```{bash, make-splicing-folder}
set -euo pipefail
mkdir -p splicing_analysis/suppa
cd splicing_analysis
# Idempotent so force-reruns don't crash: plain `gunzip -k` exits status 2 when
# the uncompressed .gtf already exists (refuses to overwrite), which halted the
# render on reruns. Only fetch+decompress when the .gtf is missing; https (not
# ftp) for reachability from SLURM compute nodes.
if [ ! -f gencode.v29.annotation.gtf ]; then
  wget -q -O gencode.v29.annotation.gtf.gz \
    https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.annotation.gtf.gz
  gunzip -kf gencode.v29.annotation.gtf.gz
fi
```

Load Isoform Quantifications:

```{r, load-isoform-quants}
# read the file table
download_metadata <- fread("processed_data/file_table.csv.gz")

# Restrict to the analysis cohort (samples with BOTH RNA-Seq and epigenomic
# data) BEFORE reading isoform files. The full ~1555-sample RNA-Seq pool was
# only used for the (unused) full isoform_quantifications.tsv.gz, an
# exploratory all-sample TPM histogram, and the SUPPA reference expressions
# (eval=FALSE) — none feed the active per-filter pipeline (02-2 reads the
# subset). Reading only rna_to_use (~441) avoids the OOM (job 6280493) and is
# ~3.5x faster. `.N >= 8` = epirrs with the full RNA+WGBS+ChIP panel.
rna_to_use <- download_metadata[
  assay_type == "RNA-Seq" &
    epirr_id_without_version %in%
      download_metadata[,
        if (.N >= 8) .SD,
        by = epirr_id_without_version
      ][, epirr_id_without_version],
  uuid
]
rna_meta <- download_metadata[assay_type == "RNA-Seq" & uuid %in% rna_to_use]
# get the isoform quantification files (cohort only)
isoform_files <- rna_meta[, file.path(rna_data_dir, basename(file_path))]
cols_to_keep <- c(
  "transcript_id",
  "gene_id",
  "effective_length",
  "expected_count",
  "TPM"
)
# load the isoform quantification files
iso_quants_list <- pbmcapply::pbmclapply(
  isoform_files,
  data.table::fread,
  stringsAsFactor = TRUE,
  fill = TRUE,
  select = cols_to_keep
)

names(iso_quants_list) <- rna_meta[, uuid]

# make one isoform quantification table
isoform_quants <- data.table::rbindlist(iso_quants_list, idcol = "uuid")
isoform_quants[, uuid := as.factor(uuid)]

# Cohort-read sanity (new): non-empty guards a mis-scoped read; the warning
# flags any rna_to_use RSEM files that failed to load (they'd silently drop
# from gene_expression_normalised downstream).
stopifnot(nrow(isoform_quants) > 0L)
n_read <- uniqueN(isoform_quants$uuid)
if (n_read < nrow(rna_meta)) {
  warning(sprintf(
    "02-1: %d/%d rna_to_use isoform files loaded (%d missing/failed to read)",
    n_read, nrow(rna_meta), nrow(rna_meta) - n_read
  ))
}
tpms <- isoform_quants[
  download_metadata[
    assay_type == "RNA-Seq",
    .(uuid, epirr_id_without_version, experiment_type)
  ],
  on = .(uuid),
  c("experiment_type", "epirr_id_without_version") := .(
    experiment_type,
    epirr_id_without_version
  )
][, .(tpm = sum(TPM)), by = .(uuid, epirr_id_without_version, experiment_type)]
ggplot(tpms, aes(x = tpm, fill = experiment_type)) +
  geom_histogram()

# isoform_quants already holds ONLY the rna_to_use cohort (filtered before the
# read above), so this IS the subset that 02-2 consumes. The old full-pool
# isoform_quantifications.tsv.gz (all ~1555 RNA-Seq samples) was never read
# downstream and is no longer written.
fwrite(
  isoform_quants,
  "splicing_analysis/isoform_quantifications_subset.tsv.gz",
  sep = "\t"
)

# bring tpms to wide format
tpm_dt <- data.table::dcast(
  isoform_quants,
  transcript_id ~ uuid,
  value.var = "TPM",
  fill = 0
)

# export expression for analysis with suppa
suppa_expr_filename <- "splicing_analysis/suppa/tpm_expressions.tsv.gz"
fileConn <- gzfile(suppa_expr_filename)
writeLines(paste(names(tpm_dt)[-1], collapse = "\t"), fileConn)
close(fileConn)
fwrite(tpm_dt, suppa_expr_filename, sep = "\t", append = TRUE)
```

```{r, filter-annotation}
# load annotation file
unfiltered_annotation <- rtracklayer::import(
  "splicing_analysis/gencode.v29.annotation.gtf"
)

confidence_tsls <- c("1", "2")
polyA_biotypes <- c(
  "protein_coding",
  "retained_intron"
  # "lincRNA", "processed_transcript",
  # "antisense", "sense_intronic", "sense_overlapping", "transcribed_unprocessed_pseudogene"
)

tx_filter <- unfiltered_annotation$type == "transcript"
tsl_tx_filter <- tx_filter &
  unfiltered_annotation$transcript_support_level %in% confidence_tsls
biotype_tx_filter <- tsl_tx_filter &
  unfiltered_annotation$transcript_type %in% polyA_biotypes

filter_list <- list(
  transcripts = tx_filter,
  tsl_filtered = tsl_tx_filter,
  biotype_filtered = biotype_tx_filter
)

saveRDS(
  sapply(
    names(filter_list),
    function(tx_filter_name) {
      unfiltered_annotation[filter_list[[tx_filter_name]]]$transcript_id
    },
    simplify = FALSE
  ),
  "splicing_analysis/filtered_transcript_ids.rds"
)

for (tx_filter_name in names(filter_list)) {
  this_filter <- filter_list[[tx_filter_name]]
  # filter the annotation
  annotation <- unfiltered_annotation[
    unfiltered_annotation$transcript_id %in%
      unfiltered_annotation[this_filter]$transcript_id |
      (unfiltered_annotation$type == "gene" &
        unfiltered_annotation$gene_id %in%
          unfiltered_annotation[this_filter]$gene_id)
  ]
  # filter autosomes
  # standard_chroms <- paste0("chr", c(1:22))
  # annotation <- annotation[seqnames(annotation) %in% standard_chroms]
  # export filtered annotation
  gtf_file <- paste0("gencode.v29.", tx_filter_name, ".gtf")
  rtracklayer::export(annotation, file.path("splicing_analysis", gtf_file))

  # # filter for transcripts
  # isoform_quants_filtered <- isoform_quants[transcript_id %in% annotation$transcript_id]
  # print(ggplot(isoform_quants_filtered[, .(tpm = sum(TPM)), by = .(uuid, epirr_id_without_version, experiment_type)], aes(x = tpm, fill = experiment_type)) +
  #   geom_histogram() +
  #   labs(title = tx_filter_name))
  #
  # # recompute TPM values on the transcripts
  # isoform_quants_filtered[effective_length > 0, RPK := 1000 * expected_count / effective_length, by = .(uuid, epirr_id_without_version)]
  # isoform_quants_filtered[, TPM_rec := 1000000 * RPK / sum(RPK, na.rm = TRUE), by = .(uuid, epirr_id_without_version)]
  # isoform_quants_filtered[is.na(TPM_rec), TPM_rec := 0]
  # isoform_quants_filtered[, TPM_scale := 1000000 * TPM/sum(TPM, na.rm = TRUE), .(uuid, epirr_id_without_version)]
  #
  # print(ggplot(isoform_quants_filtered[, .(tpm = sum(TPM_rec)), by = .(uuid, epirr_id_without_version, experiment_type)], aes(x = tpm, fill = experiment_type)) +
  #   geom_histogram() +
  #   labs(title = tx_filter_name))

  # compute and export the gene expression in TPM by summing up transcript TPMS
  # gene_quants <- isoform_quants_filtered[, .(gene_tpm = sum(TPM, na.rm = TRUE)), by = .(uuid, epirr_id_without_version, gene_id)]
  # gene_expr_file <- paste0("suppa_analysis/gene_expressions_", tx_filter_name, ".rds")
  # saveRDS(gene_quants, gene_expr_file)
  #
  # # bring tpms to wide format
  # tpm_dt <- data.table::dcast(isoform_quants_filtered, transcript_id ~ uuid, value.var = "TPM", fill = 0)
  #
  # # export expression for analysis with suppa
  # suppa_expr_filename <- paste0("suppa_analysis/tpm_expressions_", tx_filter_name, ".tsv")
  # fileConn <- file(suppa_expr_filename)
  # writeLines(paste(names(tpm_dt)[-1], collapse = "\t"), fileConn)
  # close(fileConn)
  # fwrite(tpm_dt, suppa_expr_filename, sep = "\t", append = TRUE)
}
```


SUPPA2 analysis:

```{bash, engine.opts="-l", eval=FALSE, conda-env}
conda env create -f env.yml
```

```{bash, engine.opts="-l", eval=FALSE, runSUPPA}
cd splicing_analysis
conda activate ihec-as
for suffix in transcripts tsl_filtered biotype_filtered
do
    dir_path=suppa/events_${suffix}
    mkdir -p ${dir_path}
    suppa.py generateEvents -i gencode.v29.${suffix}.gtf -o ${dir_path}/gencode.v29.${suffix}.events -f ioe -e SE SS MX RI FL #--pool-genes
    expr_file=suppa/tpm_expressions.tsv.gz #_${suffix}.tsv
    for event in SE A5 A3 MX RI AF AL
    do 
        suppa.py psiPerEvent -i ${dir_path}/gencode.v29.${suffix}.events_${event}_strict.ioe -e <(gunzip -c ${expr_file}) -o ${dir_path}/event_${event} -m INFO --save_tpm_events #-f 1
    done
done
```
