---
title: "RNA-Seq gene expression normalisation (§4.12b)"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5)
```

```{r, load-packages}
# §4.12b — properly normalised gene-level expression, replacing the raw
# per-sample-rescaled TPM (`gene_expression = sum(TPM_scale)`) used in 05.
# Raw TPM has closed-sum composition bias: a handful of highly expressed
# transcripts/genes suppress the TPM of everything else in the same sample,
# which distorts cross-sample x cross-gene comparisons (exactly the
# comparison `gene_expression` is used for downstream, e.g. §4.13d, and the
# §4.12e VST low-expression gate now applied in 02-3-rmats-event-filtering).
# Both methods are written side by side; the column-routing is DECIDED (PLAN
# §4.12b): getmm → the pooled genome-wide splicing_ml table (cross-gene
# comparable without per-gene z-scoring), vst → event-specific 07/09 + RBP
# columns AND the 02-3 event gate (single-gene, z-scored per column).
# Consumers read this file per-transcript_filter directly
# (gene_expression_normalised_{tf}.csv.gz) — 02-3 for the gate, 05 for the
# ML feature columns; there is no on/off flag.
#
# §4.12c — ComBat-seq protocol batch correction (Zhang et al. 2020, NAR
# Genom Bioinform) was attempted here 2026-07-02 and ABANDONED after running
# on real data: `qc/sanity_summary.csv`'s `combat_ri_protocol_beta_reduction`
# entry shows ComBat_seq() itself failed (NA result, caught by a tryCatch
# that has since been removed along with the rest of this branch). Root
# cause confirmed empirically, not just suspected: most `ontology` (group)
# levels are single-protocol (batch)-dominated, so ComBat-seq's
# `~ batch + group` design is degenerate for them — batch and biology aren't
# separable. See revision/file-changes/02-2-rnaseq-normalisation.Rmd.md for
# the full writeup. Decision: use raw (uncorrected) getmm/vst; rely on the
# existing protocol + protocol×gene_expression ML interaction term instead,
# which works per-sample and doesn't need balanced protocol/ontology groups.
set.seed(getOption("EpiATLAS_AS_SEED", 42L))
library(edgeR)
library(DESeq2)
library(parallel)
library(BiocParallel)

# Same standalone append_sanity() pattern as `04-2-chip-signal-sanity.Rmd` —
# each sanity-checking notebook defines its own local copy rather than
# depending on `03`'s session state (this file already avoids that
# dependency for sample_design/isoform_quants too, see `load-data` below).
sanity_csv <- "qc/sanity_summary.csv"
append_sanity <- function(check_name, status, value, threshold, notes = "") {
  fwrite(
    data.table(
      check_name = check_name,
      status = status,
      value = value,
      threshold = threshold,
      notes = notes
    ),
    sanity_csv,
    append = TRUE
  )
}
```

```{r, load-data}
# 1. Load raw transcript-level quantifications (same source 05 uses for
#    gene_expression) and the per-transcript_filter transcript ID sets.
isoform_quants <- fread(
  "splicing_analysis/isoform_quantifications_subset.tsv.gz"
)
filtered_transcript_ids <- readRDS(
  "splicing_analysis/filtered_transcript_ids.rds"
)
# Per-filter invocation (PLAN §4.12b per-filter incremental outputs): process
# only the requested transcript_filter. Snakemake passes TRANSCRIPT_FILTER as
# an env var (one rule instance per {tf}); interactive runs fall back to the
# primary filter. Output is written to a per-filter file so building one
# filter never recomputes or clobbers another.
this_transcript_filter <- Sys.getenv(
  "TRANSCRIPT_FILTER",
  unset = getOption("EpiATLAS_AS_PRIMARY_FILTER", "biotype_filtered")
)
stopifnot(this_transcript_filter %in% names(filtered_transcript_ids))

# NOTE: the "gene expression by hosted event type" diagnostic that used to
# live at the end of this file has moved to 05-create-aggregated-dt.Rmd, which
# already reads both this file's output and the event annotations — keeping
# 02-2 free of any dependency on 03 (it now runs before 03).

# sample metadata for the vst design (~ ontology + protocol): protocol comes
# from file_table (per-uuid), ontology from the sample metadata sheet
# (per-epirr). Loaded fresh here rather than depending on another script's
# session state, matching the pattern used elsewhere in the pipeline (01/03).
file_metadata <- fread("processed_data/file_table.csv.gz")
sample_metadata <- fread(sample_metadata_file)

sample_design <- unique(file_metadata[
  assay_type == "RNA-Seq",
  .(uuid, epirr_id_without_version, protocol = factor(experiment_type))
])
sample_design[
  sample_metadata,
  on = c(epirr_id_without_version = "epirr_id_without_version"),
  ontology := factor(get(..ontology_column))
]
sample_design <- sample_design[!is.na(ontology) & !is.na(protocol)]

# file_table's RNA-Seq rows are the full EpiATLAS RNA-Seq pool (1555 uuid /
# 1467 epirr) — most of which have no matching epigenomic data and are
# irrelevant to this analysis. `isoform_quants` is already restricted to the
# "both RNA-Seq and epigenomic data" cohort (441 uuid / 405 epirr, see
# `02-1-transcript-filters.Rmd`'s `rna_to_use` filter), so intersecting here keeps
# `sample_design` — and everything downstream that reads it, e.g. the
# per-ontology protocol balance a reader would compute from it — scoped to
# the samples actually used, instead of a much larger, mostly-irrelevant
# superset.
sample_design <- sample_design[uuid %in% unique(isoform_quants$uuid)]
setkey(sample_design, uuid)
```

```{r, normalise-function}
# 2. Per transcript_filter: aggregate transcript-level counts to gene level.
#    - gene count = sum of transcript expected_count (RSEM), the standard
#      tximport gene-summarisation rule
#    - gene effective length = expected_count-weighted mean of transcript
#      effective_length, so genes dominated by one long/short isoform are
#      not diluted by rarely-used isoforms of very different length

compute_getmm_and_vst <- function(count_mat, efflen_mat, col_data) {
  # GeTMM-CPM (Smid et al. 2018, BMC Bioinformatics)
  #     RPK -> TMM normalisation factors -> CPM. TMM on RPK (rather than raw
  #     counts) folds the gene-length adjustment in before the library-size
  #     adjustment, so the result is comparable both across samples (like
  #     TMM) and across genes of different length (like RPK/TPM).
  rpk_mat <- count_mat / (efflen_mat / 1000)
  dge <- DGEList(counts = rpk_mat)
  dge <- calcNormFactors(dge, method = "TMM")
  getmm_mat <- cpm(dge, normalized.lib.sizes = TRUE, log = FALSE)

  # lengthScaledTPM + DESeq2 vst (canonical Bioconductor approach)
  #     lengthScaledTPM reproduces tximport's countsFromAbundance =
  #     "lengthScaledTPM": TPM is scaled by gene length, then rescaled per
  #     sample so the column sums match the original library size (total
  #     mapped reads) — this keeps a count-like scale that DESeq2 expects,
  #     while retaining TPM's cross-sample normalisation. vst is NOT a
  #     length correction by itself (per spec note), which is why the counts
  #     fed into it must already be length-scaled.
  # `rpk_mat` (count/effLen*1000) computed above for GeTMM is byte-identical
  # to the RPK-scale numerator TPM needs — reused here instead of
  # recomputing count_mat / (efflen_mat / 1000) a second time.
  raw_tpm <- t(t(rpk_mat) / (colSums(rpk_mat) / 1e6))
  length_scaled <- raw_tpm * efflen_mat
  lib_sizes <- colSums(count_mat)
  rescale_factor <- lib_sizes / colSums(length_scaled)
  length_scaled_tpm <- t(t(length_scaled) * rescale_factor)
  length_scaled_tpm_int <- round(length_scaled_tpm)

  dds <- DESeqDataSetFromMatrix(
    countData = length_scaled_tpm_int,
    colData = col_data,
    design = ~ ontology + protocol
  )
  dds <- estimateSizeFactors(dds)
  dds <- estimateDispersions(dds, parallel = TRUE)
  vst_mat <- assay(vst(dds, blind = FALSE))

  # Mean-variance trend per gene, per representation (§4.12b column-choice
  # decision, discussed 2026-07-02): raw counts, raw TPM (pre-TMM, the
  # `raw_tpm` computed above) and GeTMM-CPM are all expected to keep a
  # mean-dependent variance (classic RNA-seq overdispersion); vst is
  # specifically designed to flatten it. Computed here, not as a separate
  # pass over `normalised_dt`, since `count_mat`/`raw_tpm`/`getmm_mat`/
  # `vst_mat` are already in scope and this is a one-line rowMeans/rowSds
  # per matrix — cheap relative to the vst() fit itself.
  gene_mean_var <- data.table(
    gene_id = rownames(count_mat),
    mean_count = rowMeans(count_mat),
    sd_count = matrixStats::rowSds(count_mat),
    mean_tpm = rowMeans(raw_tpm),
    sd_tpm = matrixStats::rowSds(raw_tpm),
    mean_getmm = rowMeans(getmm_mat),
    sd_getmm = matrixStats::rowSds(getmm_mat),
    mean_vst = rowMeans(vst_mat),
    sd_vst = matrixStats::rowSds(vst_mat)
  )

  # TMM normalisation factor spread (§4.12b column-choice decision): if
  # `calcNormFactors()`'s per-sample factors barely deviate from 1, this
  # dataset's composition bias is small and a naive per-gene z-score of raw
  # TPM might be "good enough" in practice; a wide spread means the
  # composition-bias correction TMM/GeTMM provides is doing real work.
  norm_factors <- data.table(
    uuid = colnames(count_mat),
    tmm_norm_factor = dge$samples$norm.factors
  )

  list(
    getmm = getmm_mat,
    vst = vst_mat,
    gene_mean_var = gene_mean_var,
    norm_factors = norm_factors
  )
}

normalise_one_filter <- function(this_transcript_filter) {
  quants <- isoform_quants[
    transcript_id %in%
      filtered_transcript_ids[[this_transcript_filter]] &
      uuid %in% sample_design$uuid
  ]

  gene_agg <- quants[,
    .(
      count = sum(expected_count, na.rm = TRUE),
      # weighted.mean() with all-zero weights (gene unexpressed in this sample)
      # would divide 0/0 -> NaN; fall back to the unweighted mean length there,
      # since the length estimate doesn't matter for an unexpressed gene.
      eff_length = if (sum(expected_count, na.rm = TRUE) > 0) {
        weighted.mean(effective_length, w = expected_count, na.rm = TRUE)
      } else {
        mean(effective_length, na.rm = TRUE)
      }
    ),
    by = .(uuid, gene_id)
  ]

  # gene x sample matrices (genes present in ALL samples for this filter;
  # a gene missing entirely from a sample's quantification is treated as
  # zero counts, not a missing value)
  count_wide <- dcast(gene_agg, gene_id ~ uuid, value.var = "count", fill = 0)
  efflen_wide <- dcast(gene_agg, gene_id ~ uuid, value.var = "eff_length")
  gene_ids <- count_wide$gene_id
  sample_uuids <- setdiff(names(count_wide), "gene_id")

  count_mat <- as.matrix(count_wide[, ..sample_uuids])
  efflen_mat <- as.matrix(efflen_wide[, ..sample_uuids])
  rownames(count_mat) <- rownames(efflen_mat) <- gene_ids

  # genes never expressed in ANY sample for this filter have effLen = NaN
  # everywhere (no count-weighted or unweighted mean possible if the
  # transcript's own effective_length was also NA); such genes cannot be
  # length-normalised meaningfully, so they are dropped from this filter's
  # normalisation (they contribute gene_expression = 0 in 05 either way).
  #
  # Also drop genes with effLen == 0 in any sample: RSEM legitimately
  # reports effective_length = 0 for transcripts shorter than the read
  # length (confirmed empirically: ~3.2% of all transcript-level rows in
  # isoform_quantifications_subset.tsv.gz). Dividing by 0 below
  # (count_mat / (efflen_mat / 1000)) produces Inf/NaN that poisons the
  # entire sample's colSums() in the vst rescaling step further down —
  # verified this causes DESeq2's "NA counts not allowed" error before this
  # guard was added. Same "cannot be length-normalised meaningfully"
  # reasoning as the NA case above, so the fix is the same: drop the gene.
  keep_genes <- rowSums(!is.na(efflen_mat) & efflen_mat > 0) == ncol(efflen_mat)
  count_mat <- count_mat[keep_genes, , drop = FALSE]
  efflen_mat <- efflen_mat[keep_genes, , drop = FALSE]

  col_data <- data.frame(
    uuid = sample_uuids,
    row.names = sample_uuids
  )
  col_data <- merge(col_data, sample_design, by = "uuid", sort = FALSE)
  rownames(col_data) <- col_data$uuid
  col_data <- col_data[sample_uuids, , drop = FALSE]

  raw_result <- compute_getmm_and_vst(count_mat, efflen_mat, col_data)

  list(
    main = data.table(
      transcript_filter = this_transcript_filter,
      gene_id = rep(gene_ids[keep_genes], times = length(sample_uuids)),
      uuid = rep(sample_uuids, each = sum(keep_genes)),
      gene_expression_getmm = as.vector(raw_result$getmm),
      gene_expression_vst = as.vector(raw_result$vst)
    ),
    gene_mean_var = raw_result$gene_mean_var[,
      transcript_filter := this_transcript_filter
    ],
    norm_factors = raw_result$norm_factors[,
      transcript_filter := this_transcript_filter
    ]
  )
}
```

```{r, run-normalisation}
# getmm and vst columns are written side by side; routing is decided (getmm →
# splicing_ml, vst → 07/09/RBP + the 02-3 event gate — see load-packages).
# ComBat-seq (_combat) sibling columns were also computed here through
# 2026-07-02 but removed after failing on real data — see load-packages.
#
# Per-filter Snakemake job invocation: TRANSCRIPT_FILTER env var (lines 74–79)
# routes one transcript_filter per rule execution. DESeq2's vst() is the
# bottleneck — parallelise gene-wise dispersion fitting within this filter
# via BiocParallel (estimateDispersions + vst use all 4 SLURM threads from
# config/snakemake_config.yaml line 91).
register(MulticoreParam(4))
filter_result <- normalise_one_filter(this_transcript_filter)

normalised_dt <- filter_result$main
fwrite(
  normalised_dt,
  sprintf(
    "processed_data/gene_expression_normalised_%s.csv.gz",
    this_transcript_filter
  )
)

# §4.12b column-choice diagnostics (discussed 2026-07-02): kept as separate,
# much smaller tables (one row per gene, one row per sample — not per
# gene x sample like `normalised_dt`) rather than widening the production
# output, since these columns (raw count/TPM mean+sd, TMM factors) exist to
# decide which of getmm/vst to adopt, not to ship to `05`.
gene_mean_var_dt <- filter_result$gene_mean_var
fwrite(
  gene_mean_var_dt,
  sprintf(
    "qc/gene_expression_mean_variance_%s.csv.gz",
    this_transcript_filter
  )
)

norm_factors_dt <- filter_result$norm_factors
fwrite(
  norm_factors_dt,
  sprintf(
    "qc/gene_expression_tmm_norm_factors_%s.csv.gz",
    this_transcript_filter
  )
)
```

## Column-choice diagnostics (§4.12b): mean-variance trend and TMM factor spread

Two cheap, no-model-training checks to inform the getmm-vs-vst decision
(discussed 2026-07-02, see `revision/PLAN.md` §4.12b): does vst actually
flatten the mean-variance relationship relative to count/TPM/getmm, and is
this dataset's composition bias (what TMM/GeTMM corrects for) large enough
to matter in the first place. `gene_expression_mean_variance.csv.gz` and
`gene_expression_tmm_norm_factors.csv.gz` in `qc/` carry the underlying
per-gene and per-sample values (mean/sd of count, TPM, getmm, vst; TMM
factor per sample) so these numbers are reproducible without re-running
`ComBat_seq`-scale compute.

```{r, mean-variance-trend}
mean_var_long <- rbindlist(list(
  gene_mean_var_dt[,
    .(
      gene_id,
      transcript_filter,
      representation = "count",
      mean = mean_count,
      sd = sd_count
    )
  ],
  gene_mean_var_dt[,
    .(
      gene_id,
      transcript_filter,
      representation = "tpm",
      mean = mean_tpm,
      sd = sd_tpm
    )
  ],
  gene_mean_var_dt[,
    .(
      gene_id,
      transcript_filter,
      representation = "getmm",
      mean = mean_getmm,
      sd = sd_getmm
    )
  ],
  gene_mean_var_dt[,
    .(
      gene_id,
      transcript_filter,
      representation = "vst",
      mean = mean_vst,
      sd = sd_vst
    )
  ]
))
mean_var_long[,
  transcript_filter := factor(
    transcript_filter,
    levels = c("transcripts", "tsl_filtered", "biotype_filtered")
  )
]
mean_var_long[,
  representation := factor(
    representation,
    levels = c("count", "tpm", "getmm", "vst")
  )
]

# Spearman (rank-based, robust to the different scales/transforms across
# representations) correlation between per-gene mean and sd — near 0 means
# variance no longer depends on expression level (the point of vst); count,
# TPM and getmm are all expected to still show a clear positive trend.
meanvar_corr <- mean_var_long[,
  .(
    spearman_mean_sd = cor(mean, sd, method = "spearman", use = "complete.obs")
  ),
  by = .(transcript_filter, representation)
]
meanvar_wide <- dcast(
  meanvar_corr,
  transcript_filter ~ representation,
  value.var = "spearman_mean_sd"
)
knitr::kable(
  meanvar_wide,
  caption = "Spearman correlation between per-gene mean and sd, by representation and transcript_filter (near 0 = variance-stabilised)"
)

# Decision signal: is vst's mean-sd correlation meaningfully flatter than
# getmm's for the same transcript_filter? If not, vst isn't buying the
# variance-stabilisation it's chosen for here, and the getmm/vst split
# recommendation (§4.12b) should be reconsidered.
for (tf in as.character(meanvar_wide$transcript_filter)) {
  row <- meanvar_wide[transcript_filter == tf]
  status <- if (row$vst < row$getmm * 0.5) "PASS" else "WARN"
  append_sanity(
    paste0("meanvar_trend_vst_vs_getmm__", tf),
    status,
    row$vst,
    row$getmm,
    sprintf(
      "Spearman(mean,sd): vst=%.3f vs getmm=%.3f -- PASS if vst is under half of getmm's (variance meaningfully flatter)",
      row$vst,
      row$getmm
    )
  )
}

# count/TPM/getmm are non-negative and right-skewed -- log1p both axes.
# vst is plotted separately (own chunk-local block below) since it can be
# negative and log1p would be wrong for it, same reasoning as the
# getmm-vs-vst event-type plots earlier in this file.
p_meanvar_nonneg <- ggplot(
  mean_var_long[representation != "vst"],
  aes(x = log1p(mean), y = log1p(sd), color = representation)
) +
  geom_point(alpha = 0.15, size = 0.4) +
  geom_smooth(se = FALSE, linewidth = 0.6) +
  facet_wrap(~transcript_filter, nrow = 1) +
  scale_color_viridis_d(option = "D") +
  labs(
    title = "Mean-variance trend: count, TPM, GeTMM-CPM",
    subtitle = "log1p both axes -- all three expected to show an increasing trend",
    x = "log1p(mean)",
    y = "log1p(sd)",
    color = "Representation"
  )
print(p_meanvar_nonneg)
ggsave(
  "qc/distributions/gene_expression_meanvar_count_tpm_getmm.pdf",
  p_meanvar_nonneg,
  width = 9,
  height = 5
)

p_meanvar_vst <- ggplot(
  mean_var_long[representation == "vst"],
  aes(x = mean, y = sd)
) +
  geom_point(alpha = 0.15, size = 0.4) +
  geom_smooth(se = FALSE, linewidth = 0.6, color = "black") +
  facet_wrap(~transcript_filter, nrow = 1) +
  labs(
    title = "Mean-variance trend: vst",
    subtitle = "Raw scale (vst can be negative) -- a flat trend is the intended result",
    x = "mean(vst)",
    y = "sd(vst)"
  )
print(p_meanvar_vst)
ggsave(
  "qc/distributions/gene_expression_meanvar_vst.pdf",
  p_meanvar_vst,
  width = 9,
  height = 5
)
```

```{r, tmm-factor-spread}
# If TMM factors barely deviate from 1, this dataset's composition bias is
# small and a naive per-gene z-score of raw TPM might be adequate in
# practice; a wide spread means TMM/GeTMM's correction is doing real work.
norm_factor_summary <- norm_factors_dt[,
  .(
    mean_factor = mean(tmm_norm_factor),
    cv_pct = 100 * sd(tmm_norm_factor) / mean(tmm_norm_factor),
    min_factor = min(tmm_norm_factor),
    max_factor = max(tmm_norm_factor)
  ),
  by = transcript_filter
]
knitr::kable(
  norm_factor_summary,
  caption = "TMM normalisation factor spread per transcript_filter (factors far from 1 / high CV% = real composition bias)"
)

# Note the polarity here: WARN (>10% CV) means composition bias is large
# enough to matter -- i.e. an argument FOR GeTMM/vst over a naive TPM
# z-score -- not a data-quality problem with this pipeline.
for (i in seq_len(nrow(norm_factor_summary))) {
  row <- norm_factor_summary[i]
  status <- if (row$cv_pct > 10) "WARN" else "PASS"
  append_sanity(
    paste0("tmm_factor_spread__", row$transcript_filter),
    status,
    row$cv_pct,
    10,
    "TMM norm.factor CV% across samples -- WARN (>10%) means composition bias is large enough that GeTMM/vst correction matters, not a pipeline data-quality issue"
  )
}

p_tmm_spread <- ggplot(
  norm_factors_dt,
  aes(x = transcript_filter, y = tmm_norm_factor)
) +
  geom_boxplot(width = 0.4, outlier.alpha = 0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "TMM normalisation factor spread per sample",
    subtitle = "Dashed line = no correction needed (factor = 1)",
    x = NULL,
    y = "TMM norm.factor"
  )
print(p_tmm_spread)
ggsave(
  "qc/distributions/tmm_norm_factor_spread.pdf",
  p_tmm_spread,
  width = 7,
  height = 5
)
```

## ComBat-seq protocol correction — attempted, abandoned

ComBat-seq batch correction (§4.12c) was implemented here 2026-07-02 as
`_combat` sibling columns alongside raw getmm/vst, then removed the same day
after running on real data: `ComBat_seq()` failed (its `~ batch + group`
design matrix is degenerate for most `ontology` groups, which are
single-`protocol`-dominated — confirmed via the file_table cross-tab, not
just suspected). See `revision/file-changes/02-2-rnaseq-normalisation.Rmd.md`
for the full writeup and the real sanity-check evidence
(`qc/sanity_summary.csv`'s `combat_ri_protocol_beta_reduction` row, logged
before this branch was removed). Decision: use raw getmm/vst; the existing
protocol + protocol×gene_expression ML interaction term is the mitigation
for RI protocol bias going forward, not a corrected `gene_expression` value.
