---
title: "rMATS event filtering + VST gene-expression gate (stage 02-3)"
author: "Quirin Manz"
date: "`r Sys.Date()`"
output: html_document
---

Stage 02-3 (was `02z-rmats-analysis.Rmd`). Per-`transcript_filter` (env
`TRANSCRIPT_FILTER`). Post-processes the external rMATS output into the
high-confidence event set and writes SUPPA-format `event_{et}.psi` +
`event_{et}.jc.csv.gz` under `splicing_analysis/rmats/{tf}/` (consumed by `03`).

Two filters combine into the high-confidence set:

1. **Wang et al. 2024 (Nat Protoc) Procedure 2** — average junction count ≥ 20
   AND 5th-percentile PSI ≥ 0.05 AND 95th-percentile PSI ≤ 0.95.
2. **VST gene-expression gate (§4.12e)** — per (event, sample), keep only where
   the host gene's `gene_expression_vst` (from `02-2`) is ≥ an on/off cutoff (the
   antimode of a 2-component Gaussian mixture over all genome-wide (gene, sample)
   VST for this filter). Masked cells become NA in `.psi` and are dropped from
   `.jc`. No minimum-sample floor (removed at data construction; `09-1` keeps its
   own CV-fittability guard).

Auditability: cutoff + GMM params → `qc/vst_expression_cutoff_{tf}.csv`; gate
effect (events dropped by VST vs Procedure 2, % cells masked) →
`qc/vst_gate_effect_{tf}.csv`; density plot in `qc/distributions/`.

```{r}
set.seed(getOption("EpiATLAS_AS_SEED", 42L))
rmats_command_file <- "data/rmats_split_post_commands.tsv"
command_dt <- fread(rmats_command_file)
experiment_metadata <- fread("data/EpiATLAS_experiment_metadata.csv")
experiment_metadata[, experiment_type := factor(experiment_type)]
file_metadata <- fread("processed_data/file_table.csv.gz")
rna_to_use <- file_metadata[
  assay_type == "RNA-Seq" &
    epirr_id_without_version %in%
      file_metadata[,
        if (.N >= 8) .SD,
        by = epirr_id_without_version
      ][, epirr_id_without_version],
  uuid
]
# bam_dt[experiment_metadata[, epirr := epirr_id], on = .NATURAL, uuid := uuid]

# Per-filter invocation (PLAN §4.12e per-filter incremental outputs): this
# notebook post-processes exactly one transcript_filter per run. Snakemake
# passes TRANSCRIPT_FILTER as an env var; interactive runs fall back to the
# primary filter.
this_transcript_filter <- Sys.getenv(
  "TRANSCRIPT_FILTER",
  unset = getOption("EpiATLAS_AS_PRIMARY_FILTER", "biotype_filtered")
)
```

```{r, fig.width=12, fig.height=8}
library(UpSetR)
library(grid)

# ID format for suppa: "ENSG00000000419.12;SE:chr20:50940933-50941129:50941209-50942031:-"
# Wang et al. 2024 (Nat Protoc) Procedure 2 thresholds — getOption-driven so a
# rerun can vary them without editing this file.
average_read_coverage <- getOption("EpiATLAS_AS_AVG_READ_COVERAGE", 20)
min_average_PSI <- getOption("EpiATLAS_AS_MIN_AVG_PSI", 0.05)
max_average_PSI <- 1 - min_average_PSI
min_read_coverage <- 10 # diagnostic only (keep_ids_min_sample), not applied
min_samples_coverage <- 3 # diagnostic only
splicing_dir <- file.path("splicing_analysis", "rmats")
dir.create(splicing_dir, recursive = TRUE, showWarnings = FALSE)

# --- VST low-expression gate (PLAN §4.12e) ----------------------------------
# High-confidence event set = Procedure 2 (read coverage + PSI percentiles)
# AND a per-(event,sample) gene-expression gate: a sample's PSI is trusted
# only if its host gene is expressed above an on/off cutoff. The cutoff is the
# antimode (valley) of a 2-component Gaussian mixture fit on the WHOLE
# genome-wide (gene,sample) VST distribution for this filter — an absolute
# expressed/not boundary, not one drawn from the pre-selected event genes.
gene_vst <- fread(
  sprintf(
    "processed_data/gene_expression_normalised_%s.csv.gz",
    this_transcript_filter
  ),
  select = c("gene_id", "uuid", "gene_expression_vst")
)
gene_vst[, gene_id_stripped := sub("\\.\\d+$", "", gene_id)]
# collapse any version-strip collisions (one value per (gene,sample))
gene_vst <- unique(gene_vst, by = c("gene_id_stripped", "uuid"))

compute_vst_cutoff <- function(
  vst_values,
  method = getOption("EpiATLAS_AS_VST_CUTOFF_METHOD", "gmm"),
  fixed_value = getOption("EpiATLAS_AS_VST_CUTOFF_VALUE", NA_real_),
  percentile = getOption("EpiATLAS_AS_VST_CUTOFF_PERCENTILE", 0.25)
) {
  vst_values <- vst_values[is.finite(vst_values)]
  pctl <- function() as.numeric(stats::quantile(vst_values, percentile))
  if (identical(method, "fixed")) {
    if (is.na(fixed_value)) {
      stop("VST cutoff method 'fixed' but EpiATLAS_AS_VST_CUTOFF_VALUE unset")
    }
    return(list(cutoff = fixed_value, method = "fixed"))
  }
  if (identical(method, "percentile")) {
    return(list(cutoff = pctl(), method = "percentile"))
  }
  # gmm (default): 2-component 1-D Gaussian mixture, cutoff = density antimode.
  # mclust EM over millions of (gene,sample) points is slow and can fail/OOM,
  # which silently forces the fallback — fit on a representative subsample
  # instead (seed set at notebook top → reproducible).
  n_fit <- getOption("EpiATLAS_AS_VST_GMM_MAX_N", 1e5L)
  fit_values <- if (length(vst_values) > n_fit) {
    sample(vst_values, n_fit)
  } else {
    vst_values
  }
  fit_err <- NULL
  fit <- tryCatch(
    mclust::Mclust(fit_values, G = 2, modelNames = "V", verbose = FALSE),
    error = function(e) {
      fit_err <<- conditionMessage(e)
      NULL
    }
  )
  degenerate <- is.null(fit) ||
    is.null(fit$parameters) ||
    length(unique(fit$parameters$mean)) < 2L
  if (degenerate) {
    reason <- if (!is.null(fit_err)) {
      sprintf("mclust error: %s", fit_err)
    } else if (is.null(fit)) {
      "mclust returned NULL"
    } else {
      "fit not 2-component (means collapsed)"
    }
    message(sprintf("VST GMM cutoff fell back (%s)", reason))
    if (!is.na(fixed_value)) {
      return(list(cutoff = fixed_value, method = "gmm->fixed", reason = reason))
    }
    return(list(cutoff = pctl(), method = "gmm->percentile", reason = reason))
  }
  mus <- fit$parameters$mean
  sds <- sqrt(fit$parameters$variance$sigmasq)
  if (length(sds) == 1L) sds <- rep(sds, 2L)
  pro <- fit$parameters$pro
  o <- order(mus)
  mus <- mus[o]
  sds <- sds[o]
  pro <- pro[o]
  grid <- seq(mus[1], mus[2], length.out = 2000L)
  dens <- pro[1] * dnorm(grid, mus[1], sds[1]) +
    pro[2] * dnorm(grid, mus[2], sds[2])
  # return the fitted 2-component params so the antimode placement (cutoff
  # between the two modes) is auditable straight from qc/vst_expression_cutoff_*
  list(
    cutoff = grid[which.min(dens)], method = "gmm",
    mu_low = mus[1], mu_high = mus[2],
    sd_low = sds[1], sd_high = sds[2],
    pro_low = pro[1], pro_high = pro[2],
    bic = tryCatch(as.numeric(fit$bic), error = function(e) NA_real_)
  )
}

vst_cut <- compute_vst_cutoff(gene_vst$gene_expression_vst)
vst_cutoff <- vst_cut$cutoff
vst_na_policy <- getOption("EpiATLAS_AS_VST_NA_POLICY", "keep")
dir.create("qc", showWarnings = FALSE)
dir.create("qc/distributions", recursive = TRUE, showWarnings = FALSE)
# NULL-safe getter: fallback methods (fixed/percentile) return no GMM params
.nz <- function(x) if (is.null(x)) NA_real_ else as.numeric(x)
fwrite(
  data.table(
    transcript_filter = this_transcript_filter,
    method = vst_cut$method,
    cutoff = vst_cutoff,
    na_policy = vst_na_policy,
    mu_low = .nz(vst_cut$mu_low),
    mu_high = .nz(vst_cut$mu_high),
    sd_low = .nz(vst_cut$sd_low),
    sd_high = .nz(vst_cut$sd_high),
    pro_low = .nz(vst_cut$pro_low),
    pro_high = .nz(vst_cut$pro_high),
    bic = .nz(vst_cut$bic)
  ),
  sprintf("qc/vst_expression_cutoff_%s.csv", this_transcript_filter)
)
if (vst_cut$method != "gmm") {
  warning(sprintf(
    "VST cutoff for %s used fallback method '%s' (cutoff = %.3f)%s",
    this_transcript_filter, vst_cut$method, vst_cutoff,
    if (!is.null(vst_cut$reason)) sprintf(" - reason: %s", vst_cut$reason) else ""
  ))
}
# fresh gate-effect log for this run (appended once per event type below)
unlink(sprintf("qc/vst_gate_effect_%s.csv", this_transcript_filter))
p_vst_cut <- ggplot(
  gene_vst[is.finite(gene_expression_vst)],
  aes(x = gene_expression_vst)
) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = vst_cutoff, colour = "red", linetype = "dashed") +
  labs(
    title = sprintf(
      "Genome-wide VST distribution (%s) - cutoff %.3f (%s)",
      this_transcript_filter, vst_cutoff, vst_cut$method
    ),
    x = "gene_expression_vst", y = "count (gene x sample)"
  )
ggsave(
  sprintf(
    "qc/distributions/vst_expression_cutoff_%s.pdf", this_transcript_filter
  ),
  p_vst_cut, width = 7, height = 5
)

event_type <- "SE"
junction_counts <- "MATS.JC.txt"
run_name <- "paired_fr-firststrand_75"
plot_dist <- TRUE

# restrict to the single requested transcript_filter (per-filter incremental)
rmats_dirs <- list.dirs(file.path(data_dir, "rmats"), recursive = FALSE)
rmats_dirs <- rmats_dirs[basename(rmats_dirs) == this_transcript_filter]
stopifnot(length(rmats_dirs) == 1L)
rmats_dir <- rmats_dirs

for (event_type in c("SE", "RI", "A3SS", "A5SS", "MXE")) {
  print(event_type)
  for (rmats_dir in rmats_dirs) {
    print(basename(rmats_dir))
    this_splicing_dir <- file.path(splicing_dir, basename(rmats_dir))
    dir.create(this_splicing_dir, recursive = TRUE, showWarnings = FALSE)
    for (junction_counts in c(
      "MATS.JC.txt" # , "MATS.JCEC.txt"
    )) {
      # read annotation file
      event_annotation <- fread(file.path(
        rmats_dir,
        command_dt[grep(basename(rmats_dir), out_folder), run_name][1],
        paste("fromGTF", event_type, "txt", sep = ".")
      ))
      if (event_type == "SE") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            upstreamEE,
            exonStart_0base + 1,
            exonEnd,
            downstreamES + 1,
            strand
          )
        ]
      } else if (event_type == "RI") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s:%s-%s:%s:%s",
            GeneID,
            event_type,
            chr,
            upstreamES + 1,
            upstreamEE,
            downstreamES + 1,
            downstreamEE,
            strand
          )
        ]
      } else if (event_type == "A3SS") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            ifelse(strand == "+", flankingEE, longExonEnd),
            ifelse(strand == "+", longExonStart_0base + 1, flankingES + 1),
            ifelse(strand == "+", flankingEE, shortEE),
            ifelse(strand == "+", shortES + 1, flankingES + 1),
            strand
          )
        ]
      } else if (event_type == "A5SS") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            ifelse(strand == "+", longExonEnd, flankingEE),
            ifelse(strand == "+", flankingES + 1, longExonStart_0base + 1),
            ifelse(strand == "+", shortEE, flankingEE),
            ifelse(strand == "+", flankingES + 1, shortES + 1),
            strand
          )
        ]
      } else if (event_type == "MXE") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            upstreamEE,
            `1stExonStart_0base` + 1,
            `1stExonEnd`,
            downstreamES + 1,
            upstreamEE,
            `2ndExonStart_0base` + 1,
            `2ndExonEnd`,
            downstreamES + 1,
            strand
          )
        ]
      } else {
        stop("Not implemented yet")
      }

      event_sample_data_path <- file.path(
        rmats_dir,
        paste(event_type, junction_counts, "csv.gz", sep = ".")
      )
      if (file.exists(event_sample_data_path)) {
        event_sample_data <- fread(file.path(
          rmats_dir,
          paste(event_type, junction_counts, "csv.gz", sep = ".")
        ))
      } else {
        id_vars <-
          c(
            "ID"
          )
        measure_vars <-
          c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IncLevel1")
        dt_list <- pbmclapply(
          command_dt[grep(basename(rmats_dir), out_folder), run_name],
          function(this_run_name) {
            junction_file <- file.path(
              rmats_dir,
              this_run_name,
              paste(event_type, junction_counts, sep = ".")
            )
            # read file with counts
            dt <-
              fread(junction_file, select = c(id_vars, measure_vars))
            # convert to long format

            long_dt <-
              melt(
                dt,
                id.vars = id_vars,
                measure.vars = measure_vars
              )
            # add sample names, by splitting by comma and retaining the order of which they appear in the rmats input file
            uuids <- command_dt[
              grep(basename(rmats_dir), out_folder) & run_name == this_run_name,
              strsplit(uuid_string, ",", fixed = TRUE)
            ][[1]]
            long_dt[,
              (uuids) := tstrsplit(
                value,
                ",",
                fixed = TRUE,
                type.convert = TRUE
              )
            ]
            # convert to even longer format to have one row per sample
            lapply(
              split(long_dt, long_dt[, variable]),
              melt,
              id.vars = c(id_vars, "variable"),
              measure.vars = uuids,
              variable.name = "uuid"
            )
          }
        )
        names(dt_list) <- command_dt[
          grep(basename(rmats_dir), out_folder),
          run_name
        ]
        long_dt_list <- sapply(
          measure_vars,
          function(measure_var) {
            rbindlist(
              sapply(
                dt_list,
                function(dt_split) {
                  dt_split[[measure_var]]
                },
                simplify = FALSE
              ),
              idcol = "run_name"
            )
          },
          simplify = FALSE
        )

        stopifnot(identical(
          long_dt_list$IJC_SAMPLE_1[, .(ID, uuid)],
          long_dt_list$SJC_SAMPLE_1[, .(ID, uuid)]
        ))

        stopifnot(identical(
          long_dt_list$SJC_SAMPLE_1[, .(ID, uuid)],
          long_dt_list$IncLevel1[, .(ID, uuid)]
        ))

        event_sample_data <- long_dt_list$IJC_SAMPLE_1[, .(
          ID,
          uuid,
          incl_count = value
        )]
        event_sample_data[, skip_count := long_dt_list$SJC_SAMPLE_1[, value]]
        event_sample_data[, psi := long_dt_list$IncLevel1[, value]]

        fwrite(event_sample_data, event_sample_data_path)
      }

      # event_sample_data[, total_count := incl_count + skip_count]

      # # add experiment type
      event_sample_data[
        experiment_metadata,
        on = .(uuid),
        experiment_type := experiment_type
      ]

      # compute average counts per event and per experiment type and event
      # recompute for the RNA used in this analysis as well
      # Only the epigenome cohort is written now: it is the set VST was
      # computed on and the only output 03 consumes (the old diagnostic "all"
      # variant is dropped — PLAN §4.12e).
      uuids_to_filter_on <- list(
        full_epigenomes = rna_to_use
      )
      mapply(
        as.list(names(uuids_to_filter_on)),
        uuids_to_filter_on,
        FUN = function(filter_name, uuids_to_filter) {
          print(filter_name)
          filtered_dt <- event_sample_data[uuid %in% uuids_to_filter]

          # --- VST low-expression gate (per (event,sample)) ---------------
          # Attach host gene_id (from rMATS GeneID, version-stripped to match
          # the normalisation table) and its VST; a sample passes only if its
          # host gene is expressed at/above the cutoff. NA VST (host gene
          # absent from normalisation) is handled per EpiATLAS_AS_VST_NA_POLICY.
          id_gene <- unique(event_annotation[, .(
            ID,
            gene_id_stripped = sub("\\.\\d+$", "", GeneID)
          )])
          filtered_dt[id_gene, on = .(ID), gene_id_stripped := i.gene_id_stripped]
          filtered_dt[
            gene_vst,
            on = .(gene_id_stripped, uuid),
            vst := i.gene_expression_vst
          ]
          if (identical(vst_na_policy, "drop")) {
            filtered_dt[, expr_pass := !is.na(vst) & vst >= vst_cutoff]
          } else {
            filtered_dt[, expr_pass := is.na(vst) | vst >= vst_cutoff]
          }
          # report events lost SOLELY to missing VST (every sample NA), so the
          # loss is visible rather than silent
          na_only_events <- filtered_dt[,
            .(all_na = all(is.na(vst))),
            by = ID
          ][all_na == TRUE, .N]
          message(sprintf(
            "%s %s: VST cutoff %.3f (%s); %d events have no VST for any sample (na_policy=%s)",
            event_type, basename(rmats_dir), vst_cutoff, vst_cut$method,
            na_only_events, vst_na_policy
          ))
          average_dt <- rbind(
            filtered_dt[,
              .(
                incl_sum = sum(incl_count, na.rm = TRUE),
                skip_sum = sum(skip_count, na.rm = TRUE),
                average_psi = mean(psi, na.rm = TRUE),
                n_samples = .N
              ),
              by = ID
            ],
            filtered_dt[,
              .(
                incl_sum = sum(incl_count, na.rm = TRUE),
                skip_sum = sum(skip_count, na.rm = TRUE),
                average_psi = mean(psi, na.rm = TRUE),
                n_samples = .N
              ),
              by = .(ID, experiment_type)
            ],
            fill = TRUE
          )
          average_dt[, average_count := (incl_sum + skip_sum) / n_samples]

          # scatterplot average count vs average PSI
          ggplot(
            average_dt,
            aes(x = average_count + 1, y = average_psi, color = experiment_type)
          ) +
            geom_point(alpha = .5) +
            scale_x_log10() +
            # create a line y ~ x
            geom_smooth(method = "lm", se = FALSE) +
            ggpubr::stat_cor(
              method = "spearman",
              label.x.npc = "middle",
              label.y.npc = "bottom"
            ) +
            facet_wrap(~experiment_type) +
            labs(title = sprintf("Average Count vs Average PSI:"))

          # plot average count and PSI distributions
          if (plot_dist) {
            print(
              ggplot(
                average_dt,
                aes(x = average_count + 1, color = experiment_type)
              ) +
                geom_line(aes(y = 1 - after_stat(y)), stat = "ecdf") +
                scale_x_log10() +
                geom_vline(xintercept = average_read_coverage + 1) +
                labs(
                  title = sprintf(
                    "Average Read Coverage: with line at %d; %s %s %s",
                    average_read_coverage + 1,
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  )
                )
            )
            print(
              ggplot(average_dt, aes(x = average_psi, fill = experiment_type)) +
                geom_histogram(position = "dodge") +
                geom_vline(xintercept = c(min_average_PSI, max_average_PSI)) +
                labs(
                  title = sprintf(
                    "Average PSI: %s %s %s",
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  )
                )
            )
            print(
              ggplot(
                average_dt[
                  average_psi >= min_average_PSI &
                    average_psi <= max_average_PSI
                ],
                aes(x = average_psi, fill = experiment_type)
              ) +
                geom_histogram(position = "dodge", binwidth = .05) +
                geom_vline(xintercept = c(min_average_PSI, max_average_PSI)) +
                labs(
                  title = sprintf(
                    "Average PSI: %f <= PSI < %f; %s %s %s",
                    min_average_PSI,
                    max_average_PSI,
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  )
                )
            )

            ### Check per-sample read coverage filtering ----
            min_read_cutoffs <- c(0, 1, 5, 10, 15, 20)
            min_sample_cutoffs <- c(1, 3, 5, 10)

            # Pre-compute per-threshold counts once (6 passes) rather than 24 (6x4 grid)
            count_by_thresh <- setNames(
              lapply(min_read_cutoffs, function(thresh) {
                filtered_dt[incl_count + skip_count >= thresh, .N, by = ID]
              }),
              as.character(min_read_cutoffs)
            )

            plot_dt <- CJ(
              min_read_coverage = min_read_cutoffs,
              min_samples_coverage = min_sample_cutoffs
            )

            plot_dt[,
              n_pass := count_by_thresh[[as.character(min_read_coverage)]][
                N >= min_samples_coverage,
                .N
              ],
              by = .(min_read_coverage, min_samples_coverage)
            ]

            print(
              ggplot(
                plot_dt,
                aes(
                  x = min_samples_coverage,
                  y = n_pass,
                  color = factor(min_read_coverage)
                )
              ) +
                geom_line() +
                geom_point() +
                labs(
                  title = sprintf(
                    "%s %s %s",
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  ),
                  x = "Minimum number of samples with sufficient read count",
                  y = "Number of events (ID) retained",
                  color = "Min per-sample read count"
                )
            )

            ###  Check average read coverages and PSI filtering ----

            # Step 1: Create grid of thresholds to test
            min_PSI_cutoffs <- seq(0, 0.05, by = 0.01)

            plot_dt <- CJ(
              average_read_coverage = min_read_cutoffs,
              min_average_PSI = min_PSI_cutoffs
            )

            plot_dt[,
              n_pass := average_dt[
                !is.na(experiment_type) &
                  average_count >= average_read_coverage &
                  average_psi >= min_average_PSI &
                  average_psi <= 1 - min_average_PSI,
                ID,
                by = experiment_type
              ][duplicated(ID), .N],
              by = .(average_read_coverage, min_average_PSI)
            ]

            print(
              ggplot(
                plot_dt,
                aes(
                  x = min_average_PSI,
                  y = n_pass,
                  color = factor(average_read_coverage)
                )
              ) +
                geom_line() +
                geom_point() +
                labs(
                  title = sprintf(
                    "Min averages per experiment type: %s %s %s",
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  ),
                  x = "Minimum average PSI",
                  y = "Number of events (ID) retained",
                  color = "Min average read coverage"
                )
            )
          }
          # get IDs to keep based on average PSI and average read coverage
          keep_ids_orig <-
            average_dt[
              !is.na(experiment_type) &
                average_count >= average_read_coverage &
                average_psi >= min_average_PSI &
                average_psi <= 1 - min_average_PSI,
              ID,
              by = experiment_type
            ][duplicated(ID), ID]
          length(keep_ids_orig)
          # number of unique IDs passing the filter across all experiment types
          average_dt[
            is.na(experiment_type) &
              average_count >= average_read_coverage &
              average_psi >= min_average_PSI &
              average_psi <= 1 - min_average_PSI,
            uniqueN(ID)
          ]

          # intersect(
          #   average_psi_dt[!is.na(experiment_type) &
          #     average_psi >= min_average_PSI &
          #     average_psi <= max_average_PSI, ID, by = experiment_type][duplicated(ID), ID],
          #   average_count_dt[!is.na(experiment_type) &
          #     average_count >= average_read_coverage, ID, by = experiment_type][duplicated(ID), ID]
          # )

          keep_ids_min_sample <- filtered_dt[
            incl_count + skip_count >= min_read_coverage,
            .N,
            by = ID
          ][N >= min_samples_coverage, ID]

          # Wang et al. 2024 (Nature Protocols) Procedure 2 filter, applied to
          # the VST-passing samples only: average count >= 20 + 5th/95th PSI
          # percentiles. Removes constitutively included (psi_q05 >
          # max_average_PSI) and excluded (psi_q95 < min_average_PSI) events.
          # NB: computed on expr_pass samples, so percentiles/coverage reflect
          # the high-confidence subset. No minimum-sample floor (PLAN: removed
          # at data-construction; 09-1 keeps its CV-fittability guard).
          proc2_ids <- function(dt) {
            tmp <- dt[!is.na(psi)]
            setorder(tmp, ID, psi)
            tmp[,
              .(
                psi_q05 = psi[max(1L, ceiling(0.05 * .N))],
                psi_q95 = psi[min(.N, ceiling(0.95 * .N))],
                average_count = mean(incl_count + skip_count, na.rm = TRUE)
              ),
              by = ID
            ][
              average_count >= average_read_coverage &
                psi_q95 >= min_average_PSI &
                psi_q05 <= max_average_PSI,
              ID
            ]
          }
          # applied filter: Procedure 2 on VST-passing samples
          keep_ids_percentile <- proc2_ids(filtered_dt[expr_pass == TRUE])
          # counterfactual: Procedure 2 ignoring the VST gate — isolates what
          # the VST gate removes vs what Procedure 2 removes on its own
          keep_ids_percentile_novst <- proc2_ids(filtered_dt)

          upset <- upset(
            fromList(list(
              orig = keep_ids_orig,
              percentile = keep_ids_percentile,
              min_sample = keep_ids_min_sample
            )),
            nsets = 3,
            order.by = "freq",
            mainbar.y.label = "Number of events (ID)",
            sets.x.label = "Total events (ID)"
          )
          print(upset)
          grid.text(
            sprintf("%s %s %s", event_type, basename(rmats_dir), filter_name),
            x = 0.65,
            y = 0.95,
            gp = gpar(fontsize = 20)
          )

          # --- VST gate-effect summary (auditable) -------------------------
          # Decompose event loss: Procedure 2 alone vs the added VST gate.
          #   dropped_proc2 : fail Procedure 2 even without VST
          #   dropped_vst   : pass Procedure 2 on full samples, lost after VST
          #   kept          : survive both (= keep_ids_percentile)
          n_total <- filtered_dt[!is.na(psi), uniqueN(ID)]
          n_kept <- length(keep_ids_percentile)
          n_pass_novst <- length(keep_ids_percentile_novst)
          n_dropped_proc2 <- n_total - n_pass_novst
          n_dropped_vst <- length(setdiff(
            keep_ids_percentile_novst, keep_ids_percentile
          ))
          frac_cells_masked <- filtered_dt[!is.na(psi), mean(!expr_pass)]
          gate_effect <- data.table(
            transcript_filter = this_transcript_filter,
            event_type = event_type,
            n_events_total = n_total,
            n_events_kept = n_kept,
            n_dropped_proc2_only = n_dropped_proc2,
            n_dropped_vst_gate = n_dropped_vst,
            frac_cells_vst_masked = frac_cells_masked,
            vst_cutoff = vst_cutoff,
            na_policy = vst_na_policy
          )
          print(gate_effect)
          gate_effect_file <- sprintf(
            "qc/vst_gate_effect_%s.csv", this_transcript_filter
          )
          fwrite(gate_effect, gate_effect_file, append = file.exists(gate_effect_file))

          # Emit only cells that pass BOTH the Procedure 2 event filter and the
          # per-(event,sample) VST gate. Build from filtered_dt (which carries
          # expr_pass); VST-masked cells are simply absent, so dcast fills them
          # NA in the .psi matrix, and they are dropped from the .jc file.
          filtered_dt[event_annotation, on = .(ID), suppa_id := i.suppa_id]
          write_dt <- filtered_dt[
            ID %in% keep_ids_percentile & expr_pass == TRUE
          ]
          suppa_format_dt <- dcast(
            write_dt,
            suppa_id ~ uuid,
            value.var = "psi"
          )
          suppa_file <- file.path(
            this_splicing_dir,
            paste0(
              "event_",
              substr(event_type, 1, 2),
              ".psi"
            )
          )
          write.table(
            as.data.frame(
              suppa_format_dt[, -"suppa_id"],
              row.names = suppa_format_dt[, suppa_id]
            ),
            suppa_file,
            sep = "\t",
            quote = FALSE
          )
          fwrite(
            unique(event_annotation[, .(ID, geneSymbol, suppa_id)]),
            paste0(suppa_file, ".annot")
          )
          # Write IJC/SJC junction counts for downstream coverage weighting (§4.8/§4.12f)
          fwrite(
            write_dt[, .(suppa_id, uuid, IJC = incl_count, SJC = skip_count)],
            sub("\\.psi$", ".jc.csv.gz", suppa_file)
          )
        }
      )

      ### PCA ----
      # long_long_dt[average_psi_dt, on = .NATURAL, average_psi := average_psi]
      # long_long_dt[, imputed_value := value]
      # long_long_dt[is.na(imputed_value) &
      #   variable == "IncLevel1", imputed_value := average_psi]
      # for (keep_ids_name in c("orig", "min_sample")) {
      #   keep_ids <- get(paste0("keep_ids_", keep_ids_name))
      #   for (col in c("value", "imputed_value")) {
      #     event_mat <-
      #       as.matrix(
      #         dcast(long_long_dt[ID %in% keep_ids &
      #           variable == "IncLevel1"], sample_name ~ ID, value.var = col),
      #         rownames = "sample_name"
      #       )

      #     for (i in 1:ncol(event_mat)) {
      #       event_mat[is.na(event_mat[, i]), i] <-
      #         mean(event_mat[, i], na.rm = TRUE)
      #     }
      #     event_pca <- rsvd::rpca(event_mat, 4)
      #     rsvd::ggscreeplot(event_pca)
      #     variance <- event_pca$sdev^2
      #     explained_variance_ratio <- round(variance / event_pca$var, 3) * 100
      #     plot_dt <- as.data.table(event_pca$x, keep.rownames = "uuid")
      #     setnames(plot_dt, c("V1", "V2"), c("PC1", "PC2"))
      #     print(rsvd::ggindplot(
      #       event_pca,
      #       groups = unlist(tstrsplit(rownames(event_mat), "_", keep = 2)),
      #       ind_labels = FALSE
      #     ) + labs(title = paste(junction_file, keep_ids_name, col)))
      #     print(rsvd::ggindplot(
      #       event_pca,
      #       pcs = c(1, 3),
      #       groups = unlist(tstrsplit(rownames(event_mat), "_", keep = 2)),
      #       ind_labels = FALSE
      #     ) + labs(title = paste(junction_file, keep_ids_name, col)))
      #   }
    }
  }
}
```
