---
title: "rMATS event filtering + VST gene-expression gate (stage 02-3)"
author: "Quirin Manz"
date: "`r Sys.Date()`"
output: html_document
---

Stage 02-3 (was `02z-rmats-analysis.Rmd`). Per-`transcript_filter` (env
`TRANSCRIPT_FILTER`). Post-processes the external rMATS output into the
high-confidence event set and writes SUPPA-format `event_{et}.psi` +
`event_{et}.jc.csv.gz` under `splicing_analysis/rmats/{tf}/` (consumed by `03`).

Two filters combine into the high-confidence set:

1. **Wang et al. 2024 (Nat Protoc) Procedure 2** — average junction count ≥ 20
   AND 5th-percentile PSI ≥ 0.05 AND 95th-percentile PSI ≤ 0.95.
2. **VST gene-expression gate (§4.12e)** — per (event, sample), keep only where
   the host gene's `gene_expression_vst` (from `02-2`) is ≥ an on/off cutoff (the
   antimode of a 2-component Gaussian mixture over all genome-wide (gene, sample)
   VST for this filter). Masked cells become NA in `.psi` and are dropped from
   `.jc`. No minimum-sample floor (removed at data construction; `09-1` keeps its
   own CV-fittability guard).

Auditability: cutoff + GMM params → `qc/vst_expression_cutoff_{tf}.csv`; gate
effect (events dropped by VST vs Procedure 2, % cells masked) →
`qc/vst_gate_effect_{tf}.csv`; density plot in `qc/distributions/`.

```{r}
set.seed(getOption("EpiATLAS_AS_SEED", 42L))
# NOTE: this file comes from the analysis at the GSC and has to be manually copied! Pipeline will fail otherwise
rmats_command_file <- "data/rmats_split_post_commands.tsv"
command_dt <- fread(rmats_command_file)
experiment_metadata <- fread("data/EpiATLAS_experiment_metadata.csv")
experiment_metadata[, experiment_type := factor(experiment_type)]
file_metadata <- fread("processed_data/file_table.csv.gz")
rna_to_use <- file_metadata[
  assay_type == "RNA-Seq" &
    epirr_id_without_version %in%
      file_metadata[,
        if (.N >= 8) .SD,
        by = epirr_id_without_version
      ][, epirr_id_without_version],
  uuid
]
# bam_dt[experiment_metadata[, epirr := epirr_id], on = .NATURAL, uuid := uuid]

# Per-filter invocation (PLAN §4.12e per-filter incremental outputs): this
# notebook post-processes exactly one transcript_filter per run. Snakemake
# passes TRANSCRIPT_FILTER as an env var; interactive runs fall back to the
# primary filter.
this_transcript_filter <- Sys.getenv(
  "TRANSCRIPT_FILTER",
  unset = getOption("EpiATLAS_AS_PRIMARY_FILTER", "biotype_filtered")
)

# Same standalone append_sanity() pattern as 02-2/03/04-* — append rows to the
# shared qc/sanity_summary.csv (initialised in 01; auto-created if absent).
# NB: parallel multi-filter runs interleave appends to this one file; per-filter
# sanity_summary_{tf}.csv merged by a master qc/00-sanity-checks.Rmd is the
# eventual fix.
sanity_csv <- "qc/sanity_summary.csv"
append_sanity <- function(check_name, status, value, threshold, notes = "") {
  dir.create("qc", showWarnings = FALSE)
  new_row <- data.table(check_name, status, value, threshold, notes)
  if (file.exists(sanity_csv)) {
    # upsert by check_name: drop any prior row for this check before writing
    # the new one, so sanity_summary.csv doesn't accumulate stale/duplicate
    # rows across reruns (append=TRUE never truncated -> the file only grew).
    existing <- fread(sanity_csv)
    existing <- existing[check_name != new_row$check_name]
    fwrite(rbind(existing, new_row, fill = TRUE), sanity_csv)
  } else {
    fwrite(new_row, sanity_csv)
  }
}
```

```{r, fig.width=12, fig.height=8}
library(UpSetR)
library(grid)
# mclust MUST be attached, not just ::-referenced: Mclust() calls mclustBIC()
# unqualified internally, so `mclust::Mclust` alone throws "could not find
# function mclustBIC" -> silent gmm->percentile fallback (was giving cutoff
# 6.081 = 0.25 quantile instead of the density antimode).
suppressPackageStartupMessages(library(mclust))

# ID format for suppa: "ENSG00000000419.12;SE:chr20:50940933-50941129:50941209-50942031:-"
# Wang et al. 2024 (Nat Protoc) Procedure 2 thresholds — getOption-driven so a
# rerun can vary them without editing this file.
average_read_coverage <- getOption("EpiATLAS_AS_AVG_READ_COVERAGE", 20)
min_average_PSI <- getOption("EpiATLAS_AS_MIN_AVG_PSI", 0.05)
max_average_PSI <- 1 - min_average_PSI
# Per-(event,sample) read-coverage floor — APPLIED in expr_pass (masks a cell's
# PSI to NA when IJC+SJC < this). PSI is a proportion with denominator IJC+SJC;
# below ~10 reads a single cell's PSI is sampling noise (binomial SE(Ψ=0.5,n=10)
# ≈0.16). 10 = Wang et al. 2024 "supported by ≥10 reads" floor (Procedure 1) and
# matches the WGBS ≥10× CpG rule. Complements the event-average ≥20 (Procedure 2).
min_read_coverage <- getOption("EpiATLAS_AS_MIN_READ_COVERAGE", 10)
min_samples_coverage <- 3 # diagnostic only
splicing_dir <- file.path("splicing_analysis", "rmats")
dir.create(splicing_dir, recursive = TRUE, showWarnings = FALSE)

# --- VST low-expression gate (PLAN §4.12e) ----------------------------------
# High-confidence event set = Procedure 2 (read coverage + PSI percentiles)
# AND a per-(event,sample) gene-expression gate: a sample's PSI is trusted
# only if its host gene is expressed above an on/off cutoff. The cutoff is the
# antimode (valley) of a 2-component Gaussian mixture fit on the WHOLE
# genome-wide (gene,sample) VST distribution for this filter — an absolute
# expressed/not boundary, not one drawn from the pre-selected event genes.
gene_vst <- fread(
  sprintf(
    "processed_data/gene_expression_normalised_%s.csv.gz",
    this_transcript_filter
  ),
  select = c("gene_id", "uuid", "gene_expression_vst")
)
gene_vst[, gene_id_stripped := sub("\\.\\d+$", "", gene_id)]
# collapse any version-strip collisions (one value per (gene,sample))
gene_vst <- unique(gene_vst, by = c("gene_id_stripped", "uuid"))

compute_vst_cutoff <- function(
  vst_values,
  method = getOption("EpiATLAS_AS_VST_CUTOFF_METHOD", "gmm"),
  fixed_value = getOption("EpiATLAS_AS_VST_CUTOFF_VALUE", NA_real_),
  percentile = getOption("EpiATLAS_AS_VST_CUTOFF_PERCENTILE", 0.25),
  floor_eps = getOption("EpiATLAS_AS_VST_FLOOR_EPS", 0)
) {
  vst_values <- vst_values[is.finite(vst_values)]
  pctl <- function() as.numeric(stats::quantile(vst_values, percentile))
  if (identical(method, "fixed")) {
    if (is.na(fixed_value)) {
      stop("VST cutoff method 'fixed' but EpiATLAS_AS_VST_CUTOFF_VALUE unset")
    }
    return(list(cutoff = fixed_value, method = "fixed"))
  }
  if (identical(method, "percentile")) {
    return(list(cutoff = pctl(), method = "percentile"))
  }
  # gmm (default): 2-component 1-D Gaussian mixture, cutoff = density antimode.
  # mclust EM over millions of (gene,sample) points is slow and can fail/OOM,
  # which silently forces the fallback — fit on a representative subsample
  # instead (seed set at notebook top → reproducible).
  #
  # ZERO-INFLATION (hurdle model). ~20% of (gene,sample) VST values are an EXACT
  # point mass at the distribution minimum (~5.75): structural zeros, i.e.
  # unexpressed genes clamped to the VST floor. This is a delta atom, not a
  # Gaussian. Model as a hurdle: pi0*delta(floor) + (1-pi0)*GMM2. The atom is
  # 'off' a priori, so fit the 2-component Gaussian ONLY on the continuous part
  # above the floor; the antimode of that mixture is the on/off boundary. The
  # cutoff still applies to all cells afterwards (floor values are below it ->
  # masked 'off').
  #   Leaving the atom IN the fit inflates the off component's variance (sd ~1.4)
  #   so it overlaps the on mode and pushes the antimode to ~9.0 (wrong).
  #   Removing just the atom (floor_eps = 0, strict '>') gives sd ~0.3 and an
  #   antimode ~7.15 in the true valley.
  #   floor_eps > 0 trims MORE than the atom (into the real low-expression bump)
  #   and drags the antimode up monotonically (0/0.5/1.0 -> 7.15/7.64/8.31), so
  #   keep it at 0 unless you deliberately want a stricter 'expressed' call.
  fit_pool <- vst_values[vst_values > min(vst_values) + floor_eps]
  n_fit <- getOption("EpiATLAS_AS_VST_GMM_MAX_N", 1e5L)
  fit_values <- if (length(fit_pool) > n_fit) {
    sample(fit_pool, n_fit)
  } else {
    fit_pool
  }
  fit_err <- NULL
  # Auto-select between equal ("E") and unequal ("V") variance. "V" alone
  # returns BIC = NA on this VST distribution (degenerate variance estimate) ->
  # empty fit -> fallback; "E" fits cleanly (2 well-separated modes). Letting
  # mclust pick by BIC is robust to either shape.
  fit <- tryCatch(
    mclust::Mclust(
      fit_values,
      G = 2,
      modelNames = c("E", "V"),
      verbose = FALSE
    ),
    error = function(e) {
      fit_err <<- conditionMessage(e)
      NULL
    }
  )
  degenerate <- is.null(fit) ||
    is.null(fit$parameters) ||
    length(unique(fit$parameters$mean)) < 2L
  if (degenerate) {
    reason <- if (!is.null(fit_err)) {
      sprintf("mclust error: %s", fit_err)
    } else if (is.null(fit)) {
      "mclust returned NULL"
    } else {
      "fit not 2-component (means collapsed)"
    }
    message(sprintf("VST GMM cutoff fell back (%s)", reason))
    if (!is.na(fixed_value)) {
      return(list(cutoff = fixed_value, method = "gmm->fixed", reason = reason))
    }
    return(list(cutoff = pctl(), method = "gmm->percentile", reason = reason))
  }
  mus <- fit$parameters$mean
  sds <- sqrt(fit$parameters$variance$sigmasq)
  if (length(sds) == 1L) {
    sds <- rep(sds, 2L)
  }
  pro <- fit$parameters$pro
  o <- order(mus)
  mus <- mus[o]
  sds <- sds[o]
  pro <- pro[o]
  grid <- seq(mus[1], mus[2], length.out = 2000L)
  dens <- pro[1] *
    dnorm(grid, mus[1], sds[1]) +
    pro[2] * dnorm(grid, mus[2], sds[2])
  # return the fitted 2-component params so the antimode placement (cutoff
  # between the two modes) is auditable straight from qc/vst_expression_cutoff_*
  list(
    cutoff = grid[which.min(dens)],
    method = "gmm",
    mu_low = mus[1],
    mu_high = mus[2],
    sd_low = sds[1],
    sd_high = sds[2],
    pro_low = pro[1],
    pro_high = pro[2],
    bic = tryCatch(as.numeric(fit$bic), error = function(e) NA_real_)
  )
}

vst_cut <- compute_vst_cutoff(gene_vst$gene_expression_vst)
vst_cutoff <- vst_cut$cutoff
vst_na_policy <- getOption("EpiATLAS_AS_VST_NA_POLICY", "keep")
dir.create("qc", showWarnings = FALSE)
dir.create("qc/distributions", recursive = TRUE, showWarnings = FALSE)
# NULL-safe getter: fallback methods (fixed/percentile) return no GMM params
.nz <- function(x) if (is.null(x)) NA_real_ else as.numeric(x)
fwrite(
  data.table(
    transcript_filter = this_transcript_filter,
    method = vst_cut$method,
    cutoff = vst_cutoff,
    na_policy = vst_na_policy,
    mu_low = .nz(vst_cut$mu_low),
    mu_high = .nz(vst_cut$mu_high),
    sd_low = .nz(vst_cut$sd_low),
    sd_high = .nz(vst_cut$sd_high),
    pro_low = .nz(vst_cut$pro_low),
    pro_high = .nz(vst_cut$pro_high),
    bic = .nz(vst_cut$bic)
  ),
  sprintf("qc/vst_expression_cutoff_%s.csv", this_transcript_filter)
)
if (vst_cut$method != "gmm") {
  warning(sprintf(
    "VST cutoff for %s used fallback method '%s' (cutoff = %.3f)%s",
    this_transcript_filter,
    vst_cut$method,
    vst_cutoff,
    if (!is.null(vst_cut$reason)) {
      sprintf(" - reason: %s", vst_cut$reason)
    } else {
      ""
    }
  ))
}
# fresh gate-effect log for this run (appended once per event type below)
unlink(sprintf("qc/vst_gate_effect_%s.csv", this_transcript_filter))
p_vst_cut <- ggplot(
  gene_vst[is.finite(gene_expression_vst)],
  aes(x = gene_expression_vst)
) +
  # density scale so the fitted GMM mixture (also a density) overlays directly
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 100,
    fill = "grey80",
    colour = NA
  ) +
  geom_vline(xintercept = vst_cutoff, colour = "red", linetype = "dashed") +
  labs(
    title = sprintf(
      "Genome-wide VST distribution (%s) - cutoff %.3f (%s)",
      this_transcript_filter,
      vst_cutoff,
      vst_cut$method
    ),
    subtitle = "grey = observed; black = GMM mixture;\nblue/orange = off/on components; red = antimode cutoff",
    x = "gene_expression_vst",
    y = "density (gene x sample)"
  )
# Overlay the fitted 2-component GMM (only present when method == 'gmm'; the
# antimode cutoff sits in the valley between the off/on components). Makes the
# on/off boundary visually auditable straight from the report.
if (identical(vst_cut$method, "gmm")) {
  .r <- range(gene_vst$gene_expression_vst[is.finite(
    gene_vst$gene_expression_vst
  )])
  gmm_grid <- data.table(x = seq(.r[1], .r[2], length.out = 500L))
  gmm_grid[,
    comp_off := vst_cut$pro_low * dnorm(x, vst_cut$mu_low, vst_cut$sd_low)
  ]
  gmm_grid[,
    comp_on := vst_cut$pro_high * dnorm(x, vst_cut$mu_high, vst_cut$sd_high)
  ]
  gmm_grid[, mix := comp_off + comp_on]
  p_vst_cut <- p_vst_cut +
    geom_line(
      data = gmm_grid,
      aes(x, comp_off),
      colour = "steelblue",
      linetype = "22"
    ) +
    geom_line(
      data = gmm_grid,
      aes(x, comp_on),
      colour = "darkorange",
      linetype = "22"
    ) +
    geom_line(data = gmm_grid, aes(x, mix), colour = "black", linewidth = 0.8)
}
print(p_vst_cut)
ggsave(
  sprintf(
    "qc/distributions/vst_expression_cutoff_%s.pdf",
    this_transcript_filter
  ),
  p_vst_cut,
  width = 7,
  height = 5
)

event_type <- "SE"
junction_counts <- "MATS.JC.txt"
run_name <- "paired_fr-firststrand_75"
# Diagnostic plots (avg-count-vs-PSI scatter, ECDF/histograms, per-sample and
# per-average threshold-sweep grids) are ON by default. They are the bulk of
# this notebook's runtime and produce nothing the pipeline consumes, so set
# EpiATLAS_AS_RMATS_DIAGNOSTICS=FALSE to skip them for a faster production run.
plot_dist <- isTRUE(getOption("EpiATLAS_AS_RMATS_DIAGNOSTICS", TRUE))

# restrict to the single requested transcript_filter (per-filter incremental)
rmats_dirs <- list.dirs(file.path(data_dir, "rmats"), recursive = FALSE)
rmats_dirs <- rmats_dirs[basename(rmats_dirs) == this_transcript_filter]
stopifnot(length(rmats_dirs) == 1L)
rmats_dir <- rmats_dirs

# Only the event types consumed downstream (03 reads `to_analyze` = SE, RI).
# A3SS/A5SS/MXE were computed here but never read by 03 — dropped (~60% less
# work). The suppa_id branches for them below are kept for reference/reuse.
for (event_type in to_analyze) {
  print(event_type)
  for (rmats_dir in rmats_dirs) {
    print(basename(rmats_dir))
    this_splicing_dir <- file.path(splicing_dir, basename(rmats_dir))
    dir.create(this_splicing_dir, recursive = TRUE, showWarnings = FALSE)
    for (junction_counts in c(
      "MATS.JC.txt" # , "MATS.JCEC.txt"
    )) {
      # read annotation file
      event_annotation <- fread(file.path(
        rmats_dir,
        command_dt[grep(basename(rmats_dir), out_folder), run_name][1],
        paste("fromGTF", event_type, "txt", sep = ".")
      ))
      if (event_type == "SE") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            upstreamEE,
            exonStart_0base + 1,
            exonEnd,
            downstreamES + 1,
            strand
          )
        ]
      } else if (event_type == "RI") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s:%s-%s:%s:%s",
            GeneID,
            event_type,
            chr,
            upstreamES + 1,
            upstreamEE,
            downstreamES + 1,
            downstreamEE,
            strand
          )
        ]
      } else if (event_type == "A3SS") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            ifelse(strand == "+", flankingEE, longExonEnd),
            ifelse(strand == "+", longExonStart_0base + 1, flankingES + 1),
            ifelse(strand == "+", flankingEE, shortEE),
            ifelse(strand == "+", shortES + 1, flankingES + 1),
            strand
          )
        ]
      } else if (event_type == "A5SS") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            ifelse(strand == "+", longExonEnd, flankingEE),
            ifelse(strand == "+", flankingES + 1, longExonStart_0base + 1),
            ifelse(strand == "+", shortEE, flankingEE),
            ifelse(strand == "+", flankingES + 1, shortES + 1),
            strand
          )
        ]
      } else if (event_type == "MXE") {
        event_annotation[,
          suppa_id := sprintf(
            "%s;%s:%s:%s-%s:%s-%s:%s-%s:%s-%s:%s",
            GeneID,
            event_type,
            chr,
            upstreamEE,
            `1stExonStart_0base` + 1,
            `1stExonEnd`,
            downstreamES + 1,
            upstreamEE,
            `2ndExonStart_0base` + 1,
            `2ndExonEnd`,
            downstreamES + 1,
            strand
          )
        ]
      } else {
        stop("Not implemented yet")
      }

      event_sample_data_path <- file.path(
        rmats_dir,
        paste(event_type, junction_counts, "csv.gz", sep = ".")
      )
      if (file.exists(event_sample_data_path)) {
        event_sample_data <- fread(file.path(
          rmats_dir,
          paste(event_type, junction_counts, "csv.gz", sep = ".")
        ))
      } else {
        id_vars <-
          c(
            "ID"
          )
        measure_vars <-
          c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IncLevel1")
        # MEMORY: cap fork count for the JC read. Each fork melts one run's
        # cohort-wide chunk (n_events x cohort_samples x 3 measure_vars) and
        # mclapply serialises each result back to the parent through a pipe
        # (~2x transient per in-flight result). At mc.cores=8 that peak (SE:
        # ~150k events x 441 samples) OOM-killed a child at 48G. Fewer forks =
        # lower simultaneous peak; the per-run files are I/O-light so the
        # throughput hit is small. 4 (with the 64G bump) keeps some overlap;
        # override via EpiATLAS_AS_RMATS_JC_CORES.
        dt_list <- pbmclapply(
          command_dt[grep(basename(rmats_dir), out_folder), run_name],
          mc.cores = getOption("EpiATLAS_AS_RMATS_JC_CORES", 4L),
          function(this_run_name) {
            junction_file <- file.path(
              rmats_dir,
              this_run_name,
              paste(event_type, junction_counts, sep = ".")
            )
            # read file with counts
            dt <-
              fread(junction_file, select = c(id_vars, measure_vars))
            # convert to long format

            long_dt <-
              melt(
                dt,
                id.vars = id_vars,
                measure.vars = measure_vars
              )
            # sample names, in the order they appear in the rMATS input file.
            uuids <- command_dt[
              grep(basename(rmats_dir), out_folder) & run_name == this_run_name,
              strsplit(uuid_string, ",", fixed = TRUE)
            ][[1]]
            # MEMORY: split out ONLY the analysis-cohort (rna_to_use) columns via
            # tstrsplit(keep=). rMATS quantifies far more samples per run; the old
            # unconditional tstrsplit materialised a column for EVERY sample (a
            # ~n_events x n_all_samples wide table) x 8 concurrent forks — the
            # actual OOM. `keep=` splits only the needed positions, so the wide
            # table (and the melt) is cohort-sized from the start. The sole
            # written output is the full_epigenomes / rna_to_use set anyway.
            keep_idx <- which(uuids %in% rna_to_use)
            # a run with no cohort samples contributes nothing — skip it
            # (tstrsplit/melt over zero positions would otherwise error).
            if (length(keep_idx) == 0L) {
              return(NULL)
            }
            keep_uuids <- uuids[keep_idx]
            long_dt[,
              (keep_uuids) := tstrsplit(
                value,
                ",",
                fixed = TRUE,
                keep = keep_idx,
                type.convert = TRUE
              )
            ]
            # convert to even longer format to have one row per (cohort) sample
            lapply(
              split(long_dt, long_dt[, variable]),
              melt,
              id.vars = c(id_vars, "variable"),
              measure.vars = keep_uuids,
              variable.name = "uuid"
            )
          }
        )
        names(dt_list) <- command_dt[
          grep(basename(rmats_dir), out_folder),
          run_name
        ]
        # drop runs that had no cohort samples; require at least one remains
        dt_list <- dt_list[!vapply(dt_list, is.null, logical(1))]
        stopifnot(length(dt_list) > 0L)
        long_dt_list <- sapply(
          measure_vars,
          function(measure_var) {
            rbindlist(
              sapply(
                dt_list,
                function(dt_split) {
                  dt_split[[measure_var]]
                },
                simplify = FALSE
              ),
              idcol = "run_name"
            )
          },
          simplify = FALSE
        )
        # dt_list (per-run 3-way split, all runs) is fully consumed into
        # long_dt_list now — free it before the next allocation. Peak here is
        # the triple copy dt_list + long_dt_list + event_sample_data held at
        # once (SE: ~66M rows x3); dropping each as it's consumed is the real
        # lever (fork-count tuning didn't move MaxRSS).
        rm(dt_list)
        gc(verbose = FALSE)

        stopifnot(identical(
          long_dt_list$IJC_SAMPLE_1[, .(ID, uuid)],
          long_dt_list$SJC_SAMPLE_1[, .(ID, uuid)]
        ))

        stopifnot(identical(
          long_dt_list$SJC_SAMPLE_1[, .(ID, uuid)],
          long_dt_list$IncLevel1[, .(ID, uuid)]
        ))

        event_sample_data <- long_dt_list$IJC_SAMPLE_1[, .(
          ID,
          uuid,
          incl_count = value
        )]
        event_sample_data[, skip_count := long_dt_list$SJC_SAMPLE_1[, value]]
        event_sample_data[, psi := long_dt_list$IncLevel1[, value]]
        # long_dt_list (3 full copies) consumed into event_sample_data — free.
        rm(long_dt_list)
        gc(verbose = FALSE)

        fwrite(event_sample_data, event_sample_data_path)
      }

      # Restrict to the analysis cohort (covers a stale all-sample cache; fresh
      # builds are already cohort-only from the melt). All downstream work —
      # averages, VST gate, writers — is cohort-only.
      event_sample_data <- event_sample_data[uuid %in% rna_to_use]

      # Cohort-coverage sanity (new, because we now hard-scope to rna_to_use):
      # non-empty guards an empty rna_to_use x rMATS-uuid intersection (a wiring
      # bug → cryptic downstream crash otherwise); the warning flags cohort
      # samples rMATS never quantified — they silently drop, leaving fewer
      # samples per event for Procedure 2 and the VST gate.
      # EXPECTED GAP: 26 rna_to_use samples (all AMED-CREST mRNA-Seq) are absent
      # from rMATS output — their data-access agreement expired, so no BAM files
      # were available to run rMATS on. Coverage settles at 415/441; this WARN
      # is informational, not a pipeline fault.
      stopifnot(nrow(event_sample_data) > 0L)
      n_cohort_present <- uniqueN(event_sample_data$uuid)
      if (n_cohort_present < length(rna_to_use)) {
        warning(sprintf(
          "%s %s: %d/%d rna_to_use samples present in rMATS output (%d missing)",
          event_type,
          basename(rmats_dir),
          n_cohort_present,
          length(rna_to_use),
          length(rna_to_use) - n_cohort_present
        ))
      }
      append_sanity(
        sprintf(
          "rmats_cohort_coverage_%s__%s",
          event_type,
          this_transcript_filter
        ),
        if (n_cohort_present == length(rna_to_use)) "PASS" else "WARN",
        n_cohort_present,
        length(rna_to_use),
        "rna_to_use samples present in rMATS output"
      )

      # # add experiment type
      event_sample_data[
        experiment_metadata,
        on = .(uuid),
        experiment_type := experiment_type
      ]

      # compute average counts per event and per experiment type and event
      # recompute for the RNA used in this analysis as well
      # Only the epigenome cohort is written now: it is the set VST was
      # computed on and the only output 03 consumes (the old diagnostic "all"
      # variant is dropped — PLAN §4.12e).
      uuids_to_filter_on <- list(
        full_epigenomes = rna_to_use
      )
      mapply(
        as.list(names(uuids_to_filter_on)),
        uuids_to_filter_on,
        FUN = function(filter_name, uuids_to_filter) {
          print(filter_name)
          filtered_dt <- event_sample_data[uuid %in% uuids_to_filter]

          # --- VST low-expression gate (per (event,sample)) ---------------
          # Attach host gene_id (from rMATS GeneID, version-stripped to match
          # the normalisation table) and its VST; a sample passes only if its
          # host gene is expressed at/above the cutoff. NA VST (host gene
          # absent from normalisation) is handled per EpiATLAS_AS_VST_NA_POLICY.
          id_gene <- unique(event_annotation[, .(
            ID,
            gene_id_stripped = sub("\\.\\d+$", "", GeneID)
          )])
          filtered_dt[
            id_gene,
            on = .(ID),
            gene_id_stripped := i.gene_id_stripped
          ]
          filtered_dt[
            gene_vst,
            on = .(gene_id_stripped, uuid),
            vst := i.gene_expression_vst
          ]
          # expr_pass = VST host-gene gate AND per-cell read-coverage floor.
          # VST honours the NA policy (keep|drop); the read-coverage gate is
          # ALWAYS applied (a cell with < min_read_coverage junction reads has an
          # unreliable PSI regardless of host-gene expression). NA counts → fail.
          # Persist both component flags so the gate-effect report can decompose.
          filtered_dt[, read_cov_pass :=
            !is.na(incl_count + skip_count) &
              (incl_count + skip_count) >= min_read_coverage]
          if (identical(vst_na_policy, "drop")) {
            filtered_dt[, vst_pass := !is.na(vst) & vst >= vst_cutoff]
          } else {
            filtered_dt[, vst_pass := is.na(vst) | vst >= vst_cutoff]
          }
          filtered_dt[, expr_pass := vst_pass & read_cov_pass]
          # report events lost SOLELY to missing VST (every sample NA), so the
          # loss is visible rather than silent
          na_only_events <- filtered_dt[,
            .(all_na = all(is.na(vst))),
            by = ID
          ][all_na == TRUE, .N]
          message(sprintf(
            "%s %s: VST cutoff %.3f (%s); %d events have no VST for any sample (na_policy=%s)",
            event_type,
            basename(rmats_dir),
            vst_cutoff,
            vst_cut$method,
            na_only_events,
            vst_na_policy
          ))
          average_dt <- rbind(
            filtered_dt[,
              .(
                incl_sum = sum(incl_count, na.rm = TRUE),
                skip_sum = sum(skip_count, na.rm = TRUE),
                average_psi = mean(psi, na.rm = TRUE),
                n_samples = .N
              ),
              by = ID
            ],
            filtered_dt[,
              .(
                incl_sum = sum(incl_count, na.rm = TRUE),
                skip_sum = sum(skip_count, na.rm = TRUE),
                average_psi = mean(psi, na.rm = TRUE),
                n_samples = .N
              ),
              by = .(ID, experiment_type)
            ],
            fill = TRUE
          )
          average_dt[, average_count := (incl_sum + skip_sum) / n_samples]

          # scatterplot average count vs average PSI
          ggplot(
            average_dt,
            aes(x = average_count + 1, y = average_psi, color = experiment_type)
          ) +
            geom_point(alpha = .5) +
            scale_x_log10() +
            # create a line y ~ x
            geom_smooth(method = "lm", se = FALSE) +
            ggpubr::stat_cor(
              method = "spearman",
              label.x.npc = "middle",
              label.y.npc = "bottom"
            ) +
            facet_wrap(~experiment_type) +
            labs(title = sprintf("Average Count vs Average PSI:"))

          # plot average count and PSI distributions
          if (plot_dist) {
            print(
              ggplot(
                average_dt,
                aes(x = average_count + 1, color = experiment_type)
              ) +
                geom_line(aes(y = 1 - after_stat(y)), stat = "ecdf") +
                scale_x_log10() +
                geom_vline(xintercept = average_read_coverage + 1) +
                labs(
                  title = sprintf(
                    "Average Read Coverage: with line at %d; %s %s %s",
                    average_read_coverage + 1,
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  )
                )
            )
            print(
              ggplot(average_dt, aes(x = average_psi, fill = experiment_type)) +
                geom_histogram(position = "dodge") +
                geom_vline(xintercept = c(min_average_PSI, max_average_PSI)) +
                labs(
                  title = sprintf(
                    "Average PSI: %s %s %s",
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  )
                )
            )
            print(
              ggplot(
                average_dt[
                  average_psi >= min_average_PSI &
                    average_psi <= max_average_PSI
                ],
                aes(x = average_psi, fill = experiment_type)
              ) +
                geom_histogram(position = "dodge", binwidth = .05) +
                geom_vline(xintercept = c(min_average_PSI, max_average_PSI)) +
                labs(
                  title = sprintf(
                    "Average PSI: %f <= PSI < %f; %s %s %s",
                    min_average_PSI,
                    max_average_PSI,
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  )
                )
            )

            ### Check per-sample read coverage filtering ----
            min_read_cutoffs <- c(0, 1, 5, 10, 15, 20)
            min_sample_cutoffs <- c(1, 3, 5, 10)

            # Pre-compute per-threshold counts once (6 passes) rather than 24 (6x4 grid)
            count_by_thresh <- setNames(
              lapply(min_read_cutoffs, function(thresh) {
                filtered_dt[incl_count + skip_count >= thresh, .N, by = ID]
              }),
              as.character(min_read_cutoffs)
            )

            plot_dt <- CJ(
              min_read_coverage = min_read_cutoffs,
              min_samples_coverage = min_sample_cutoffs
            )

            plot_dt[,
              n_pass := count_by_thresh[[as.character(min_read_coverage)]][
                N >= min_samples_coverage,
                .N
              ],
              by = .(min_read_coverage, min_samples_coverage)
            ]

            print(
              ggplot(
                plot_dt,
                aes(
                  x = min_samples_coverage,
                  y = n_pass,
                  color = factor(min_read_coverage)
                )
              ) +
                geom_line() +
                geom_point() +
                labs(
                  title = sprintf(
                    "%s %s %s",
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  ),
                  subtitle = sprintf(
                    "APPLIED per-cell gate: IJC+SJC >= %d (no min-samples floor); this sweep is exploratory",
                    min_read_coverage
                  ),
                  x = "Minimum number of samples with sufficient read count",
                  y = "Number of events (ID) retained",
                  color = "Min per-sample read count"
                )
            )

            ###  Check average read coverages and PSI filtering ----

            # Step 1: Create grid of thresholds to test
            min_PSI_cutoffs <- seq(0, 0.05, by = 0.01)

            plot_dt <- CJ(
              average_read_coverage = min_read_cutoffs,
              min_average_PSI = min_PSI_cutoffs
            )

            plot_dt[,
              n_pass := average_dt[
                !is.na(experiment_type) &
                  average_count >= average_read_coverage &
                  average_psi >= min_average_PSI &
                  average_psi <= 1 - min_average_PSI,
                ID,
                by = experiment_type
              ][duplicated(ID), .N],
              by = .(average_read_coverage, min_average_PSI)
            ]

            print(
              ggplot(
                plot_dt,
                aes(
                  x = min_average_PSI,
                  y = n_pass,
                  color = factor(average_read_coverage)
                )
              ) +
                geom_line() +
                geom_point() +
                labs(
                  title = sprintf(
                    "Min averages per experiment type: %s %s %s",
                    event_type,
                    basename(rmats_dir),
                    filter_name
                  ),
                  x = "Minimum average PSI",
                  y = "Number of events (ID) retained",
                  color = "Min average read coverage"
                )
            )
          }
          # get IDs to keep based on average PSI and average read coverage
          keep_ids_orig <-
            average_dt[
              !is.na(experiment_type) &
                average_count >= average_read_coverage &
                average_psi >= min_average_PSI &
                average_psi <= 1 - min_average_PSI,
              ID,
              by = experiment_type
            ][duplicated(ID), ID]
          length(keep_ids_orig)
          # number of unique IDs passing the filter across all experiment types
          average_dt[
            is.na(experiment_type) &
              average_count >= average_read_coverage &
              average_psi >= min_average_PSI &
              average_psi <= 1 - min_average_PSI,
            uniqueN(ID)
          ]

          # intersect(
          #   average_psi_dt[!is.na(experiment_type) &
          #     average_psi >= min_average_PSI &
          #     average_psi <= max_average_PSI, ID, by = experiment_type][duplicated(ID), ID],
          #   average_count_dt[!is.na(experiment_type) &
          #     average_count >= average_read_coverage, ID, by = experiment_type][duplicated(ID), ID]
          # )

          keep_ids_min_sample <- filtered_dt[
            incl_count + skip_count >= min_read_coverage,
            .N,
            by = ID
          ][N >= min_samples_coverage, ID]

          # Wang et al. 2024 (Nature Protocols) Procedure 2 filter, applied to
          # the VST-passing samples only: average count >= 20 + 5th/95th PSI
          # percentiles. Removes constitutively included (psi_q05 >
          # max_average_PSI) and excluded (psi_q95 < min_average_PSI) events.
          # NB: computed on expr_pass samples, so percentiles/coverage reflect
          # the high-confidence subset. No minimum-sample floor (PLAN: removed
          # at data-construction; 09-1 keeps its CV-fittability guard).
          #
          # Percentile direction matches the Xinglab rmats-turbo-tutorial CCLE
          # implementation (plot_Fig_CCLE.ipynb): `nanquantile(PSI,0.05)<=maxPSI
          # AND nanquantile(PSI,0.95)>=minPSI`, i.e. q95>=0.05 & q05<=0.95 ->
          # keep everything NOT constitutive (incl. switch-like 0<->1 events).
          # This is the OPPOSITE of the paper's prose ("5% pctl >=0.05 and 95%
          # pctl <=0.95", which would drop switch-like events). The code is
          # authoritative; the prose is imprecise. Do not "fix" to the prose.
          proc2_ids <- function(dt) {
            tmp <- dt[!is.na(psi)]
            setorder(tmp, ID, psi)
            res <- tmp[,
              .(
                psi_q05 = psi[max(1L, ceiling(0.05 * .N))],
                psi_q95 = psi[min(.N, ceiling(0.95 * .N))],
                average_count = mean(incl_count + skip_count, na.rm = TRUE)
              ),
              by = ID
            ][
              average_count >= average_read_coverage &
                psi_q95 >= min_average_PSI &
                psi_q05 <= max_average_PSI,
              ID
            ]
            rm(tmp)
            return(res)
          }
          # applied filter: Procedure 2 on VST-passing samples
          keep_ids_percentile <- proc2_ids(filtered_dt[expr_pass == TRUE])
          # counterfactual: Procedure 2 ignoring the VST gate — isolates what
          # the VST gate removes vs what Procedure 2 removes on its own
          keep_ids_percentile_novst <- proc2_ids(filtered_dt)

          if (plot_dist) {
            ttl <- sprintf(
              "%s %s %s",
              event_type,
              basename(rmats_dir),
              filter_name
            )

            # --- UpSet 1: EVENT-level filters -> "how many EVENTS survive" ---
            # Sets of event IDs, each an independent Procedure-2 criterion
            # (computed on all samples, pre-VST, so each filter's own effect is
            # visible; the VST event-loss is quantified separately in the
            # gate-effect summary below).
            #   avg_count       : avg (incl+skip) >= average_read_coverage
            #   psi_percentile  : Wang 5th/95th-percentile PSI test (same logic
            #                     as the applied proc2_ids: not constitutive)
            #   min_sample_read : >= min_samples_coverage samples with
            #                     incl+skip >= min_read_coverage (keep_ids_min_sample)
            ev_avg_count <- filtered_dt[,
              .(ac = mean(incl_count + skip_count, na.rm = TRUE)),
              by = ID
            ][ac >= average_read_coverage, ID]
            ev_psi_pctl <- {
              tmp <- filtered_dt[!is.na(psi)]
              setorder(tmp, ID, psi)
              tmp[,
                .(
                  q05 = psi[max(1L, ceiling(0.05 * .N))],
                  q95 = psi[min(.N, ceiling(0.95 * .N))]
                ),
                by = ID
              ][q95 >= min_average_PSI & q05 <= max_average_PSI, ID]
            }
            upset_events <- upset(
              fromList(list(
                avg_count = ev_avg_count,
                psi_percentile = ev_psi_pctl,
                min_sample_read = keep_ids_min_sample
              )),
              nsets = 3,
              order.by = "freq",
              mainbar.y.label = "Number of events (ID)",
              sets.x.label = "Total events (ID)"
            )
            print(upset_events)
            grid.text(
              paste("EVENT-level (events surviving):", ttl),
              x = 0.65,
              y = 0.95,
              gp = gpar(fontsize = 14)
            )

            # --- UpSet 2: (EVENT,SAMPLE)-level -> "how many ROWS survive" ----
            # Sets of (event,sample) cells, the two per-cell masks:
            #   read_cov : incl+skip >= min_read_coverage in this cell
            #   vst      : host-gene VST gate passed in this cell (vst_pass)
            # The intersection read_cov & vst == expr_pass (the applied gate); the
            # single-set regions show what each gate removes on its own. Use the
            # standalone vst_pass column (NOT expr_pass, which already ANDs both).
            # Cell count is ~n_events x n_samples (millions) -> fromList's
            # membership matrix would blow up; count the 2^2 disjoint regions
            # and feed fromExpression instead (exact intersection sizes).
            cell_flags <- filtered_dt[, .(
              read_cov = read_cov_pass == TRUE,
              vst = vst_pass == TRUE
            )]
            combo <- cell_flags[, .N, by = .(read_cov, vst)]
            .combo_name <- function(r, v) {
              paste(c(if (r) "read_cov", if (v) "vst"), collapse = "&")
            }
            expr_vec <- combo[, setNames(
              N,
              mapply(.combo_name, read_cov, vst)
            )]
            expr_vec <- expr_vec[names(expr_vec) != ""] # drop none-of-them
            upset_cells <- upset(
              fromExpression(as.list(expr_vec)),
              nsets = 2,
              order.by = "freq",
              mainbar.y.label = "Number of (event,sample) rows",
              sets.x.label = "Total rows"
            )
            print(upset_cells)
            grid.text(
              paste("(EVENT,SAMPLE)-level (rows surviving):", ttl),
              x = 0.65,
              y = 0.95,
              gp = gpar(fontsize = 14)
            )
          }

          # --- gate-effect summary (auditable) -----------------------------
          # Decompose event loss: Procedure 2 alone vs the added cell-level gates
          # (VST host-gene expression + per-cell read coverage, folded into expr_pass).
          #   dropped_proc2 : fail Procedure 2 even without the gates
          #   dropped_gates : pass Procedure 2 on full samples, lost after gating
          #   kept          : survive both (= keep_ids_percentile)
          n_total <- filtered_dt[!is.na(psi), uniqueN(ID)]
          n_kept <- length(keep_ids_percentile)
          n_pass_novst <- length(keep_ids_percentile_novst)
          n_dropped_proc2 <- n_total - n_pass_novst
          n_dropped_vst <- length(setdiff(
            keep_ids_percentile_novst,
            keep_ids_percentile
          ))
          # Cell-masking, decomposed over cells that HAVE a PSI value:
          #   total   = fail either gate (= !expr_pass)
          #   vst     = fail the VST gate
          #   lowcov  = fail the read-coverage gate (IJC+SJC < min_read_coverage)
          # (vst + lowcov overlap where a cell fails both, so they need not sum to total)
          frac_cells_masked <- filtered_dt[!is.na(psi), mean(!expr_pass)]
          frac_cells_vst_masked <- filtered_dt[!is.na(psi), mean(!vst_pass)]
          frac_cells_lowcov_masked <- filtered_dt[!is.na(psi), mean(!read_cov_pass)]
          gate_effect <- data.table(
            transcript_filter = this_transcript_filter,
            event_type = event_type,
            n_events_total = n_total,
            n_events_kept = n_kept,
            n_dropped_proc2_only = n_dropped_proc2,
            n_dropped_gates = n_dropped_vst,
            frac_cells_masked_total = frac_cells_masked,
            frac_cells_vst_masked = frac_cells_vst_masked,
            frac_cells_lowcov_masked = frac_cells_lowcov_masked,
            vst_cutoff = vst_cutoff,
            min_read_coverage = min_read_coverage,
            na_policy = vst_na_policy
          )
          print(gate_effect)
          gate_effect_file <- sprintf(
            "qc/vst_gate_effect_%s.csv",
            this_transcript_filter
          )
          fwrite(
            gate_effect,
            gate_effect_file,
            append = file.exists(gate_effect_file)
          )
          append_sanity(
            sprintf(
              "expr_gate_frac_cells_masked_%s__%s",
              event_type,
              this_transcript_filter
            ),
            "INFO",
            frac_cells_masked,
            NA_real_,
            "fraction of (event,sample) cells masked by expr_pass (VST OR low read coverage)"
          )
          append_sanity(
            sprintf(
              "lowcov_gate_frac_cells_masked_%s__%s",
              event_type,
              this_transcript_filter
            ),
            "INFO",
            frac_cells_lowcov_masked,
            NA_real_,
            sprintf("fraction of (event,sample) cells with IJC+SJC < %d (unreliable PSI)", min_read_coverage)
          )

          # Emit only cells that pass BOTH the Procedure 2 event filter and the
          # per-(event,sample) expr_pass gate (VST host-gene expression AND read
          # coverage). Build from filtered_dt (which carries expr_pass); masked
          # cells are simply absent, so dcast fills them NA in the .psi matrix,
          # and they are dropped from the .jc file.
          filtered_dt[event_annotation, on = .(ID), suppa_id := i.suppa_id]
          write_dt <- filtered_dt[
            ID %in% keep_ids_percentile & expr_pass == TRUE
          ]
          suppa_format_dt <- dcast(
            write_dt,
            suppa_id ~ uuid,
            value.var = "psi"
          )
          suppa_file <- file.path(
            this_splicing_dir,
            paste0(
              "event_",
              substr(event_type, 1, 2),
              ".psi"
            )
          )
          write.table(
            as.data.frame(
              suppa_format_dt[, -"suppa_id"],
              row.names = suppa_format_dt[, suppa_id]
            ),
            suppa_file,
            sep = "\t",
            quote = FALSE
          )
          fwrite(
            unique(event_annotation[, .(ID, geneSymbol, suppa_id)]),
            paste0(suppa_file, ".annot")
          )
          # Write IJC/SJC junction counts for downstream coverage weighting (§4.8/§4.12f)
          fwrite(
            write_dt[, .(suppa_id, uuid, IJC = incl_count, SJC = skip_count)],
            sub("\\.psi$", ".jc.csv.gz", suppa_file)
          )

          # --- NA-fraction diagnostics on the written PSI matrix -----------
          # The .psi output is (kept events) x (cohort samples); a cell is NA
          # whenever that (event,sample) was masked (VST gate) or had no valid
          # PSI. These NA patterns drive downstream fittability (an event with
          # few valid samples can't be modelled), so summarise them explicitly.
          # CRUCIAL: this is the shape of the data every later stage inherits.
          kept_events <- keep_ids_percentile
          all_uuids <- filtered_dt[, unique(uuid)]
          n_samp <- length(all_uuids)
          n_ev <- length(kept_events)
          # valid = a written, non-NA PSI cell
          valid <- write_dt[!is.na(psi), .(ID, uuid)]
          # per-event: number of samples with a valid PSI (0-filled for events
          # that survived the filter but ended with no valid cell)
          per_event <- merge(
            data.table(ID = kept_events),
            valid[, .(n_valid = uniqueN(uuid)), by = ID],
            by = "ID",
            all.x = TRUE
          )
          per_event[is.na(n_valid), n_valid := 0L]
          per_event[, na_frac := 1 - n_valid / n_samp]
          # per-sample: number of events with a valid PSI
          per_sample <- merge(
            data.table(uuid = all_uuids),
            valid[, .(n_valid = uniqueN(ID)), by = uuid],
            by = "uuid",
            all.x = TRUE
          )
          per_sample[is.na(n_valid), n_valid := 0L]
          per_sample[, na_frac := 1 - n_valid / n_ev]
          overall_na <- 1 - nrow(valid) / (as.numeric(n_ev) * n_samp)

          na_stats <- data.table(
            transcript_filter = this_transcript_filter,
            event_type = event_type,
            n_events = n_ev,
            n_samples = n_samp,
            overall_na_frac = overall_na,
            event_na_median = median(per_event$na_frac),
            event_na_q90 = as.numeric(quantile(per_event$na_frac, 0.9)),
            sample_na_median = median(per_sample$na_frac),
            sample_na_q90 = as.numeric(quantile(per_sample$na_frac, 0.9)),
            event_min_valid_samples = min(per_event$n_valid),
            event_median_valid_samples = median(per_event$n_valid),
            events_lt10_valid = per_event[n_valid < 10L, .N],
            events_lt20_valid = per_event[n_valid < 20L, .N]
          )
          print(na_stats)
          na_stats_file <- sprintf(
            "qc/psi_na_stats_%s.csv",
            this_transcript_filter
          )
          fwrite(na_stats, na_stats_file, append = file.exists(na_stats_file))
          append_sanity(
            sprintf(
              "psi_overall_na_frac_%s__%s",
              event_type,
              this_transcript_filter
            ),
            "INFO",
            overall_na,
            NA_real_,
            "NA fraction of written PSI matrix (kept events x cohort samples)"
          )
          append_sanity(
            sprintf(
              "psi_event_na_median_%s__%s",
              event_type,
              this_transcript_filter
            ),
            "INFO",
            median(per_event$na_frac),
            NA_real_,
            "median per-event NA fraction across samples"
          )
          append_sanity(
            sprintf(
              "psi_events_lt10_valid_%s__%s",
              event_type,
              this_transcript_filter
            ),
            "INFO",
            per_event[n_valid < 10L, .N],
            NA_real_,
            "kept events with < 10 valid-PSI samples (fittability risk)"
          )

          if (plot_dist) {
            # a cell is "valid" iff it survived BOTH the Procedure 2 event filter
            # AND the per-(event,sample) VST gate and has a non-NA written PSI.
            print(
              ggplot(per_event, aes(na_frac)) +
                geom_histogram(bins = 50) +
                labs(
                  title = sprintf(
                    "Per-EVENT NA fraction (%s %s)",
                    event_type,
                    this_transcript_filter
                  ),
                  subtitle = sprintf(
                    "per kept event: 1 - (samples with valid PSI) / %d present cohort samples; NA = VST-masked or no PSI. %d events, cutoff %.3f (%s)",
                    n_samp, n_ev, vst_cutoff, vst_cut$method
                  ),
                  x = "fraction of samples NA",
                  y = "events"
                )
            )
            print(
              ggplot(per_sample, aes(na_frac)) +
                geom_histogram(bins = 50) +
                labs(
                  title = sprintf(
                    "Per-SAMPLE NA fraction (%s %s)",
                    event_type,
                    this_transcript_filter
                  ),
                  subtitle = sprintf(
                    "per cohort sample: 1 - (kept events with valid PSI) / %d kept events; NA = VST-masked or no PSI. %d samples",
                    n_ev, n_samp
                  ),
                  x = "fraction of events NA",
                  y = "samples"
                )
            )
            print(
              ggplot(per_event, aes(n_valid)) +
                geom_histogram(bins = 50) +
                labs(
                  title = sprintf(
                    "Valid samples per EVENT (%s %s)",
                    event_type,
                    this_transcript_filter
                  ),
                  subtitle = sprintf(
                    "per kept event: # samples with a written non-NA PSI (max %d present cohort samples); drives downstream fittability",
                    n_samp
                  ),
                  x = "# samples with valid PSI",
                  y = "events"
                )
            )
          }
        }
      )

      ### PCA ----
      # long_long_dt[average_psi_dt, on = .NATURAL, average_psi := average_psi]
      # long_long_dt[, imputed_value := value]
      # long_long_dt[is.na(imputed_value) &
      #   variable == "IncLevel1", imputed_value := average_psi]
      # for (keep_ids_name in c("orig", "min_sample")) {
      #   keep_ids <- get(paste0("keep_ids_", keep_ids_name))
      #   for (col in c("value", "imputed_value")) {
      #     event_mat <-
      #       as.matrix(
      #         dcast(long_long_dt[ID %in% keep_ids &
      #           variable == "IncLevel1"], sample_name ~ ID, value.var = col),
      #         rownames = "sample_name"
      #       )

      #     for (i in 1:ncol(event_mat)) {
      #       event_mat[is.na(event_mat[, i]), i] <-
      #         mean(event_mat[, i], na.rm = TRUE)
      #     }
      #     event_pca <- rsvd::rpca(event_mat, 4)
      #     rsvd::ggscreeplot(event_pca)
      #     variance <- event_pca$sdev^2
      #     explained_variance_ratio <- round(variance / event_pca$var, 3) * 100
      #     plot_dt <- as.data.table(event_pca$x, keep.rownames = "uuid")
      #     setnames(plot_dt, c("V1", "V2"), c("PC1", "PC2"))
      #     print(rsvd::ggindplot(
      #       event_pca,
      #       groups = unlist(tstrsplit(rownames(event_mat), "_", keep = 2)),
      #       ind_labels = FALSE
      #     ) + labs(title = paste(junction_file, keep_ids_name, col)))
      #     print(rsvd::ggindplot(
      #       event_pca,
      #       pcs = c(1, 3),
      #       groups = unlist(tstrsplit(rownames(event_mat), "_", keep = 2)),
      #       ind_labels = FALSE
      #     ) + labs(title = paste(junction_file, keep_ids_name, col)))
      #   }
    }
  }
}
```
