---
title: "ML Global Comparison (splicing_ml)"
output: html_document
---

Recreates the comparison originally made in `08-2-ml-global.Rmd` (caret-based RF/LASSO/OLS on
`processed_data/global_models/*.rds`, matching the preprint's Fig 2B), but reads results from the
current `splicing_ml` Python pipeline instead. `08-2-ml-global.Rmd`'s own data dependency
(`processed_data/global_models/`) no longer exists on disk -- that notebook is a frozen historical
record of the pre-rewrite approach, not something this one edits or replaces in place.

Numbered `07` in anticipation of `splicing_ml/` eventually being renamed `07_splicing_ml/` (07 is
currently the one unused slot in the 01-10 sequence, sitting exactly where this belongs: after
`06-correlation`'s genome-wide association work, before `08-2`'s frozen old-approach comparison and
`09`'s event-local models). `-2` mirrors `08-1`/`08-2`'s fit/analyze split -- the "07-1" equivalent
is the `splicing_ml` package + its Snakemake rules themselves, not a single script.

Known differences from the original Fig 2B, all deliberate (see conversation this notebook was
written in for the full reasoning):

- No "Chr & Cell" (both held out at once) condition -- that CV mode doesn't exist in the current
  pipeline. Only the two that do (`seqnames`="Chr", `ontology`="Cell") are shown.
- The ablation battery (feature-set breakdown) only ever ran with `group_col="seqnames"` (see
  `Snakefile`: `wildcard_constraints: group_col = "seqnames"` on `splicing_ml_ablation_classification`
  -- deliberately excludes `ontology`, known to leak event identity, see `CLAUDE.md` Known
  Pitfalls). So the "Cell" x-position only ever has an "All features" point; the Non-Epigenetic/
  Epigenetic lines simply don't reach it.
- Every "Cell" (`ontology`) point is ringed in red: `group_col="ontology"` is confirmed (`CLAUDE.md`
  Known Pitfalls) to leak event identity for tree models like xgb -- near-ceiling AUROC there is
  not real generalization. Included for completeness, not to be read as a trustworthy number.

```{r, setup}
splicing_ml_root <- "splicing_ml"
model <- "xgb" # single model for the feature-set comparison, matching the preprint's tree-based
# (random forest) comparison in spirit -- xgb is this project's current default tree model (rf
# itself is excluded from defaults, see CLAUDE.md "Default Models": "redundant with XGB; XGB
# consistently outperforms it").

metrics <- c("balanced_accuracy", "auroc", "auprc", "mcc")
metric_labels <- c(
  balanced_accuracy = "Balanced\nAccuracy", auroc = "AUROC", auprc = "AUPRC", mcc = "MCC"
)
variabilities <- c("High", "both", "Low")
variability_labels <- c(High = "High", both = "All", Low = "Low") # matches variability_colors names
group_cols <- c("seqnames", "ontology")
group_col_labels <- c(seqnames = "Chr", ontology = "Cell")
feature_sets <- list(
  list(key = "all", label = "All"),
  list(key = "sequence", label = "Non Epigenetic"),
  list(key = "histone+dnam", label = "Epigenetic")
)
# Distinct from variability_colors (Low/All/High) so the two color legends never clash when both
# appear near each other -- picked from the remaining unused Okabe-Ito colors.
feature_colors <- c(
  "All" = "#0072B2", "Non Epigenetic" = "#009E73", "Epigenetic" = "#CC79A7"
)
```

```{r, load-results}
# Deliberately read the small `reports/subset_report_*.json` payload (already
# written by splicing_ml's reporting/payload.py for the HTML report), NOT the
# raw `splicing_ml_results_classification.json.gz` next to it. The raw file
# carries full per-fold diagnostics arrays (y_true/y_pred/shap/
# transformed_features) -- confirmed real sizes 186MB-824MB *compressed* (RI/
# High/seqnames, SE/Low/seqnames respectively), several GB uncompressed.
# `jsonlite::fromJSON()` on that (even piping through `jq` first) measured out
# to 60GB+ RAM and 30+ minutes for a single file just to read `model_name`/
# `scores` -- unusable across the 24 files this notebook reads. The report
# payload already has exactly what's needed: a fold-averaged
# `metric_heatmap.z[model][metric]`, max 5.7MB even at SE scale, since
# payload.py never puts the raw per-fold arrays in it. This is why the rule is
# resourced at only 8GB/30min in config/snakemake_config.yaml.
result_path <- function(event_type, variability, group_col, feature_key) {
  subset_name <- sprintf("%s_biotype_filtered_%s_%s", event_type, variability, group_col)
  dir <- if (feature_key == "all") {
    file.path(splicing_ml_root, "output", subset_name)
  } else {
    file.path(splicing_ml_root, "output_ablation", feature_key, subset_name)
  }
  file.path(dir, "reports", sprintf("subset_report_%s.json", subset_name))
}

read_metric <- function(path, model, metric) {
  # simplifyVector=FALSE: jsonlite's default auto-simplification turns a
  # uniform numeric matrix like `z` (3 models x 5 metrics here) into a real R
  # matrix rather than nested lists, silently breaking `z[[mi]][[ti]]`-style
  # indexing (double-bracket on a matrix flattens to vector indexing instead
  # of row/col lookup) -- confirmed via a real crash ("subscript out of
  # bounds") that only triggered for ti > 1.
  d <- jsonlite::fromJSON(path, simplifyVector = FALSE)[[1]]
  mh <- d$metric_heatmap
  mi <- match(model, unlist(mh$models))
  ti <- match(metric, unlist(mh$metrics))
  if (is.na(mi) || is.na(ti)) NA_real_ else as.numeric(mh$z[[mi]][[ti]])
}

rows <- list()
missing_configs <- character(0)
for (event_type in to_analyze) {
  for (variability in variabilities) {
    for (group_col in group_cols) {
      for (fs in feature_sets) {
        path <- result_path(event_type, variability, group_col, fs$key)
        if (!file.exists(path)) {
          missing_configs <- c(missing_configs, path)
          next
        }
        for (metric in metrics) {
          v <- read_metric(path, model, metric)
          if (!is.na(v)) {
            rows[[length(rows) + 1]] <- data.table::data.table(
              `Event Type` = event_type, Variability = variability, `Group Col` = group_col,
              Features = fs$label, Metric = metric, Value = v
            )
          }
        }
      }
    }
  }
}

if (length(missing_configs) > 0) {
  message(sprintf(
    "%d config(s) not yet available (plotted with gaps): \n%s",
    length(missing_configs), paste(missing_configs, collapse = "\n")
  ))
}
stopifnot("No splicing_ml result files found -- has splicing_ml_classification been run yet?" = length(rows) > 0)

comparison_dt <- data.table::rbindlist(rows)
comparison_dt[, `Test on` := factor(group_col_labels[`Group Col`], levels = group_col_labels[group_cols])]
comparison_dt[, Variability := factor(variability_labels[Variability], levels = variability_labels[variabilities])]
comparison_dt[, Measure := factor(metric_labels[Metric], levels = metric_labels[metrics])]
comparison_dt[, Features := factor(Features, levels = vapply(feature_sets, function(f) f$label, character(1)))]
comparison_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
```

```{r, plot-comparison, fig.width=14, fig.height=9}
# Ontology ("Cell") points: known event-identity CV leak for tree models -- ring in red so the
# plot itself carries the caveat, not just this notebook's prose.
leak_points <- comparison_dt[`Test on` == "Cell"]

p <- ggplot(comparison_dt, aes(x = `Test on`, y = Value, color = Features, group = Features)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  geom_point(
    data = leak_points, aes(x = `Test on`, y = Value),
    shape = 21, color = "red", fill = NA, size = 4.5, stroke = 1.1, inherit.aes = FALSE
  ) +
  ggh4x::facet_grid2(
    Measure ~ `Event Type` + Variability,
    scales = "free_y", strip = ggh4x::strip_nested()
  ) +
  scale_color_manual(values = feature_colors) +
  labs(
    x = "Test on Unseen Chromosome (Chr) or Cell Type (Cell)",
    y = "Value",
    title = sprintf("model=%s performance by held-out condition and feature set", model),
    subtitle = paste0(
      "Red-ringed points = group_col=\"ontology\" (\"Cell\") -- confirmed to leak event identity ",
      "for tree models\n(CLAUDE.md Known Pitfalls); do not read as real generalization."
    )
  ) +
  theme(legend.position = "bottom", plot.subtitle = element_text(size = 8, color = "gray30"))
print(p)
ggsave(filename = file.path(plot_dir, "07-2_ml_global_comparison.pdf"), plot = p, width = 14, height = 9)

data.table::fwrite(comparison_dt, file.path(plot_dir, "07-2_ml_global_comparison_data.csv"))
```

```{r, upload-wandb}
# No native R wandb client in this project -- shell out to a small standalone Python script
# instead of adding a splicing_ml/*.py dependency here (that would sit inside SPLICING_ML_SRC and
# retrigger every splicing_ml_classification/ablation Snakemake rule on next invocation).
# Pushes the tidy comparison data as a native wandb Table + bar charts (built by wandb itself,
# interactive/filterable in the UI) rather than uploading a static PNG -- avoids the PIL
# dependency wandb.Image() needs (not installed in this project's env) and gives a richer result.
wandb_upload_script <- "scripts/upload_comparison_to_wandb.py"
csv_path <- file.path(plot_dir, "07-2_ml_global_comparison_data.csv")

if (file.exists(wandb_upload_script)) {
  status <- system2(
    "mamba", c("run", "-n", "ihec-as", "python", wandb_upload_script, csv_path),
    stdout = TRUE, stderr = TRUE
  )
  message(paste(status, collapse = "\n"))
} else {
  warning(sprintf("%s not found -- skipping W&B upload.", wandb_upload_script))
}
```
