---
title: "ML event models"
output: pdf_document
---

# Epigenetic Prediction of Alternative Splicing: Per-Event Models

Each AS event is modelled independently with elastic net regression (glmnet) using IHEC epigenetic features. Predictive performance is evaluated against 10 **rotated negative controls** — models trained on PSI values from unrelated events that preserve data structure while breaking the epigenetic–splicing link.

**Feature sets**: `local` (event-overlapping regions only), `short` (±50 kb), `long` (±500 kb).
**Lambda selection**: `min` (lowest CV error) vs `1se` (most regularised model within 1 SE of min).
**Expression bias**: a real-PSI model is labelled *Epigenetic Only* if no `gene_expression`-containing feature survives robust cross-fold selection; otherwise *Expression Bias*.

```{r setup}
library(tidymodels)
library(glmnet)
library(data.table)
library(ggplot2)
library(pbmcapply)
```

```{r load-session}
# Per-filter (matches 09-1 and everything from 02-2 on): one transcript_filter
# per invocation, chosen by TRANSCRIPT_FILTER (standalone falls back to the
# primary filter).
tf <- Sys.getenv(
  "TRANSCRIPT_FILTER",
  getOption("EpiATLAS_AS_PRIMARY_FILTER", "biotype_filtered")
)

# Load session objects produced by 09-1-ml-local.R (same tf)
sess <- readRDS(sprintf("processed_data/session_09_1_ml_local_%s.rds", tf))
event_dt <- sess$event_dt # trimmed: ID, Event Type, seqnames, Variability, transcript_filter
keep_rows <- sess$keep_rows_manual
rm(sess)
gc()

# `aggregating.rda` is GONE (removed by the per-filter refactor). aggregated_dt
# is per-filter; activeChromHMM (only other object 09-2 needs from that old
# workspace — see the chromhmm-region-annotation chunk below) is now persisted
# directly by 03 (same run as keep_rows_manual, so index-consistent).
aggregated_dt <- fread(
  sprintf("processed_data/aggregated_dt_filtered_%s.csv.gz", tf),
  stringsAsFactors = TRUE
)
activeChromHMM <- readRDS(sprintf("processed_data/activeChromHMM_%s.rds", tf))

# NB: no "restrict to biotype_filtered" subset needed anymore — event_dt (from
# 09-1's session, built off event_annotations_dt_{tf}) and aggregated_dt are
# already scoped to exactly `tf`.
```

## Hyperparameter diagnostics

```{r load-data, cache = TRUE}
event_dir <- file.path("processed_data", "event_models", tf)

plot_dir_ev <- file.path(plot_dir, "event_models", tf)
dir.create(plot_dir_ev, showWarnings = FALSE, recursive = TRUE)

# Helper: aggregate per-event CSVs saved by worker. Robust to (a) no matching
# files and (b) individual unreadable/truncated files (a cancelled/partial run
# can leave a half-written .csv.gz) — pbmclapply otherwise returns error strings
# in place of tables and rbindlist(fill=TRUE) then errors with "Input is
# character…".
#
# When nothing readable is found it returns an empty data.table WITH THE EXPECTED
# COLUMNS for that file type (schema matches 09zz's writers). This is what keeps
# the report renderable at 0 Tier-1 hits AND keeps Positron "Run Above" working:
# knitr chunk options (eval = en_available) are ignored by interactive execution,
# so the EN chunks run regardless; giving them a 0-ROW-but-well-formed table lets
# their `[, := ]` transforms and ggplot() calls no-op on empty data instead of
# crashing on missing columns (e.g. `ID not found`).
.event_csv_schema <- list(
  best_params = c("penalty", "mixture", ".config", "cv_mean_ccc", "cv_mean_rmse",
    "cv_mean_rsq_trad", "cv_std_err_ccc", "cv_std_err_rmse", "cv_std_err_rsq_trad",
    "cv_n_ccc", "cv_n_rmse", "cv_n_rsq_trad", "wflow_id", "lambda_selection", "ID"),
  fit_metrics = c(".metric", ".estimator", ".estimate", "wflow_id", "lambda_selection", "ID"),
  all_metrics = c("id", "penalty", "mixture", ".metric", ".estimator", ".estimate",
    ".config", "wflow_id", "ID"),
  event_summary = c("wflow_id", "model_type", "n_robust", "has_model",
    "expression_bias", "lambda_selection", "ID"),
  nonzero_coefs = c("wflow_id", "feature", "coef", "lambda_selection", "ID"),
  fold_features = c("wflow_id", "fold", "feature", "lambda_selection", "ID")
)
.event_csv_numeric <- c("penalty", "mixture", ".estimate", "coef", "n_robust",
  "cv_mean_ccc", "cv_mean_rmse", "cv_mean_rsq_trad", "cv_std_err_ccc",
  "cv_std_err_rmse", "cv_std_err_rsq_trad", "cv_n_ccc", "cv_n_rmse", "cv_n_rsq_trad")
.empty_event_dt <- function(pattern) {
  key <- names(.event_csv_schema)[
    vapply(names(.event_csv_schema), function(k) grepl(k, pattern, fixed = TRUE), logical(1))
  ]
  if (length(key) != 1L) {
    return(data.table())
  }
  cols <- .event_csv_schema[[key]]
  data.table::as.data.table(setNames(
    lapply(cols, function(c) if (c %in% .event_csv_numeric) numeric(0) else character(0)),
    cols
  ))
}
read_event_csvs <- function(pattern) {
  files <- list.files(event_dir, pattern = pattern, full.names = TRUE)
  if (length(files) == 0L) {
    return(.empty_event_dt(pattern))
  }
  parts <- pbmclapply(files, function(f) tryCatch(fread(f), error = function(e) NULL))
  parts <- parts[vapply(parts, is.data.frame, logical(1))]
  if (length(parts) == 0L) {
    return(.empty_event_dt(pattern))
  }
  rbindlist(parts, fill = TRUE)
}

# Tier-1 ridge-screen results — significance (is the event epigenetically
# predictable beyond confounds?) is now decided HERE, not by the elastic-net
# rotation test. One row per event: screen_R2/CCC, matched feature-rotation null
# summary, empirical p, effect (= real − null mean), and FDR q. `sig := q < 0.1`
# drives every downstream significance split. The 6 per-event elastic-net tables
# below are HITS-ONLY (Tier-2) and all response_type == "real".
# Robust load so 09-2 renders on PARTIAL results too (screen still running / not
# yet aggregated). Prefer the aggregated screen_results.csv.gz (real Storey q from
# 09s-aggregate.R over ALL events); if it is absent, assemble a provisional
# screen_dt from whatever per-event screen rows exist and approximate q with BH —
# the final report always uses the true aggregate, this is only so intermediate
# renders don't hard-fail.
.screen_results_file <- file.path(event_dir, "screen_results.csv.gz")
if (file.exists(.screen_results_file)) {
  screen_dt <- fread(.screen_results_file)
} else {
  .screen_rows <- list.files(
    file.path(event_dir, "screen"),
    pattern = "^\\d+_screen\\.csv\\.gz$", full.names = TRUE  # excludes _screen_null
  )
  if (length(.screen_rows) == 0L) {
    warning(
      "No screen_results.csv.gz and no per-event screen rows — Tier-1 sections ",
      "will be empty. Run the ridge screen (09s) first."
    )
    screen_dt <- data.table(
      ID = integer(), n_samples = integer(), R_used = integer(),
      screen_R2 = numeric(), screen_CCC = numeric(), null_R2_mean = numeric(),
      null_R2_sd = numeric(), p_emp = numeric(), effect = numeric(),
      note = character()
    )
  } else {
    warning(sprintf(
      "screen_results.csv.gz absent — assembling provisional screen_dt from %d ",
      length(.screen_rows)
    ), "per-event rows; q via BH proxy (PARTIAL, not the final Storey FDR).")
    screen_dt <- rbindlist(pbmclapply(.screen_rows, fread), fill = TRUE)
  }
  screen_dt[, q := NA_real_]
  .ok <- is.finite(screen_dt$p_emp)
  if (any(.ok)) screen_dt$q[.ok] <- p.adjust(screen_dt$p_emp[.ok], method = "BH")
}
screen_dt[, ID := as.integer(ID)]
screen_q_threshold <- 0.1
screen_dt[, sig := is.finite(q) & q < screen_q_threshold & is.finite(effect) & effect > 0]

# en_available — is there USABLE Tier-2/EN data? Gates every EN-dependent chunk
# (via `eval = en_available`) so the Tier-1 (screen) report always renders, even
# at 0 hits or on a partial run. DATA-driven, not just file-presence: read the
# all_metrics (CV-metrics) tables robustly and require ≥1 row with a real metric
# column. This is deliberately stricter than "a file exists" — a run can leave
# only truncated files or skip-stubs (`data.table(ID=…)` with no `.metric`), and
# gating on mere presence would then let build-cv-tables run and crash on the
# missing ID/.metric columns.
.en_all_metrics <- read_event_csvs("^\\d+_all_metrics\\.csv\\.gz$")
en_available <- nrow(.en_all_metrics) > 0L && ".metric" %in% names(.en_all_metrics)
```

```{r load-en-tables, eval = en_available, cache = TRUE, dependson = "load-data"}
# Tier-2 elastic-net per-event tables (HITS ONLY). This whole chunk is skipped
# when en_available is FALSE (no Tier-1 hits) — see load-data — so downstream
# EN chunks (also eval = en_available) never touch undefined/empty tables.

# Best hyperparameters per (event, workflow)
best_params_table <- read_event_csvs("^\\d+_best_params\\.csv\\.gz$")
best_params_table[, ID := as.integer(ID)]
best_params_table[,
  c("feature_set", "response") := if (.N > 0L) {
    tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
  } else {
    list(character(), character())  # tstrsplit() errors on 0-row input
  }
]
best_params_table[,
  response_type := fifelse(response == "PSI", "Real", "Rotated")
]
best_params_table[,
  lambda_selection := factor(lambda_selection, levels = c("min", "1se"))
]

# In-sample fit metrics per (event, workflow)
fit_metrics_table <- read_event_csvs("^\\d+_fit_metrics\\.csv\\.gz$")
fit_metrics_table[, ID := as.integer(ID)]
fit_metrics_table[,
  c("feature_set", "response") := if (.N > 0L) {
    tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
  } else {
    list(character(), character())  # tstrsplit() errors on 0-row input
  }
]
```

```{r exploratory-plots, eval = en_available}
# How do hyperparameter choices vary across feature sets and real vs rotated response?
# A systematic difference in penalty/mixture suggests the model uses different
# regularisation regimes for real vs random targets.
ggplot(
  best_params_table[!is.na(lambda_selection)],
  aes(
    x = lambda_selection,
    y = -log10(penalty),
    color = as.factor(mixture),
    group = interaction(ID, wflow_id)
  )
) +
  geom_line(alpha = 0.2) +
  geom_boxplot(
    aes(group = lambda_selection),
    color = "grey30",
    fill = NA,
    outliers = FALSE
  ) +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_grid(response_type ~ feature_set) +
  scale_color_brewer(palette = "Set1", name = "mixture") +
  labs(
    title = "Selected penalty by lambda rule, feature set, and response type",
    x = "Lambda selection rule",
    y = "-log10(penalty)"
  ) +
  theme_bw()

ggplot(
  best_params_table,
  aes(x = as.character(mixture), y = -log10(penalty), color = feature_set)
) +
  geom_boxplot(alpha = .5) +
  facet_grid(. ~ response == "PSI") +
  labs(
    title = "Penalty distribution by mixture and response type",
    x = "Mixture (0 = ridge, 1 = lasso)",
    y = "-log10(penalty)"
  ) +
  theme_bw()

# In-sample R² will be inflated by overfitting; compare to CV in paired-rotation-analysis.
ggplot(
  fit_metrics_table,
  aes(x = response == "PSI", y = .estimate, color = feature_set)
) +
  geom_violin(alpha = .5, position = position_dodge(width = 0.8)) +
  geom_jitter(
    alpha = .2,
    position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.5),
    size = .5
  ) +
  facet_grid(.metric ~ lambda_selection, scales = "free_y") +
  labs(
    title = "In-sample fit metrics: real PSI vs rotated (overfitting diagnostic)",
    x = "Real PSI response",
    y = "Fit estimate (PSI scale)"
  ) +
  main_theme(16) +
  coord_cartesian(ylim = c(0, NA))
```

**What to look for:** penalty should be similar for real and rotated responses — systematic differences would suggest the CV procedure is leaking information. Mixture near 1 (lasso) implies sparse solutions; near 0 (ridge) implies dense regularisation.

## Data tables

```{r build-cv-tables, eval = en_available, cache = TRUE, dependson = c("load-data", "load-en-tables")}
# cv_metrics_table: per-fold CV metrics at selected config, tagged by lambda_selection
# cv_rsq: mean CV R² per (event, workflow) — used downstream by paired-rotation-analysis
#         and n-robust-significance
cv_metrics_table <- read_event_csvs("^\\d+_all_metrics\\.csv\\.gz$")
cv_metrics_table[, ID := as.integer(ID)]
cv_metrics_table <- cv_metrics_table[
  best_params_table[, .(ID, wflow_id, .config, lambda_selection)],
  on = .(ID, wflow_id, .config),
  nomatch = NULL
]
cv_metrics_table[,
  c("feature_set", "response") := if (.N > 0L) {
    tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
  } else {
    list(character(), character())  # tstrsplit() errors on 0-row input
  }
]
cv_metrics_table[,
  response_type := fifelse(response == "PSI", "real", "rotated")
]
cv_metrics_table[,
  feature_set := factor(feature_set, levels = c("local", "short", "long"))
]

cv_rsq <- cv_metrics_table[
  .metric == "rsq_trad",
  .(mean_cv_rsq = mean(.estimate, na.rm = TRUE)),
  by = .(ID, wflow_id, lambda_selection)
]
```

```{r build-summary-tables, eval = en_available, cache = TRUE, dependson = c("load-data", "load-en-tables", "build-cv-tables")}
# event_summary_table: one row per (event, workflow, lambda_selection).
# Includes n_robust, expression_bias, mean CV R², rotated_hit flag.
event_summary_table <- read_event_csvs("^\\d+_event_summary\\.csv\\.gz$")
event_summary_table[, ID := as.integer(ID)]
event_summary_table[,
  c("feature_set", "response") := if (.N > 0L) {
    tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
  } else {
    list(character(), character())  # tstrsplit() errors on 0-row input
  }
]
event_summary_table[,
  response_type := fifelse(response == "PSI", "real", "rotated")
]
event_summary_table[
  event_dt,
  on = "ID",
  c("Event Type", "Variability", "seqnames") := .(
    `Event Type`,
    Variability,
    seqnames
  )
]
event_summary_table[
  cv_rsq,
  on = .(ID, wflow_id, lambda_selection),
  mean_cv_rsq := mean_cv_rsq
]

# "Rotated Hit": did this event's FEATURES get borrowed as a background control
# for another event? (Rotation is now FEATURE-rotation — 09zz fits this event's
# PSI against other matched events' real feature matrices — so the rotated
# `response` token is the control event whose FEATURES were used, not whose PSI.
# Semantics flipped from PSI-borrowing to feature-borrowing; the ID set is
# computed the same way.)
rotation_ids <- event_summary_table[
  response_type == "rotated",
  unique(as.integer(response))
]
event_summary_table[
  response_type == "real",
  rotated_hit := fifelse(ID %in% rotation_ids, "Hit by Rotation", "Not Hit")
]

# rotation_rank: 1..n_rotations per (event, feature_set, lambda), 0 for real.
# Used to group rotated models by slot rather than source event ID.
# Source IDs are unique per event (independent sampling), so grouping by source
# gives counts ≈ 1. Grouping by rank gives uniqueN(ID) ≈ N_events per slot.
event_summary_table[
  response_type == "rotated",
  rotation_rank := frankv(response, ties.method = "first"),
  by = .(ID, feature_set, lambda_selection)
]
event_summary_table[response_type == "real", rotation_rank := 0L]

event_summary_table[,
  feature_set := factor(feature_set, levels = c("local", "short", "long"))
]
event_summary_table[,
  lambda_selection := factor(lambda_selection, levels = c("min", "1se"))
]
event_summary_table[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
event_summary_table[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
```

```{r build-feature-coef-tables, eval = en_available, cache = TRUE, dependson = c("load-data", "load-en-tables", "build-summary-tables")}
# feature_table: features selected per CV fold per workflow.
# coef_table: robust non-zero coefficients per workflow.
# expression_bias is propagated from the real-PSI model for each event via the
# ID + lambda_selection join, so rotated coef rows also carry the real model's bias label.

feature_table <- read_event_csvs("^\\d+_fold_features\\.csv\\.gz$")
feature_table[, ID := as.integer(ID)]
feature_table[,
  c("feature_set", "response") := if (.N > 0L) {
    tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
  } else {
    list(character(), character())  # tstrsplit() errors on 0-row input
  }
]
feature_table[, response_type := fifelse(response == "PSI", "real", "rotated")]
feature_table[
  event_dt,
  on = "ID",
  c("Event Type", "Variability") := .(`Event Type`, Variability)
]
feature_table[
  event_summary_table[
    response_type == "real",
    .(ID, lambda_selection, expression_bias)
  ],
  on = .(ID, lambda_selection),
  expression_bias := expression_bias
]

coef_table <- read_event_csvs("^\\d+_nonzero_coefs\\.csv\\.gz$")
coef_table[, ID := as.integer(ID)]
coef_table[,
  c("feature_set", "response") := if (.N > 0L) {
    tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
  } else {
    list(character(), character())  # tstrsplit() errors on 0-row input
  }
]
coef_table[, response_type := fifelse(response == "PSI", "real", "rotated")]
coef_table[
  event_dt,
  on = "ID",
  c("Event Type", "Variability") := .(`Event Type`, Variability)
]
# expression_bias from real-PSI model propagated to rotated rows via ID+lambda_selection join
coef_table[
  event_summary_table[
    response_type == "real",
    .(ID, lambda_selection, expression_bias)
  ],
  on = .(ID, lambda_selection),
  expression_bias := expression_bias
]
coef_table[, let(c("mark", "region"), if (.N > 0L) {
  tstrsplit(feature, ";", fixed = TRUE)
} else {
  list(character(), character())  # tstrsplit() errors on 0-row input
})]
coef_table[,
  coefficient_sign := fifelse(
    coef > 0,
    "positive coefficient",
    "negative coefficient"
  )
]
coef_table[,
  Response := factor(
    fifelse(response_type == "real", "Real", "Rotated"),
    levels = c("Real", "Rotated")
  )
]
coef_table[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
coef_table[
  response_type == "rotated",
  rotation_rank := match(response, unique(response)),
  by = .(ID, feature_set, lambda_selection)
]
coef_table[response_type == "real", rotation_rank := 0L]

for (.dt in list(feature_table, coef_table)) {
  .dt[,
    feature_set := factor(feature_set, levels = c("local", "short", "long"))
  ]
  .dt[, lambda_selection := factor(lambda_selection, levels = c("min", "1se"))]
  .dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  .dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
}
```

## 1. Event counts and filtering

How many events survive each filtering step, and how many receive a model?
A model requires at least one feature to be robustly selected across all CV folds (`n_robust > 0`).
`lambda.min` produces weaker regularisation — more features selected, more events modelled.
`lambda.1se` is stricter — fewer events have any robust features (`has_model == FALSE`).

```{r counting-events, eval = en_available}
event_counts <- rbindlist(
  lapply(unique(event_summary_table$lambda_selection), function(sel) {
    rbindlist(
      lapply(unique(event_summary_table$feature_set), function(fs) {
        this_summary <- event_summary_table[
          lambda_selection == sel & feature_set == fs
        ]
        modeled_ids <- this_summary[
          response_type == "real" & has_model == TRUE,
          unique(ID)
        ]
        rbindlist(
          list(
            Quantified = aggregated_dt[
              ID %in% keep_rows,
              .(N = uniqueN(ID)),
              by = `Event Type`
            ],
            Filtered = event_dt[ID %in% keep_rows, .(.N), by = .(`Event Type`)],
            `No Model` = aggregated_dt[
              !ID %in% modeled_ids & ID %in% keep_rows,
              .(N = uniqueN(ID)),
              by = `Event Type`
            ],
            `Model` = this_summary[
              response_type == "real" & has_model == TRUE,
              .(.N),
              by = .(`Event Type`)
            ],
            `Epigenetic\nOnly Model` = this_summary[
              response_type == "real" &
                expression_bias == "Epigenetic Only" &
                has_model == TRUE,
              .(.N),
              by = .(`Event Type`)
            ]
          ),
          idcol = "filtering"
        )[, c("lambda_selection", "feature_set") := .(sel, fs)]
      }),
      fill = TRUE
    )
  }),
  fill = TRUE
)

print(event_summary_table[
  response_type == "real",
  table(expression_bias, `Event Type`, lambda_selection, feature_set)
])

event_counts[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
event_counts[,
  feature_set := factor(feature_set, levels = c("long", "short", "local"))
]

ggplot(
  event_counts,
  aes(x = N, y = reorder(filtering, N), fill = lambda_selection)
) +
  geom_col(position = "dodge") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "white"),
    text = element_text(size = 20),
    axis.text.x = element_blank()
  ) +
  facet_grid(feature_set ~ `Event Type`, scales = "free_x") +
  geom_text(
    aes(label = N),
    position = position_dodge(width = 0.9),
    vjust = 1.1,
    size = 4,
    angle = 90
  ) +
  labs(
    title = "Event waterfall: counts at each filtering stage",
    x = "Number of Events",
    y = NULL
  )
ggsave(file.path(plot_dir_ev, "event_counts.pdf"), width = 12, height = 5)
```

Events without any robust features (`has_model == FALSE`) are excluded from all downstream analyses.
Expression bias is classified per real-PSI model and propagated to its associated rotated controls.

```{r generate-count-plots, eval = en_available}
event_summary_table[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
event_summary_table[,
  `Event Type` := factor(`Event Type`, levels = to_analyze)
]

event_count_dt <- event_summary_table[
  has_model == TRUE,
  {
    # Group by rotation_rank (0 for real, 1-10 for rotated) and count unique events.
    # For real: one rank (0), uniqueN(ID) = N events. mean_count = N.
    # For rotated: ranks 1-10, uniqueN(ID) per rank = events in that slot. mean/SD across slots.
    counts <- .SD[, .(count = uniqueN(ID)), by = rotation_rank][, count]
    .(
      mean_count = mean(counts),
      sd_count = sd(counts) / sqrt(uniqueN(rotation_rank))
    )
  },
  by = .(
    `Event Type`,
    feature_set,
    response_type,
    expression_bias,
    Variability,
    lambda_selection
  )
]

event_count_dt[response_type == "real", response_type := "Real"]
event_count_dt[response_type == "rotated", response_type := "Rotated"]
event_count_dt[,
  response_type := factor(response_type, levels = c("Real", "Rotated"))
]
event_count_dt[
  expression_bias == "Expression Bias",
  expression_bias := "Expression\nBias"
]
event_count_dt[
  expression_bias == "Epigenetic Only",
  expression_bias := "Epigenetic\nOnly"
]
event_count_dt[
  order(Variability, decreasing = TRUE),
  cumsum_mean_count := cumsum(mean_count),
  by = .(
    `Event Type`,
    feature_set,
    response_type,
    expression_bias,
    lambda_selection
  )
]

ggplot(
  event_count_dt,
  aes(y = mean_count, x = response_type, fill = Variability)
) +
  geom_col() +
  ggh4x::facet_nested_wrap(
    `Event Type` ~ feature_set + lambda_selection + expression_bias
  ) +
  theme_bw() +
  theme(
    legend.box = "vertical",
    strip.background = element_rect(fill = "white"),
    legend.position = "bottom",
    axis.title.x = element_blank()
  ) +
  scale_fill_manual(values = variability_colors) +
  labs(
    title = "Model counts by expression bias, variability, and response type",
    y = "Number of Events"
  ) +
  geom_errorbar(
    aes(
      y = cumsum_mean_count,
      ymin = cumsum_mean_count - sd_count,
      ymax = cumsum_mean_count + sd_count
    ),
    width = .2
  )
ggsave(
  file.path(plot_dir_ev, "events_rotation.png"),
  width = 16,
  height = 3,
  dpi = 600
)
```
```{r generate-proportion-plots, eval = en_available}
event_count_dt[
  response_type == "Real",
  expression_bias := sub("\n", " ", expression_bias, fixed = TRUE)
]
ggplot(
  event_count_dt[response_type == "Real"],
  aes(y = mean_count, x = Variability, fill = expression_bias)
) +
  geom_col(position = "fill") +
  geom_text(aes(label = mean_count), position = position_fill(vjust = 0.5)) +
  facet_grid(`Event Type` ~ feature_set + lambda_selection) +
  theme_bw() +
  theme(
    legend.box = "vertical",
    strip.background = element_rect(fill = "white"),
    legend.position = "bottom"
  ) +
  scale_fill_manual(
    values = c(`Epigenetic Only` = "orange", `Expression Bias` = "steelblue")
  ) +
  labs(
    title = "Proportion of real-PSI models by variability: Epigenetic Only vs Expression Bias",
    y = "Proportion of Events",
    fill = NULL
  )
ggsave(file.path(plot_dir_ev, "events_hit.png"), width = 10, height = 3)
```

## 2. Tier-1 ridge screen: does epigenetic signal exceed chance?

Significance is decided by the **Tier-1 fit-free ridge screen** (`09s-ridge-screen.R` → `screen_results.csv.gz`), run on *every* event, not by the elastic-net. For each event it computes a closed-form group-leave-one-ontology-out ridge **partial-R²** (epigenetics beyond protocol/expression/TSS) and tests it against a **matched feature-rotation null** (hundreds of unrelated control loci, each a cheap closed-form re-fit) → empirical p → FDR **q**. An event is significant when **q < threshold AND effect > 0**, where `effect = real R² − mean(null R²)`.

Key point (fingerprint unmasking): **high raw R² is not sufficient**. Cell-type-fingerprint events have high R² but `effect ≈ 0` — a *random* control locus's chromatin predicts their PSI just as well, because both track cell-type identity. Only `effect > 0 & q < threshold` events carry genuinely event-specific epigenetic signal. The elastic-net below (Tier 2) then runs **only on these hits**, real PSI only, for feature interpretation.

Plot 1 (real in-sample vs CV R²) is retained as an overfitting diagnostic; Plots 2–3 show the screen's p-value distribution and the effect-vs-R² unmasking.

```{r paired-rotation-analysis, eval = en_available}
# EN-only (Plot 1 real-R2 diagnostic + emp_p_dt broadcast for downstream
# feature-set analyses). Gated on en_available — the screen significance plots
# were moved to the always-on `screen-significance-plots` chunk below.
# Defensive re-assignment (already set in build chunks)
fit_metrics_table[,
  response_type := fifelse(response == "PSI", "real", "rotated")
]
cv_metrics_table[,
  response_type := fifelse(response == "PSI", "real", "rotated")
]

# Average CV folds to one value per workflow
cv_wflow <- cv_metrics_table[
  .metric == "rsq_trad",
  .(.estimate = mean(.estimate, na.rm = TRUE)),
  by = .(ID, wflow_id, feature_set, lambda_selection, response_type)
]

# Combine in-sample and CV for overfitting diagnostic (Plot 1 only).
# emp_p and t-statistic use CV only — in-sample R² is biased by overfitting.
wflow_fit <- rbindlist(
  list(
    `In-sample` = fit_metrics_table[
      .metric == "rsq_trad",
      .(ID, wflow_id, feature_set, lambda_selection, response_type, .estimate)
    ],
    CV = cv_wflow
  ),
  idcol = "metric_type"
)
wflow_fit[event_dt, on = "ID", `Event Type` := `Event Type`]
wflow_fit[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
wflow_fit[, metric_type := factor(metric_type, levels = c("In-sample", "CV"))]

# ------------------------------------------------------------------
# Plot 1: Real R² distribution — overfitting diagnostic only.
# In-sample R² is inflated; the rotation test below uses CV only.
# ------------------------------------------------------------------
real_rsq_dt <- wflow_fit[
  response_type == "real",
  .(
    ID,
    `Event Type`,
    feature_set,
    lambda_selection,
    metric_type,
    real_rsq = .estimate
  )
]

ggplot(real_rsq_dt, aes(x = feature_set, y = real_rsq, color = feature_set)) +
  geom_violin(alpha = .5) +
  geom_jitter() +
  facet_grid(`Event Type` ~ metric_type + lambda_selection) +
  coord_cartesian(ylim = c(0, 1)) +
  main_theme(base_size = 16) +
  labs(
    title = "Real PSI R2: in-sample vs CV (in-sample inflated by overfitting)",
    x = NULL,
    y = "R2 (real PSI)"
  )
ggsave(file.path(plot_dir_ev, "real_rsq.pdf"), width = 10, height = 4)

# ------------------------------------------------------------------
# Plot 2: Empirical p-value (in-sample and CV)
# For each event: p = (# rotations with R² >= real R² + 1) / (# rotations + 1)
# With 10 rotations: min achievable p = 1/11 ≈ 0.091.
# Conventional p < 0.05 requires >= 19 rotations.
# Computed for both in-sample and CV; see rationale below.
# ------------------------------------------------------------------
# ------------------------------------------------------------------
# Significance now comes from the Tier-1 ridge SCREEN (screen_dt), not from
# elastic-net rotations — Tier-2 (09zz) is hits-only, real PSI, so there are no
# rotated rows here. screen_dt is per-EVENT: a closed-form group-leave-one-
# ontology-out ridge partial-R² (epigenetics beyond protocol/expression/TSS)
# tested against a matched FEATURE-rotation null (hundreds of control loci) →
# empirical p → FDR q; `sig := q < threshold & effect > 0` (effect = real R² −
# mean null R²). Broadcast the event-level significance across the EN
# (feature_set × lambda_selection) rows so the downstream per-feature-set chunks
# keep their join shape; the value is identical across an event's rows.
#
# NOTE (fingerprint unmasking): high raw screen_R2 is NOT sufficient — cell-type
# fingerprint events have high R² but effect ≈ 0 (a random control locus predicts
# their PSI just as well). Only effect > 0 & q < threshold events are genuinely
# event-specific. See screen_effect_vs_r2.pdf below.
fs_lambda_combos <- unique(
  best_params_table[, .(ID, feature_set, lambda_selection)]
)
emp_p_dt <- merge(
  fs_lambda_combos,
  screen_dt[, .(
    ID, emp_p = p_emp, q, effect, screen_R2, screen_CCC,
    null_R2_mean, sig
  )],
  by = "ID", all.x = TRUE
)
emp_p_dt[event_dt, on = "ID", `Event Type` := `Event Type`]
emp_p_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
emp_p_dt[,
  feature_set := factor(feature_set, levels = c("local", "short", "long"))
]
```

```{r screen-significance-plots}
# Screen significance plots — built DIRECTLY from screen_dt (one row per event),
# so they always render, even at 0 hits / on a partial run (independent of the EN
# tables). Previously these were computed off emp_p_dt, which broadcasts screen
# values across the EN (feature_set × lambda) rows and is therefore empty when
# there are no hits.
screen_plot_dt <- screen_dt[, .(ID, emp_p = p_emp, q, effect, screen_R2, sig)]
screen_plot_dt[event_dt, on = "ID", `Event Type` := `Event Type`]
screen_plot_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# Per-event significance summary by event type.
print(screen_plot_dt[,
  .(
    n_sig = sum(sig, na.rm = TRUE),
    n_total = sum(!is.na(q)),
    pct_sig = round(100 * mean(sig, na.rm = TRUE), 1)
  ),
  by = `Event Type`
])

# Plot 2: screen empirical-p per event, faceted by event type.
ggplot(screen_plot_dt[is.finite(emp_p)], aes(x = emp_p, fill = `Event Type`)) +
  geom_histogram(bins = 40, alpha = .7, position = "identity") +
  geom_vline(xintercept = screen_q_threshold, linetype = "dashed", color = "red") +
  facet_wrap(~`Event Type`, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Tier-1 ridge-screen empirical p-value (per event)",
    x = "Empirical p (matched feature-rotation null)",
    y = "Number of events",
    caption = "Significant = FDR q < threshold AND effect (real - mean null R2) > 0."
  )
ggsave(file.path(plot_dir_ev, "screen_pvalue.pdf"), width = 10, height = 4)

# Plot 3: effect vs raw R² — the fingerprint unmasking. High raw R² with effect
# ≈ 0 = cell-type fingerprint (a random locus predicts as well); only effect > 0
# & q < threshold events pass.
ggplot(
  screen_plot_dt[is.finite(effect) & is.finite(screen_R2)],
  aes(x = screen_R2, y = effect, color = sig)
) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_point(alpha = .4, size = .8) +
  facet_wrap(~`Event Type`) +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey60")) +
  theme_bw() +
  labs(
    title = "Screen effect (real - null) vs raw R2: fingerprint unmasking",
    x = "Raw group-LOO screen R2",
    y = "Effect = real R2 - mean null R2",
    color = "Significant\n(q < thresh, effect > 0)",
    caption = paste(
      "High raw R2 with effect near 0 = cell-type fingerprint (random control",
      "loci predict as well). Only effect > 0 & q < threshold = event-specific."
    )
  )
ggsave(file.path(plot_dir_ev, "screen_effect_vs_r2.pdf"), width = 10, height = 4)
```

Plot 4 shows, per event, exactly *where* the real partial-R² sits inside its own matched feature-rotation null distribution — the raw material behind the empirical p. This reads the full null vectors persisted by `09s-ridge-screen.R` (`screen/<ID>_screen_null.csv.gz`, one row per control locus), which the summary `screen_results.csv.gz` collapses to mean/sd. Only a manageable subset of events is drawn (all hits, plus the top events by effect) — the full event set is far too many to facet.

```{r screen-null-histograms}
# Per-event null-distribution histograms: real screen_R2 (vertical line) over the
# ~200 matched feature-rotation null R2 values. The nulls are the SAME quantity
# the empirical p is computed from (p = (1 + #{null >= real}) / (1 + R_used)), so
# a real line far in the right tail = small p = event-specific epigenetic signal;
# a real line buried in the bulk = fingerprint / no signal.
null_files <- list.files(
  file.path(event_dir, "screen"),
  pattern = "_screen_null\\.csv\\.gz$",
  full.names = TRUE
)

if (length(null_files) == 0L) {
  message("No screen_null sidecars found — skipping per-event null histograms. ",
          "(Re-run 09s-ridge-screen.R after the null-vector write to populate them.)")
} else {
  null_long <- rbindlist(lapply(null_files, fread), fill = TRUE)
  null_long[, ID := as.integer(ID)]
  null_long <- null_long[is.finite(null_R2)]

  # Choose which events to draw: every hit, plus the top events by effect (most
  # interesting even if not significant). Capped so the full-run event set can't
  # explode the facet grid.
  max_facets <- 24L
  hit_ids <- screen_dt[sig == TRUE, ID]
  top_ids <- screen_dt[is.finite(effect)][order(-effect), ID]
  plot_ids <- head(unique(c(hit_ids, top_ids)), max_facets)
  n_omitted <- length(unique(null_long$ID)) - length(plot_ids)
  if (n_omitted > 0L) {
    message(sprintf(
      "Null histograms: drawing %d event(s) (%d hit + top-by-effect); %d omitted.",
      length(plot_ids), length(hit_ids), n_omitted
    ))
  }

  nd <- null_long[ID %in% plot_ids]
  # Per-event real value + significance + a compact facet label.
  ev_lab <- screen_dt[
    ID %in% plot_ids,
    .(ID, screen_R2, p_emp, q, sig)
  ]
  ev_lab[event_dt, on = "ID", `Event Type` := `Event Type`]
  ev_lab[, facet := sprintf(
    "ID %d (%s)\nR2=%.3f p=%.2g q=%.2g%s",
    ID, `Event Type`, screen_R2, p_emp, q,
    fifelse(sig %in% TRUE, " *", "")
  )]
  nd[ev_lab, on = "ID", facet := facet]
  vlines <- ev_lab[, .(ID, facet, screen_R2, sig)]

  ggplot(nd, aes(x = null_R2)) +
    geom_histogram(bins = 30, fill = "grey70", color = NA) +
    geom_vline(
      data = vlines,
      aes(xintercept = screen_R2, color = sig %in% TRUE),
      linewidth = 0.7
    ) +
    facet_wrap(~facet, scales = "free") +
    scale_color_manual(
      values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"),
      labels = c(`TRUE` = "significant", `FALSE` = "not sig"),
      name = "real R2 (event)"
    ) +
    theme_bw(base_size = 10) +
    labs(
      title = "Tier-1: real partial-R2 vs its matched feature-rotation null (per event)",
      x = "R2 (grey = null control loci; line = this event's real R2)",
      y = "Number of null control loci",
      caption = paste(
        "Each grey histogram = the ~200 matched-control null R2 for one event;",
        "the vertical line is that event's own real R2. Right-tail line = small",
        "empirical p = event-specific signal. * = FDR-significant (q < threshold",
        "& effect > 0)."
      )
    )
  ggsave(
    file.path(plot_dir_ev, "screen_null_histograms.pdf"),
    width = 14, height = 10
  )
}
```

Finally, screen QC / calibration diagnostics: is the feature-rotation null honest (empirical-p roughly uniform, not piled toward 0), is the fingerprint pattern present (raw R² rises but effect stays ≈0), and are the very-negative group-LOO R² just small-n instability rather than a bug? These are the checks to eyeball before trusting the hit list.

```{r screen-diagnostics}
sd_fin <- screen_dt[is.finite(p_emp) & is.finite(screen_R2) & is.finite(effect)]

# Coverage: rotations actually used + degenerate-event notes (NA = fit fine).
cat("Screen coverage:\n")
print(screen_dt[, .(
  n_events        = .N,
  finite_p        = sum(is.finite(p_emp)),
  R_used_median   = median(R_used, na.rm = TRUE),
  R_used_below_max = sum(R_used < max(R_used, na.rm = TRUE), na.rm = TRUE)
)])
cat("Degenerate-event notes (NA = screened cleanly):\n")
print(screen_dt[, .N, by = note][order(-N)])

# (1) NULL CALIBRATION — p_emp should be ~flat under the null. A broken/optimistic
# null piles p toward 0 across all bins; genuine signal shows only as a modest
# excess in the leftmost bin over the uniform (dashed) line.
n_bins <- 10L
ggplot(sd_fin, aes(x = p_emp)) +
  geom_histogram(
    breaks = seq(0, 1, 1 / n_bins), fill = "grey70", color = "white"
  ) +
  geom_hline(
    yintercept = nrow(sd_fin) / n_bins, linetype = "dashed", color = "firebrick"
  ) +
  theme_bw() +
  labs(
    title = "Tier-1 null calibration: empirical p-value distribution",
    x = "Empirical p (matched feature-rotation null)",
    y = "Number of events",
    caption = paste(
      "Dashed = uniform expectation. Flat ≈ well-calibrated null; excess in the",
      "leftmost bin = genome-wide signal; a monotone pile-up toward 0 would",
      "indicate an optimistic (broken) null."
    )
  )
ggsave(file.path(plot_dir_ev, "screen_pvalue_calibration.pdf"), width = 8, height = 4)

# (2) FINGERPRINT — effect & p within raw-R2 strata. High raw R2 with effect ≈ 0
# and non-small p = cell-type fingerprint, not event-specific signal (binned
# complement to screen_effect_vs_r2.pdf).
sd_fin[, r2_stratum := cut(
  screen_R2, c(-Inf, 0, 0.1, 0.2, 0.5, Inf),
  labels = c("<=0", "(0,0.1]", "(0.1,0.2]", "(0.2,0.5]", ">0.5")
)]
cat("Fingerprint check — effect & p within raw-R2 strata:\n")
print(sd_fin[, .(
  n              = .N,
  median_R2      = round(median(screen_R2), 3),
  median_effect  = round(median(effect), 4),
  pct_effect_pos = round(100 * mean(effect > 0), 1),
  median_p       = round(median(p_emp), 3)
), by = r2_stratum][order(r2_stratum)])

# (3) SMALL-n INSTABILITY — very-negative group-LOO R2 concentrates in small
# events (expected), not a bug. y clipped at the 1st percentile for readability.
ggplot(sd_fin, aes(x = n_samples, y = screen_R2, color = sig)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_point(alpha = .3, size = .7) +
  scale_color_manual(
    values = c(`TRUE` = "firebrick", `FALSE` = "grey60"),
    labels = c(`TRUE` = "significant", `FALSE` = "not sig"), name = NULL
  ) +
  coord_cartesian(
    ylim = c(quantile(sd_fin$screen_R2, 0.01, na.rm = TRUE), 1)
  ) +
  theme_bw() +
  labs(
    title = "Screen R2 vs sample size (very-negative R2 = small-n LOO instability)",
    x = "n samples (event)",
    y = "Group-LOO screen R2"
  )
ggsave(file.path(plot_dir_ev, "screen_r2_vs_nsamples.pdf"), width = 8, height = 4)
```

## 3. Biological context of significant events

The following analyses ask *which events* show the strongest epigenetic signal and *why*, probing whether significance associates with biologically meaningful properties.

```{r variability-significance, eval = en_available}
# Research question: do High-Variability events show a higher rotation-test
# significance rate than Low-Variability events? This directly tests whether
# epigenetic regulation is stronger for dynamically regulated splicing.

emp_p_var_dt <- copy(emp_p_dt)
emp_p_var_dt[
  event_dt[, .(ID, Variability)],
  on = "ID",
  Variability := Variability
]
emp_p_var_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
emp_p_var_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

var_sig_dt <- emp_p_var_dt[
  !is.na(emp_p) & !is.na(Variability),
  .(
    n_sig = sum(sig, na.rm = TRUE),
    n_total = .N,
    pct_sig = 100 * mean(sig, na.rm = TRUE)
  ),
  by = .(Variability, `Event Type`, feature_set, lambda_selection)
]

ggplot(var_sig_dt, aes(x = Variability, y = pct_sig, fill = feature_set)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = sprintf("%d/%d", n_sig, n_total)),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  facet_grid(`Event Type` ~ lambda_selection) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Rotation test significance rate by splicing variability group",
    x = "Variability",
    y = "% events with emp p <= 1/11",
    fill = "Feature set",
    caption = "CV R2 only. emp p <= 1/11 is the minimum achievable with 10 rotations. Positive result: High-Variability bars taller than Low-Variability, supporting epigenetic control of dynamic splicing."
  )
ggsave(
  file.path(plot_dir_ev, "variability_significance.pdf"),
  width = 8,
  height = 4
)
```

```{r n-robust-significance, eval = en_available}
# Research question: do events with more robustly selected features (n_robust)
# tend to pass the rotation test? n_robust counts features present in all CV
# folds; higher n_robust = more stable, confident model.

nrob_sig_dt <- copy(emp_p_dt)
nrob_sig_dt[
  event_summary_table[
    response_type == "real",
    .(ID, feature_set, lambda_selection, n_robust)
  ],
  on = .(ID, feature_set, lambda_selection),
  n_robust := n_robust
]
nrob_sig_dt[
  event_dt[, .(ID, Variability)],
  on = "ID",
  Variability := Variability
]
nrob_sig_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
nrob_sig_dt[
  !is.na(emp_p),
  sig_label := fifelse(sig, "Significant (p <= 1/11)", "Not significant")
]
nrob_sig_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# Events with n_robust == 0 (no model) will not appear in emp_p_dt; filter is defensive
ggplot(
  nrob_sig_dt[!is.na(n_robust) & n_robust > 0 & !is.na(sig_label)],
  aes(x = sig_label, y = n_robust, fill = sig_label)
) +
  geom_violin(alpha = .5, quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(alpha = .15, width = 0.2, size = .5) +
  facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
  scale_fill_manual(
    values = c(
      "Significant (p <= 1/11)" = "#D55E00",
      "Not significant" = "#56B4E9"
    ),
    guide = "none"
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Robust feature count by rotation test outcome",
    x = NULL,
    y = "n_robust (features selected in all CV folds)",
    caption = "CV R2 only. Events with n_robust = 0 (no model) excluded. Positive result: Significant events shifted toward higher n_robust, confirming that more stable models pass the rotation test."
  )
ggsave(
  file.path(plot_dir_ev, "nrobust_significance.pdf"),
  width = 16,
  height = 6
)
```

```{r feature-consistency, eval = en_available}
# Research question: how consistently is each feature selected across CV folds?
# Computed on real-PSI models only; fold "0" (full-training fit) excluded since
# it is not a cross-validation fold.
# High mean consistency = features robustly selected regardless of held-out tissue.

feat_cons_dt <- feature_table[
  response_type == "real" & fold != "0",
  .(n_folds_selected = uniqueN(fold)),
  by = .(ID, wflow_id, feature_set, lambda_selection, feature, `Event Type`)
]
total_folds_dt <- feature_table[
  response_type == "real" & fold != "0",
  .(total_folds = uniqueN(fold)),
  by = .(ID, wflow_id, lambda_selection)
]
feat_cons_dt[
  total_folds_dt,
  on = .(ID, wflow_id, lambda_selection),
  frac_folds := n_folds_selected / total_folds
]

per_event_cons_dt <- feat_cons_dt[,
  .(mean_consistency = mean(frac_folds, na.rm = TRUE)),
  by = .(ID, feature_set, lambda_selection, `Event Type`)
]
per_event_cons_dt[
  event_dt[, .(ID, Variability)],
  on = "ID",
  Variability := Variability
]
per_event_cons_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
per_event_cons_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

ggplot(
  per_event_cons_dt[!is.na(mean_consistency)],
  aes(x = mean_consistency, fill = feature_set)
) +
  geom_histogram(bins = 20, position = "dodge", alpha = .7) +
  facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "CV fold consistency of selected features (real PSI models)",
    x = "Mean fraction of folds selecting each feature",
    y = "Number of events",
    fill = "Feature set",
    caption = "Fold 0 (full-training fit) excluded. Mean computed over all features per event. Events concentrated near 1.0 = features selected in nearly every fold (robust); near 0 = unstable, fold-sensitive selection."
  )
ggsave(
  file.path(plot_dir_ev, "feature_consistency.pdf"),
  width = 16,
  height = 6
)
```

If significant events consistently have more robust features, this suggests that the rotation test is not just detecting noise but events where epigenetic marks provide a stable, cross-tissue predictive signal.

```{r real-vs-rotated-n-robust, eval = en_available}
# Research question: do real PSI models robustly select more features than
# rotated (permutation) controls? More robust features in real models supports
# genuine epigenetic signal rather than overfitting to PSI structure.

nrob_comp_dt <- event_summary_table[
  !is.na(n_robust),
  .(
    ID,
    feature_set,
    lambda_selection,
    `Event Type`,
    Variability,
    response_type,
    n_robust
  )
]
nrob_comp_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
nrob_comp_dt[,
  response_label := factor(
    fifelse(response_type == "real", "Real PSI", "Rotated PSI"),
    levels = c("Real PSI", "Rotated PSI")
  )
]
nrob_comp_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

ggplot(
  nrob_comp_dt,
  aes(x = response_label, y = n_robust, fill = response_label)
) +
  geom_violin(alpha = .5, quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(alpha = .1, width = 0.2, size = .3) +
  facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
  scale_fill_manual(
    values = c("Real PSI" = "#D55E00", "Rotated PSI" = "#56B4E9"),
    guide = "none"
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Robust feature count: real PSI vs rotated controls",
    x = NULL,
    y = "n_robust (features selected in all CV folds)",
    caption = "Includes events with n_robust = 0 (no model built). Positive result: Real PSI distribution shifted right vs Rotated, indicating more features are robustly selected when predicting genuine splicing."
  )
ggsave(
  file.path(plot_dir_ev, "nrobust_real_vs_rotated.pdf"),
  width = 16,
  height = 6
)
```

Feature selection consistency across folds measures whether the same marks are chosen regardless of which tissue group is held out. Events with high consistency provide more reliable targets for follow-up validation.

```{r lambda-switching, eval = en_available}
# Research question: are rotation-test results stable across lambda selection rules?
# Concordance (both sig or both non-sig) suggests robust signal.
# "min only" events = sensitive to regularisation; "1se only" = rare edge case.

lambda_wide_dt <- dcast(
  emp_p_dt[!is.na(emp_p)],
  ID + feature_set + `Event Type` ~ lambda_selection,
  value.var = "sig"
)
setnames(lambda_wide_dt, c("min", "1se"), c("sig_min", "sig_1se"))
lambda_wide_dt[
  event_dt[, .(ID, Variability)],
  on = "ID",
  Variability := Variability
]
lambda_wide_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]

lambda_wide_dt[,
  lambda_category := fcase(
    sig_min & sig_1se   , "Both significant"    ,
    sig_min & !sig_1se  , "min only"            ,
    !sig_min & sig_1se  , "1se only"            ,
    !sig_min & !sig_1se , "Neither significant"
  )
]
lambda_wide_dt[,
  lambda_category := factor(
    lambda_category,
    levels = c(
      "Both significant",
      "min only",
      "1se only",
      "Neither significant"
    )
  )
]
lambda_wide_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

lambda_summary <- lambda_wide_dt[
  !is.na(lambda_category),
  .(n = .N),
  by = .(`Event Type`, Variability, feature_set, lambda_category)
]
lambda_summary[,
  pct := 100 * n / sum(n),
  by = .(`Event Type`, Variability, feature_set)
]

ggplot(lambda_summary, aes(x = feature_set, y = pct, fill = lambda_category)) +
  geom_col(position = "stack") +
  geom_text(
    aes(label = sprintf("%.0f%%\n(n=%d)", pct, n)),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  facet_grid(Variability ~ `Event Type`) +
  scale_fill_manual(
    values = c(
      "Both significant" = "#D55E00",
      "min only" = "#E69F00",
      "1se only" = "#56B4E9",
      "Neither significant" = "#999999"
    )
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Lambda rule concordance in rotation test significance",
    x = "Feature set",
    y = "% of events",
    fill = "Lambda significance",
    caption = "CV R2 only, emp p <= 1/11. Large 'Both significant' segment = results robust to regularisation choice; large 'min only' segment = signal is sensitive to regularisation and should be interpreted cautiously."
  )
ggsave(file.path(plot_dir_ev, "lambda_switching.pdf"), width = 8, height = 4)
```

## 4. Epigenetic mark-level analysis

Which specific histone marks drive prediction in Epigenetic Only models? Comparing real vs rotated mark usage tests whether the mark composition reflects genuine epigenetic–splicing associations or chance feature selection. Events where the same marks are selected for both real and rotated PSI are less convincing.

```{r plot-mark-counts, eval = en_available}
# activeChromHMM loaded per-filter in the load-session chunk above.
# Research question: which epigenetic marks are robustly selected in Epigenetic Only models?
# Does mark usage differ between real PSI and rotated controls?
# Similar mark profiles for real and rotated would imply chance selection.
activeChromHMM_dt <- data.table(
  region = activeChromHMM$name,
  chrom_state = activeChromHMM$chrom_state,
  region_width = width(activeChromHMM),
  region_ID = seq_along(activeChromHMM)
)
coef_table[
  activeChromHMM_dt,
  on = "region",
  c("chrom_state", "region_width", "region_ID") := .(
    chrom_state,
    region_width,
    region_ID
  )
]

# Both real and rotated rows included; expression_bias was propagated from the
# real-PSI model for each event, so the Epigenetic Only filter applies to both.
# feature_set kept to distinguish local (event-overlapping) vs short/long window marks.
err_dt <- coef_table[
  mark != "(Intercept)" & expression_bias == "Epigenetic Only",
  .(mark = unique(mark)),
  by = .(
    ID,
    rotation_rank,
    feature_set,
    response_type,
    `Event Type`,
    Variability,
    lambda_selection
  )
][,
  {
    counts <- .SD[, .(count = uniqueN(ID)), by = rotation_rank][, count]
    .(
      mean_count = mean(counts),
      sd_count = sd(counts) / sqrt(uniqueN(rotation_rank))
    )
  },
  by = .(
    `Event Type`,
    feature_set,
    response_type,
    mark,
    Variability,
    lambda_selection
  )
]

err_dt[,
  Response := factor(
    fifelse(response_type == "real", "Real", "Rotated"),
    levels = c("Real", "Rotated")
  )
]
err_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
err_dt[, `Event Type` := factor(`Event Type`, levels = c("SE", "RI"))]
err_dt[, mark := tools::toTitleCase(gsub("_", " ", mark))]

hex_colors <- c(
  mark_hex_colors,
  summed_enhancer = "darkorange",
  max_promoter = "tomato"
)
plot_colors <- setNames(
  rep("dark grey", err_dt[, uniqueN(mark)]),
  err_dt[, unique(mark)]
)
names(hex_colors) <- tools::toTitleCase(gsub("_", " ", names(hex_colors)))
for (name in names(hex_colors)) {
  inds <- startsWith(names(plot_colors), name)
  plot_colors[inds] <- hex_colors[name]
}

ggplot(err_dt, aes(x = Response, y = mean_count, fill = mark)) +
  geom_col(position = "dodge") +
  geom_errorbar(
    aes(ymin = mean_count - sd_count, ymax = mean_count + sd_count),
    width = .7,
    position = position_dodge(.9)
  ) +
  facet_grid(
    Variability + `Event Type` ~ feature_set + lambda_selection,
    scales = "free_y"
  ) +
  labs(
    title = "Epigenetic mark usage in Epigenetic Only models: real vs rotated",
    y = "Mean number of events selecting mark",
    x = NULL,
    fill = "Mark",
    caption = "Positive result: Real bar taller than Rotated for specific marks. Compare local vs short/long: local marks are event-overlapping, long marks include distal regulatory elements."
  ) +
  theme_bw() +
  scale_fill_manual(values = plot_colors) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "white")
  )
ggsave(file.path(plot_dir_ev, "mark_counts_error.pdf"), width = 18, height = 10)
```

### Rotation-test stratified mark analysis

The overall Real > Rotated gap in mark counts conflates two groups: events that **pass** the rotation test (genuine epigenetic–splicing signal) and events that **fail** (likely fingerprint confound). Only the passing group has trustworthy feature associations. If mark profiles are identical between the two groups, all selection is fingerprint-driven. Different profiles support causal interpretation for the significant subset.

```{r mark-counts-rotation-stratified, eval = en_available}
# Research question: do events passing the rotation test select different marks
# than failing events?
#
# Interpretation guide:
#   POSITIVE — distinct profiles, mark enriched in passing group:
#     Specific marks drive genuine epigenetic-splicing associations beyond
#     fingerprint confound. Feature interpretation trustworthy for the
#     significant subset.
#
#   NULL — similar profiles between groups:
#     Two indistinguishable explanations at this n:
#     (a) Fingerprint confound: all mark selection reflects cell-type identity;
#         passing events beat the test by chance (min achievable p = 1/11 ≈ 9%,
#         observed 14-15%, only borderline above chance rate).
#     (b) Genuine signal uses the SAME marks as the fingerprint — mark profile
#         comparison cannot separate the two; larger n or different design needed.
#
#   CONFOUND signal — Protocol/RNA-seq marks enriched in passing group:
#     Technical RNA-seq protocol effects drive apparent signal rather than
#     epigenetic regulation. Flag for investigation.

# Collapse rotation test significance to (ID, lambda_selection): significant if
# any feature_set passes emp_p <= 1/11 (CV R2 only).
sig_collapse_dt <- emp_p_dt[,
  .(sig_any = any(sig, na.rm = TRUE)),
  by = .(ID, lambda_selection)
]
sig_collapse_dt[,
  rotation_outcome := factor(
    fifelse(sig_any, "Passes rotation test", "Fails rotation test"),
    levels = c("Passes rotation test", "Fails rotation test")
  )
]

coef_real_sig <- coef_table[
  response_type == "real" &
    mark != "(Intercept)" &
    expression_bias == "Epigenetic Only"
]
coef_real_sig[
  sig_collapse_dt,
  on = .(ID, lambda_selection),
  rotation_outcome := rotation_outcome
]

# Per event per feature_set: unique marks selected.
# Keeping feature_set separates local (event-overlapping) from short/long window marks.
event_mark_dt <- coef_real_sig[
  !is.na(rotation_outcome),
  .(mark = unique(mark)),
  by = .(
    ID,
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    rotation_outcome
  )
]

# n_group: total events per stratum.
n_group_dt <- event_mark_dt[,
  .(n_group = uniqueN(ID)),
  by = .(
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    rotation_outcome
  )
]

err_sig_dt <- event_mark_dt[,
  .(n_events = uniqueN(ID)),
  by = .(
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    rotation_outcome,
    mark
  )
]
err_sig_dt[
  n_group_dt,
  on = .(
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    rotation_outcome
  ),
  c("n_group", "pct_events") := .(n_group, 100 * n_events / n_group)
]

err_sig_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
err_sig_dt[, `Event Type` := factor(`Event Type`, levels = c("SE", "RI"))]
err_sig_dt[, mark := tools::toTitleCase(gsub("_", " ", mark))]

# Fisher's exact test per mark per stratum.
# Fisher preferred over chi-squared: passing group sizes are small.
# Null: profiles indistinguishable → fingerprint OR same-mark genuine signal.
# Significant enrichment in passing group: mark associated beyond fingerprint.
fisher_dt <- err_sig_dt[,
  {
    pass_row <- .SD[rotation_outcome == "Passes rotation test"]
    fail_row <- .SD[rotation_outcome == "Fails rotation test"]
    if (nrow(pass_row) > 0L && nrow(fail_row) > 0L) {
      mat <- matrix(
        c(
          pass_row$n_events[1L],
          pass_row$n_group[1L] - pass_row$n_events[1L],
          fail_row$n_events[1L],
          fail_row$n_group[1L] - fail_row$n_events[1L]
        ),
        nrow = 2,
        dimnames = list(c("selected", "not"), c("passes", "fails"))
      )
      res <- fisher.test(mat)
      .(p_value = res$p.value, odds_ratio = unname(res$estimate))
    } else {
      .(p_value = NA_real_, odds_ratio = NA_real_)
    }
  },
  by = .(`Event Type`, Variability, feature_set, lambda_selection, mark)
]
fisher_dt[, p_adj := p.adjust(p_value, method = "BH")]
fisher_dt[,
  sig_label := fcase(
    p_adj < 0.001 , "***" ,
    p_adj < 0.01  , "**"  ,
    p_adj < 0.05  , "*"   ,
    p_adj < 0.1   , "."   ,
    default = ""
  )
]
print(fisher_dt[order(
  `Event Type`,
  Variability,
  feature_set,
  lambda_selection,
  p_adj
)])

# Annotate passing-group bars with significance label (stars above bar).
plot_sig_dt <- err_sig_dt[rotation_outcome == "Passes rotation test"][
  fisher_dt[, .(
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    mark,
    sig_label
  )],
  on = .(`Event Type`, Variability, feature_set, lambda_selection, mark)
]

sig_plot_colors <- setNames(
  rep("dark grey", err_sig_dt[, uniqueN(mark)]),
  err_sig_dt[, unique(mark)]
)
for (name in names(hex_colors)) {
  inds <- startsWith(names(sig_plot_colors), name)
  sig_plot_colors[inds] <- hex_colors[name]
}

ggplot(err_sig_dt, aes(x = rotation_outcome, y = pct_events, fill = mark)) +
  geom_col(position = "dodge") +
  geom_text(
    data = plot_sig_dt[sig_label != ""],
    aes(label = sig_label, y = pct_events + 3),
    position = position_dodge(0.9),
    size = 3,
    vjust = 0
  ) +
  facet_grid(
    Variability + `Event Type` ~ feature_set + lambda_selection,
    scales = "free_y"
  ) +
  scale_fill_manual(values = sig_plot_colors) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "white"),
    axis.text.x = element_text(angle = 20, hjust = 1)
  ) +
  labs(
    title = "Mark composition by rotation test outcome (Epigenetic Only, real PSI)",
    x = NULL,
    y = "% of events in group selecting mark",
    fill = "Mark",
    caption = paste(
      "Stars above Passes bars: Fisher's exact test, BH-adjusted (. p<0.1, * p<0.05, ** p<0.01).",
      "Null (no stars, similar profiles): fingerprint confound explains selection, OR genuine",
      "signal uses same marks as fingerprint -- indistinguishable at this n.",
      "Significant enrichment in passing group: specific mark beyond fingerprint confound."
    )
  )
ggsave(
  file.path(plot_dir_ev, "mark_counts_rotation_stratified.pdf"),
  width = 18,
  height = 10
)
```

### TSS proximity of promoter-mark-selected events

H3K4me3 and H3K27ac are promoter-associated marks. If events where these marks are robustly
selected are systematically closer to the TSS than unselected events, the mark selection may
reflect promoter chromatin contaminating the feature window rather than event-level regulation.
The `local` feature set (event-overlapping only) is the critical comparator: if H3K4me3 is
selected even in `local`, the mark genuinely overlaps the exon/intron body. Selection in
`short`/`long` windows near TSS is more likely window contamination.

```{r tss-proximal, eval = en_available}
# distance_TSS: genomic distance from event to nearest TSS.
# Event-level property — same across all samples. Take first non-NA per event.
tss_dist_dt <- aggregated_dt[
  !is.na(distance_TSS),
  .(
    distance_TSS = distance_TSS[1L],
    abs_distance_TSS = abs(distance_TSS[1L]),
    `Event Type` = `Event Type`[1L]
  ),
  by = ID
]
tss_dist_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# Identify events with H3K4me3 and/or H3K27ac robustly selected.
# Real PSI models only; all expression_bias categories (H3K4me3 may drive expression bias).
# Stratify by feature_set and lambda_selection.
promoter_events <- coef_table[
  response_type == "real" &
    mark %in% c("H3K4me3", "H3K27ac"),
  .(
    has_H3K4me3 = any(mark == "H3K4me3"),
    has_H3K27ac = any(mark == "H3K27ac"),
    has_either = TRUE,
    expression_bias = expression_bias[1L]
  ),
  by = .(ID, feature_set, lambda_selection)
]

# Reference: all events with a fitted real-PSI model — one set per feature_set x lambda_selection.
# Use cross-join to avoid missing events that appear in coef_table for some combos but not all.
modeled_ids <- event_summary_table[
  response_type == "real" & has_model == TRUE,
  unique(ID)
]
fs_ls_combos <- unique(promoter_events[, .(feature_set, lambda_selection)])

combined_dt <- rbindlist(
  lapply(seq_len(nrow(fs_ls_combos)), function(i) {
    fs <- fs_ls_combos$feature_set[i]
    ls <- fs_ls_combos$lambda_selection[i]
    sel_ids <- promoter_events[
      feature_set == fs & lambda_selection == ls & has_either == TRUE,
      unique(ID)
    ]
    tss_dist_dt[
      ID %in% modeled_ids,
      .(
        ID,
        `Event Type`,
        abs_distance_TSS,
        distance_TSS,
        feature_set = fs,
        lambda_selection = ls,
        group = fifelse(
          ID %in% sel_ids,
          "H3K4me3 or H3K27ac selected",
          "Not selected"
        )
      )
    ]
  })
)
combined_dt[,
  group := factor(
    group,
    levels = c("H3K4me3 or H3K27ac selected", "Not selected")
  )
]
combined_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# --- Plot 1: Density of abs(distance_TSS), log-scale ---
ggplot(
  combined_dt[!is.na(abs_distance_TSS)],
  aes(x = abs_distance_TSS, fill = group, color = group)
) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  scale_fill_manual(
    values = c(
      "H3K4me3 or H3K27ac selected" = "#D55E00",
      "Not selected" = "#56B4E9"
    ),
    name = NULL
  ) +
  scale_color_manual(
    values = c(
      "H3K4me3 or H3K27ac selected" = "#D55E00",
      "Not selected" = "#56B4E9"
    ),
    guide = "none"
  ) +
  facet_grid(`Event Type` ~ feature_set + lambda_selection) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "TSS proximity: events with promoter marks selected vs not",
    x = "abs(distance_TSS) [bp, log10]",
    y = "Density",
    caption = paste(
      "H3K4me3 and H3K27ac are promoter-associated marks.",
      "Left shift = selected events closer to TSS than unselected.",
      "local feature set: mark overlaps exon body — strongest evidence of event-level regulation.",
      "short/long: flanking window may capture nearby promoter, not the event itself.",
      "RI events are often first-intron; TSS proximity expected and not necessarily a confound."
    )
  )
ggsave(
  file.path(plot_dir_ev, "tss_proximal_density.pdf"),
  width = 16,
  height = 5
)

# --- Plot 2: % of events within distance thresholds ---
thresholds <- c(1000L, 2000L, 5000L, 10000L)
thresh_dt <- rbindlist(lapply(thresholds, function(th) {
  combined_dt[
    !is.na(abs_distance_TSS),
    .(
      pct_proximal = 100 * mean(abs_distance_TSS <= th),
      n_events = .N,
      threshold_bp = th
    ),
    by = .(`Event Type`, feature_set, lambda_selection, group)
  ]
}))
thresh_dt[, threshold_label := paste0("<= ", threshold_bp / 1000L, " kb")]
thresh_dt[,
  threshold_label := factor(
    threshold_label,
    levels = paste0("<= ", thresholds / 1000L, " kb")
  )
]

ggplot(
  thresh_dt,
  aes(x = threshold_label, y = pct_proximal, fill = group)
) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = sprintf("%.0f%%\n(n=%d)", pct_proximal, n_events)),
    position = position_dodge(width = 0.9),
    vjust = -0.2,
    size = 2.3
  ) +
  facet_grid(`Event Type` ~ feature_set + lambda_selection) +
  scale_fill_manual(
    values = c(
      "H3K4me3 or H3K27ac selected" = "#D55E00",
      "Not selected" = "#56B4E9"
    ),
    name = NULL
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "% of events within distance threshold of TSS",
    x = "Distance threshold",
    y = "% of events"
  )
ggsave(
  file.path(plot_dir_ev, "tss_proximal_thresholds.pdf"),
  width = 16,
  height = 5
)

# --- Wilcoxon test: are selected events closer to TSS? ---
# One-tailed: alternative = "less" tests whether selected events are *closer*.
wilcox_dt <- combined_dt[
  !is.na(abs_distance_TSS),
  {
    sel <- .SD[group == "H3K4me3 or H3K27ac selected", abs_distance_TSS]
    nosel <- .SD[group == "Not selected", abs_distance_TSS]
    if (length(sel) >= 3L && length(nosel) >= 3L) {
      wt <- wilcox.test(sel, nosel, alternative = "less")
      .(
        p_value = wt$p.value,
        median_sel_kb = round(median(sel) / 1000, 1),
        median_ref_kb = round(median(nosel) / 1000, 1),
        n_sel = length(sel),
        n_ref = length(nosel)
      )
    } else {
      .(
        p_value = NA_real_,
        median_sel_kb = NA_real_,
        median_ref_kb = NA_real_,
        n_sel = length(sel),
        n_ref = length(nosel)
      )
    }
  },
  by = .(`Event Type`, feature_set, lambda_selection)
]
wilcox_dt[, p_adj := p.adjust(p_value, method = "BH")]
wilcox_dt[,
  sig_label := fcase(
    p_adj < 0.001 , "***" ,
    p_adj < 0.01  , "**"  ,
    p_adj < 0.05  , "*"   ,
    p_adj < 0.1   , "."   ,
    default = "ns"
  )
]
print(wilcox_dt[order(`Event Type`, feature_set, lambda_selection)])
```

Interpretation guide:
- **`local` feature set, H3K4me3, no TSS enrichment**: mark genuinely overlaps exon/intron; event-level regulation plausible.
- **`short`/`long` feature set, significant TSS enrichment**: window captures nearby promoter; selection may be a positional confound. Add as Discussion note; consider TSS-distance filter (e.g., exclude events within 5 kb of TSS from H3K4me3/H3K27ac interpretation).
- **RI events near TSS**: first-intron retention is common and biologically expected near promoters; TSS proximity alone does not disqualify RI H3K4me3 associations.

## 5. PSI variability and model performance

Does the magnitude of PSI change across samples predict model quality? Events with larger PSI range provide more signal for feature learning but may also show stronger expression bias (gene expression co-varies with large PSI shifts).

```{r dpsi-analysis, eval = en_available}
# Research question: do high-dPSI events yield better models and more expression bias?
#
# NOTE: aggregated_dt is pre-filtered to variable splicing events — all events have
# PSI range > 0.25. Absolute cutoffs (e.g., "stable <= 0.1") produce empty bins.
# Use quantile-based tertile bins to compare within the observed range.
#
# Interpretation guide:
#   HIGH dPSI → more models built + higher CV R²:
#     Signal strength drives model quality. Supports genuine epigenetic regulation.
#
#   NO dPSI effect on R² (flat slope in scatter):
#     Marks capture static cell-type identity, not dynamic regulation intensity.
#     All events are already "high change"; variation within that range is irrelevant.
#
#   HIGH dPSI → more expression bias:
#     Gene expression co-varies with large PSI shifts (differentiation state), confounding
#     epigenetic attribution. "Epigenetic Only" label may be unreliable for top-dPSI events.
#
#   HIGH and LOW Variability groups show similar PSI range:
#     The "High/Low Variability" classification is NOT based on PSI range magnitude —
#     it likely reflects relative between-tissue vs. within-tissue variation. Both groups
#     fully overlap in absolute PSI range; the label captures a different biological axis.

# Per-event PSI statistics.
psi_var_dt <- aggregated_dt[
  !is.na(PSI),
  .(
    sd_psi = sd(PSI),
    range_psi = max(PSI) - min(PSI),
    n_samples = .N,
    `Event Type` = `Event Type`[1L],
    Variability = Variability[1L]
  ),
  by = ID
]
psi_var_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# Variability in aggregated_dt uses "High"/"Low" (not "High Variability"/"Low Variability").
ggplot(
  psi_var_dt[!is.na(`Event Type`)],
  aes(x = Variability, y = range_psi, fill = Variability)
) +
  geom_violin(alpha = 0.5, quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(alpha = 0.1, width = 0.2, size = 0.3) +
  facet_wrap(~`Event Type`) +
  scale_fill_manual(
    values = c("High" = "#D55E00", "Low" = "#56B4E9"),
    guide = "none"
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "PSI range by variability group (alignment check)",
    x = NULL,
    y = "PSI range (max - min)",
    caption = paste(
      "Range = max(PSI) - min(PSI) across IHEC samples ~ max pairwise dPSI.",
      "All events have range > 0.25 (pre-filtered). High/Low Variability groups overlap",
      "almost completely -- the Variability label is not based on PSI range magnitude."
    )
  )
ggsave(
  file.path(plot_dir_ev, "psi_range_distribution.pdf"),
  width = 6,
  height = 4
)

# Compute tertile bins on psi_var_dt (one row per event) BEFORE cartesian join.
# Use -Inf/Inf as outer boundaries to avoid "breaks not unique" from probs=c(0,1) matching
# the min/max values. Epsilon guard ensures strict monotonicity if q1==q2.
{
  dpsi_q <- quantile(
    psi_var_dt$range_psi,
    probs = c(1 / 3, 2 / 3),
    na.rm = TRUE
  )
  if (dpsi_q[[1L]] >= dpsi_q[[2L]]) {
    dpsi_q[[2L]] <- dpsi_q[[1L]] +
      .Machine$double.eps * max(1, abs(dpsi_q[[1L]])) * 1e6
  }
  psi_var_dt[,
    dpsi_bin := cut(
      range_psi,
      breaks = c(-Inf, dpsi_q[[1L]], dpsi_q[[2L]], Inf),
      labels = c(
        sprintf("Low tertile\n(<= %.2f)", dpsi_q[[1L]]),
        sprintf("Mid tertile\n(%.2f-%.2f)", dpsi_q[[1L]], dpsi_q[[2L]]),
        sprintf("High tertile\n(> %.2f)", dpsi_q[[2L]])
      ),
      include.lowest = TRUE
    )
  ]
}

# Join with all feature_set x lambda_selection combinations for full stratification.
# allow.cartesian: one psi_var_dt row fans out to (n_feature_sets x n_lambda) rows.
# dpsi_bin carried through join from psi_var_dt.
dpsi_model_dt <- psi_var_dt[
  event_summary_table[
    response_type == "real",
    .(
      ID,
      feature_set,
      lambda_selection,
      has_model,
      expression_bias,
      n_robust,
      mean_cv_rsq
    )
  ],
  on = "ID",
  allow.cartesian = TRUE,
  nomatch = NULL
]
dpsi_model_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
dpsi_model_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# Model-building rate by dPSI tertile, stratified by Event Type, Variability,
# feature_set, and lambda_selection.
model_rate_dt <- dpsi_model_dt[
  !is.na(dpsi_bin),
  .(
    n_model = sum(has_model, na.rm = TRUE),
    n_total = .N,
    pct_model = 100 * mean(has_model, na.rm = TRUE)
  ),
  by = .(`Event Type`, Variability, feature_set, lambda_selection, dpsi_bin)
]
ggplot(model_rate_dt, aes(x = dpsi_bin, y = pct_model, fill = dpsi_bin)) +
  geom_col() +
  geom_text(
    aes(label = sprintf("%d/%d", n_model, n_total)),
    vjust = -0.3,
    size = 2.5
  ) +
  facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
  scale_fill_brewer(palette = "OrRd", guide = "none") +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Model-building rate by PSI range tertile",
    x = "PSI range tertile",
    y = "% events with model (n_robust > 0)"
  )
ggsave(file.path(plot_dir_ev, "dpsi_model_rate.pdf"), width = 16, height = 8)

# Continuous scatter: range_psi vs mean CV R² with loess trend.
# More informative than bins given the pre-filtered range (all > 0.25).
# Null: flat or noisy slope — PSI range within this range doesn't predict model quality.
# Positive: upward slope — higher absolute change → better epigenetic predictability.
ggplot(
  dpsi_model_dt[has_model == TRUE & !is.na(mean_cv_rsq)],
  aes(x = range_psi, y = mean_cv_rsq, color = Variability)
) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(aes(group = Variability), method = "loess", se = FALSE) +
  facet_grid(`Event Type` ~ feature_set + lambda_selection) +
  scale_color_manual(values = c("High" = "#D55E00", "Low" = "#56B4E9")) +
  coord_cartesian(ylim = c(0, NA)) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "PSI range vs CV R2",
    x = "PSI range (max - min)",
    y = "Mean CV R2",
    caption = paste(
      "Loess trend per Variability group. Null: flat slope within the pre-filtered high-change range.",
      "All events have range > 0.25; variation above that threshold may not drive model quality."
    )
  )
ggsave(file.path(plot_dir_ev, "dpsi_cv_rsq.pdf"), width = 16, height = 5)

# Expression bias by dPSI tertile.
# High-dPSI events may show more expression bias if gene expression co-varies
# with large PSI shifts (e.g., differentiation).
bias_dt <- dpsi_model_dt[
  has_model == TRUE & !is.na(dpsi_bin) & !is.na(expression_bias),
  .(n = .N),
  by = .(
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    dpsi_bin,
    expression_bias
  )
][,
  pct := 100 * n / sum(n),
  by = .(`Event Type`, Variability, feature_set, lambda_selection, dpsi_bin)
]
ggplot(bias_dt, aes(x = dpsi_bin, y = pct, fill = expression_bias)) +
  geom_col(position = "stack") +
  geom_text(
    aes(label = sprintf("%.0f%%\n(n=%d)", pct, n)),
    position = position_stack(vjust = 0.5),
    size = 2.5
  ) +
  facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
  scale_fill_manual(
    values = c("Epigenetic Only" = "#56B4E9", "Expression Bias" = "#E69F00"),
    name = "Expression bias"
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Expression bias by PSI range tertile (events with models)",
    x = "PSI range tertile",
    y = "% of events",
    caption = "Expression Bias = gene_expression feature in all CV folds."
  )
ggsave(
  file.path(plot_dir_ev, "dpsi_expression_bias.pdf"),
  width = 16,
  height = 8
)
```

## 6. Inter-model feature variation

Why do individual event models differ from each other? Coherent groups of events sharing the same marks would imply shared epigenetic regulatory programs. Idiosyncratic, uncorrelated mark selection would support fingerprint confound.

```{r feature-heatmap, eval = en_available}
# Research question: is mark selection structured across events, or random?
#
# Interpretation guide:
#   COHERENT CLUSTERS (events group by mark pattern in heatmap):
#     Shared epigenetic programs drive splicing prediction for groups of events.
#     Supports genuine regulatory logic: different chromatin contexts predict
#     splicing in different genomic contexts (e.g., promoter vs. enhancer regulated).
#
#   NO CLUSTER STRUCTURE (scattered heatmap):
#     Feature selection is idiosyncratic — no shared regulatory logic across events.
#     Consistent with fingerprint confound: marks selected via cell-type correlation,
#     not because of event-specific epigenetic regulation.
#
#   ALL-POSITIVE CO-OCCURRENCE MATRIX (observed result):
#     No negative correlations even between biologically antagonistic marks
#     (H3K27ac active vs H3K27me3 repressive, r=0.65). This is a hallmark of
#     fingerprint confound: all marks correlate with cell-type identity, so
#     events where the model learns cell-type structure select ALL marks regardless
#     of their biological antagonism. A genuine regulatory signal would show
#     negative correlation between active and repressive marks (mutually exclusive
#     chromatin states at the same locus).
#     Protocol mRNA.Seq = Protocol total.RNA.Seq at r=1.00: these are duplicate
#     features (likely same numerical values), not independent marks.
#
#   MIXED COEFFICIENT SIGN (same mark → + in some events, − in others):
#     Mark effect is context-dependent. H3K27ac might activate exon inclusion in
#     one locus and repress it in another, reflecting local regulatory architecture.
#     Consistent sign = universal association; mixed sign = locus-specific effect.

# All real Epigenetic Only models — stratified by Event Type, feature_set, lambda_selection.
hm_coef_dt <- coef_table[
  response_type == "real" &
    expression_bias == "Epigenetic Only" &
    mark != "(Intercept)"
]
hm_coef_dt[,
  Variability := factor(
    sub(" Variability", "", Variability, fixed = TRUE),
    levels = c("High", "Low")
  )
]
hm_coef_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

# --- Coefficient sign distribution -------------------------------------------
# Stratified by Event Type (SE vs RI), feature_set (local/short/long),
# lambda_selection (min/1se), Variability (High/Low).
sign_dt <- hm_coef_dt[,
  .(n_pos = sum(coef > 0), n_neg = sum(coef < 0)),
  by = .(mark, `Event Type`, Variability, feature_set, lambda_selection)
]
sign_dt[, `:=`(
  pct_pos = 100 * n_pos / (n_pos + n_neg),
  pct_neg = 100 * n_neg / (n_pos + n_neg)
)]
sign_dt[, mark := tools::toTitleCase(gsub("_", " ", mark))]
sign_long <- melt(
  sign_dt,
  id.vars = c(
    "mark",
    "Event Type",
    "Variability",
    "feature_set",
    "lambda_selection"
  ),
  measure.vars = c("pct_pos", "pct_neg"),
  variable.name = "direction",
  value.name = "pct"
)
sign_long[,
  direction := fifelse(
    direction == "pct_pos",
    "Positive (+PSI)",
    "Negative (-PSI)"
  )
]

ggplot(
  sign_long,
  aes(x = reorder(mark, pct), y = pct, fill = direction)
) +
  geom_col(position = "stack") +
  coord_flip() +
  facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
  scale_fill_manual(
    values = c("Positive (+PSI)" = "#D55E00", "Negative (-PSI)" = "#56B4E9"),
    name = "Coefficient sign"
  ) +
  theme_bw() +
  theme(strip.background = element_rect(fill = NA)) +
  labs(
    title = "Coefficient sign per mark (Epigenetic Only)",
    x = NULL,
    y = "% of coefficients",
    caption = paste(
      "Positive = mark associated with higher PSI (active chromatin -> exon inclusion).",
      "Negative = mark associated with lower PSI (repressive chromatin -> exon skipping).",
      "Mixed sign = context-dependent effect; consistent sign = universal association.",
      "Compare SE vs RI: RI marks may differ (intron retention context differs from exon skipping).",
      "Compare local vs long: local = event-overlapping marks; long = 500kb window (distal regulatory).",
      "High vs Low Variability: if sign profiles differ, epigenetic regulation may vary by event dynamics."
    )
  )
ggsave(file.path(plot_dir_ev, "mark_coef_sign.pdf"), width = 18, height = 10)

# --- Mark co-occurrence correlation and sig-event heatmap --------------------
# Loop over Event Type x feature_set x lambda_selection.
# For each: binary presence matrix -> correlation heatmap + sig-event heatmap.
# Expected: all-positive co-occurrence (fingerprint). Any negative correlation
# between antagonistic marks (H3K27ac/H3K27me3) = evidence of genuine signal.

for (et in to_analyze) {
  for (fs in c("local", "short", "long")) {
    for (ls in c("min", "1se")) {
      hm_sub <- hm_coef_dt[
        `Event Type` == et & feature_set == fs & lambda_selection == ls
      ]
      if (nrow(hm_sub) < 5L) {
        next
      }

      hm_dt_sub <- hm_sub[, .(mean_coef = mean(coef)), by = .(ID, mark)]
      all_ids_sub <- hm_dt_sub[, unique(ID)]
      all_marks_sub <- hm_dt_sub[, unique(mark)]
      pres_grid_sub <- CJ(ID = all_ids_sub, mark = all_marks_sub)
      pres_grid_sub[hm_dt_sub, on = .(ID, mark), selected := 1L]
      pres_grid_sub[is.na(selected), selected := 0L]

      pres_wide_sub <- dcast(
        pres_grid_sub,
        mark ~ ID,
        value.var = "selected",
        fill = 0L
      )
      pres_mat_sub <- as.matrix(pres_wide_sub[, -1L, with = FALSE])
      rownames(pres_mat_sub) <- pres_wide_sub$mark

      valid_marks_sub <- rownames(pres_mat_sub)[
        apply(pres_mat_sub, 1L, var) > 0
      ]
      if (length(valid_marks_sub) < 2L) {
        next
      }

      cor_mat_sub <- cor(t(pres_mat_sub[valid_marks_sub, ]))
      hc_cor_sub <- hclust(as.dist(1 - cor_mat_sub), method = "ward.D2")
      mark_ord_sub <- tools::toTitleCase(
        gsub("_", " ", rownames(cor_mat_sub)[hc_cor_sub$order])
      )
      cor_long_sub <- melt(
        data.table(mark1 = rownames(cor_mat_sub), cor_mat_sub),
        id.vars = "mark1",
        variable.name = "mark2",
        value.name = "correlation"
      )
      cor_long_sub[, `:=`(
        mark1 = factor(
          tools::toTitleCase(gsub("_", " ", mark1)),
          levels = mark_ord_sub
        ),
        mark2 = factor(
          tools::toTitleCase(gsub("_", " ", mark2)),
          levels = mark_ord_sub
        )
      )]

      p_cor_sub <- ggplot(
        cor_long_sub,
        aes(x = mark1, y = mark2, fill = correlation)
      ) +
        geom_tile(color = "white") +
        geom_text(aes(label = sprintf("%.2f", correlation)), size = 2.5) +
        scale_fill_gradient2(
          low = "#56B4E9",
          mid = "white",
          high = "#D55E00",
          midpoint = 0,
          limits = c(-1, 1),
          name = "Pearson r"
        ) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(
          title = sprintf(
            "Mark co-occurrence: %s | %s | lambda.%s (n=%d)",
            et,
            fs,
            ls,
            length(all_ids_sub)
          ),
          x = NULL,
          y = NULL,
          caption = paste(
            "All-positive = fingerprint confound signature.",
            "Negative r between active/repressive marks = genuine regulatory signal.",
            "Protocol mRNA.Seq = Protocol total.RNA.Seq (r=1.00): duplicate features."
          )
        )
      print(p_cor_sub)
      ggsave(
        file.path(
          plot_dir_ev,
          sprintf("mark_cooccurrence_%s_%s_%s.pdf", tolower(et), fs, ls)
        ),
        plot = p_cor_sub,
        width = 7,
        height = 6
      )

      # Sig-event heatmap.
      sig_ids_sub <- emp_p_dt[
        sig == TRUE & lambda_selection == ls & ID %in% all_ids_sub,
        unique(ID)
      ]
      pres_sig_sub <- pres_grid_sub[ID %in% sig_ids_sub]
      if (
        pres_sig_sub[, uniqueN(ID)] < 3L || pres_sig_sub[, uniqueN(mark)] < 3L
      ) {
        next
      }

      pres_sig_wide_sub <- dcast(
        pres_sig_sub,
        mark ~ ID,
        value.var = "selected",
        fill = 0L
      )
      pres_sig_mat_sub <- as.matrix(pres_sig_wide_sub[, -1L, with = FALSE])
      rownames(pres_sig_mat_sub) <- pres_sig_wide_sub$mark

      hc_ev_sub <- hclust(
        dist(t(pres_sig_mat_sub), method = "binary"),
        method = "ward.D2"
      )
      hc_mark_sub <- hclust(
        dist(pres_sig_mat_sub, method = "binary"),
        method = "ward.D2"
      )

      sig_long_sub <- melt(
        data.table(mark = rownames(pres_sig_mat_sub), pres_sig_mat_sub),
        id.vars = "mark",
        variable.name = "ID",
        value.name = "selected"
      )
      sig_long_sub[,
        ID := factor(
          ID,
          levels = colnames(pres_sig_mat_sub)[hc_ev_sub$order]
        )
      ]
      sig_long_sub[,
        mark := factor(
          tools::toTitleCase(gsub("_", " ", mark)),
          levels = tools::toTitleCase(
            gsub("_", " ", rownames(pres_sig_mat_sub)[hc_mark_sub$order])
          )
        )
      ]
      # Annotate Variability per event.
      sig_annot_sub <- hm_sub[
        ID %in% sig_ids_sub,
        .(ID = as.character(ID), Variability = as.character(Variability))
      ][, .SD[1L], by = ID]
      sig_long_sub[sig_annot_sub, on = "ID", Variability := Variability]

      p_hm_sub <- ggplot(
        sig_long_sub,
        aes(x = ID, y = mark, fill = factor(selected))
      ) +
        geom_tile(color = "grey85", linewidth = 0.3) +
        scale_fill_manual(
          values = c("0" = "white", "1" = "#D55E00"),
          labels = c("0" = "Not selected", "1" = "Selected"),
          name = NULL
        ) +
        theme_bw() +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
        labs(
          title = sprintf(
            "%d sig events: %s | %s | lambda.%s",
            pres_sig_sub[, uniqueN(ID)],
            et,
            fs,
            ls
          ),
          x = "Events (hierarchical clustering, binary distance)",
          y = NULL,
          caption = paste(
            "Coherent column blocks = shared epigenetic predictor program.",
            "Scattered = idiosyncratic selection.",
            "Clusters likely reflect cell-type fingerprint subgroups."
          )
        )
      print(p_hm_sub)
      ggsave(
        file.path(
          plot_dir_ev,
          sprintf("sig_event_mark_heatmap_%s_%s_%s.pdf", tolower(et), fs, ls)
        ),
        plot = p_hm_sub,
        width = 9,
        height = 4
      )
    }
  }
}
```
