---
title: "ML event models"
output: pdf_document
---

# Epigenetic Prediction of Alternative Splicing: Per-Event Models

## What this report asks, and what would count as an answer

One question, asked per alternative-splicing event: **can that event's PSI be predicted from the
epigenetic state of its own locus, beyond what a matched but unrelated locus would predict?**

The "beyond a matched locus" clause carries the whole argument. Chromatin marks, DNA methylation and
gene expression all track **cell-type identity**, and PSI does too. So a model can predict an event's
PSI very well while learning nothing about that event: it learns *which cell type this sample is* and
exploits the fact that PSI differs between cell types. This is the **cell-type fingerprint confound**,
and it is why a high R² is not evidence of epigenetic control of splicing. What separates the two is a
**matched null**: refit the same model on a *different, unrelated* locus's features while keeping this
event's PSI. If the unrelated locus predicts just as well, the signal was fingerprint, not regulation.

This is the analysis the manuscript's "revisiting" framing rests on, so the report is deliberately
ordered so that *null-calibration* evidence comes before any hit list.

## Two tiers, and why the work is split

Significance and interpretation are different jobs wanting different models, so they are separated:

|                | **Tier 1 — screen**                                   | **Tier 2 — elastic net**              |
| -------------- | ----------------------------------------------------- | ------------------------------------- |
| Question       | *Is* this event predictable beyond its matched null?  | *Which features* carry the signal?    |
| Model          | closed-form ridge partial-R², group-LOO by ontology   | tuned elastic net (glmnet/tidymodels) |
| Runs on        | **every** event (~34k)                                | **hits only**                         |
| Null           | matched control loci (RI 818, SE 200) → p → FDR q     | none (real PSI only)                  |
| Script         | `09s-ridge-screen.R` → `09s-aggregate.R`              | `09-1-ml-local.R` → `09zz-…glmnet….R` |

Tier 1 wants an **omnibus, high-power, closed-form** detector: ridge's L2 penalty aggregates diffuse,
correlated chromatin signal instead of arbitrarily picking one of several collinear windows, and being
closed-form makes a several-hundred-locus null essentially free. Tier 2 wants **sparsity**, because its
deliverable is a short, readable feature list — but sparse selection among collinear features is
unstable, which is exactly why it must not be the thing that decides significance.

A consequence worth stating plainly: **if Tier 1 returns no hits, Tier 2 has nothing to run on** and
every elastic-net section renders empty. That is a real, expected production state, not a failure — see
*How to read a report with no hits* below.

## Feature spaces (and why they are tested separately)

Both tiers use the same three nested spaces:

- **`local`** — event-overlapping regions only
- **`short`** — plus near-vicinity chromHMM windows
- **`long`** — plus far-vicinity chromHMM windows; i.e. all epigenetic features

The two tiers share the *structure* of this partition (the same nested chromHMM split, derived from the
same `chromhmm_hits_smaller` vicinity map) but **not the identical column sets**. Tier 1 partials
protocol, gene expression and TSS distance out as *unpenalised confounds*, while Tier 2 carries them as
ordinary penalised predictors (plus a protocol × expression interaction); and Tier 1 additionally excludes
provenance (`H3K*_source`, i.e. observed-vs-imputed) and QC columns that Tier 2 currently admits. So
compare *rankings* between tiers, not coefficient-for-coefficient membership.

Tier 1 screens **each space separately** and computes FDR **within each
(`transcript_filter`, `Event Type`, `feature_set`) group**. Reason: a signal confined to the exon's
immediate neighbourhood is *diluted* when hundreds of far chromHMM windows go into one omnibus ridge,
so an event with genuine local signal can be missed by `long` and found by `local`.

Two rules follow from the nesting (`local ⊂ short ⊂ long`) and must be respected when reading any table
below:

1. **Never sum per-set hit counts.** The sets are not independent tests of different things; they are
   three views of the same event. An event is a Tier-2 **hit if significant in *any* set**.
2. **Never compare raw R² across sets.** Dropping chromHMM lowers R² mechanically. `effect`
   (`real R² − mean null R²`) *is* comparable, because the matched null drops with it.

## Other conventions

- **Lambda selection** (Tier 2): `min` (lowest CV error) vs `1se` (most regularised model within 1 SE
  of the minimum). `1se` is the conservative read.
- **Expression bias** (Tier 2): a real-PSI model is *Epigenetic Only* if no `gene_expression`-derived
  feature survives robust cross-fold selection, else *Expression Bias*. Gene expression is a confound
  here, not a feature of interest — it co-varies with differentiation state, so an *Expression Bias*
  model's chromatin coefficients cannot be attributed to splicing regulation.
- **CV groups are ontology supergroups**, never random folds — a random split would let the same cell
  type appear in train and test and re-import the fingerprint confound through the back door.
- **`n_robust`** = features selected in *every* CV fold. `has_model == FALSE` means `n_robust == 0`:
  nothing survived, so the event has no interpretable model.

## Document structure (mirrors the order the code runs)

0. **Setup and table construction** — session, Tier-1 screen, Tier-2 per-event tables, and every
   derived table used later. No analysis, no plots.
1. **Tier 1 — the screen.** QC/calibration *first* (is the null honest?), then the significance
   results, then the per-event raw material behind them, and finally the **detection floor** (1.5):
   what effect size this screen could have found. A null result is not interpretable without it.
2. **Tier 2 — elastic net on hits.** How many events got models, which hyperparameters were chosen,
   and fit quality.
3. **What distinguishes significant events?** Variability, feature-count robustness, fold-to-fold
   consistency, lambda stability.
4. **Mark-level analysis.** Which marks are selected, stratified by significance; TSS-proximity
   control.
5. **PSI variability vs model quality.**
6. **Inter-model feature variation** — is mark selection structured across events, or idiosyncratic?

## How to read a report with no hits

**Numbers below are `stat_version` 5** (re-screened 2026-08-21–25 after `09-ml-shared.R`'s ridge λ
selection changed from GCV to exact LOOCV + a 1SE rule — see the version history at
`SCREEN_STAT_VERSION` in that file). They supersede an earlier `stat_version` 4 pass whose figures
this guide previously quoted; re-run `.claude/scratch/verify_09_2_reported_numbers.R` after any future
re-screen; every figure below is checked against it.

If Tier 1 yields no significant events, sections 2–6 are empty by construction (`en_available` is
`FALSE`, those chunks are skipped). The report is still **complete and interpretable** — the evidence
lives in section 1, and it rests on two things:

- a **flat** empirical-p histogram says the null is calibrated, so "no hits" is a real negative result
  rather than a broken test. On the completed screen `p_emp <= 0.05` runs **4.39–5.97%** per
  `(Event Type × feature set)` family against 5% expected, and the floor-free `p_pooled_z` runs
  3.77–6.20% (and 0.056–0.169% at the 0.001 level, vs 0.1% expected) — calibrated under both
  constructions in aggregate, so there is no *reservoir* of signal being lost to multiple testing. A
  p-histogram **piled up at 0** would instead mean the null is optimistic and nothing could be
  concluded. This calibration claim is about the bulk of each family, not about individual events —
  see the pooled-z caveat below;
- the **p gradient across signed-R² strata** (§1.1 check 2's table, cut at 0 / 0.1 / 0.2 / 0.5) is the
  fingerprint signature, and it is cleanest in the two large-*n* families: `median_p` falls
  monotonically as R² rises for **SE/local (0.716 → 0.030) and SE/long (0.522 → 0.127)**. It is
  noisier elsewhere — RI's three families split only ~1,775 tested events across 5 strata (the top
  stratum holds as few as **1** event for RI/short and **0** for RI/long), and SE/short dips
  non-monotonically at its second-highest stratum — so only 2 of the 6 families are strictly monotone
  under this statistic version. Even in the clean SE families the top (`> 0.5`) stratum never
  approaches significance (family min q 0.796 / 0.842). High R² buys almost nothing once the
  comparison is against a matched locus, which is what "fingerprint, not regulation" means
  quantitatively — the claim just now rests on SE's larger families rather than being uniform across
  all six.

**Do not use `effect` (or `mean(screen_R2)`) as that signature.** An earlier version of this guide said
the no-hit fingerprint was `effect ≈ 0` in every raw-R² stratum; the completed screen contradicts it.
`effect = screen_R2 − mean(null_R2)` is mean-based, so the heavy-tailed R² blow-ups documented in §1
corrupt it: it is positive for **79.3%** of rows (vs ~50% under a symmetric null), and the blow-up
fraction itself (`|screen_R2| > 1`) is now **50.1%** of all rows — up from 29.6% under the prior λ
selection, an unexplained side effect of the LOOCV+1SE change worth investigating separately (more
regularization did not reduce out-of-support extrapolation blow-ups; if anything the rate roughly
doubled). Use the rank-based `q` for inference, and `effect_bounded` / `effect_ccc` if a bounded effect
size is wanted.

**The primary hit rule (`q < 0.1`) still gives 0 hits, min q 0.551 (RI/local).** The floor-free
pooled-z sensitivity, which previously agreed with the primary rule at 0 hits, now surfaces **7
candidate events** (min `q_pooled_z` 0.0145) — the two constructions no longer agree, which is exactly
the scenario §1.3 below says to treat as candidates, not hits, pending a calibration check. Of the 7,
2 (small-*n*, `long`-space, `r2_confounds` in the hundreds-to-thousands range) look like the same
confound-projection instability tracked in `09-confound-projection-instability.md`; the other 5 (all
`SE/local`, n = 188–409, self-consistent `r2_confounds`) remain open and are not yet resolved one way
or the other. Do not read "0 primary hits" as "the sensitivity analysis also found nothing" — it no
longer does, and that gap is the most interesting open thread in this run.
**ID 38354's `local`-space result is the strongest of the 5** — beats all 200/200 of its own matched
controls, non-degenerate sample composition (409/415 cohort samples, 34 ontologies, no single-ontology
dominance), well-behaved regularization (no out-of-support blow-up, unlike its own `long`/`short`
spaces for the same event), and persisted across the GCV→LOOCV+1SE change (q_pooled_z 0.242 → 0.0145).
Survives every internal-consistency check run so far but is **not validated as a real hit** — that was
one event examined after seeing the ranking, not a pre-registered test, and no biological look at what
gene/exon it is has been done. Full write-up:
`.claude/scratch/investigate_pooled_z_candidates_v5_findings.md`.

That combination — a null calibrated in aggregate under both p constructions, no surviving event under
the primary rule, a handful of pooled-z candidates still under investigation, and a fingerprint p
gradient that is clean for SE and noisy for RI — is close to, but not as clean as, a textbook
confirmation of the manuscript's claim that reported epigenetic control of splicing is largely a
cell-type-identity confound. The qualitative conclusion is unchanged from the prior statistic version;
the margin by which it holds is smaller and the pooled-z candidates are unresolved.

**The positive counterpart is cohort-level, not event-level**, and the two do not contradict each other.
Under clean `seqnames`-grouped CV (`splicing_ml`, test AUROC, linear/xgb), sequence features alone
recover nearly the whole model — RI 0.809/0.784 and SE 0.950/0.954, against 0.871/0.855 and 0.957/0.963
for all features — while histone+DNAm alone reach only 0.591/0.587 and 0.641/0.659 over a 0.500 prior
baseline. Epigenetic marks therefore do carry *some* cohort-level signal; what this screen rules out is
that the signal is **locus-specific**. A matched null is blind to regulation shared across loci, so a
diffuse mechanism cancels in the null and reads here as "not locus-specific" — see
`revision/file-changes/09-psi-independent-controls.md` (OUTCOME + RESULT boxes) for the full numbers and
for what the design cannot see.

# 0. Setup and table construction

Loading and derived-table construction only, kept free of analysis so the analytical sections below
read top-to-bottom without side effects. Nothing here produces a plot.

`load-tier1-screen` also sets **`en_available`**, the gate for every Tier-2 chunk. It is deliberately
*data*-driven (are there readable elastic-net metric rows?) rather than file-driven (does a file
exist?), because a cancelled run leaves truncated files and skip-stubs behind, and gating on mere
presence would let the build chunks run and then crash on missing columns.

```{r setup}
library(tidymodels)
library(glmnet)
library(data.table)
library(ggplot2)
library(pbmcapply)
```

```{r load-session}
# Per-filter (matches 09-1 and everything from 02-2 on): one transcript_filter
# per invocation, chosen by TRANSCRIPT_FILTER (standalone falls back to the
# primary filter).
tf <- Sys.getenv(
  "TRANSCRIPT_FILTER",
  getOption("EpiATLAS_AS_PRIMARY_FILTER", "biotype_filtered")
)

# Load session objects produced by 09-1-ml-local.R (same tf)
sess <- readRDS(sprintf("processed_data/session_09_1_ml_local_%s.rds", tf))
event_dt <- sess$event_dt # trimmed: ID, Event Type, seqnames, Variability, transcript_filter
keep_rows <- sess$keep_rows_manual
rm(sess)
gc()

# `aggregating.rda` is GONE (removed by the per-filter refactor). aggregated_dt
# is per-filter; activeChromHMM (only other object 09-2 needs from that old
# workspace — see the chromhmm-region-annotation chunk below) is now persisted
# directly by 03 (same run as keep_rows_manual, so index-consistent).
aggregated_dt <- fread(
  sprintf("processed_data/aggregated_dt_filtered_%s.csv.gz", tf),
  stringsAsFactors = TRUE
)
activeChromHMM <- readRDS(sprintf("processed_data/activeChromHMM_%s.rds", tf))

# NB: no "restrict to biotype_filtered" subset needed anymore — event_dt (from
# 09-1's session, built off event_annotations_dt_{tf}) and aggregated_dt are
# already scoped to exactly `tf`.
```

```{r load-tier1-screen, cache = TRUE}
event_dir <- file.path("processed_data", "event_models", tf)

plot_dir_ev <- file.path(plot_dir, "event_models", tf)
dir.create(plot_dir_ev, showWarnings = FALSE, recursive = TRUE)

# Helper: aggregate per-event CSVs saved by worker. Robust to (a) no matching
# files and (b) individual unreadable/truncated files (a cancelled/partial run
# can leave a half-written .csv.gz) — pbmclapply otherwise returns error strings
# in place of tables and rbindlist(fill=TRUE) then errors with "Input is
# character…".
#
# When nothing readable is found it returns an empty data.table WITH THE EXPECTED
# COLUMNS for that file type (schema matches 09zz's writers). This is what keeps
# the report renderable at 0 Tier-1 hits AND keeps Positron "Run Above" working:
# knitr chunk options (eval = en_available) are ignored by interactive execution,
# so the EN chunks run regardless; giving them a 0-ROW-but-well-formed table lets
# their `[, := ]` transforms and ggplot() calls no-op on empty data instead of
# crashing on missing columns (e.g. `ID not found`).
.event_csv_schema <- list(
  best_params = c(
    "penalty",
    "mixture",
    ".config",
    "cv_mean_ccc",
    "cv_mean_rmse",
    "cv_mean_rsq_trad",
    "cv_std_err_ccc",
    "cv_std_err_rmse",
    "cv_std_err_rsq_trad",
    "cv_n_ccc",
    "cv_n_rmse",
    "cv_n_rsq_trad",
    "wflow_id",
    "lambda_selection",
    "ID"
  ),
  fit_metrics = c(
    ".metric",
    ".estimator",
    ".estimate",
    "wflow_id",
    "lambda_selection",
    "ID"
  ),
  all_metrics = c(
    "id",
    "penalty",
    "mixture",
    ".metric",
    ".estimator",
    ".estimate",
    ".config",
    "wflow_id",
    "ID"
  ),
  event_summary = c(
    "wflow_id",
    "model_type",
    "n_robust",
    "has_model",
    "expression_bias",
    "lambda_selection",
    "ID"
  ),
  nonzero_coefs = c("wflow_id", "feature", "coef", "lambda_selection", "ID"),
  fold_features = c("wflow_id", "fold", "feature", "lambda_selection", "ID")
)
.event_csv_numeric <- c(
  "penalty",
  "mixture",
  ".estimate",
  "coef",
  "n_robust",
  "cv_mean_ccc",
  "cv_mean_rmse",
  "cv_mean_rsq_trad",
  "cv_std_err_ccc",
  "cv_std_err_rmse",
  "cv_std_err_rsq_trad",
  "cv_n_ccc",
  "cv_n_rmse",
  "cv_n_rsq_trad"
)
.empty_event_dt <- function(pattern) {
  key <- names(.event_csv_schema)[
    vapply(
      names(.event_csv_schema),
      function(k) grepl(k, pattern, fixed = TRUE),
      logical(1)
    )
  ]
  if (length(key) != 1L) {
    return(data.table())
  }
  cols <- .event_csv_schema[[key]]
  data.table::as.data.table(setNames(
    lapply(cols, function(c) {
      if (c %in% .event_csv_numeric) numeric(0) else character(0)
    }),
    cols
  ))
}
read_event_csvs <- function(pattern) {
  files <- list.files(event_dir, pattern = pattern, full.names = TRUE)
  if (length(files) == 0L) {
    return(.empty_event_dt(pattern))
  }
  parts <- pbmclapply(files, function(f) {
    tryCatch(fread(f), error = function(e) NULL)
  })
  parts <- parts[vapply(parts, is.data.frame, logical(1))]
  if (length(parts) == 0L) {
    return(.empty_event_dt(pattern))
  }
  rbindlist(parts, fill = TRUE)
}

# Tier-1 ridge-screen results — significance (is the event epigenetically
# predictable beyond confounds?) is now decided HERE, not by the elastic-net
# rotation test. One row per event PER FEATURE SET (long/short/local): screen_R2/CCC,
# matched feature-rotation null summary, empirical p, effect (= real − null mean),
# and FDR q computed WITHIN each (transcript_filter, Event Type, feature_set) group
# by 09s-aggregate.R. So screen_dt is keyed by (ID, feature_set), NOT by ID alone —
# every join/facet below must carry feature_set or the 3 rows per event collide.
# `sig` is per (event, feature_set); an event is a Tier-2 hit if sig in ANY set.
# The 6 per-event elastic-net tables below are HITS-ONLY (Tier-2), all real PSI.
# Robust load so 09-2 renders on PARTIAL results too (screen still running / not
# yet aggregated). Prefer the aggregated screen_results.csv.gz (real Storey q from
# 09s-aggregate.R over ALL events); if it is absent, assemble a provisional
# screen_dt from whatever per-event screen rows exist and approximate q with BH —
# the final report always uses the true aggregate, this is only so intermediate
# renders don't hard-fail.
.screen_results_file <- file.path(event_dir, "screen_results.csv.gz")
if (file.exists(.screen_results_file)) {
  screen_dt <- fread(.screen_results_file)
} else {
  .screen_rows <- list.files(
    file.path(event_dir, "screen"),
    pattern = "^\\d+_screen\\.csv\\.gz$",
    full.names = TRUE # excludes _screen_null
  )
  if (length(.screen_rows) == 0L) {
    warning(
      "No screen_results.csv.gz and no per-event screen rows — Tier-1 sections ",
      "will be empty. Run the ridge screen (09s) first."
    )
    screen_dt <- data.table(
      ID = integer(),
      feature_set = character(),
      `Event Type` = character(),
      transcript_filter = character(),
      n_samples = integer(),
      R_used = integer(),
      screen_R2 = numeric(),
      screen_CCC = numeric(),
      null_R2_mean = numeric(),
      null_R2_sd = numeric(),
      p_emp = numeric(),
      effect = numeric(),
      note = character()
    )
  } else {
    warning(
      sprintf(
        "screen_results.csv.gz absent — assembling provisional screen_dt from %d ",
        length(.screen_rows)
      ),
      "per-event rows; q via BH proxy (PARTIAL, not the final Storey FDR)."
    )
    screen_dt <- rbindlist(pbmclapply(.screen_rows, fread), fill = TRUE)
    # STAMP GATE. This fallback path is used precisely while a screen is PARTIAL,
    # which is exactly when the directory can hold a mix of statistic versions —
    # so it must filter, unlike the aggregate path (09s-aggregate.R already
    # refuses to mix). Unstamped rows are version 1 (the pre-2026-07-29 leaky
    # statistic). Keep only the newest version present and say what was dropped.
    if (!"stat_version" %in% names(screen_dt)) {
      screen_dt[, stat_version := 1L]
    }
    screen_dt[, stat_version := suppressWarnings(as.integer(stat_version))]
    screen_dt[is.na(stat_version), stat_version := 1L]
    if (uniqueN(screen_dt$stat_version) > 1L) {
      .keep_v <- max(screen_dt$stat_version)
      warning(
        sprintf(
          "per-event screen rows span stat_version %s — keeping only v%d (%d of %d rows)",
          paste(sort(unique(screen_dt$stat_version)), collapse = "/"),
          .keep_v,
          sum(screen_dt$stat_version == .keep_v),
          nrow(screen_dt)
        ),
        immediate. = TRUE
      )
      screen_dt <- screen_dt[stat_version == .keep_v]
    }
  }
  # BH proxy mirrors the aggregator's GROUPED FDR (per tf × Event Type × feature_set),
  # so a provisional render doesn't silently pool the feature sets into one test.
  if (!"feature_set" %in% names(screen_dt)) {
    screen_dt[, feature_set := "long"]
  }
  if (!"transcript_filter" %in% names(screen_dt)) {
    screen_dt[, transcript_filter := tf]
  }
  screen_dt[, q := NA_real_]
  if (nrow(screen_dt) > 0L) {
    if (!"Event Type" %in% names(screen_dt)) {
      screen_dt[event_dt, on = "ID", `Event Type` := `Event Type`]
    }
    screen_dt[
      is.finite(p_emp),
      q := p.adjust(p_emp, method = "BH"),
      by = .(transcript_filter, `Event Type`, feature_set)
    ]
  }
}
screen_dt[, ID := as.integer(ID)]
# Pre-2026-07-29 aggregates are the single omnibus ridge over all epigenetic X —
# that IS the `long` set, so tag rather than drop (keeps old results renderable).
if (!"feature_set" %in% names(screen_dt)) {
  screen_dt[, feature_set := "long"]
}
# Ordered long → short → local (decreasing feature count) for stable facet order.
screen_feature_sets <- intersect(
  c("long", "short", "local"),
  unique(as.character(screen_dt$feature_set))
)
screen_dt[,
  feature_set := factor(
    as.character(feature_set),
    levels = c(
      screen_feature_sets,
      setdiff(unique(as.character(feature_set)), screen_feature_sets)
    )
  )
]
if (!"Event Type" %in% names(screen_dt) && nrow(screen_dt) > 0L) {
  screen_dt[event_dt, on = "ID", `Event Type` := `Event Type`]
}
screen_q_threshold <- 0.1
# Per (event, feature_set) significance. This MUST mirror 09s-aggregate.R's hit rule,
# which is `q < q_threshold` ALONE since 2026-08-01: the one-sided exceedance p already
# encodes direction, so the former `& effect > 0` clause was redundant, and on the raw
# statistic it was outlier-dominated (it passed ~3/4 of events for numerical reasons —
# see QC check 5). The aggregator's `EpiATLAS_AS_SCREEN_MIN_EFFECT` survives as an
# opt-in variance-collapse filter acting on `effect_bounded`, and defaults to no filter.
screen_dt[,
  sig := is.finite(q) & q < screen_q_threshold
]
# … and the per-EVENT roll-up (sig in ANY feature set) — this is what defines a
# Tier-2 hit in 09s-aggregate.R, and what the biology sections downstream split on.
screen_dt[, sig_any := any(sig %in% TRUE), by = ID]
screen_sig_ids <- sort(unique(screen_dt[sig %in% TRUE, ID]))

# en_available — is there USABLE Tier-2/EN data? Gates every EN-dependent chunk
# (via `eval = en_available`) so the Tier-1 (screen) report always renders, even
# at 0 hits or on a partial run. DATA-driven, not just file-presence: read the
# all_metrics (CV-metrics) tables robustly and require ≥1 row with a real metric
# column. This is deliberately stricter than "a file exists" — a run can leave
# only truncated files or skip-stubs (`data.table(ID=…)` with no `.metric`), and
# gating on mere presence would then let build-cv-tables run and crash on the
# missing ID/.metric columns.
.en_all_metrics <- read_event_csvs("^\\d+_all_metrics\\.csv\\.gz$")
en_available <- nrow(.en_all_metrics) > 0L &&
  ".metric" %in% names(.en_all_metrics)
```

# 1. Tier 1 — the ridge screen: is any event predictable beyond its matched null?

This section decides significance for **every** event; nothing downstream can add an event the screen
did not select. `09s-ridge-screen.R` computes, per event and per feature space, a closed-form
group-leave-one-ontology-out ridge **partial-R²** — the variance in PSI explained by epigenetics *after*
protocol, gene expression and TSS distance are absorbed as unpenalised confounds — then re-computes that
same statistic on **matched control loci** — same event type, same variability tier, different
chromosome — while keeping this event's PSI. `screen_rotations` is per-Event-Type: **818 for RI**
(its full eligible pool) and **200 for SE**. That gives an empirical p, and `09s-aggregate.R` turns the
p-values into FDR **q** within each (`transcript_filter`, `Event Type`, `feature_set`) group.

**An event is significant when `q < threshold`** — that single condition, nothing else. The one-sided
exceedance p already encodes *direction* (it is the mass of the null at or above the real R²), so the
`effect > 0` clause this rule used to carry was redundant; it was also actively misleading on the raw
statistic, where `effect = real R² − mean(null R²)` is mean-based and passed ~3/4 of events for purely
numerical reasons (§1.1 check 5). Dropped 2026-08-01; see `09s-aggregate.R`.

### The order of this section is the argument

The QC/calibration chunk comes **first, on purpose**. A hit list from a miscalibrated null is
meaningless, so the null is audited before any result is read:

1. **`screen-diagnostics`** — is the null honest, is the fingerprint pattern present, is the
   negative-R² tail explained?
2. **`screen-significance-plots`** — the actual per-group results and hit counts.
3. **`screen-floor-sensitivity`** — could the test have rejected anything at all? The p-floor
   arithmetic and the floor-free pooled-z sensitivity p (§1.3).
4. **`screen-null-histograms`** — the raw material: where each real R² falls inside its own null.

### Hypotheses this section discriminates between

- **H1 — genuine event-specific epigenetic control.** Some events' local chromatin predicts their PSI
  better than unrelated loci do. *Signature:* an excess of events in the leftmost p bin over uniform,
  and survival after FDR.
- **H2 — cell-type fingerprint only.** All apparent predictability is shared cell-type identity.
  *Signature:* flat p histogram and no survivors after FDR, with p never approaching significance even
  in the highest raw-R² stratum. High-R² events are the clearest fingerprint cases, not the best hits.
  (Read this off `median_p` per R² stratum, **not** off `effect` — see check 5.)
- **H3 — broken/optimistic null** (a nuisance hypothesis that must be excluded before H1 or H2 can be
  read). *Signature:* p piling up toward 0 monotonically across all bins. Note that an implausibly
  large fraction of `effect > 0` indicates a corrupted `effect`, not an optimistic null — check 5.

H2 vs H1 is the manuscript's central question; H3 is why the diagnostics run first.

## 1.1 Screen QC: is the null honest? (read this before the results)

Five checks, each with an expected shape and a specific failure mode:

**(1) Null calibration — empirical-p distribution, per (Event Type x feature set).**
*Expected under no genome-wide signal:* flat (uniform), because a matched control locus should beat the
real locus about as often as chance dictates.
*Expected under genuine signal:* flat, **plus** a bump in the leftmost bin — the hits.
*Failure:* a monotone pile-up toward 0 across all bins means the null is systematically too weak
(controls handicapped somehow — e.g. mismatched sample sets or fewer usable features), and **no hit list
from this run can be trusted**. Checked per **family** — `(Event Type, feature_set)`, the unit the FDR is actually computed in — not
per feature set alone. SE outnumbers RI ~18:1, so a pooled panel is an SE panel and an RI-specific
failure would be invisible in it; the two types' nulls are also measurably different (KS p = 0.0072 on
their standardised null z). Every table and panel in this subsection is grouped that way.

**(2) Fingerprint check — `effect` and p within raw-R² strata.**
*Hypothesis:* if predictability were event-specific, the best-predicted events would also be the ones
that most beat their null, so `effect` should rise with raw R².
*Fingerprint prediction (H2):* the top stratum still fails to reject — median p well above the FDR
threshold and no survivor after BH — even though it holds the best-predicted events in the genome.
**Read that off `median_p`, not off `effect`.** On the completed screen `effect` is positive in
**80–100%** of events in every stratum above the `<=0` baseline, in every family, with exactly one
exception: RI/short's top (`> 0.5`) bucket holds a single event with a negative effect — illustrating
how noisy a 1-event stratum is, not a reversal of the pattern. Its *magnitude* is unstable across
strata (from near 0 to several tens), which is check 5's corruption signature, not evidence for H1: raw
`effect` is mean-based and the R² blow-ups drag it. `median_p` behaves better: it falls cleanly with R²
in the two large families, SE/local (0.716 → 0.030) and SE/long (0.522 → 0.127), and the `> 0.5`
stratum never approaches significance in any family (family min q 0.551–0.842).
*Deduction from that pattern:* raw R² is measuring cell-type identity, and any published ranking of
"most epigenetically predictable events" built on R² alone is a ranking of fingerprint strength.

**(3) Negative-R² tail vs sample size.**
Group-LOO R² can go sharply negative; the question is whether that is a bug or expected instability.
*Expected:* the negative tail concentrates in **small-n** events, where holding out a whole ontology
group removes a large share of the data. It should also shrink from `long` → `local` as the p ≫ n
pressure eases.
*Deduction if very negative R² appears at large n, or does not ease as features are dropped:* suspect a
data or confound-partition problem rather than small-sample noise.

**(4) Silently voided events (`stat_version` ≥ 2 only).** The leak-fixed statistic fits the confound design
per fold on training rows, but its constant-column guard is evaluated over *all* rows — so a fold whose
training subset makes a confound constant yields a rank-deficient design, an `NA` out-of-fold residual, and
an `NA` R² for the whole event. `screen_CCC` does **not** fail with it: `.ccc()` filters non-finite pairs
and returns a value computed on the survivors, which looks valid.
*Signature:* `screen_R2` NA with a **finite** `screen_CCC`, `R_used == 0`, and no `note`.
*Expected:* a small percentage, concentrated at low `n_samples`.
*Deduction:* those events are **excluded, not tested** — so they cannot produce false positives, but the
loss is non-random (it hits exactly the low-*n* stratum check 3 examines), and any `screen_CCC` reported for
them is computed on an unknown subset and must not be quoted.

**(5) Is `effect` still a usable effect size?** `effect = screen_R2 − mean(null_R2)` is **mean**-based, so a
single exploded null value corrupts it; `p_emp` is **rank**-based and does not care.
*Expected if healthy:* the fraction with `effect > 0` is broadly consistent with the fraction at small
`p_emp`, and median `effect` does not flip sign across `|screen_R2|` strata.
Note that `effect > 0` at ~50% alongside `p_emp ≤ 0.05` at ~5% is **healthy**, not contradictory —
`effect` is a symmetric difference, so half of it is positive under the null. The pathology is `effect > 0`
drifting far from 50% (**79.3%** on the completed screen) *together with* a material fraction of
`|screen_R2| > 1` (**50.1%** — over half of all rows, up from 29.6% under the prior GCV λ selection;
the LOOCV+1SE change made this worse, not better, and that is not yet explained), which means the
null mean is being dragged by exploded out-of-fold predictions.
*Deduction if that happens:* `effect` has stopped carrying information about direction and is positive for
numerical reasons — which is why it is no longer part of the hit rule at all. Use `q` alone for inference,
treat raw `effect` as unreportable (`effect_bounded` / `effect_ccc` are the bounded substitutes), and see
`revision/file-changes/09-event-models-redesign.md` TODO-2.

```{r screen-diagnostics}
sd_fin <- screen_dt[is.finite(p_emp) & is.finite(screen_R2) & is.finite(effect)]
sd_fin[event_dt, Variability := `Variability`, on = "ID"]

# EVERY diagnostic below is grouped by (Event Type x feature_set) — the FDR family,
# NOT feature_set alone.
#
# This matters more than it looks. SE outnumbers RI ~18:1 (32,368 vs 1,776 events),
# so anything pooled across event types is an SE diagnostic with RI invisible: RI's
# null could be badly miscalibrated and the pooled p-histogram would still look
# flat. And the two event types demonstrably differ — a KS test on their
# leave-one-out standardised null z gives p = 0.0072, with q99.99 of 9.73 (SE) vs
# 5.93 (RI). This is the same trap that made a pooled-z reference read as calibrated
# at ratio 1.01 while actually being 1.43x anti-conservative for SE and 0.41x
# conservative for RI: majority-class dominance hiding a minority-class failure.
# Diagnose in the unit the inference is done in.
.diag_by <- c("Event Type", "feature_set")

# Coverage per family: rotations used, n_features (the point of the split), and
# degenerate-event notes (NA = fit fine).
cat("Screen coverage per Event Type x feature set:\n")
# NB every summary must have a TYPE-STABLE result across groups: median() on an
# integer column returns integer for odd-length groups but double for even-length
# ones, which makes data.table abort with "Column N ... is type 'double' but
# expecting type 'integer'" as soon as the groups differ in parity. Hence the
# explicit as.double()/as.integer() casts — do not drop them.
print(screen_dt[,
  .(
    n_events = .N,
    finite_p = sum(is.finite(p_emp)),
    median_features = as.double(median(n_features, na.rm = TRUE)),
    R_used_median = as.double(median(R_used, na.rm = TRUE)),
    R_used_below_max = sum(R_used < max(R_used, na.rm = TRUE), na.rm = TRUE)
  ),
  by = .diag_by
][order(`Event Type`, feature_set)])
cat("Degenerate-event notes (NA = screened cleanly):\n")
print(screen_dt[, .N, by = c(.diag_by, "note")][order(
  `Event Type`,
  feature_set,
  -N
)])

# (1) NULL CALIBRATION — p_emp should be ~flat under the null. A broken/optimistic
# null piles p toward 0 across all bins; genuine signal shows only as a modest
# excess in the leftmost bin over the uniform (dashed) line.
# Calibration must be checked PER FAMILY (Event Type x feature set) — each is its
# own set of tests with its own FDR, so one family's null can be broken while
# another's is fine, and a pooled panel would hide it (see .diag_by above).
n_bins <- 100L
.unif_dt <- sd_fin[, .(y = .N / n_bins), by = .diag_by]
print(
  ggplot(sd_fin, aes(x = p_emp)) +
    geom_histogram(
      breaks = seq(0, 1, 1 / n_bins),
      fill = "grey70",
      color = "white"
    ) +
    geom_hline(
      data = .unif_dt,
      aes(yintercept = y),
      linetype = "dashed",
      color = "firebrick"
    ) +
    facet_grid(
      factor(`Event Type`, levels = to_analyze) + Variability ~ feature_set,
      scales = "free_y"
    ) +
    theme_bw() +
    labs(
      title = "Tier-1 null calibration: empirical p per Event Type x feature set",
      x = "Empirical p (matched feature-rotation null)",
      y = "Number of events",
      caption = paste(
        "Dashed = uniform expectation (per panel). Flat = well-calibrated null; excess",
        "in the leftmost bin = genome-wide signal; a monotone pile-up toward 0 would",
        "indicate an optimistic (broken) null. Read RI on its own: it is ~5% of events,",
        "so pooling it with SE would hide an RI-specific failure entirely."
      )
    )
)
ggsave(
  file.path(plot_dir_ev, "screen_pvalue_calibration.pdf"),
  width = 3 + 3 * max(1L, uniqueN(sd_fin$`Event Type`)),
  height = 2 + 2.2 * max(1L, uniqueN(sd_fin$feature_set)),
  limitsize = FALSE
)

# (2) FINGERPRINT — effect & p within raw-R2 strata. High raw R2 with effect ≈ 0
# and non-small p = cell-type fingerprint, not event-specific signal (binned
# complement to screen_effect_vs_r2.pdf). READ THE `median_p` COLUMN, not the effect
# columns: on the real screen `effect` is outlier-corrupted (check 5 below), so the
# "effect ≈ 0" half of that signature is not observable and `median_p` carries the
# argument on its own. NB the cut here is on SIGNED screen_R2 (0/0.1/0.2/0.5), which is
# NOT check 5's abs(screen_R2) magnitude cut — the two give opposite-looking p gradients
# and must not be compared to each other.
sd_fin[,
  r2_stratum := cut(
    screen_R2,
    c(-Inf, 0, 0.1, 0.2, 0.5, Inf),
    labels = c("<=0", "(0,0.1]", "(0.1,0.2]", "(0.2,0.5]", ">0.5")
  )
]
cat(
  "Fingerprint check — effect & p within raw-R2 strata, per Event Type x feature set:\n"
)
print(sd_fin[,
  .(
    n = .N,
    median_R2 = round(median(screen_R2), 3),
    median_effect = round(median(effect), 4),
    pct_effect_pos = round(100 * mean(effect > 0), 1),
    median_p = round(median(p_emp), 3)
  ),
  by = c(.diag_by, "r2_stratum")
][order(`Event Type`, feature_set, r2_stratum)])

# (3) SMALL-n INSTABILITY — very-negative group-LOO R2 concentrates in small
# events (expected), not a bug. y clipped at the 1st percentile for readability.
print(
  ggplot(sd_fin, aes(x = n_samples, y = screen_R2, color = sig)) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
    geom_point(alpha = .3, size = .7) +
    scale_color_manual(
      values = c(`TRUE` = "firebrick", `FALSE` = "grey60"),
      labels = c(`TRUE` = "significant", `FALSE` = "not sig"),
      name = NULL
    ) +
    coord_cartesian(
      ylim = c(quantile(sd_fin$screen_R2, 0.01, na.rm = TRUE), 1)
    ) +
    facet_grid(feature_set ~ `Event Type`) +
    theme_bw() +
    labs(
      title = "Screen R2 vs sample size (very-negative R2 = small-n LOO instability)",
      x = "n samples (event)",
      y = "Group-LOO screen R2",
      caption = paste(
        "Per feature set: the p >> n regime eases as features are dropped, so the",
        "negative-R2 tail should shrink from long -> short -> local."
      )
    )
)
ggsave(
  file.path(plot_dir_ev, "screen_r2_vs_nsamples.pdf"),
  width = 3 + 3 * max(1L, uniqueN(sd_fin$`Event Type`)),
  height = 2 + 2.2 * max(1L, uniqueN(sd_fin$feature_set)),
  limitsize = FALSE
)

# (4) SILENTLY VOIDED EVENTS. See revision/file-changes/09-event-models-redesign.md
# TODO-1. A fold whose train rows make a confound constant gives a rank-deficient
# design -> NA out-of-fold residual -> NA screen_R2 for the WHOLE event, while
# .ccc() filters non-finite pairs and still returns a value. So the signature of the
# defect is: screen_R2 NA, screen_CCC FINITE, R_used == 0, note absent.
cat("\n=== (4) event losses and their attribution ===\n")
screen_dt[, `:=`(
  .r2_na = !is.finite(screen_R2),
  .ccc_ok = is.finite(screen_CCC)
)]
loss_tab <- screen_dt[,
  .(
    n_rows = .N,
    finite_p = sum(is.finite(p_emp)),
    pct_finite_p = round(100 * mean(is.finite(p_emp)), 2),
    voided_silently = sum(.r2_na & .ccc_ok), # R2 NA but CCC finite = the defect
    voided_both = sum(.r2_na & !.ccc_ok),
    with_note = sum(!is.na(note))
  ),
  by = .diag_by
][order(`Event Type`, feature_set)]
print(loss_tab)
if (sum(loss_tab$voided_silently) > 0L) {
  cat(sprintf(
    "!! %d row(s) have screen_R2 = NA with a FINITE screen_CCC and no note.\n   Those events were EXCLUDED, not tested; their screen_CCC is computed on an\n   unknown subset of samples and must not be quoted. n_samples of affected rows:\n",
    sum(loss_tab$voided_silently)
  ))
  print(summary(screen_dt[.r2_na & .ccc_ok, n_samples]))
  cat("   (a low-n concentration is the expected signature; see TODO-1)\n")
}
cat("Degenerate `note` values actually recorded:\n")
print(screen_dt[!is.na(note), .N, by = note][order(-N)])

# (5) IS `effect` STILL A USABLE EFFECT SIZE? See TODO-2. `effect` is mean-based
# (screen_R2 - mean(null_R2)) so one exploded null corrupts it; p_emp is rank-based
# and immune. When the out-of-fold R2 blows up, `effect > 0` stops carrying
# information — it holds for most events for numerical reasons. This is why the hit
# rule no longer uses it (dropped 2026-08-01); the check stays, as a health signal.
cat(
  "\n=== (5) numeric health of screen_R2, and whether `effect` still means anything ===\n"
)
mag <- rbindlist(lapply(c(1, 10, 100, 1000), function(t) {
  data.table(
    threshold = t,
    n_rows = sum(abs(sd_fin$screen_R2) > t),
    pct = round(100 * mean(abs(sd_fin$screen_R2) > t), 2)
  )
}))
cat(
  "Rows with |screen_R2| above each threshold (an honest out-of-fold R2 should sit\nnear or below 1; large values are prediction blow-ups, not signal):\n"
)
print(mag)
sd_fin[, .blow := abs(screen_R2) > 10]
cat(
  "\nDo blow-ups track sample size / lambda? (small n + small lambda = the known cause)\n"
)
print(sd_fin[,
  .(
    n_rows = .N,
    median_n_samples = as.double(median(n_samples, na.rm = TRUE)),
    median_lambda = signif(median(lambda, na.rm = TRUE), 3)
  ),
  by = .blow
][order(.blow)])
.pct_eff_pos <- 100 * mean(sd_fin$effect > 0, na.rm = TRUE)
.pct_p_small <- 100 * mean(sd_fin$p_emp <= 0.05, na.rm = TRUE)
cat(sprintf(
  "\neffect > 0: %.1f%% of rows | p_emp <= 0.05: %.2f%%\n",
  .pct_eff_pos,
  .pct_p_small
))
sd_fin[,
  .r2_mag := cut(
    abs(screen_R2),
    c(-Inf, 1, 10, 100, Inf),
    labels = c("<=1", "(1,10]", "(10,100]", ">100")
  )
]
print(sd_fin[,
  .(
    n = .N,
    median_effect = signif(median(effect), 3),
    median_null_mean = signif(median(null_R2_mean), 3),
    pct_effect_pos = round(100 * mean(effect > 0), 1),
    median_p = round(median(p_emp), 3)
  ),
  by = .r2_mag
][order(.r2_mag)])
# Trigger on the RIGHT signature. Under a well-behaved null, `effect > 0` sits near
# 50% (it is a symmetric difference) while p_emp <= 0.05 sits near 5% — those two are
# consistent, so their ratio is NOT evidence of anything. The pathology is `effect > 0`
# drifting far from 50% (the null MEAN being dragged by exploded values) together with
# a material blow-up fraction.
.blow_pct <- 100 * mean(abs(sd_fin$screen_R2) > 1, na.rm = TRUE)
.eff_skew <- abs(.pct_eff_pos - 50)
if (.blow_pct > 5 || .eff_skew > 15) {
  cat(sprintf(
    "!! %.1f%% of rows have |screen_R2| > 1 and `effect > 0` holds for %.1f%% (expected\n   ~50%% under a symmetric null). `effect` is outlier-corrupted and is NOT a usable\n   effect size here -- which is why it is no longer part of the hit rule at all (the\n   `effect > 0` clause was dropped 2026-08-01). Use q alone for inference; do not quote\n   `effect` or mean(screen_R2); `effect_bounded`/`effect_ccc` are the bounded\n   substitutes. See revision/file-changes/09-event-models-redesign.md TODO-2.\n",
    .blow_pct,
    .pct_eff_pos
  ))
} else {
  cat(sprintf(
    "effect looks usable: %.1f%% of rows |screen_R2| > 1, `effect > 0` at %.1f%% (~50%%\nexpected under a symmetric null).\n",
    .blow_pct,
    .pct_eff_pos
  ))
}
# p_emp is rank-based, so it should survive the blow-ups — verify, since the whole
# FDR rests on it.
.pmin <- min(sd_fin$p_emp, na.rm = TRUE)
.rmax <- max(sd_fin$R_used, na.rm = TRUE)
cat(sprintf(
  "\nmin p_emp = %.6g; expected floor 1/(R_used+1) = %.6g -> %s\n",
  .pmin,
  1 / (.rmax + 1),
  if (abs(.pmin - 1 / (.rmax + 1)) < 1e-9) "OK" else "UNEXPECTED, investigate"
))
screen_dt[, c(".r2_na", ".ccc_ok") := NULL]
sd_fin[, c(".blow", ".r2_mag") := NULL]
```

## 1.2 Screen results: hit counts, p-values, and fingerprint unmasking

Only read this once section 1.1 has established the null is calibrated.

**Per-group counts.** Reported by `(Event Type, feature_set)` — the exact grouping the FDR was computed
within, so these *are* the per-family hit counts. The roll-up table beside it gives `n_sig_any_set`
(the Tier-2 hit definition) and `n_sig_all_sets`. Because the spaces are nested, expect
`n_sig_all_sets ≤ n_sig_any_set` and expect the two to be *close* if signal is broad, far apart if
signal is space-specific.

**Plot 2 — empirical p per event.** Same calibration logic as §1.1, now split by event type as well, so
a signal restricted to one event type is visible. SE and RI differ in sample size and in baseline
predictability, so their p distributions should be read separately rather than pooled.

**Plot 3 — `effect` vs raw R²: the fingerprint unmasking.** The single most important plot in the
report.
*Expected under H2 (fingerprint):* a horizontal band at `effect ≈ 0` that extends all the way to the
highest R², with significant points absent or scattered.
*Expected under H1 (real signal):* points lifting off the `effect = 0` line, and — importantly — the
lift-off need **not** be at high R². Genuine event-specific signal is expected at *modest* R² with a
clearly-beaten null, precisely because it is not riding the cell-type axis.
*Read this plot with check 5 in hand.* On the completed screen neither expectation is legible, because
the y-axis quantity is corrupted: raw `effect` is mean-based, 79.3% of rows are positive against ~50%
under a symmetric null, and the clipping to `|R²|, |effect| ≤ 1` now drops **over half** of all rows
(the panel's printed count gives the exact figure for this run — check it, since the blow-up rate moved
from 29.6% to 50.1% between statistic versions and this clip fraction moves with it). So the panel
shows where the surviving points sit, not whether anything lifted off.
*Deduction:* **`q` identifies a hit, not `effect`** — `effect` left the hit rule on 2026-08-01. Use this
plot as a shape diagnostic and §1.2's hit table plus §1.3's floor arithmetic as the verdict; if a bounded
version of the vertical axis is wanted, `effect_bounded` / `effect_ccc` exist for it.

**Plot 4 (new) — does narrowing the feature space recover diluted signal?** Per event, `effect` in the
narrower space vs `effect` in `long`, with the identity line drawn.
*Hypothesis being tested:* the omnibus `long` ridge dilutes exon-local signal across hundreds of far
chromHMM windows, so genuinely local signal should sit **above** the identity line.
*Deduction if a substantial fraction lies above the line:* the per-feature-set split is doing real work
and a single omnibus screen under-reports; the accompanying `median_delta` / `pct_improved` table
quantifies it. **Caveat, same as Plot 3:** these are differences of the *raw* `effect`, so they inherit
its outlier corruption and 63% of the points fall outside the clip window. On the completed screen the
narrower spaces do sit above the line (`pct_improved` 61–85%), but no family gained a single hit from it
— so read this as "the split changes the statistic", not "the split recovers signal".
*Deduction if points cluster on the line:* the far chromHMM windows were contributing neither signal nor
meaningful dilution, and the three spaces are largely redundant for significance purposes.

```{r screen-significance-plots}
# Screen significance plots — built DIRECTLY from screen_dt (one row per event),
# so they always render, even at 0 hits / on a partial run (independent of the EN
# tables). Previously these were computed off emp_p_dt, which broadcasts screen
# values across the EN (feature_set × lambda) rows and is therefore empty when
# there are no hits.
screen_plot_dt <- screen_dt[, .(
  ID,
  feature_set,
  emp_p = p_emp,
  q,
  effect,
  screen_R2,
  sig
)]
screen_plot_dt[event_dt, on = "ID", `Event Type` := `Event Type`]
screen_plot_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
.n_fs <- max(1L, uniqueN(screen_plot_dt$feature_set))

# Significance summary per (Event Type × feature_set) — the exact grouping the FDR
# was computed within, so these counts are the per-group hit counts.
print(screen_plot_dt[,
  .(
    n_sig = sum(sig, na.rm = TRUE),
    n_total = sum(!is.na(q)),
    pct_sig = round(100 * mean(sig, na.rm = TRUE), 1)
  ),
  by = .(`Event Type`, feature_set)
][order(`Event Type`, feature_set)])

# Roll-up: an event counts as a hit if significant in ANY feature set (Tier-2's
# definition). Reported alongside the per-set counts because the sets are nested
# (local ⊂ short ⊂ long), so per-set counts are not independent and don't add up.
print(screen_plot_dt[,
  .(n_sig_sets = sum(sig, na.rm = TRUE)),
  by = .(`Event Type`, ID)
][,
  .(
    n_events = uniqueN(ID),
    n_sig_any_set = uniqueN(ID[n_sig_sets > 0L]),
    n_sig_all_sets = uniqueN(ID[n_sig_sets == .n_fs])
  ),
  by = `Event Type`
][order(`Event Type`)])

# Plot 2: screen empirical-p per event, event type × feature set.
print(
  ggplot(
    screen_plot_dt[is.finite(emp_p)],
    aes(x = emp_p, fill = `Event Type`)
  ) +
    geom_histogram(bins = 100, alpha = .7, position = "identity") +
    geom_vline(
      xintercept = screen_q_threshold,
      linetype = "dashed",
      color = "red"
    ) +
    facet_grid(`Event Type` ~ feature_set, scales = "free_y") +
    theme_bw() +
    labs(
      title = "Tier-1 ridge-screen empirical p-value (per event, per feature set)",
      x = "Empirical p (matched feature-rotation null)",
      y = "Number of events",
      caption = paste(
        "Rows = feature set (long = all epigenetic; short = drop far chromHMM;",
        "local = drop all chromHMM). FDR is computed within each row × column cell.",
        "Significant = q < threshold AND effect (real - mean null R2) > 0."
      )
    )
)
ggsave(
  file.path(plot_dir_ev, "screen_pvalue.pdf"),
  width = 10,
  height = 2 + 2.2 * .n_fs,
  limitsize = FALSE
)

# Plot 3: effect vs raw R² — the fingerprint unmasking. High raw R² with effect
# ≈ 0 = cell-type fingerprint (a random locus predicts as well). Colour is the real
# hit rule, q < threshold alone — `effect` is diagnostic here, not a criterion.
# CLIPPED, deliberately. `effect = screen_R2 - mean(null_R2)`, so while the v2
# out-of-fold R2 blows up (see QC check 5) the y-axis is dominated by its own x-axis:
# cor(effect, screen_R2) measured 0.988 across all rows vs -0.05 restricted to
# |R2| <= 1. Unclipped, this panel is one dot plus a diagonal streak and shows NO
# fingerprint structure. The excluded count is printed, not hidden.
.p3 <- screen_plot_dt[is.finite(effect) & is.finite(screen_R2)]
.p3_lim <- 1
.p3_out <- .p3[abs(screen_R2) > .p3_lim | abs(effect) > .p3_lim, .N]
cat(sprintf(
  "Plot 3 clipped to |R2| and |effect| <= %g: %d of %d rows (%.1f%%) outside, not drawn.\n",
  .p3_lim,
  .p3_out,
  nrow(.p3),
  100 * .p3_out / max(1L, nrow(.p3))
))
print(
  ggplot(.p3, aes(x = screen_R2, y = effect, color = sig)) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
    geom_point(alpha = .4, size = .8) +
    coord_cartesian(
      xlim = c(-.p3_lim, .p3_lim),
      ylim = c(-.p3_lim, .p3_lim)
    ) +
    facet_grid(`Event Type` ~ feature_set, scales = "free_y") +
    scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey60")) +
    theme_bw() +
    labs(
      title = "Screen effect (real - null) vs raw R2: fingerprint unmasking",
      x = "Raw group-LOO screen R2",
      y = "Effect = real R2 - mean null R2",
      color = "Significant\n(q < threshold)",
      caption = paste(
        "High raw R2 with effect near 0 = cell-type fingerprint (random control",
        "loci predict as well). CLIPPED axes -- see the printed count of excluded",
        "rows. While the out-of-fold R2 blows up, `effect` is outlier-corrupted and",
        "is NOT comparable across rows; read q, not effect. See TODO-2 in",
        "revision/file-changes/09-event-models-redesign.md."
      )
    )
)
ggsave(
  file.path(plot_dir_ev, "screen_effect_vs_r2.pdf"),
  width = 10,
  height = 2 + 2.2 * .n_fs,
  limitsize = FALSE
)

# Does narrowing the feature space actually recover diluted local signal? Per event,
# compare effect across sets — the motivation for splitting the screen at all. A
# positive local-minus-long effect means the omnibus ridge was diluting that event.
if (.n_fs > 1L) {
  eff_wide <- dcast(
    screen_plot_dt[is.finite(effect)],
    ID + `Event Type` ~ feature_set,
    value.var = "effect"
  )
  .pairs <- intersect(c("short", "local"), names(eff_wide))
  if ("long" %in% names(eff_wide) && length(.pairs) > 0L) {
    eff_delta <- melt(
      eff_wide,
      id.vars = c("ID", "Event Type", "long"),
      measure.vars = .pairs,
      variable.name = "feature_set",
      value.name = "effect"
    )[is.finite(effect) & is.finite(long)]
    eff_delta[, delta := effect - long]
    cat("Effect gain from narrowing the feature space (vs `long`):\n")
    print(eff_delta[,
      .(
        n = .N,
        median_delta = round(median(delta), 4),
        pct_improved = round(100 * mean(delta > 0), 1)
      ),
      by = .(`Event Type`, feature_set)
    ][order(`Event Type`, feature_set)])

    .ed_lim <- 1
    .ed_out <- eff_delta[abs(long) > .ed_lim | abs(effect) > .ed_lim, .N]
    cat(sprintf(
      "Clipped to |effect| <= %g on both axes: %d of %d points (%.1f%%) not drawn.\n",
      .ed_lim,
      .ed_out,
      nrow(eff_delta),
      100 * .ed_out / max(1L, nrow(eff_delta))
    ))
    print(
      ggplot(eff_delta, aes(x = long, y = effect, color = `Event Type`)) +
        coord_cartesian(
          xlim = c(-.ed_lim, .ed_lim),
          ylim = c(-.ed_lim, .ed_lim)
        ) +
        geom_abline(
          slope = 1,
          intercept = 0,
          linetype = "dashed",
          color = "grey40"
        ) +
        geom_hline(yintercept = 0, linetype = "dotted", color = "grey60") +
        geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
        geom_point(alpha = .3, size = .7) +
        facet_grid(feature_set ~ `Event Type`) +
        theme_bw() +
        labs(
          title = "Does a narrower feature space recover diluted local signal?",
          x = "Effect in `long` (all epigenetic features)",
          y = "Effect in the narrower feature set",
          caption = paste(
            "Dashed = equality; points above it = the narrower space beats the omnibus",
            "ridge. NB the dilution rationale predicts local >= short >= long, but the",
            "measured ordering is short > long > local (local is the only set that gets",
            "WORSE), which argues against dilution as the mechanism. Axes clipped; see",
            "the printed excluded count, and TODO-2 on effect being outlier-corrupted."
          )
        )
    )
    ggsave(
      file.path(plot_dir_ev, "screen_effect_by_feature_set.pdf"),
      width = 10,
      height = 2 + 2.2 * length(.pairs),
      limitsize = FALSE
    )
  }
}
```

## 1.3 Could the test have found anything? The p-floor and the floor-free sensitivity p

The exceedance p cannot go below **`1/(R_used + 1)`** — 0.00498 at SE's 200 rotations, 0.00122 at RI's
818. BH at `q < threshold` only passes a p-value at rank *k* if `p ≤ k·q/m`. Those two facts can conflict:
if a family's floor sits *above* what BH allows, **no event in it can pass FDR regardless of how strong
its signal is**, and a 0-hit result there says nothing on its own. On the completed screen that is the
situation in **every one of the six families** (`floor_can_pass = FALSE` throughout): SE needed ≥ 1,608
events tied at the floor and has 191 / 208 / 199 (long / short / local), RI needed ≥ 22 and has
2 / 2 / 2 — the same count in every one of RI's three feature spaces, unlike the earlier statistic
version where RI's counts differed sharply across spaces (0 / 3 / 1). Whether that is the same 2 events
recurring in all three spaces or 2 different events per space has not been checked.

This is a distinct nuisance hypothesis from the three in §1, and it must be excluded before the negative
result can be read:

> **H4 — the test lacked the resolution to reject anything.** *Signature:* every candidate event tied at
> exactly `1/(R_used + 1)`, and `floor_can_pass = FALSE` in the table below. A flat p-histogram does
> **not** exclude H4 — it only excludes a broken null (H3).

Raising the rotation count cannot fix it, and the reason is **not** simply that the pool is too small —
SE's matched pool runs to ~15,000 loci. `k_at_floor` counts events that beat **every** one of their `R`
controls, so a larger pool is a *higher* bar: if an event sits at quantile *u* among its controls,
P(beat all R) ≈ *u*^R decays geometrically, while the BH requirement `p_floor·m/q` eases only like 1/R.
Measured across this project's own rotation counts (under the prior, `stat_version` 4 statistic), the
requirement eased 9× (198.5 → 22) while RI/local's `k_at_floor` fell 25× (25 → 1) — the gap got ~3×
worse. Under the current `stat_version` 5 statistic at the same R = 818, RI/local's `k_at_floor` is 2,
not 1 — the λ-selection change moved this specific count, though the conclusion it supports (RI's ~818
control ceiling is far below what its k needs) is unaffected either way. RI is additionally capped: its
~818 eligible controls are set by the RI event count itself, so no rotation setting reaches the
thousands its measured *k* would need. Dropping low-variance events does not help either (the BH
threshold `k·q/m` is scale-invariant, since *k* shrinks with *m*). What does fix it is a second,
**floor-free** p-value: standardise each event's statistic by its own matched null
(`z = (R² − mean null)/sd null`) and read the p off a reference of leave-one-out-standardised **null** z
values pooled **within the FDR family** `(transcript_filter, Event Type, feature_set)`. Only the
dimensionless deviation *shape* is borrowed — each event's own R² level and scale stay local, so a
cell-type-fingerprint event (high null **mean**) still gets a small z and is not rescued by this.

The family stratification is not cosmetic: pooling across Event Types measures as 1.43× anti-conservative
for SE and 0.41× conservative for RI (α = 10⁻⁴), errors that cancel to a deceptive 1.01 when averaged over
a 95%-SE sample. Within-family it is 1.00 (SE) / 0.91–1.05 (RI). A median/MAD scale is 2.6–5.2×
anti-conservative and must not be substituted.

**This is a sensitivity analysis, not the primary result.** `sig` and the Tier-2 hit list stay on the
exceedance p, which assumes only that an event's own controls are exchangeable with it; the pooled-z p
additionally assumes a family-common deviation shape. Two independent constructions agreeing is the
claim; the pooled-z alone is the weaker of the two.

*Deduction if both give 0 hits (the current state):* the negative result is not an artifact of resolution —
it survives a construction with ~4 orders more of it.
*Deduction if pooled-z yields hits where the exceedance p could not:* those are candidates the floor was
hiding, but they inherit the extra assumption, so check the calibration ratio for that family and the
effect-size guard below before treating any as real.

```{r screen-floor-sensitivity}
# (a) p-FLOOR ARITHMETIC per FDR family -- can the primary statistic pass at all?
screen_min_effect <- getOption("EpiATLAS_AS_SCREEN_MIN_EFFECT", 0)
.has_tf <- "transcript_filter" %in% names(screen_dt)
.by <- if (.has_tf) {
  c("transcript_filter", "Event Type", "feature_set")
} else {
  c("Event Type", "feature_set")
}
floor_tab <- screen_dt[
  is.finite(p_emp),
  {
    m <- .N
    fl <- 1 / (max(R_used, na.rm = TRUE) + 1)
    k <- sum(p_emp <= fl + 1e-12 & effect > screen_min_effect, na.rm = TRUE)
    .(
      m_tests = m,
      p_floor = signif(fl, 3),
      k_at_floor = k,
      bh_allows = if (k > 0L) {
        signif(k * screen_q_threshold / m, 3)
      } else {
        NA_real_
      },
      floor_can_pass = k > 0L && fl <= k * screen_q_threshold / m,
      k_needed = ceiling(fl * m / screen_q_threshold)
    )
  },
  by = .by
]
cat("Can the primary (exceedance) p clear FDR in each family?\n")
print(floor_tab[order(`Event Type`, feature_set)])
if (any(!floor_tab$floor_can_pass)) {
  cat(
    "\n!! floor_can_pass = FALSE above: in those families NO event could pass FDR\n",
    "   via the exceedance p at this rotation count, whatever its signal. A 0-hit\n",
    "   result there is uninterpretable WITHOUT the sensitivity p below.\n"
  )
}

# (b) PRIMARY vs SENSITIVITY, side by side.
if (!"p_pooled_z" %in% names(screen_dt)) {
  cat("\nNo `p_pooled_z` column. Which of the two reasons applies:\n")
  if (!file.exists(.screen_results_file)) {
    cat(
      "  -> screen_results.csv.gz does NOT exist, so screen_dt came from the PER-EVENT\n",
      "     fallback path, which computes only a BH-proxy q. The sensitivity p needs the\n",
      "     null sidecars AND the family-stratified reference, and only 09s-aggregate.R\n",
      "     builds those. Run it once the screen completes -- nothing is stale here.\n"
    )
  } else {
    cat(
      "  -> the aggregate exists but predates the sensitivity p; re-run 09s-aggregate.R\n",
      "     (needs the screen/*_screen_null.csv.gz sidecars).\n"
    )
  }
} else if (!any(is.finite(screen_dt$p_pooled_z))) {
  cat(
    "\n`p_pooled_z` present but all NA -- the null sidecars were missing or unreadable\n",
    "when 09s-aggregate.R ran.\n"
  )
} else {
  cmp <- screen_dt[
    is.finite(p_emp),
    .(
      n_tested = .N,
      min_p_primary = signif(min(p_emp, na.rm = TRUE), 3),
      min_p_pooled_z = signif(min(p_pooled_z, na.rm = TRUE), 3),
      hits_primary = sum(
        is.finite(q) & q < screen_q_threshold & effect > screen_min_effect,
        na.rm = TRUE
      ),
      hits_pooled_z = sum(
        is.finite(q_pooled_z) &
          q_pooled_z < screen_q_threshold &
          effect > screen_min_effect,
        na.rm = TRUE
      )
    ),
    by = .by
  ]
  cat("\nHit counts and resolution, primary vs floor-free sensitivity p:\n")
  print(cmp[order(`Event Type`, feature_set)])

  # (c) EFFECT-SIZE GUARD. A large z can come from a collapsed denominator rather
  # than signal: null_R2_sd is never 0 (min ~1e-5) but can be proportionally tiny,
  # so an event with R2 = 0.005 and null sd = 3e-4 scores z = 17 on no real signal.
  # Effect size -- not null sd -- is the instrument that separates these.
  top <- screen_dt[is.finite(p_pooled_z)][order(p_pooled_z)][
    seq_len(min(15L, .N)),
    .(
      ID,
      `Event Type`,
      feature_set,
      n_samples,
      screen_R2 = signif(screen_R2, 3),
      null_R2_sd = signif(null_R2_sd, 3),
      z = round(z_screen, 1),
      effect = signif(effect, 3),
      p_pooled_z = signif(p_pooled_z, 3),
      q_pooled_z = signif(q_pooled_z, 3),
      artifact_suspect = effect < 0.01
    )
  ]
  cat(
    "\nTop events by the sensitivity p (artifact_suspect = effect < 0.01,\n",
    "i.e. z inflated by a small null sd rather than by signal):\n"
  )
  print(top)
  cat(sprintf(
    "\n%d of the top %d have effect < 0.01. screen_min_effect is currently %g;\n raise EpiATLAS_AS_SCREEN_MIN_EFFECT to exclude them from hit calls.\n",
    sum(top$artifact_suspect),
    nrow(top),
    screen_min_effect
  ))

  print(
    ggplot(
      screen_dt[is.finite(p_pooled_z) & is.finite(p_emp)],
      aes(x = p_emp, y = p_pooled_z)
    ) +
      geom_point(alpha = .2, size = .6) +
      scale_y_log10() +
      geom_hline(
        yintercept = floor_tab$bh_allows[1],
        linetype = "dashed",
        color = "firebrick"
      ) +
      facet_grid(feature_set ~ `Event Type`) +
      theme_bw() +
      labs(
        title = "Floor-free sensitivity p vs floored primary p",
        x = "Exceedance p (vertical stripe at its 1/(R+1) floor)",
        y = "Pooled-z p (log scale)",
        caption = paste(
          "The primary p piles into a vertical stripe at its floor; the sensitivity p",
          "spreads those events out. Dashed = the BH threshold the primary p could not",
          "reach. Points below it are events the floor was hiding."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "screen_floor_sensitivity.pdf"),
    width = 10,
    height = 2 + 2.2 * max(1L, uniqueN(screen_dt$feature_set)),
    limitsize = FALSE
  )
}
```

## 1.4 Per-event null distributions: the raw material behind each p-value

Everything above summarises the null by its mean and SD. This shows the null itself: for a selected
`(event, feature set)` pair, the histogram of that event's `R_used` matched-control R² values (818 for
RI, 200 for SE) with the event's own real R² drawn as a vertical line. The empirical p *is* the tail mass to the right of that line
(`p = (1 + #{null ≥ real}) / (1 + R_used)`), so the plot and the p-value are the same fact shown two ways.

*Expected for a hit:* the line sits in the far right tail, clear of the bulk.
*Expected for a fingerprint event:* the line sits inside the bulk — often at a **high absolute R²**,
which is exactly the point: the controls reach the same R².
*Deduction if a null distribution is bimodal or extremely wide:* the matched-control pool is
heterogeneous for that event (e.g. controls with very different feature availability), so its p-value is
noisier than the summary table suggests — worth checking `R_used` for that event.

A panel is an `(event, feature set)` **pair**, not an event, since the same event has a different real R²
and a different null in each space; seeing one event across its three spaces side by side is the useful
comparison. Panels are capped (all significant pairs first, then top-by-effect) and only the selected
events' sidecar files are read — the full set is tens of thousands of files.

```{r screen-null-histograms}
# Per-event null-distribution histograms: real screen_R2 (vertical line) over the
# R_used matched feature-rotation null R2 values (818 for RI, 200 for SE). These are
# the SAME quantity
# the empirical p is computed from (p = (1 + #{null >= real}) / (1 + R_used)), so
# a real line far in the right tail = small p = event-specific epigenetic signal;
# a real line buried in the bulk = fingerprint / no signal.
# Which panels to draw is decided from screen_dt ALONE, so the file selection happens
# FIRST: the full sidecar set is ~34k files / hundreds of MB at genome scale, and
# reading all of it to then plot <=24 panels dominated the render. Read only the
# events actually drawn.
#
# A facet is an (event, feature_set) PAIR, not an event — the same event has a
# different real R2 and a different null in each feature space, which is exactly the
# comparison worth seeing. Draw every significant pair, then top-by-effect, capped so
# the full-run set can't explode the grid.
max_facets <- 24L
.key <- c("ID", "feature_set")
screen_pairs <- copy(screen_dt)[, feature_set := as.character(feature_set)]
hit_pairs <- screen_pairs[sig %in% TRUE, .(ID, feature_set)]
top_pairs <- screen_pairs[is.finite(effect)][order(-effect), .(ID, feature_set)]
plot_pairs <- unique(rbindlist(list(hit_pairs, top_pairs)))
plot_pairs <- plot_pairs[seq_len(min(max_facets, nrow(plot_pairs)))]

null_files <- file.path(
  event_dir,
  "screen",
  sprintf("%d_screen_null.csv.gz", unique(plot_pairs$ID))
)
null_files <- null_files[file.exists(null_files)]

if (nrow(plot_pairs) == 0L || length(null_files) == 0L) {
  message(
    "No screen_null sidecars for the selected events — skipping per-event ",
    "null histograms. (Re-run 09s-ridge-screen.R to populate them.)"
  )
} else {
  null_long <- rbindlist(lapply(null_files, fread), fill = TRUE)
  null_long[, ID := as.integer(ID)]
  # Sidecars written before the per-feature-set screen hold the omnibus null = `long`.
  if (!"feature_set" %in% names(null_long)) {
    null_long[, feature_set := "long"]
  }
  null_long[, feature_set := as.character(feature_set)]
  null_long <- null_long[is.finite(null_R2)]

  n_omitted <- nrow(unique(screen_pairs[
    is.finite(effect),
    .(ID, feature_set)
  ])) -
    nrow(plot_pairs)
  if (n_omitted > 0L) {
    message(sprintf(
      "Null histograms: drawing %d (event × feature set) panel(s) (%d significant + top-by-effect); %d omitted.",
      nrow(plot_pairs),
      nrow(hit_pairs),
      n_omitted
    ))
  }

  nd <- null_long[plot_pairs, on = .key, nomatch = NULL]
  # Per-(event, feature_set) real value + significance + a compact facet label.
  ev_lab <- screen_pairs[
    plot_pairs,
    on = .key,
    nomatch = NULL
  ][, .(ID, feature_set, screen_R2, p_emp, q, sig)]
  ev_lab[event_dt, on = "ID", `Event Type` := `Event Type`]
  ev_lab[,
    facet := sprintf(
      "ID %d (%s) [%s]\nR2=%.3f p=%.2g q=%.2g%s",
      ID,
      `Event Type`,
      feature_set,
      screen_R2,
      p_emp,
      q,
      fifelse(sig %in% TRUE, " *", "")
    )
  ]
  nd[ev_lab, on = .key, facet := facet]
  vlines <- ev_lab[, .(ID, feature_set, facet, screen_R2, sig)]

  print(
    ggplot(nd, aes(x = null_R2)) +
      geom_histogram(bins = 30, fill = "grey70", color = NA) +
      geom_vline(
        data = vlines,
        aes(xintercept = screen_R2, color = sig %in% TRUE),
        linewidth = 0.7
      ) +
      facet_wrap(~facet, scales = "free") +
      scale_color_manual(
        values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"),
        labels = c(`TRUE` = "significant", `FALSE` = "not sig"),
        name = "real R2 (event)"
      ) +
      theme_bw(base_size = 10) +
      labs(
        title = "Tier-1: real partial-R2 vs its matched feature-rotation null (per event × feature set)",
        x = "R2 (grey = null control loci; line = this event's real R2)",
        y = "Number of null control loci",
        caption = paste(
          "One panel per (event, feature set): each grey histogram = that event's",
          "matched-control null R2 computed in THAT feature space, with the vertical",
          "line that event's own real R2 in the same space. Right-tail line = small",
          "empirical p = event-specific signal. * = FDR-significant (q < threshold",
          "FDR within tf x Event Type x feature set)."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "screen_null_histograms.pdf"),
    width = 14,
    height = 10
  )
}
```

```{r load-en-tables, eval = en_available, cache = TRUE, dependson = "load-tier1-screen"}
if (en_available) {
  # Tier-2 elastic-net per-event tables (HITS ONLY). This whole chunk is skipped
  # when en_available is FALSE (no Tier-1 hits) — see load-data — so downstream
  # EN chunks (also eval = en_available) never touch undefined/empty tables.

  # Best hyperparameters per (event, workflow)
  best_params_table <- read_event_csvs("^\\d+_best_params\\.csv\\.gz$")
  best_params_table[, ID := as.integer(ID)]
  best_params_table[,
    c("feature_set", "response") := if (.N > 0L) {
      tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  ]
  best_params_table[,
    response_type := fifelse(response == "PSI", "Real", "Rotated")
  ]
  best_params_table[,
    lambda_selection := factor(lambda_selection, levels = c("min", "1se"))
  ]

  # In-sample fit metrics per (event, workflow)
  fit_metrics_table <- read_event_csvs("^\\d+_fit_metrics\\.csv\\.gz$")
  fit_metrics_table[, ID := as.integer(ID)]
  fit_metrics_table[,
    c("feature_set", "response") := if (.N > 0L) {
      tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  ]
}
```

```{r build-cv-tables, eval = en_available, cache = TRUE, dependson = c("load-tier1-screen", "load-en-tables")}
if (en_available) {
  # cv_metrics_table: per-fold CV metrics at selected config, tagged by lambda_selection
  # cv_rsq: mean CV R² per (event, workflow) — used downstream by paired-rotation-analysis
  #         and n-robust-significance
  cv_metrics_table <- read_event_csvs("^\\d+_all_metrics\\.csv\\.gz$")
  cv_metrics_table[, ID := as.integer(ID)]
  cv_metrics_table <- cv_metrics_table[
    best_params_table[, .(ID, wflow_id, .config, lambda_selection)],
    on = .(ID, wflow_id, .config),
    nomatch = NULL
  ]
  cv_metrics_table[,
    c("feature_set", "response") := if (.N > 0L) {
      tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  ]
  cv_metrics_table[,
    response_type := fifelse(response == "PSI", "real", "rotated")
  ]
  cv_metrics_table[,
    feature_set := factor(feature_set, levels = c("local", "short", "long"))
  ]

  cv_rsq <- cv_metrics_table[
    .metric == "rsq_trad",
    .(mean_cv_rsq = mean(.estimate, na.rm = TRUE)),
    by = .(ID, wflow_id, lambda_selection)
  ]
}
```

```{r build-summary-tables, eval = en_available, cache = TRUE, dependson = c("load-tier1-screen", "load-en-tables", "build-cv-tables")}
if (en_available) {
  # event_summary_table: one row per (event, workflow, lambda_selection).
  # Includes n_robust, expression_bias, mean CV R², rotated_hit flag.
  event_summary_table <- read_event_csvs("^\\d+_event_summary\\.csv\\.gz$")
  event_summary_table[, ID := as.integer(ID)]
  event_summary_table[,
    c("feature_set", "response") := if (.N > 0L) {
      tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  ]
  event_summary_table[,
    response_type := fifelse(response == "PSI", "real", "rotated")
  ]
  event_summary_table[
    event_dt,
    on = "ID",
    c("Event Type", "Variability", "seqnames") := .(
      `Event Type`,
      Variability,
      seqnames
    )
  ]
  event_summary_table[
    cv_rsq,
    on = .(ID, wflow_id, lambda_selection),
    mean_cv_rsq := mean_cv_rsq
  ]

  # "Rotated Hit": did this event's FEATURES get borrowed as a background control
  # for another event? (Rotation is now FEATURE-rotation — 09zz fits this event's
  # PSI against other matched events' real feature matrices — so the rotated
  # `response` token is the control event whose FEATURES were used, not whose PSI.
  # Semantics flipped from PSI-borrowing to feature-borrowing; the ID set is
  # computed the same way.)
  rotation_ids <- event_summary_table[
    response_type == "rotated",
    unique(as.integer(response))
  ]
  event_summary_table[
    response_type == "real",
    rotated_hit := fifelse(ID %in% rotation_ids, "Hit by Rotation", "Not Hit")
  ]

  # `rotation_rank` (slot index within an event's rotated models) was computed here
  # and in coef_table below. REMOVED 2026-07-30: Tier-2 fits real PSI only — the
  # rotated null moved to Tier-1 — so it was constant 0 and its only consumers
  # reduced to a plain count with an NA standard error. Nothing reads it now.

  event_summary_table[,
    feature_set := factor(feature_set, levels = c("local", "short", "long"))
  ]
  event_summary_table[,
    lambda_selection := factor(lambda_selection, levels = c("min", "1se"))
  ]
  event_summary_table[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  event_summary_table[,
    `Event Type` := factor(`Event Type`, levels = to_analyze)
  ]
}
```

```{r build-feature-coef-tables, eval = en_available, cache = TRUE, dependson = c("load-tier1-screen", "load-en-tables", "build-summary-tables")}
if (en_available) {
  # feature_table: features selected per CV fold per workflow.
  # coef_table: robust non-zero coefficients per workflow.
  # expression_bias is propagated from the real-PSI model for each event via the
  # ID + lambda_selection join, so rotated coef rows also carry the real model's bias label.

  feature_table <- read_event_csvs("^\\d+_fold_features\\.csv\\.gz$")
  feature_table[, ID := as.integer(ID)]
  feature_table[,
    c("feature_set", "response") := if (.N > 0L) {
      tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  ]
  feature_table[,
    response_type := fifelse(response == "PSI", "real", "rotated")
  ]
  feature_table[
    event_dt,
    on = "ID",
    c("Event Type", "Variability") := .(`Event Type`, Variability)
  ]
  feature_table[
    event_summary_table[
      response_type == "real",
      .(ID, lambda_selection, expression_bias)
    ],
    on = .(ID, lambda_selection),
    expression_bias := expression_bias
  ]

  coef_table <- read_event_csvs("^\\d+_nonzero_coefs\\.csv\\.gz$")
  coef_table[, ID := as.integer(ID)]
  coef_table[,
    c("feature_set", "response") := if (.N > 0L) {
      tstrsplit(wflow_id, "_", fixed = TRUE, keep = c(1L, 2L))
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  ]
  coef_table[, response_type := fifelse(response == "PSI", "real", "rotated")]
  coef_table[
    event_dt,
    on = "ID",
    c("Event Type", "Variability") := .(`Event Type`, Variability)
  ]
  # expression_bias from real-PSI model propagated to rotated rows via ID+lambda_selection join
  coef_table[
    event_summary_table[
      response_type == "real",
      .(ID, lambda_selection, expression_bias)
    ],
    on = .(ID, lambda_selection),
    expression_bias := expression_bias
  ]
  coef_table[, let(
    c("mark", "region"),
    if (.N > 0L) {
      tstrsplit(feature, ";", fixed = TRUE)
    } else {
      list(character(), character()) # tstrsplit() errors on 0-row input
    }
  )]
  coef_table[,
    coefficient_sign := fifelse(
      coef > 0,
      "positive coefficient",
      "negative coefficient"
    )
  ]
  coef_table[,
    Response := factor(
      fifelse(response_type == "real", "Real", "Rotated"),
      levels = c("Real", "Rotated")
    )
  ]
  coef_table[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
  for (.dt in list(feature_table, coef_table)) {
    .dt[,
      feature_set := factor(feature_set, levels = c("local", "short", "long"))
    ]
    .dt[,
      lambda_selection := factor(lambda_selection, levels = c("min", "1se"))
    ]
    .dt[,
      Variability := factor(
        sub(" Variability", "", Variability, fixed = TRUE),
        levels = c("High", "Low")
      )
    ]
    .dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
  }
}
```

```{r build-emp-p-table, eval = en_available, cache = TRUE, dependson = c("load-tier1-screen", "load-en-tables", "build-summary-tables")}
if (en_available) {
  # emp_p_dt — the bridge table between Tier 1 and Tier 2: it attaches each event's
  # SCREEN significance to each elastic-net workflow that was fit for it. Every
  # "do significant events differ in X?" analysis downstream reads this table, so it
  # is built once here rather than inside an analysis chunk.
  #
  # KEY: the join is on (ID, feature_set), NOT on ID. The screen runs in the same
  # three feature spaces the elastic-net uses, so each EN workflow gets the
  # significance of the space it was actually fit in. Joining on ID alone would both
  # triplicate every EN row (3 screen rows per event) and cross feature spaces —
  # attributing, say, a `local` screen result to a `long` model. Only
  # lambda_selection is broadcast, since the screen has no lambda analogue (it is a
  # single closed-form ridge with lambda chosen by GCV, not a regularisation path).
  #
  # `sig` = significant in THIS feature space; `sig_any` = significant in any space
  # (the Tier-2 hit definition). Use `sig` when the analysis is per-feature-set,
  # `sig_any` when it is per-event.
  fit_metrics_table[,
    response_type := fifelse(response == "PSI", "real", "rotated")
  ]
  cv_metrics_table[,
    response_type := fifelse(response == "PSI", "real", "rotated")
  ]

  fs_lambda_combos <- unique(
    best_params_table[, .(ID, feature_set, lambda_selection)]
  )
  .screen_join <- screen_dt[, .(
    ID,
    feature_set = as.character(feature_set),
    emp_p = p_emp,
    q,
    effect,
    screen_R2,
    screen_CCC,
    null_R2_mean,
    sig,
    sig_any
  )]
  emp_p_dt <- merge(
    fs_lambda_combos[, feature_set := as.character(feature_set)][],
    .screen_join,
    by = c("ID", "feature_set"),
    all.x = TRUE
  )
  emp_p_dt[event_dt, on = "ID", `Event Type` := `Event Type`]
  emp_p_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
  # NB: EN-side ordering (local -> long, increasing window size), the reverse of
  # screen_dt's (long -> local, decreasing feature count). Downstream EN plots were
  # built against this ordering; screen plots keep theirs.
  emp_p_dt[,
    feature_set := factor(feature_set, levels = c("local", "short", "long"))
  ]
  cat(sprintf(
    "emp_p_dt: %d EN workflow rows, %d unique events, %d rows significant in their own feature space\n",
    nrow(emp_p_dt),
    uniqueN(emp_p_dt$ID),
    sum(emp_p_dt$sig %in% TRUE)
  ))
}
```

## 1.5 Detection floor: what effect size could this screen have found?

A null result is only interpretable against a stated sensitivity. Section 1.3 establishes that the
*exceedance* p is floored by the rotation count and therefore cannot reject a single event at FDR 0.1
in either event type — that is a property of the test's resolution, not of the data. This section
answers the complementary and harder question: **holding the resolution fixed, how large would a real
epigenetic effect have had to be for this screen to see it?**

**How the floor is measured.** The pipeline re-runs the *real* Tier-1 worker on a stratified sample of
real events, with the response replaced by a synthetic one built from that event's own feature matrix:

$$y_{\text{lat}} = \rho \cdot \text{(standardised linear combination of } s \text{ real feature columns)} + \sqrt{1-\rho^{2}}\cdot\varepsilon$$

Everything else — the confound set, the leave-one-ontology-supergroup-out folds, the λ rule, the matched
control rotations — is untouched, because the injected run *is* the production worker (`09s-ridge-screen.R`
under `EPIATLAS_AS_SCREEN_INJECT`), not a re-implementation. A separate implementation would drift from
`prep_event()`'s fold logic with nothing to detect it.

Three things about this design decide how the numbers below should be read:

- **ρ is a knob, not the reported quantity.** The map from ρ to achieved out-of-fold R² is strongly
  nonlinear and mode-dependent (at ρ = 1: 1.000 under a raw Gaussian response, 0.809 after a logistic
  squash, 0.518 after rank-mapping). **Always quote the achieved R², never ρ.**
- **The response is rank-mapped onto the real PSI marginal.** The injected `PSI` is the event's own
  observed PSI values re-ordered to follow $y_{\text{lat}}$, so the boundary atoms (39.3% of PSI values
  are exactly 1.0) and the bounded support are preserved exactly. This costs real recoverable signal —
  the ceiling is ≈ 0.52, not 1.0 — and that is the point: an unbounded Gaussian response would report a
  floor for a response shape PSI never has.
- **Detection is scored on `p_pooled_z`, not `p_emp`.** The exceedance p is floored, so at the sample
  size the floor sweep can afford it would measure the rotation count rather than the signal. Each
  injected event is standardised against the **real** screen's pooled-z reference — every real event's
  leave-one-out standardised null z, within its own `(Event Type × feature_set)` family — and spliced
  into that family's real p-vector to get a genuine BH rank.

**Scope.** The sweep runs on `local` only. In the `long` space the injection fails to recover a signal
that is by construction present in that space's own columns, so `long` has no measurable floor to
report — see the anchor table. That is itself a finding: the screen's power lives in the
event-overlapping features.

```{r tier1-detection-floor, echo=FALSE, message=FALSE, warning=FALSE}
.floor_file <- file.path(event_dir, "floor", "floor_results.csv.gz")
floor_dt <- if (file.exists(.floor_file)) fread(.floor_file) else data.table()

if (!nrow(floor_dt)) {
  cat("Detection floor not available: ", .floor_file, " is missing.\n",
    "Run `snakemake ", .floor_file, "` (the floor is in `rule all`).\n", sep = "")
} else {
  # `local` only, and this is a hard filter, not a default view: at rho = 1 the injection
  # fails to recover a signal that is by construction present in `long`'s own columns, so
  # `long` has no measurable floor to report. Pooling the spaces would average a real floor
  # against an undefined one. See the "Scope" paragraph above.
  floor_dt <- floor_dt[feature_set == "local"]
  cat(sprintf("floor sweep: %d rows, %d events, rho in {%s}\n\n",
    nrow(floor_dt), uniqueN(floor_dt$ID),
    paste(sort(unique(floor_dt$inject_rho)), collapse = ", ")))

  # ANCHORS FIRST. rho = 0 is a pure null: its detection rate must sit near the nominal
  # 10%, or the pooled-z reference is miscalibrated and nothing below it is trustworthy.
  # rho = 1 must be detected: if the strongest injectable signal is invisible, the sweep
  # is measuring a broken harness, not the screen's sensitivity.
  anch <- floor_dt[inject_rho %in% c(0, 1), .(
    n = .N,
    med_R2 = round(median(screen_R2, na.rm = TRUE), 4),
    med_p_pooled_z = signif(median(p_pooled_z, na.rm = TRUE), 3),
    det_rate_pct = round(100 * mean(detected, na.rm = TRUE), 1)
  ), by = .(inject_rho)][order(inject_rho)]
  cat("=== anchors (rho = 0 must be near-null; rho = 1 must be detected) ===\n")
  print(anch)

  cat("\n=== detection rate vs achieved out-of-fold R2, by stratum ===\n")
  print(floor_dt[, .(
    n = .N,
    med_R2 = round(median(screen_R2, na.rm = TRUE), 4),
    det_rate_pct = round(100 * mean(detected, na.rm = TRUE), 1)
  ), by = .(`Event Type`, Variability, inject_rho)][order(`Event Type`, Variability, inject_rho)])

  # Power is NOT uniform across events. A blown-up statistic (|R2| > 1, 50.1% of real
  # rows) means the leave-one-group-out fit extrapolated catastrophically on at least one
  # fold; such an event cannot be rescued by a stronger signal, so pooling it with the
  # well-behaved events understates the floor for the events that actually have power.
  cat("\n=== same, stratified by blow-up |R2| > 1 (power is not uniform) ===\n")
  print(floor_dt[, .(
    n = .N,
    med_R2 = round(median(screen_R2, na.rm = TRUE), 4),
    det_rate_pct = round(100 * mean(detected, na.rm = TRUE), 1)
  ), by = .(`Event Type`, blowup, inject_rho)][order(`Event Type`, blowup, inject_rho)])

  cat("\n=== THE FLOOR: smallest achieved R2 detected at q_pooled_z < 0.1 ===\n")
  print(floor_dt[detected == TRUE, .(
    min_detected_R2 = round(min(screen_R2, na.rm = TRUE), 4),
    med_detected_R2 = round(median(screen_R2, na.rm = TRUE), 4),
    n_detected = .N
  ), by = .(`Event Type`, Variability)][order(`Event Type`, Variability)])
}
```

```{r tier1-detection-floor-curve, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.2, fig.width=8}
if (nrow(floor_dt)) {
  # Bin on the ACHIEVED R2, not on rho -- rho is the knob and the rho -> R2 map is
  # nonlinear, so a curve against rho would be a curve against an arbitrary parameter.
  .fc <- floor_dt[is.finite(screen_R2) & !is.na(detected)]
  .fc[, r2_bin := cut(screen_R2,
    breaks = c(-Inf, 0, 0.05, 0.1, 0.2, 0.3, 0.5, Inf),
    labels = c("<=0", "0-0.05", "0.05-0.1", "0.1-0.2", "0.2-0.3", "0.3-0.5", ">0.5"))]
  .fcs <- .fc[, .(det = mean(detected), n = .N), by = .(`Event Type`, r2_bin)][n >= 5L]
  if (nrow(.fcs)) {
    print(
      ggplot(.fcs, aes(x = r2_bin, y = 100 * det, group = `Event Type`, colour = `Event Type`)) +
        geom_line() +
        geom_point(aes(size = n)) +
        geom_hline(yintercept = 100 * screen_q_threshold, linetype = "dashed", colour = "grey40") +
        scale_size_continuous(range = c(1, 4)) +
        labs(
          x = "achieved out-of-fold R² (binned)", y = "detected at q_pooled_z < 0.1 (%)",
          title = "Tier-1 detection floor (local features, rank-mapped response)",
          subtitle = sprintf("dashed line = nominal %g; bins with n < 5 dropped", screen_q_threshold)
        ) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1))
    )
  } else {
    cat("too few injected rows per bin to draw the detection curve\n")
  }
}
```

**How to state this in the manuscript.** The floor is the smallest achieved out-of-fold R² that this
screen detects at FDR 0.1, per event type and variability tier, on the `local` feature space, under a
response that preserves PSI's own marginal distribution. Real events fall *below* that floor. The
correct claim is therefore: *no event's epigenetic signal exceeds the matched-locus null by more than
R² ≈ (floor), which the screen would have detected*. It is **not** *epigenetics does not influence
splicing* — a diffuse mechanism shared across loci cancels in a matched-locus null by construction
(section 1.1), and effects below the floor remain entirely possible.

# 2. Tier 2 — elastic net on the hits: which features, and how well?

Everything from here on runs **only on Tier-1 hits**, on **real PSI only** (no rotated responses), and is
gated by `en_available`. If section 1 produced no hits, the rest of this report is empty — that is the
designed behaviour, not a failure.

**What Tier 2 is and is not.** It is a *description* of the signal Tier 1 already certified: which
features, with what sign, how stably across folds. It is **not** a second significance test. Reading an
elastic-net CV R² as evidence of epigenetic control would reintroduce exactly the error Tier 1 exists to
prevent, because a good CV R² is just as achievable by fingerprint as by regulation.

**Why coefficients here need care even for genuine hits.** The features are strongly collinear
(neighbouring windows of the same mark, and all marks with cell-type identity). The elastic net will pick
one of a correlated group and zero the rest, and which one it picks can change with the fold — so
*absence* of a mark is weak evidence, and the `n_robust` / fold-consistency diagnostics in section 3 are
the guard against over-reading any single coefficient.

## 2.1 How many events have an interpretable model?

A model "exists" only if at least one feature is robustly selected in **every** CV fold
(`n_robust > 0`, i.e. `has_model == TRUE`). Events failing that are excluded from all downstream
feature-level analysis.

*Expected:* `lambda.min` (weaker regularisation) yields more selected features and therefore more events
with a model; `lambda.1se` is stricter and drops more events to `has_model == FALSE`.
*Deduction if the `min` → `1se` attrition is severe:* the selected features are only marginally
informative — they survive at the CV optimum but not once regularisation is tightened by one standard
error, so their interpretation should be treated as provisional.
*Deduction if attrition is mild:* the selected features carry enough signal to survive conservative
regularisation, which strengthens their interpretation.

Note on `rotated_hit` in these tables: it flags whether an event's **features** were borrowed as a
matched control for some other event (Tier 1's rotation is a *feature* rotation, not a PSI permutation).
It is bookkeeping about control reuse, not a result about that event.

```{r counting-events, eval = en_available}
if (en_available) {
  event_counts <- rbindlist(
    lapply(unique(event_summary_table$lambda_selection), function(sel) {
      rbindlist(
        lapply(unique(event_summary_table$feature_set), function(fs) {
          this_summary <- event_summary_table[
            lambda_selection == sel & feature_set == fs
          ]
          modeled_ids <- this_summary[
            response_type == "real" & has_model == TRUE,
            unique(ID)
          ]
          rbindlist(
            list(
              Quantified = aggregated_dt[
                ID %in% keep_rows,
                .(N = uniqueN(ID)),
                by = `Event Type`
              ],
              Filtered = event_dt[
                ID %in% keep_rows,
                .(.N),
                by = .(`Event Type`)
              ],
              `No Model` = aggregated_dt[
                !ID %in% modeled_ids & ID %in% keep_rows,
                .(N = uniqueN(ID)),
                by = `Event Type`
              ],
              `Model` = this_summary[
                response_type == "real" & has_model == TRUE,
                .(.N),
                by = .(`Event Type`)
              ],
              `Epigenetic\nOnly Model` = this_summary[
                response_type == "real" &
                  expression_bias == "Epigenetic Only" &
                  has_model == TRUE,
                .(.N),
                by = .(`Event Type`)
              ]
            ),
            idcol = "filtering"
          )[, c("lambda_selection", "feature_set") := .(sel, fs)]
        }),
        fill = TRUE
      )
    }),
    fill = TRUE
  )

  print(event_summary_table[
    response_type == "real",
    table(expression_bias, `Event Type`, lambda_selection, feature_set)
  ])

  event_counts[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
  event_counts[,
    feature_set := factor(feature_set, levels = c("long", "short", "local"))
  ]

  print(
    ggplot(
      event_counts,
      aes(x = N, y = reorder(filtering, N), fill = lambda_selection)
    ) +
      geom_col(position = "dodge") +
      theme_bw() +
      theme(
        strip.background = element_rect(fill = "white"),
        text = element_text(size = 20),
        axis.text.x = element_blank()
      ) +
      facet_grid(feature_set ~ `Event Type`, scales = "free_x") +
      geom_text(
        aes(label = N),
        position = position_dodge(width = 0.9),
        vjust = 1.1,
        size = 4,
        angle = 90
      ) +
      labs(
        title = "Event waterfall: counts at each filtering stage",
        x = "Number of Events",
        y = NULL
      )
  )
  ggsave(file.path(plot_dir_ev, "event_counts.pdf"), width = 12, height = 5)
}
```

Events without any robust features (`has_model == FALSE`) are excluded from every downstream analysis.
Expression bias is classified on the real-PSI model and propagated to that event's other rows, so the
*Epigenetic Only* filter means the same thing everywhere it is applied.

*Deduction to draw from the proportion plots:* a large *Expression Bias* share means gene expression —
not chromatin — is carrying much of the fit for those events. Since expression tracks differentiation
state, those events cannot support a chromatin-regulation claim even if they passed Tier 1; the
*Epigenetic Only* subset is the one to interpret.

```{r generate-count-plots, eval = en_available}
if (en_available) {
  event_summary_table[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  event_summary_table[,
    `Event Type` := factor(`Event Type`, levels = to_analyze)
  ]

  # Plain per-group event count. The old rotation_rank aggregation (mean/SE of the
  # per-rotation-slot counts) was removed 2026-07-30: Tier-2 fits real PSI only, so
  # every row has rotation_rank == 0, `counts` was a single value, mean() returned
  # that value unchanged and sd() returned NA — i.e. the machinery computed the plain
  # count via a detour and produced an all-NA error bar.
  event_count_dt <- event_summary_table[
    has_model == TRUE,
    .(n_events = uniqueN(ID)),
    by = .(
      `Event Type`,
      feature_set,
      response_type,
      expression_bias,
      Variability,
      lambda_selection
    )
  ]

  event_count_dt[response_type == "real", response_type := "Real"]
  event_count_dt[response_type == "rotated", response_type := "Rotated"]
  event_count_dt[,
    response_type := factor(response_type, levels = c("Real", "Rotated"))
  ]
  event_count_dt[
    expression_bias == "Expression Bias",
    expression_bias := "Expression\nBias"
  ]
  event_count_dt[
    expression_bias == "Epigenetic Only",
    expression_bias := "Epigenetic\nOnly"
  ]
  event_count_dt[
    order(Variability, decreasing = TRUE),
    cumsum_n_events := cumsum(n_events),
    by = .(
      `Event Type`,
      feature_set,
      response_type,
      expression_bias,
      lambda_selection
    )
  ]

  print(
    ggplot(
      event_count_dt,
      aes(y = n_events, x = response_type, fill = Variability)
    ) +
      geom_col() +
      ggh4x::facet_nested_wrap(
        `Event Type` ~ feature_set + lambda_selection + expression_bias
      ) +
      theme_bw() +
      theme(
        legend.box = "vertical",
        strip.background = element_rect(fill = "white"),
        legend.position = "bottom",
        axis.title.x = element_blank()
      ) +
      scale_fill_manual(values = variability_colors) +
      labs(
        title = "Model counts by expression bias and variability",
        y = "Number of Events"
      ) +
      NULL
  )
  ggsave(
    file.path(plot_dir_ev, "events_rotation.png"),
    width = 16,
    height = 3,
    dpi = 600
  )
}
```

```{r generate-proportion-plots, eval = en_available}
if (en_available) {
  event_count_dt[
    response_type == "Real",
    expression_bias := sub("\n", " ", expression_bias, fixed = TRUE)
  ]
  print(
    ggplot(
      event_count_dt[response_type == "Real"],
      aes(y = n_events, x = Variability, fill = expression_bias)
    ) +
      geom_col(position = "fill") +
      geom_text(aes(label = n_events), position = position_fill(vjust = 0.5)) +
      facet_grid(`Event Type` ~ feature_set + lambda_selection) +
      theme_bw() +
      theme(
        legend.box = "vertical",
        strip.background = element_rect(fill = "white"),
        legend.position = "bottom"
      ) +
      scale_fill_manual(
        values = c(
          `Epigenetic Only` = "orange",
          `Expression Bias` = "steelblue"
        )
      ) +
      labs(
        title = "Proportion of real-PSI models by variability: Epigenetic Only vs Expression Bias",
        y = "Proportion of Events",
        fill = NULL
      )
  )
  ggsave(file.path(plot_dir_ev, "events_hit.png"), width = 10, height = 3)
}
```

## 2.2 Hyperparameters actually selected

A sanity check on the tuning itself, before any coefficient is read.

*Expected:* `mixture` near 1 (lasso-like) gives sparse solutions, near 0 (ridge-like) dense ones; with
strongly collinear features the CV optimum often drifts toward the ridge end, because spreading weight
across correlated windows predicts better than committing to one.
*Deduction if `mixture` sits near 0 for most events:* the data are telling you the signal is diffuse and
not attributable to individual windows — so treat the per-feature coefficient lists in section 4 as
"which correlated group", not "which window".
*Deduction if the selected penalty is at the edge of the search grid:* the grid is truncating the
optimum and the fits are not converged in the tuning sense; widen the grid before interpreting.

The `response_type` facet is a historical artifact: Tier 2 has been real-PSI-only since the screen took
over significance, so the "rotated" panel is empty by construction.

```{r en-hyperparameters, eval = en_available}
if (en_available) {
  # How do hyperparameter choices vary across event type and feature set?
  #
  # Faceted by Event Type as well (2026-07-30): SE outnumbers RI ~18:1, so a pooled
  # panel is an SE panel and an RI-specific tuning pathology would be invisible in it.
  # The two types differ in sample size and baseline predictability, so there is no
  # reason to expect the same penalty regime. `response_type` is NOT used as a facet
  # any more — Tier-2 fits real PSI only, so it is constant and its panel was empty.
  #
  # best_params_table / fit_metrics_table come straight from 09zz's per-event CSVs and
  # carry no event metadata, so Event Type is joined on here. (`i.` cannot prefix a
  # backticked column name, hence the k_et alias.)
  .et_map <- event_dt[, .(ID = as.integer(ID), k_et = `Event Type`)]
  .bp <- copy(best_params_table)[, ID := as.integer(ID)]
  .bp[.et_map, on = "ID", `Event Type` := i.k_et]
  .bp[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
  .fm <- copy(fit_metrics_table)[, ID := as.integer(ID)]
  .fm[.et_map, on = "ID", `Event Type` := i.k_et]
  .fm[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  print(
    ggplot(
      .bp[!is.na(lambda_selection)],
      aes(
        x = lambda_selection,
        y = -log10(penalty),
        color = as.factor(mixture),
        group = interaction(ID, wflow_id)
      )
    ) +
      geom_line(alpha = 0.2) +
      geom_boxplot(
        aes(group = lambda_selection),
        color = "grey30",
        fill = NA,
        outliers = FALSE
      ) +
      geom_point(alpha = 0.4, size = 0.8) +
      facet_grid(`Event Type` ~ feature_set) +
      scale_color_brewer(palette = "Set1", name = "mixture") +
      labs(
        title = "Selected penalty by lambda rule, feature set, and event type",
        x = "Lambda selection rule",
        y = "-log10(penalty)"
      ) +
      theme_bw()
  )

  print(
    ggplot(
      .bp,
      aes(x = as.character(mixture), y = -log10(penalty), color = feature_set)
    ) +
      geom_boxplot(alpha = .5) +
      facet_grid(. ~ `Event Type`) +
      labs(
        title = "Penalty distribution by mixture and event type",
        x = "Mixture (0 = ridge, 1 = lasso)",
        y = "-log10(penalty)"
      ) +
      theme_bw()
  )

  # In-sample R² will be inflated by overfitting; compare to CV in paired-rotation-analysis.
  print(
    ggplot(
      .fm,
      aes(x = `Event Type`, y = .estimate, color = feature_set)
    ) +
      geom_violin(alpha = .5, position = position_dodge(width = 0.8)) +
      geom_jitter(
        alpha = .2,
        position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.5),
        size = .5
      ) +
      facet_grid(.metric ~ lambda_selection, scales = "free_y") +
      labs(
        title = "In-sample fit metrics by event type (overfitting diagnostic)",
        x = "Event Type",
        y = "Fit estimate (PSI scale)"
      ) +
      main_theme(16) +
      coord_cartesian(ylim = c(0, NA))
  )
}
```

## 2.3 Fit quality: in-sample vs cross-validated R²

An overfitting diagnostic, not a significance test.

*Expected:* in-sample R² well above CV R², since the model was tuned on the folds it reports. The gap is
the overfitting penalty, and `lambda.1se` should show a smaller gap than `lambda.min` by construction.
*Deduction if CV R² is ≈ 0 or negative while in-sample R² is high:* the fit memorises its training folds
and transfers nothing — expected for events that only just cleared the Tier-1 threshold, and a reason not
to interpret their coefficients.
*Deduction if the gap is small and CV R² is decent:* the fit is stable across held-out ontology groups,
so its coefficients are worth interpreting.
*Deduction if `1se` does not shrink the gap:* the CV curve is flat, so the lambda choice is not doing
meaningful work and `min` vs `1se` differences elsewhere are noise.

```{r en-fit-quality, eval = en_available}
if (en_available) {
  # In-sample vs cross-validated R2 for the Tier-2 elastic-net fits — an OVERFITTING
  # diagnostic, not a significance test (significance was already settled in Tier 1).
  #
  # EXPECTATION: in-sample R2 sits well above CV R2, because the EN was tuned on the
  # same folds it reports; the gap is the overfitting penalty. Read only the CV panel
  # as a performance estimate.
  # DEDUCTION if CV R2 is near zero or negative while in-sample R2 is high: the model
  # is memorising its training folds and carries no transferable signal — expected for
  # events that only just cleared the Tier-1 threshold.
  # DEDUCTION if the gap is small AND CV R2 is decent: the fit is genuinely stable
  # across held-out ontology groups, so its coefficients are worth interpreting.
  # Note lambda.1se should show a smaller gap than lambda.min by construction (more
  # regularisation, less capacity to overfit) — if it does not, the CV curve is flat
  # and the lambda choice is not doing meaningful work.

  # Average CV folds to one value per workflow
  cv_wflow <- cv_metrics_table[
    .metric == "rsq_trad",
    .(.estimate = mean(.estimate, na.rm = TRUE)),
    by = .(ID, wflow_id, feature_set, lambda_selection, response_type)
  ]

  wflow_fit <- rbindlist(
    list(
      `In-sample` = fit_metrics_table[
        .metric == "rsq_trad",
        .(ID, wflow_id, feature_set, lambda_selection, response_type, .estimate)
      ],
      CV = cv_wflow
    ),
    idcol = "metric_type"
  )
  wflow_fit[event_dt, on = "ID", `Event Type` := `Event Type`]
  wflow_fit[, `Event Type` := factor(`Event Type`, levels = to_analyze)]
  wflow_fit[, metric_type := factor(metric_type, levels = c("In-sample", "CV"))]

  real_rsq_dt <- wflow_fit[
    response_type == "real",
    .(
      ID,
      `Event Type`,
      feature_set,
      lambda_selection,
      metric_type,
      real_rsq = .estimate
    )
  ]

  print(
    ggplot(
      real_rsq_dt,
      aes(x = feature_set, y = real_rsq, color = feature_set)
    ) +
      geom_violin(alpha = .5) +
      geom_jitter() +
      facet_grid(`Event Type` ~ metric_type + lambda_selection) +
      coord_cartesian(ylim = c(0, 1)) +
      main_theme(base_size = 16) +
      labs(
        title = "Tier-2 elastic net, real PSI: in-sample vs CV R2",
        x = NULL,
        y = "R2 (real PSI)",
        caption = paste(
          "In-sample is inflated by tuning on the same folds; only the CV panel is a",
          "performance estimate. The in-sample-minus-CV gap is the overfitting penalty."
        )
      )
  )
  ggsave(file.path(plot_dir_ev, "real_rsq.pdf"), width = 10, height = 4)
}
```

# 3. What distinguishes the significant events?

If Tier-1 significance reflects genuine event-specific regulation, it should **correlate with independent
properties** of the events — properties the screen never saw. If instead it is residual noise on top of a
fingerprint, significance should look random with respect to everything else. That is the shared logic of
this section: each analysis is a weak external validation of the hit list.

Every analysis here reads `emp_p_dt`, so `sig` is significance **in that workflow's own feature space**
(use `sig_any` for per-event statements).

**Caveat that applies throughout:** with few hits these comparisons have little power, and a null result
is then uninformative rather than evidence against H1. Read the group sizes before the p-values.

## 3.1 Do High-Variability events pass more often?

*Hypothesis:* epigenetic regulation should be most detectable where splicing actually changes across
samples, so High-Variability events should pass at a higher rate than Low-Variability ones.
*Deduction if High > Low:* consistent with regulation tracking dynamic splicing — though note it is also
consistent with a purely statistical explanation, since more PSI variance means more signal to fit.
*Deduction if the rates are equal:* significance is not tracking the dynamic range of splicing at all,
which is what one expects if it tracks cell-type identity instead.
*Caveat:* "High/Low Variability" is a between- vs within-tissue variance classification, **not** a PSI
range cutoff — both groups span the full PSI range (see section 5), so this is not a proxy for effect
size.

```{r variability-significance, eval = en_available}
if (en_available) {
  # Research question: do High-Variability events show a higher rotation-test
  # significance rate than Low-Variability events? This directly tests whether
  # epigenetic regulation is stronger for dynamically regulated splicing.

  emp_p_var_dt <- copy(emp_p_dt)
  emp_p_var_dt[
    event_dt[, .(ID, Variability)],
    on = "ID",
    Variability := Variability
  ]
  emp_p_var_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  emp_p_var_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  var_sig_dt <- emp_p_var_dt[
    !is.na(emp_p) & !is.na(Variability),
    .(
      n_sig = sum(sig, na.rm = TRUE),
      n_total = .N,
      pct_sig = 100 * mean(sig, na.rm = TRUE)
    ),
    by = .(Variability, `Event Type`, feature_set, lambda_selection)
  ]

  print(
    ggplot(var_sig_dt, aes(x = Variability, y = pct_sig, fill = feature_set)) +
      geom_col(position = "dodge") +
      geom_text(
        aes(label = sprintf("%d/%d", n_sig, n_total)),
        position = position_dodge(width = 0.9),
        vjust = -0.3,
        size = 3
      ) +
      facet_grid(`Event Type` ~ lambda_selection) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Tier-1 screen significance rate by splicing variability group",
        x = "Variability",
        y = "% events significant (FDR within tf x Event Type x feature set)",
        fill = "Feature set",
        caption = paste(
          "Significance is the Tier-1 ridge screen's (q < threshold), not",
          "the retired 10-rotation elastic-net p <= 1/11 test. Positive result:",
          "High-Variability bars taller than Low-Variability, consistent with epigenetic",
          "signal tracking dynamic splicing -- though more PSI variance also means more",
          "variance to fit, so this is not by itself evidence of regulation."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "variability_significance.pdf"),
    width = 8,
    height = 4
  )
}
```

## 3.2 Do significant events have more robustly selected features?

`n_robust` counts features selected in **every** CV fold, i.e. features whose usefulness survives holding
out each ontology group in turn. It is an independent stability measure: Tier 1 never saw it.

*Hypothesis:* genuine signal should be stable across held-out cell types, so significant events should
carry higher `n_robust`.
*Deduction if significant events have higher `n_robust`:* the two methods agree using different
machinery (closed-form ridge vs sparse EN across folds) — mutual corroboration, and the strongest cheap
evidence available here for H1.
*Deduction if `n_robust` is unrelated to significance:* the features driving the ridge screen are not the
ones the elastic net can pin down stably — consistent with a diffuse, collinear (fingerprint-like) signal
rather than a few specific regulatory marks.
*Deduction if significant events have systematically **lower** `n_robust`:* suspect the hits are
small-n/high-variance events where both the screen statistic and the fold-wise selection are unstable —
check against the R²-vs-n diagnostic in §1.1.

```{r n-robust-significance, eval = en_available}
if (en_available) {
  # Research question: do events with more robustly selected features (n_robust)
  # tend to pass the rotation test? n_robust counts features present in all CV
  # folds; higher n_robust = more stable, confident model.

  nrob_sig_dt <- copy(emp_p_dt)
  nrob_sig_dt[
    event_summary_table[
      response_type == "real",
      .(ID, feature_set, lambda_selection, n_robust)
    ],
    on = .(ID, feature_set, lambda_selection),
    n_robust := n_robust
  ]
  nrob_sig_dt[
    event_dt[, .(ID, Variability)],
    on = "ID",
    Variability := Variability
  ]
  nrob_sig_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  # Label reflects the Tier-1 screen's criterion (FDR within tf × Event Type ×
  # feature_set), not the retired 10-rotation elastic-net p-floor of 1/11.
  # Built once and reused by the fill scale below — a hardcoded label string there
  # silently desynchronises from this one (it did: the scale still keyed on the
  # retired "Significant (p <= 1/11)" text, so every significant violin fell through
  # to NA fill).
  .sig_lab <- sprintf("Significant (q < %g)", screen_q_threshold)
  nrob_sig_dt[
    !is.na(emp_p),
    sig_label := fifelse(sig, .sig_lab, "Not significant")
  ]
  nrob_sig_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  # Events with n_robust == 0 (no model) will not appear in emp_p_dt; filter is defensive
  print(
    ggplot(
      nrob_sig_dt[!is.na(n_robust) & n_robust > 0 & !is.na(sig_label)],
      aes(x = sig_label, y = n_robust, fill = sig_label)
    ) +
      geom_violin(alpha = .5, quantiles = c(0.25, 0.5, 0.75)) +
      geom_jitter(alpha = .15, width = 0.2, size = .5) +
      facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
      scale_fill_manual(
        values = setNames(
          c("#D55E00", "#56B4E9"),
          c(.sig_lab, "Not significant")
        ),
        guide = "none"
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Robust feature count by Tier-1 screen outcome",
        x = NULL,
        y = "n_robust (features selected in all CV folds)",
        caption = paste(
          "Events with n_robust = 0 (no model) excluded. Significance is the Tier-1",
          "screen's (FDR within tf x Event Type x feature_set), not the retired",
          "10-rotation elastic-net test. Positive result: significant events shifted",
          "toward higher n_robust, i.e. the two methods agree via different machinery."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "nrobust_significance.pdf"),
    width = 16,
    height = 6
  )
}
```

## 3.3 Fold-to-fold selection consistency

The same idea as `n_robust` but per *feature* rather than per event: how often is a given feature chosen
when a different ontology group is held out? Fold `0` (the full-training fit) is excluded, as it is not a
held-out fold.

*Expected under genuine locus-specific regulation:* a subset of features with high consistency — the same
marks are chosen no matter which cell types are held out.
*Expected under the fingerprint confound:* consistency spread broadly and low, because the model can
substitute any cell-type-correlated feature for any other, and which one wins depends on the fold.
*Deduction:* only high-consistency features are defensible candidates for follow-up; a feature selected
in one fold is indistinguishable from a collinear stand-in.

```{r feature-consistency, eval = en_available}
if (en_available) {
  # Research question: how consistently is each feature selected across CV folds?
  # Computed on real-PSI models only; fold "0" (full-training fit) excluded since
  # it is not a cross-validation fold.
  # High mean consistency = features robustly selected regardless of held-out tissue.

  feat_cons_dt <- feature_table[
    response_type == "real" & fold != "0",
    .(n_folds_selected = uniqueN(fold)),
    by = .(ID, wflow_id, feature_set, lambda_selection, feature, `Event Type`)
  ]
  total_folds_dt <- feature_table[
    response_type == "real" & fold != "0",
    .(total_folds = uniqueN(fold)),
    by = .(ID, wflow_id, lambda_selection)
  ]
  feat_cons_dt[
    total_folds_dt,
    on = .(ID, wflow_id, lambda_selection),
    frac_folds := n_folds_selected / total_folds
  ]

  per_event_cons_dt <- feat_cons_dt[,
    .(mean_consistency = mean(frac_folds, na.rm = TRUE)),
    by = .(ID, feature_set, lambda_selection, `Event Type`)
  ]
  per_event_cons_dt[
    event_dt[, .(ID, Variability)],
    on = "ID",
    Variability := Variability
  ]
  per_event_cons_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  per_event_cons_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  print(
    ggplot(
      per_event_cons_dt[!is.na(mean_consistency)],
      aes(x = mean_consistency, fill = feature_set)
    ) +
      geom_histogram(bins = 20, position = "dodge", alpha = .7) +
      facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "CV fold consistency of selected features (real PSI models)",
        x = "Mean fraction of folds selecting each feature",
        y = "Number of events",
        fill = "Feature set",
        caption = "Fold 0 (full-training fit) excluded. Mean computed over all features per event. Events concentrated near 1.0 = features selected in nearly every fold (robust); near 0 = unstable, fold-sensitive selection."
      )
  )
  ggsave(
    file.path(plot_dir_ev, "feature_consistency.pdf"),
    width = 16,
    height = 6
  )
}
```

## 3.4 `n_robust`: real PSI vs rotated controls — **retired comparison**

**This analysis no longer has a comparison arm and is kept only for provenance.** It was written for the
old design, where each event was re-fit with the elastic net against ~10 rotated responses and
significance came from that comparison. Under the two-tier design, Tier 2 fits **real PSI only** — the
null lives entirely in Tier 1, where it is far cheaper and uses hundreds rather than ten controls — so
there are no `response_type == "rotated"` rows and the "Rotated PSI" violin is empty by construction.

The chunk therefore reports the real-PSI `n_robust` distribution only, and prints the rotated row count so
the absence is explicit rather than a silently half-drawn plot. The equivalent question — *does the real
locus beat unrelated loci?* — is answered properly in §1.2 and §1.3.

```{r n-robust-real-vs-null, eval = en_available}
if (en_available) {
  # RETIRED COMPARISON — kept for provenance, no longer has a second arm.
  # Original question: do real-PSI models robustly select more features than rotated
  # (permutation) controls? Under the two-tier design that question is answered in
  # Tier 1 (§1.2/§1.3) against HUNDREDS of matched control loci, not 10 rotated
  # responses, and Tier 2 fits real PSI ONLY — so `response_type == "rotated"` rows do
  # not exist and the "Rotated PSI" violin is empty by construction.
  # The row count is printed so the absence is explicit rather than a silently
  # half-drawn two-arm plot. Do NOT read a one-sided version of this as a result.

  nrob_comp_dt <- event_summary_table[
    !is.na(n_robust),
    .(
      ID,
      feature_set,
      lambda_selection,
      `Event Type`,
      Variability,
      response_type,
      n_robust
    )
  ]
  nrob_comp_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  nrob_comp_dt[,
    response_label := factor(
      fifelse(response_type == "real", "Real PSI", "Rotated PSI"),
      levels = c("Real PSI", "Rotated PSI")
    )
  ]
  cat(sprintf(
    "n_robust real-vs-rotated: %d real rows, %d rotated rows%s\n",
    sum(nrob_comp_dt$response_type == "real"),
    sum(nrob_comp_dt$response_type == "rotated"),
    if (sum(nrob_comp_dt$response_type == "rotated") == 0L) {
      " -- rotated arm EMPTY as expected (Tier 2 is real-PSI only); the plot below shows the real distribution only, it is NOT a comparison"
    } else {
      ""
    }
  ))
  nrob_comp_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  print(
    ggplot(
      nrob_comp_dt,
      aes(x = response_label, y = n_robust, fill = response_label)
    ) +
      geom_violin(alpha = .5, quantiles = c(0.25, 0.5, 0.75)) +
      geom_jitter(alpha = .1, width = 0.2, size = .3) +
      facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
      scale_fill_manual(
        values = c("Real PSI" = "#D55E00", "Rotated PSI" = "#56B4E9"),
        guide = "none"
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Robust feature count: real PSI vs rotated controls",
        x = NULL,
        y = "n_robust (features selected in all CV folds)",
        caption = "Includes events with n_robust = 0 (no model built). Positive result: Real PSI distribution shifted right vs Rotated, indicating more features are robustly selected when predicting genuine splicing."
      )
  )
  ggsave(
    file.path(plot_dir_ev, "nrobust_real_vs_rotated.pdf"),
    width = 16,
    height = 6
  )
}
```

## 3.5 Is the result stable across the lambda rule?

A robustness check on the regularisation choice, not a biological question.

*Expected:* high concordance — events should be significant (or not) under both `min` and `1se`.
*Deduction for "min only" events:* significance depends on the weaker regularisation, so it rests on
features that do not survive a one-standard-error tightening — treat as fragile.
*Deduction for "1se only" events:* rare and slightly odd (the stricter model passing where the looser one
fails); usually a sign of an unstable CV curve for that event rather than a stronger result.
*Deduction if concordance is low overall:* the feature selection is dominated by the penalty setting
rather than by the data, and no per-event coefficient list should be reported without stating which
lambda rule produced it.

Note that Tier-1 significance itself has **no** lambda dimension (the screen's ridge λ is set by GCV, not
by a path), so any discordance here reflects the elastic-net side only.

```{r lambda-switching, eval = en_available}
if (en_available) {
  # Research question: are rotation-test results stable across lambda selection rules?
  # Concordance (both sig or both non-sig) suggests robust signal.
  # "min only" events = sensitive to regularisation; "1se only" = rare edge case.

  lambda_wide_dt <- dcast(
    emp_p_dt[!is.na(emp_p)],
    ID + feature_set + `Event Type` ~ lambda_selection,
    value.var = "sig"
  )
  setnames(lambda_wide_dt, c("min", "1se"), c("sig_min", "sig_1se"))
  lambda_wide_dt[
    event_dt[, .(ID, Variability)],
    on = "ID",
    Variability := Variability
  ]
  lambda_wide_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]

  lambda_wide_dt[,
    lambda_category := fcase(
      sig_min & sig_1se   , "Both significant"    ,
      sig_min & !sig_1se  , "min only"            ,
      !sig_min & sig_1se  , "1se only"            ,
      !sig_min & !sig_1se , "Neither significant"
    )
  ]
  lambda_wide_dt[,
    lambda_category := factor(
      lambda_category,
      levels = c(
        "Both significant",
        "min only",
        "1se only",
        "Neither significant"
      )
    )
  ]
  lambda_wide_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  lambda_summary <- lambda_wide_dt[
    !is.na(lambda_category),
    .(n = .N),
    by = .(`Event Type`, Variability, feature_set, lambda_category)
  ]
  lambda_summary[,
    pct := 100 * n / sum(n),
    by = .(`Event Type`, Variability, feature_set)
  ]

  print(
    ggplot(
      lambda_summary,
      aes(x = feature_set, y = pct, fill = lambda_category)
    ) +
      geom_col(position = "stack") +
      geom_text(
        aes(label = sprintf("%.0f%%\n(n=%d)", pct, n)),
        position = position_stack(vjust = 0.5),
        size = 3
      ) +
      facet_grid(Variability ~ `Event Type`) +
      scale_fill_manual(
        values = c(
          "Both significant" = "#D55E00",
          "min only" = "#E69F00",
          "1se only" = "#56B4E9",
          "Neither significant" = "#999999"
        )
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Lambda rule concordance in Tier-1 screen significance",
        x = "Feature set",
        y = "% of events",
        fill = "Lambda significance",
        caption = paste(
          "Significance is the Tier-1 screen's. NB the screen itself has NO lambda",
          "dimension (its ridge lambda is set by GCV, not a path), so discordance here",
          "reflects the elastic-net side only. Large 'Both significant' = robust to the",
          "regularisation choice; large 'min only' = signal depends on the weaker",
          "penalty and should be treated as fragile."
        )
      )
  )
  ggsave(file.path(plot_dir_ev, "lambda_switching.pdf"), width = 8, height = 4)
}
```

# 4. Mark-level analysis: which epigenetic features get selected?

The question shifts from *which events* to *which marks*. Feature names encode both a mark and a position
(which window relative to the event), and `feature_set` distinguishes marks that overlap the event itself
(`local`) from window marks further out (`short`/`long`) — a distinction that matters for every
interpretation below, because a mark selected only in a wide window may simply be reporting a neighbouring
promoter rather than anything about the exon.

## 4.1 Which marks are selected in Epigenetic Only models?

*Hypothesis under genuine regulation:* a few marks with known splicing links should dominate —
canonically H3K36me3 (co-transcriptional, exon-biased) and DNA methylation over exon bodies.
*Hypothesis under the fingerprint confound:* selection should look like a generic "active chromatin"
mixture with no particular splicing logic, and should be broadly similar whichever events you look at.
*Deduction if selection is dominated by promoter marks (H3K4me3/H3K27ac) in wide windows only:* the models
are reading promoter/expression state, which §4.3's TSS control tests directly.
*Deduction if `local` selection is negligible while `short`/`long` selection is rich:* nothing at the exon
itself is informative, and the apparent signal comes from genomic neighbourhood — evidence against
event-level epigenetic control specifically.

```{r plot-mark-counts, eval = en_available}
if (en_available) {
  # activeChromHMM loaded per-filter in the load-session chunk above.
  # Research question: which epigenetic marks are robustly selected in Epigenetic Only models?
  # Does mark usage differ between real PSI and rotated controls?
  # Similar mark profiles for real and rotated would imply chance selection.
  activeChromHMM_dt <- data.table(
    region = activeChromHMM$name,
    chrom_state = activeChromHMM$chrom_state,
    region_width = width(activeChromHMM),
    region_ID = seq_along(activeChromHMM)
  )
  coef_table[
    activeChromHMM_dt,
    on = "region",
    c("chrom_state", "region_width", "region_ID") := .(
      chrom_state,
      region_width,
      region_ID
    )
  ]

  # Both real and rotated rows included; expression_bias was propagated from the
  # real-PSI model for each event, so the Epigenetic Only filter applies to both.
  # feature_set kept to distinguish local (event-overlapping) vs short/long window marks.
  # rotation_rank dropped 2026-07-30: Tier-2 is real-PSI only, so it was constant at 0
  # and the mean/SE across "rotation slots" reduced to the plain count with an NA SD.
  err_dt <- coef_table[
    mark != "(Intercept)" & expression_bias == "Epigenetic Only",
    .(mark = unique(mark)),
    by = .(
      ID,
      feature_set,
      response_type,
      `Event Type`,
      Variability,
      lambda_selection
    )
  ][,
    .(n_events = uniqueN(ID)),
    by = .(
      `Event Type`,
      feature_set,
      response_type,
      mark,
      Variability,
      lambda_selection
    )
  ]

  err_dt[,
    Response := factor(
      fifelse(response_type == "real", "Real", "Rotated"),
      levels = c("Real", "Rotated")
    )
  ]
  err_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  err_dt[, `Event Type` := factor(`Event Type`, levels = c("SE", "RI"))]
  err_dt[, mark := tools::toTitleCase(gsub("_", " ", mark))]

  hex_colors <- c(
    mark_hex_colors,
    summed_enhancer = "darkorange",
    max_promoter = "tomato"
  )
  plot_colors <- setNames(
    rep("dark grey", err_dt[, uniqueN(mark)]),
    err_dt[, unique(mark)]
  )
  names(hex_colors) <- tools::toTitleCase(gsub("_", " ", names(hex_colors)))
  for (name in names(hex_colors)) {
    inds <- startsWith(names(plot_colors), name)
    plot_colors[inds] <- hex_colors[name]
  }

  print(
    ggplot(err_dt, aes(x = Response, y = n_events, fill = mark)) +
      geom_col(position = "dodge") +
      facet_grid(
        Variability + `Event Type` ~ feature_set + lambda_selection,
        scales = "free_y"
      ) +
      labs(
        title = "Epigenetic mark usage in Epigenetic Only models",
        y = "Number of events selecting mark",
        x = NULL,
        fill = "Mark",
        caption = paste(
          "Tier-2 is real-PSI only, so there is no rotated comparison arm here; the",
          "real-vs-null question is answered in section 1. Compare local vs short/long:",
          "local marks are event-overlapping, long marks include distal elements."
        )
      ) +
      theme_bw() +
      scale_fill_manual(values = plot_colors) +
      theme(
        legend.position = "bottom",
        strip.background = element_rect(fill = "white")
      )
  )
  ggsave(
    file.path(plot_dir_ev, "mark_counts_error.pdf"),
    width = 18,
    height = 10
  )
}
```

## 4.2 Mark profiles stratified by Tier-1 significance

The sharpest version of the mark question: do events that **pass** the screen select *different* marks
from those that fail? Per mark, a Fisher's exact test compares selection frequency between the passing and
failing strata (Fisher rather than chi-squared because the passing group is small).

Three outcomes, with what each licenses:

- **POSITIVE — distinct profiles, specific marks enriched in the passing group.** Particular marks drive
  genuine epigenetic–splicing associations beyond the fingerprint, and feature interpretation is
  trustworthy for the significant subset. This is the result that would support a mechanistic claim.
- **NULL — indistinguishable profiles.** Two explanations that this comparison **cannot** separate:
  (a) all mark selection reflects cell-type identity and the passing events cleared the threshold by
  chance; or (b) genuine signal uses the *same* marks as the fingerprint, so profile comparison is blind
  to the difference. Distinguishing them needs a different design (e.g. perturbation, or matched-cell-type
  replicates), not more of this analysis.
- **CONFOUND — protocol/RNA-seq features enriched in the passing group.** Technical protocol effects, not
  epigenetics, are driving apparent signal. This is a flag to investigate, not a result.

*Power caveat, stated before the p-values:* with a small passing group, Fisher has little power, so a NULL
here is weak evidence. Check `n_group` per stratum first.

```{r mark-counts-screen-stratified, eval = en_available}
if (en_available) {
  # Research question: do events passing the rotation test select different marks
  # than failing events?
  #
  # Interpretation guide:
  #   POSITIVE — distinct profiles, mark enriched in passing group:
  #     Specific marks drive genuine epigenetic-splicing associations beyond
  #     fingerprint confound. Feature interpretation trustworthy for the
  #     significant subset.
  #
  #   NULL — similar profiles between groups:
  #     Two indistinguishable explanations at this n:
  #     (a) Fingerprint confound: all mark selection reflects cell-type identity;
  #         passing events beat the test by chance. (The old rotation-era numbers
  #         here -- min achievable p = 1/11 ≈ 9%, observed 14-15% -- no longer
  #         apply: significance is now the Tier-1 screen's FDR-controlled q.)
  #     (b) Genuine signal uses the SAME marks as the fingerprint — mark profile
  #         comparison cannot separate the two; larger n or different design needed.
  #
  #   CONFOUND signal — Protocol/RNA-seq marks enriched in passing group:
  #     Technical RNA-seq protocol effects drive apparent signal rather than
  #     epigenetic regulation. Flag for investigation.

  # Collapse Tier-1 screen significance to (ID, lambda_selection): significant if
  # ANY feature set passes (q < threshold). `sig` comes from the screen
  # via emp_p_dt, so there is no 1/11 p-floor involved any more.
  sig_collapse_dt <- emp_p_dt[,
    .(sig_any = any(sig, na.rm = TRUE)),
    by = .(ID, lambda_selection)
  ]
  sig_collapse_dt[,
    rotation_outcome := factor(
      fifelse(sig_any, "Passes Tier-1 screen", "Fails Tier-1 screen"),
      levels = c("Passes Tier-1 screen", "Fails Tier-1 screen")
    )
  ]

  coef_real_sig <- coef_table[
    response_type == "real" &
      mark != "(Intercept)" &
      expression_bias == "Epigenetic Only"
  ]
  coef_real_sig[
    sig_collapse_dt,
    on = .(ID, lambda_selection),
    rotation_outcome := rotation_outcome
  ]

  # Per event per feature_set: unique marks selected.
  # Keeping feature_set separates local (event-overlapping) from short/long window marks.
  event_mark_dt <- coef_real_sig[
    !is.na(rotation_outcome),
    .(mark = unique(mark)),
    by = .(
      ID,
      `Event Type`,
      Variability,
      feature_set,
      lambda_selection,
      rotation_outcome
    )
  ]

  # n_group: total events per stratum.
  n_group_dt <- event_mark_dt[,
    .(n_group = uniqueN(ID)),
    by = .(
      `Event Type`,
      Variability,
      feature_set,
      lambda_selection,
      rotation_outcome
    )
  ]

  err_sig_dt <- event_mark_dt[,
    .(n_events = uniqueN(ID)),
    by = .(
      `Event Type`,
      Variability,
      feature_set,
      lambda_selection,
      rotation_outcome,
      mark
    )
  ]
  err_sig_dt[
    n_group_dt,
    on = .(
      `Event Type`,
      Variability,
      feature_set,
      lambda_selection,
      rotation_outcome
    ),
    c("n_group", "pct_events") := .(n_group, 100 * n_events / n_group)
  ]

  err_sig_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  err_sig_dt[, `Event Type` := factor(`Event Type`, levels = c("SE", "RI"))]
  err_sig_dt[, mark := tools::toTitleCase(gsub("_", " ", mark))]

  # Fisher's exact test per mark per stratum.
  # Fisher preferred over chi-squared: passing group sizes are small.
  # Null: profiles indistinguishable → fingerprint OR same-mark genuine signal.
  # Significant enrichment in passing group: mark associated beyond fingerprint.
  fisher_dt <- err_sig_dt[,
    {
      pass_row <- .SD[rotation_outcome == "Passes Tier-1 screen"]
      fail_row <- .SD[rotation_outcome == "Fails Tier-1 screen"]
      if (nrow(pass_row) > 0L && nrow(fail_row) > 0L) {
        mat <- matrix(
          c(
            pass_row$n_events[1L],
            pass_row$n_group[1L] - pass_row$n_events[1L],
            fail_row$n_events[1L],
            fail_row$n_group[1L] - fail_row$n_events[1L]
          ),
          nrow = 2,
          dimnames = list(c("selected", "not"), c("passes", "fails"))
        )
        res <- fisher.test(mat)
        .(p_value = res$p.value, odds_ratio = unname(res$estimate))
      } else {
        .(p_value = NA_real_, odds_ratio = NA_real_)
      }
    },
    by = .(`Event Type`, Variability, feature_set, lambda_selection, mark)
  ]
  fisher_dt[, p_adj := p.adjust(p_value, method = "BH")]
  fisher_dt[,
    sig_label := fcase(
      p_adj < 0.001 , "***" ,
      p_adj < 0.01  , "**"  ,
      p_adj < 0.05  , "*"   ,
      p_adj < 0.1   , "."   ,
      default = ""
    )
  ]
  print(fisher_dt[order(
    `Event Type`,
    Variability,
    feature_set,
    lambda_selection,
    p_adj
  )])

  # Annotate passing-group bars with significance label (stars above bar).
  plot_sig_dt <- err_sig_dt[rotation_outcome == "Passes Tier-1 screen"][
    fisher_dt[, .(
      `Event Type`,
      Variability,
      feature_set,
      lambda_selection,
      mark,
      sig_label
    )],
    on = .(`Event Type`, Variability, feature_set, lambda_selection, mark)
  ]

  sig_plot_colors <- setNames(
    rep("dark grey", err_sig_dt[, uniqueN(mark)]),
    err_sig_dt[, unique(mark)]
  )
  for (name in names(hex_colors)) {
    inds <- startsWith(names(sig_plot_colors), name)
    sig_plot_colors[inds] <- hex_colors[name]
  }

  print(
    ggplot(err_sig_dt, aes(x = rotation_outcome, y = pct_events, fill = mark)) +
      geom_col(position = "dodge") +
      geom_text(
        data = plot_sig_dt[sig_label != ""],
        aes(label = sig_label, y = pct_events + 3),
        position = position_dodge(0.9),
        size = 3,
        vjust = 0
      ) +
      facet_grid(
        Variability + `Event Type` ~ feature_set + lambda_selection,
        scales = "free_y"
      ) +
      scale_fill_manual(values = sig_plot_colors) +
      theme_bw() +
      theme(
        legend.position = "bottom",
        strip.background = element_rect(fill = "white"),
        axis.text.x = element_text(angle = 20, hjust = 1)
      ) +
      labs(
        title = "Mark composition by rotation test outcome (Epigenetic Only, real PSI)",
        x = NULL,
        y = "% of events in group selecting mark",
        fill = "Mark",
        caption = paste(
          "Stars above Passes bars: Fisher's exact test, BH-adjusted (. p<0.1, * p<0.05, ** p<0.01).",
          "Null (no stars, similar profiles): fingerprint confound explains selection, OR genuine",
          "signal uses same marks as fingerprint -- indistinguishable at this n.",
          "Significant enrichment in passing group: specific mark beyond fingerprint confound."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "mark_counts_rotation_stratified.pdf"),
    width = 18,
    height = 10
  )
}
```

## 4.3 TSS-proximity control for promoter-mark selection

H3K4me3 and H3K27ac are promoter marks. If events whose models select them are systematically **closer to
the TSS** than unselected events, then the selection plausibly reflects promoter chromatin leaking into the
feature window — not event-level regulation. This is a targeted confound test, not an exploratory look.

The `local` feature set is the critical comparator, and the logic is a genuine dissociation:

- **`local`, H3K4me3 selected, no TSS enrichment** → the mark genuinely overlaps the exon/intron body;
  promoter contamination is not the explanation.
- **`short`/`long`, significant TSS enrichment** → the window is capturing a nearby promoter; the mark's
  coefficient should not be read as an exon-level effect.
- **RI events near the TSS** → expected biologically (first-intron retention is common), so RI enrichment
  is not by itself evidence of contamination — interpret RI and SE separately.

The Wilcoxon test is one-tailed (`alternative = "less"`), i.e. it asks specifically whether selected events
are *closer* to the TSS, which is the direction the confound would produce.

```{r tss-proximal, eval = en_available}
if (en_available) {
  # distance_TSS: genomic distance from event to nearest TSS.
  # Event-level property — same across all samples. Take first non-NA per event.
  tss_dist_dt <- aggregated_dt[
    !is.na(distance_TSS),
    .(
      distance_TSS = distance_TSS[1L],
      abs_distance_TSS = abs(distance_TSS[1L]),
      `Event Type` = `Event Type`[1L]
    ),
    by = ID
  ]
  tss_dist_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  # Identify events with H3K4me3 and/or H3K27ac robustly selected.
  # Real PSI models only; all expression_bias categories (H3K4me3 may drive expression bias).
  # Stratify by feature_set and lambda_selection.
  promoter_events <- coef_table[
    response_type == "real" &
      mark %in% c("H3K4me3", "H3K27ac"),
    .(
      has_H3K4me3 = any(mark == "H3K4me3"),
      has_H3K27ac = any(mark == "H3K27ac"),
      has_either = TRUE,
      expression_bias = expression_bias[1L]
    ),
    by = .(ID, feature_set, lambda_selection)
  ]

  # Reference: all events with a fitted real-PSI model — one set per feature_set x lambda_selection.
  # Use cross-join to avoid missing events that appear in coef_table for some combos but not all.
  modeled_ids <- event_summary_table[
    response_type == "real" & has_model == TRUE,
    unique(ID)
  ]
  fs_ls_combos <- unique(promoter_events[, .(feature_set, lambda_selection)])

  combined_dt <- rbindlist(
    lapply(seq_len(nrow(fs_ls_combos)), function(i) {
      fs <- fs_ls_combos$feature_set[i]
      ls <- fs_ls_combos$lambda_selection[i]
      sel_ids <- promoter_events[
        feature_set == fs & lambda_selection == ls & has_either == TRUE,
        unique(ID)
      ]
      tss_dist_dt[
        ID %in% modeled_ids,
        .(
          ID,
          `Event Type`,
          abs_distance_TSS,
          distance_TSS,
          feature_set = fs,
          lambda_selection = ls,
          group = fifelse(
            ID %in% sel_ids,
            "H3K4me3 or H3K27ac selected",
            "Not selected"
          )
        )
      ]
    })
  )
  combined_dt[,
    group := factor(
      group,
      levels = c("H3K4me3 or H3K27ac selected", "Not selected")
    )
  ]
  combined_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  # --- Plot 1: Density of abs(distance_TSS), log-scale ---
  print(
    ggplot(
      combined_dt[!is.na(abs_distance_TSS)],
      aes(x = abs_distance_TSS, fill = group, color = group)
    ) +
      geom_density(alpha = 0.4) +
      scale_x_log10() +
      scale_fill_manual(
        values = c(
          "H3K4me3 or H3K27ac selected" = "#D55E00",
          "Not selected" = "#56B4E9"
        ),
        name = NULL
      ) +
      scale_color_manual(
        values = c(
          "H3K4me3 or H3K27ac selected" = "#D55E00",
          "Not selected" = "#56B4E9"
        ),
        guide = "none"
      ) +
      facet_grid(`Event Type` ~ feature_set + lambda_selection) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "TSS proximity: events with promoter marks selected vs not",
        x = "abs(distance_TSS) [bp, log10]",
        y = "Density",
        caption = paste(
          "H3K4me3 and H3K27ac are promoter-associated marks.",
          "Left shift = selected events closer to TSS than unselected.",
          "local feature set: mark overlaps exon body — strongest evidence of event-level regulation.",
          "short/long: flanking window may capture nearby promoter, not the event itself.",
          "RI events are often first-intron; TSS proximity expected and not necessarily a confound."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "tss_proximal_density.pdf"),
    width = 16,
    height = 5
  )

  # --- Plot 2: % of events within distance thresholds ---
  thresholds <- c(1000L, 2000L, 5000L, 10000L)
  thresh_dt <- rbindlist(lapply(thresholds, function(th) {
    combined_dt[
      !is.na(abs_distance_TSS),
      .(
        pct_proximal = 100 * mean(abs_distance_TSS <= th),
        n_events = .N,
        threshold_bp = th
      ),
      by = .(`Event Type`, feature_set, lambda_selection, group)
    ]
  }))
  thresh_dt[, threshold_label := paste0("<= ", threshold_bp / 1000L, " kb")]
  thresh_dt[,
    threshold_label := factor(
      threshold_label,
      levels = paste0("<= ", thresholds / 1000L, " kb")
    )
  ]

  print(
    ggplot(
      thresh_dt,
      aes(x = threshold_label, y = pct_proximal, fill = group)
    ) +
      geom_col(position = "dodge") +
      geom_text(
        aes(label = sprintf("%.0f%%\n(n=%d)", pct_proximal, n_events)),
        position = position_dodge(width = 0.9),
        vjust = -0.2,
        size = 2.3
      ) +
      facet_grid(`Event Type` ~ feature_set + lambda_selection) +
      scale_fill_manual(
        values = c(
          "H3K4me3 or H3K27ac selected" = "#D55E00",
          "Not selected" = "#56B4E9"
        ),
        name = NULL
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "% of events within distance threshold of TSS",
        x = "Distance threshold",
        y = "% of events"
      )
  )
  ggsave(
    file.path(plot_dir_ev, "tss_proximal_thresholds.pdf"),
    width = 16,
    height = 5
  )

  # --- Wilcoxon test: are selected events closer to TSS? ---
  # One-tailed: alternative = "less" tests whether selected events are *closer*.
  wilcox_dt <- combined_dt[
    !is.na(abs_distance_TSS),
    {
      sel <- .SD[group == "H3K4me3 or H3K27ac selected", abs_distance_TSS]
      nosel <- .SD[group == "Not selected", abs_distance_TSS]
      if (length(sel) >= 3L && length(nosel) >= 3L) {
        wt <- wilcox.test(sel, nosel, alternative = "less")
        .(
          p_value = wt$p.value,
          median_sel_kb = round(median(sel) / 1000, 1),
          median_ref_kb = round(median(nosel) / 1000, 1),
          n_sel = length(sel),
          n_ref = length(nosel)
        )
      } else {
        .(
          p_value = NA_real_,
          median_sel_kb = NA_real_,
          median_ref_kb = NA_real_,
          n_sel = length(sel),
          n_ref = length(nosel)
        )
      }
    },
    by = .(`Event Type`, feature_set, lambda_selection)
  ]
  wilcox_dt[, p_adj := p.adjust(p_value, method = "BH")]
  wilcox_dt[,
    sig_label := fcase(
      p_adj < 0.001 , "***" ,
      p_adj < 0.01  , "**"  ,
      p_adj < 0.05  , "*"   ,
      p_adj < 0.1   , "."   ,
      default = "ns"
    )
  ]
  print(wilcox_dt[order(`Event Type`, feature_set, lambda_selection)])
}
```

# 5. PSI variability vs model quality

Does the *magnitude* of PSI change across samples predict how well an event can be modelled?

**Read the binning caveat first — and note it was previously stated wrongly.** An older comment here
claimed `aggregated_dt` is "pre-filtered to variable events, all with PSI range > 0.25". That is **false**:
there is no variance or range filter anywhere in the pipeline. The only PSI-based event filter is Wang
et al. 2024 Procedure 2 as implemented upstream in `02-3` (`psi_q95 >= 0.05 & psi_q05 <= 0.95`), which is
a **non-constitutive** test — it removes only always-excluded and always-included events, and says nothing
about how much PSI varies in between. Measured on the real `biotype_filtered` set: PSI range is ≤ 0.25 for
**33%** of events and ≤ 0.15 for 11%, and **39% of SE events have `q05 > 0.90`**, i.e. they sit hard against
the constitutive-inclusion ceiling and were retained by a margin of ≤ 0.05.

Quantile tertiles are therefore used because they are robust, **not** because the low bin is
"variable-but-less-so": the low tertile genuinely contains near-constant events, whose PSI variance is
small enough that little can be explained in principle. That matters for reading everything below —
`Variability = Low` SE events have a median range of only 0.23 — and it matters for Tier 1, where a
near-constant response is a plausible source of the wild negative group-LOO R² values seen in §1.1.

*Hypothesis (signal-strength):* larger PSI swings give more variance to explain, so high-dPSI events
should more often get a model and reach higher CV R².
*Deduction if the R²-vs-range slope is flat:* within this range, how much an event's PSI moves does not
predict how well chromatin explains it — consistent with the marks capturing static cell-type identity
rather than the intensity of dynamic regulation.
*Deduction if high-dPSI events show more Expression Bias:* gene expression co-varies with large PSI shifts
(differentiation state), so the *Epigenetic Only* label is least reliable exactly where the effect sizes are
biggest — an important caveat for any "top events" list.
*Deduction from High/Low Variability groups spanning the same PSI range:* confirms that the Variability
label captures between- vs within-tissue variance structure, **not** effect size, so §3.1 and this section
are testing different things and should not be conflated.

```{r dpsi-analysis, eval = en_available}
if (en_available) {
  # Research question: do high-dPSI events yield better models and more expression bias?
  #
  # CORRECTED 2026-07-29 — the old note here ("pre-filtered to variable splicing
  # events, all events have PSI range > 0.25") was WRONG. There is no variance/range
  # filter in the pipeline. `02-3`'s only PSI-based event filter is Wang Procedure 2,
  # `psi_q95 >= 0.05 & psi_q05 <= 0.95`, which merely drops CONSTITUTIVE events
  # (always ~0 or always ~1) and constrains variability not at all.
  # Real numbers (biotype_filtered): 33% of events have PSI range <= 0.25, 11% <= 0.15,
  # and 39% of SE events have q05 > 0.90 (retained by <= 0.05 of margin).
  # Tertile bins are used because they are robust to that skew — NOT because the low
  # bin is merely "less variable". The low bin contains genuinely near-constant events
  # with almost no variance to explain, which is a confound for any
  # "does dPSI predict model quality?" reading below.
  #
  # Interpretation guide:
  #   HIGH dPSI → more models built + higher CV R²:
  #     Signal strength drives model quality. Supports genuine epigenetic regulation.
  #
  #   NO dPSI effect on R² (flat slope in scatter):
  #     Marks capture static cell-type identity, not dynamic regulation intensity.
  #     All events are already "high change"; variation within that range is irrelevant.
  #
  #   HIGH dPSI → more expression bias:
  #     Gene expression co-varies with large PSI shifts (differentiation state), confounding
  #     epigenetic attribution. "Epigenetic Only" label may be unreliable for top-dPSI events.
  #
  #   HIGH and LOW Variability groups show similar PSI range:
  #     The "High/Low Variability" classification is NOT based on PSI range magnitude —
  #     it likely reflects relative between-tissue vs. within-tissue variation. Both groups
  #     fully overlap in absolute PSI range; the label captures a different biological axis.

  # Per-event PSI statistics.
  psi_var_dt <- aggregated_dt[
    !is.na(PSI),
    .(
      sd_psi = sd(PSI),
      range_psi = max(PSI) - min(PSI),
      n_samples = .N,
      `Event Type` = `Event Type`[1L],
      Variability = Variability[1L]
    ),
    by = ID
  ]
  psi_var_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  # Variability in aggregated_dt uses "High"/"Low" (not "High Variability"/"Low Variability").
  print(
    ggplot(
      psi_var_dt[!is.na(`Event Type`)],
      aes(x = Variability, y = range_psi, fill = Variability)
    ) +
      geom_violin(alpha = 0.5, quantiles = c(0.25, 0.5, 0.75)) +
      geom_jitter(alpha = 0.1, width = 0.2, size = 0.3) +
      facet_wrap(~`Event Type`) +
      scale_fill_manual(
        values = c("High" = "#D55E00", "Low" = "#56B4E9"),
        guide = "none"
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "PSI range by variability group (alignment check)",
        x = NULL,
        y = "PSI range (max - min)",
        caption = paste(
          "Range = max(PSI) - min(PSI) across IHEC samples ~ max pairwise dPSI.",
          "NB: there is NO range pre-filter -- 33% of events have range <= 0.25. High/Low Variability groups overlap",
          "almost completely -- the Variability label is not based on PSI range magnitude."
        )
      )
  )
  ggsave(
    file.path(plot_dir_ev, "psi_range_distribution.pdf"),
    width = 6,
    height = 4
  )

  # Compute tertile bins on psi_var_dt (one row per event) BEFORE cartesian join.
  # Use -Inf/Inf as outer boundaries to avoid "breaks not unique" from probs=c(0,1) matching
  # the min/max values. Epsilon guard ensures strict monotonicity if q1==q2.
  {
    dpsi_q <- quantile(
      psi_var_dt$range_psi,
      probs = c(1 / 3, 2 / 3),
      na.rm = TRUE
    )
    if (dpsi_q[[1L]] >= dpsi_q[[2L]]) {
      dpsi_q[[2L]] <- dpsi_q[[1L]] +
        .Machine$double.eps * max(1, abs(dpsi_q[[1L]])) * 1e6
    }
    psi_var_dt[,
      dpsi_bin := cut(
        range_psi,
        breaks = c(-Inf, dpsi_q[[1L]], dpsi_q[[2L]], Inf),
        labels = c(
          sprintf("Low tertile\n(<= %.2f)", dpsi_q[[1L]]),
          sprintf("Mid tertile\n(%.2f-%.2f)", dpsi_q[[1L]], dpsi_q[[2L]]),
          sprintf("High tertile\n(> %.2f)", dpsi_q[[2L]])
        ),
        include.lowest = TRUE
      )
    ]
  }

  # Join with all feature_set x lambda_selection combinations for full stratification.
  # allow.cartesian: one psi_var_dt row fans out to (n_feature_sets x n_lambda) rows.
  # dpsi_bin carried through join from psi_var_dt.
  dpsi_model_dt <- psi_var_dt[
    event_summary_table[
      response_type == "real",
      .(
        ID,
        feature_set,
        lambda_selection,
        has_model,
        expression_bias,
        n_robust,
        mean_cv_rsq
      )
    ],
    on = "ID",
    allow.cartesian = TRUE,
    nomatch = NULL
  ]
  dpsi_model_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  dpsi_model_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  # Model-building rate by dPSI tertile, stratified by Event Type, Variability,
  # feature_set, and lambda_selection.
  model_rate_dt <- dpsi_model_dt[
    !is.na(dpsi_bin),
    .(
      n_model = sum(has_model, na.rm = TRUE),
      n_total = .N,
      pct_model = 100 * mean(has_model, na.rm = TRUE)
    ),
    by = .(`Event Type`, Variability, feature_set, lambda_selection, dpsi_bin)
  ]
  print(
    ggplot(model_rate_dt, aes(x = dpsi_bin, y = pct_model, fill = dpsi_bin)) +
      geom_col() +
      geom_text(
        aes(label = sprintf("%d/%d", n_model, n_total)),
        vjust = -0.3,
        size = 2.5
      ) +
      facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
      scale_fill_brewer(palette = "OrRd", guide = "none") +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Model-building rate by PSI range tertile",
        x = "PSI range tertile",
        y = "% events with model (n_robust > 0)"
      )
  )
  ggsave(file.path(plot_dir_ev, "dpsi_model_rate.pdf"), width = 16, height = 8)

  # Continuous scatter: range_psi vs mean CV R² with loess trend.
  # More informative than bins given the pre-filtered range (all > 0.25).
  # Null: flat or noisy slope — PSI range within this range doesn't predict model quality.
  # Positive: upward slope — higher absolute change → better epigenetic predictability.
  print(
    ggplot(
      dpsi_model_dt[has_model == TRUE & !is.na(mean_cv_rsq)],
      aes(x = range_psi, y = mean_cv_rsq, color = Variability)
    ) +
      geom_point(alpha = 0.15, size = 0.5) +
      geom_smooth(aes(group = Variability), method = "loess", se = FALSE) +
      facet_grid(`Event Type` ~ feature_set + lambda_selection) +
      scale_color_manual(values = c("High" = "#D55E00", "Low" = "#56B4E9")) +
      coord_cartesian(ylim = c(0, NA)) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "PSI range vs CV R2",
        x = "PSI range (max - min)",
        y = "Mean CV R2",
        caption = paste(
          "Loess trend per Variability group. Null: flat slope within the pre-filtered high-change range.",
          "No range pre-filter exists (33% of events have range <= 0.25); a flat slope means PSI range does not drive model quality."
        )
      )
  )
  ggsave(file.path(plot_dir_ev, "dpsi_cv_rsq.pdf"), width = 16, height = 5)

  # Expression bias by dPSI tertile.
  # High-dPSI events may show more expression bias if gene expression co-varies
  # with large PSI shifts (e.g., differentiation).
  bias_dt <- dpsi_model_dt[
    has_model == TRUE & !is.na(dpsi_bin) & !is.na(expression_bias),
    .(n = .N),
    by = .(
      `Event Type`,
      Variability,
      feature_set,
      lambda_selection,
      dpsi_bin,
      expression_bias
    )
  ][,
    pct := 100 * n / sum(n),
    by = .(`Event Type`, Variability, feature_set, lambda_selection, dpsi_bin)
  ]
  print(
    ggplot(bias_dt, aes(x = dpsi_bin, y = pct, fill = expression_bias)) +
      geom_col(position = "stack") +
      geom_text(
        aes(label = sprintf("%.0f%%\n(n=%d)", pct, n)),
        position = position_stack(vjust = 0.5),
        size = 2.5
      ) +
      facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
      scale_fill_manual(
        values = c(
          "Epigenetic Only" = "#56B4E9",
          "Expression Bias" = "#E69F00"
        ),
        name = "Expression bias"
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Expression bias by PSI range tertile (events with models)",
        x = "PSI range tertile",
        y = "% of events",
        caption = "Expression Bias = gene_expression feature in all CV folds."
      )
  )
  ggsave(
    file.path(plot_dir_ev, "dpsi_expression_bias.pdf"),
    width = 16,
    height = 8
  )
}
```

# 6. Inter-model feature variation: is mark selection structured, or idiosyncratic?

Final question: do event models share regulatory logic, or is each one its own story? Two views —
coefficient *signs* per mark, and mark *co-occurrence* across events.

*Hypothesis A — coherent clusters.* Groups of events share mark patterns, i.e. distinct chromatin contexts
predict splicing in distinct genomic settings (promoter-proximal vs enhancer-associated, say). Supports
genuine regulatory logic.
*Hypothesis B — no cluster structure.* Selection is idiosyncratic; no shared logic. Consistent with the
fingerprint confound, where marks enter via cell-type correlation rather than event-specific regulation.

**The co-occurrence matrix is the decisive panel, and its expected shape is the confound's fingerprint.**
If marks entered models because of local regulatory state, biologically antagonistic marks should be
*negatively* correlated — active (H3K27ac) and repressive (H3K27me3) states are largely mutually exclusive
at the same locus. An **all-positive** matrix, including for antagonistic pairs, means selection is driven
by something all marks share: cell-type identity. This mirrors the sample-axis correlation result reported
elsewhere in the project (H3K27ac vs H3K27me3 r ≈ 0.65 across samples, while the *spatial* window-axis
correlation for the same pair is negative, as biology predicts) — so the all-positive pattern here is
expected and is evidence about the confound, not a bug.

*Deduction from mixed coefficient signs for the same mark* (positive in some events, negative in others):
the mark's effect is context-dependent — e.g. H3K27ac coinciding with inclusion at one locus and exclusion
at another. Consistent sign across events would instead indicate a universal association. Mixed signs are
also, less charitably, what unstable selection among collinear features produces, so cross-check against
the fold-consistency results in §3.3 before claiming context dependence.

*Watch for r = 1.00 pairs:* these are duplicate features (identical values under two names), not two marks
that happen to agree perfectly — e.g. the `Protocol mRNA-Seq` / `Protocol total-RNA-Seq` pair.

```{r feature-heatmap, eval = en_available}
if (en_available) {
  # Research question: is mark selection structured across events, or random?
  #
  # Interpretation guide:
  #   COHERENT CLUSTERS (events group by mark pattern in heatmap):
  #     Shared epigenetic programs drive splicing prediction for groups of events.
  #     Supports genuine regulatory logic: different chromatin contexts predict
  #     splicing in different genomic contexts (e.g., promoter vs. enhancer regulated).
  #
  #   NO CLUSTER STRUCTURE (scattered heatmap):
  #     Feature selection is idiosyncratic — no shared regulatory logic across events.
  #     Consistent with fingerprint confound: marks selected via cell-type correlation,
  #     not because of event-specific epigenetic regulation.
  #
  #   ALL-POSITIVE CO-OCCURRENCE MATRIX (observed result):
  #     No negative correlations even between biologically antagonistic marks
  #     (H3K27ac active vs H3K27me3 repressive, r=0.65). This is a hallmark of
  #     fingerprint confound: all marks correlate with cell-type identity, so
  #     events where the model learns cell-type structure select ALL marks regardless
  #     of their biological antagonism. A genuine regulatory signal would show
  #     negative correlation between active and repressive marks (mutually exclusive
  #     chromatin states at the same locus).
  #     Protocol mRNA.Seq = Protocol total.RNA.Seq at r=1.00: these are duplicate
  #     features (likely same numerical values), not independent marks.
  #
  #   MIXED COEFFICIENT SIGN (same mark → + in some events, − in others):
  #     Mark effect is context-dependent. H3K27ac might activate exon inclusion in
  #     one locus and repress it in another, reflecting local regulatory architecture.
  #     Consistent sign = universal association; mixed sign = locus-specific effect.

  # All real Epigenetic Only models — stratified by Event Type, feature_set, lambda_selection.
  hm_coef_dt <- coef_table[
    response_type == "real" &
      expression_bias == "Epigenetic Only" &
      mark != "(Intercept)"
  ]
  hm_coef_dt[,
    Variability := factor(
      sub(" Variability", "", Variability, fixed = TRUE),
      levels = c("High", "Low")
    )
  ]
  hm_coef_dt[, `Event Type` := factor(`Event Type`, levels = to_analyze)]

  # --- Coefficient sign distribution -------------------------------------------
  # Stratified by Event Type (SE vs RI), feature_set (local/short/long),
  # lambda_selection (min/1se), Variability (High/Low).
  sign_dt <- hm_coef_dt[,
    .(n_pos = sum(coef > 0), n_neg = sum(coef < 0)),
    by = .(mark, `Event Type`, Variability, feature_set, lambda_selection)
  ]
  sign_dt[, `:=`(
    pct_pos = 100 * n_pos / (n_pos + n_neg),
    pct_neg = 100 * n_neg / (n_pos + n_neg)
  )]
  sign_dt[, mark := tools::toTitleCase(gsub("_", " ", mark))]
  sign_long <- melt(
    sign_dt,
    id.vars = c(
      "mark",
      "Event Type",
      "Variability",
      "feature_set",
      "lambda_selection"
    ),
    measure.vars = c("pct_pos", "pct_neg"),
    variable.name = "direction",
    value.name = "pct"
  )
  sign_long[,
    direction := fifelse(
      direction == "pct_pos",
      "Positive (+PSI)",
      "Negative (-PSI)"
    )
  ]

  print(
    ggplot(
      sign_long,
      aes(x = reorder(mark, pct), y = pct, fill = direction)
    ) +
      geom_col(position = "stack") +
      coord_flip() +
      facet_grid(Variability + `Event Type` ~ feature_set + lambda_selection) +
      scale_fill_manual(
        values = c(
          "Positive (+PSI)" = "#D55E00",
          "Negative (-PSI)" = "#56B4E9"
        ),
        name = "Coefficient sign"
      ) +
      theme_bw() +
      theme(strip.background = element_rect(fill = NA)) +
      labs(
        title = "Coefficient sign per mark (Epigenetic Only)",
        x = NULL,
        y = "% of coefficients",
        caption = paste(
          "Positive = mark associated with higher PSI (active chromatin -> exon inclusion).",
          "Negative = mark associated with lower PSI (repressive chromatin -> exon skipping).",
          "Mixed sign = context-dependent effect; consistent sign = universal association.",
          "Compare SE vs RI: RI marks may differ (intron retention context differs from exon skipping).",
          "Compare local vs long: local = event-overlapping marks; long = 500kb window (distal regulatory).",
          "High vs Low Variability: if sign profiles differ, epigenetic regulation may vary by event dynamics."
        )
      )
  )
  ggsave(file.path(plot_dir_ev, "mark_coef_sign.pdf"), width = 18, height = 10)

  # --- Mark co-occurrence correlation and sig-event heatmap --------------------
  # Loop over Event Type x feature_set x lambda_selection.
  # For each: binary presence matrix -> correlation heatmap + sig-event heatmap.
  # Expected: all-positive co-occurrence (fingerprint). Any negative correlation
  # between antagonistic marks (H3K27ac/H3K27me3) = evidence of genuine signal.

  for (et in to_analyze) {
    for (fs in c("local", "short", "long")) {
      for (ls in c("min", "1se")) {
        hm_sub <- hm_coef_dt[
          `Event Type` == et & feature_set == fs & lambda_selection == ls
        ]
        if (nrow(hm_sub) < 5L) {
          next
        }

        hm_dt_sub <- hm_sub[, .(mean_coef = mean(coef)), by = .(ID, mark)]
        all_ids_sub <- hm_dt_sub[, unique(ID)]
        all_marks_sub <- hm_dt_sub[, unique(mark)]
        pres_grid_sub <- CJ(ID = all_ids_sub, mark = all_marks_sub)
        pres_grid_sub[hm_dt_sub, on = .(ID, mark), selected := 1L]
        pres_grid_sub[is.na(selected), selected := 0L]

        pres_wide_sub <- dcast(
          pres_grid_sub,
          mark ~ ID,
          value.var = "selected",
          fill = 0L
        )
        pres_mat_sub <- as.matrix(pres_wide_sub[, -1L, with = FALSE])
        rownames(pres_mat_sub) <- pres_wide_sub$mark

        valid_marks_sub <- rownames(pres_mat_sub)[
          apply(pres_mat_sub, 1L, var) > 0
        ]
        if (length(valid_marks_sub) < 2L) {
          next
        }

        cor_mat_sub <- cor(t(pres_mat_sub[valid_marks_sub, ]))
        hc_cor_sub <- hclust(as.dist(1 - cor_mat_sub), method = "ward.D2")
        mark_ord_sub <- tools::toTitleCase(
          gsub("_", " ", rownames(cor_mat_sub)[hc_cor_sub$order])
        )
        cor_long_sub <- melt(
          data.table(mark1 = rownames(cor_mat_sub), cor_mat_sub),
          id.vars = "mark1",
          variable.name = "mark2",
          value.name = "correlation"
        )
        cor_long_sub[, `:=`(
          mark1 = factor(
            tools::toTitleCase(gsub("_", " ", mark1)),
            levels = mark_ord_sub
          ),
          mark2 = factor(
            tools::toTitleCase(gsub("_", " ", mark2)),
            levels = mark_ord_sub
          )
        )]

        p_cor_sub <- ggplot(
          cor_long_sub,
          aes(x = mark1, y = mark2, fill = correlation)
        ) +
          geom_tile(color = "white") +
          geom_text(aes(label = sprintf("%.2f", correlation)), size = 2.5) +
          scale_fill_gradient2(
            low = "#56B4E9",
            mid = "white",
            high = "#D55E00",
            midpoint = 0,
            limits = c(-1, 1),
            name = "Pearson r"
          ) +
          theme_bw() +
          theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
          labs(
            title = sprintf(
              "Mark co-occurrence: %s | %s | lambda.%s (n=%d)",
              et,
              fs,
              ls,
              length(all_ids_sub)
            ),
            x = NULL,
            y = NULL,
            caption = paste(
              "All-positive = fingerprint confound signature.",
              "Negative r between active/repressive marks = genuine regulatory signal.",
              "Protocol mRNA.Seq = Protocol total.RNA.Seq (r=1.00): duplicate features."
            )
          )
        print(p_cor_sub)
        ggsave(
          file.path(
            plot_dir_ev,
            sprintf("mark_cooccurrence_%s_%s_%s.pdf", tolower(et), fs, ls)
          ),
          plot = p_cor_sub,
          width = 7,
          height = 6
        )

        # Sig-event heatmap.
        sig_ids_sub <- emp_p_dt[
          sig == TRUE & lambda_selection == ls & ID %in% all_ids_sub,
          unique(ID)
        ]
        pres_sig_sub <- pres_grid_sub[ID %in% sig_ids_sub]
        if (
          pres_sig_sub[, uniqueN(ID)] < 3L || pres_sig_sub[, uniqueN(mark)] < 3L
        ) {
          next
        }

        pres_sig_wide_sub <- dcast(
          pres_sig_sub,
          mark ~ ID,
          value.var = "selected",
          fill = 0L
        )
        pres_sig_mat_sub <- as.matrix(pres_sig_wide_sub[, -1L, with = FALSE])
        rownames(pres_sig_mat_sub) <- pres_sig_wide_sub$mark

        hc_ev_sub <- hclust(
          dist(t(pres_sig_mat_sub), method = "binary"),
          method = "ward.D2"
        )
        hc_mark_sub <- hclust(
          dist(pres_sig_mat_sub, method = "binary"),
          method = "ward.D2"
        )

        sig_long_sub <- melt(
          data.table(mark = rownames(pres_sig_mat_sub), pres_sig_mat_sub),
          id.vars = "mark",
          variable.name = "ID",
          value.name = "selected"
        )
        sig_long_sub[,
          ID := factor(
            ID,
            levels = colnames(pres_sig_mat_sub)[hc_ev_sub$order]
          )
        ]
        sig_long_sub[,
          mark := factor(
            tools::toTitleCase(gsub("_", " ", mark)),
            levels = tools::toTitleCase(
              gsub("_", " ", rownames(pres_sig_mat_sub)[hc_mark_sub$order])
            )
          )
        ]
        # Annotate Variability per event.
        sig_annot_sub <- hm_sub[
          ID %in% sig_ids_sub,
          .(ID = as.character(ID), Variability = as.character(Variability))
        ][, .SD[1L], by = ID]
        sig_long_sub[sig_annot_sub, on = "ID", Variability := Variability]

        p_hm_sub <- ggplot(
          sig_long_sub,
          aes(x = ID, y = mark, fill = factor(selected))
        ) +
          geom_tile(color = "grey85", linewidth = 0.3) +
          scale_fill_manual(
            values = c("0" = "white", "1" = "#D55E00"),
            labels = c("0" = "Not selected", "1" = "Selected"),
            name = NULL
          ) +
          theme_bw() +
          theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
          labs(
            title = sprintf(
              "%d sig events: %s | %s | lambda.%s",
              pres_sig_sub[, uniqueN(ID)],
              et,
              fs,
              ls
            ),
            x = "Events (hierarchical clustering, binary distance)",
            y = NULL,
            caption = paste(
              "Coherent column blocks = shared epigenetic predictor program.",
              "Scattered = idiosyncratic selection.",
              "Clusters likely reflect cell-type fingerprint subgroups."
            )
          )
        print(p_hm_sub)
        ggsave(
          file.path(
            plot_dir_ev,
            sprintf("sig_event_mark_heatmap_%s_%s_%s.pdf", tolower(et), fs, ls)
          ),
          plot = p_hm_sub,
          width = 9,
          height = 4
        )
      }
    }
  }
}
```
