---
title: "Paper figures, tables and numbers"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

<!--
THIS FILE IS THE SOLE SOURCE FOR ANYTHING THE MANUSCRIPT QUOTES.

`09-2-ml-local-new.Rmd` and the `splicing_ml` HTML reports are INTERNAL DIAGNOSTICS and are never
cited. If a figure appears in both, this file wins.

Spec: revision/file-changes/11-paper-figures.Rmd.md

Two hard rules, both load-bearing:

1. DO NOT RECOMPUTE. Read aggregates, reshape, plot. Any new computation belongs upstream as its own
   stage with its own declared output. Recomputing here lets the paper figure and the pipeline disagree,
   which is the same drift this file exists to kill, only in the other direction.
2. MUST STAY FAST, because it re-runs on every figure tweak. No chip_matrix (13.8 GB), no 9 M-row
   aggregated_dt scan, no sweep over the ~68 k per-event screen files. Anything needing those gets a
   small precomputed summary upstream.

Being a LEAF of the DAG is the point: nothing downstream, so editing it re-runs only itself. Contrast
`01`-`05`, where a one-character edit cascades through the 117 GB feature-table build and 34,146 screen
jobs (see revision/deferred-cosmetic-edits.md).
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE, dpi = 150)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(jsonlite)
})
data.table::setDTthreads(2L)

tf <- Sys.getenv(
  "TRANSCRIPT_FILTER",
  getOption("EpiATLAS_AS_PRIMARY_FILTER", "biotype_filtered")
)
event_dir <- file.path("processed_data", "event_models", tf)
fig_dir <- file.path("reports", "paper_figures", tf)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## 0. Conventions

```{r conventions}
# ---- journal figure theme -------------------------------------------------------------------
# Nature: 5-8 pt fonts at print size, one typeface, NO rainbow scale, NO red/green contrast.
# Defined once here and used nowhere else in the project -- the Plotly heatmaps in 09-2 and
# splicing_ml violate all three and are diagnostics, not paper figures.
PAPER_BASE_PT <- 7
theme_paper <- function(base_size = PAPER_BASE_PT) {
  theme_bw(base_size = base_size, base_family = "sans") +
    theme(
      panel.grid.minor = element_blank(),
      panel.border = element_rect(linewidth = 0.3),
      axis.ticks = element_line(linewidth = 0.3),
      strip.background = element_rect(fill = "grey92", linewidth = 0),
      strip.text = element_text(size = base_size, margin = margin(2, 2, 2, 2)),
      legend.key.size = unit(3, "mm"),
      legend.margin = margin(0, 0, 0, 0),
      plot.title = element_text(size = base_size + 1, face = "bold")
    )
}
theme_set(theme_paper())

# Colour-blind-safe, no red/green pairing, no rainbow. Okabe-Ito subset.
PAL <- c(
  "#0072B2",
  "#E69F00",
  "#009E73",
  "#CC79A7",
  "#56B4E9",
  "#D55E00",
  "#666666"
)
scale_fill_paper <- function(...) scale_fill_manual(values = PAL, ...)
scale_colour_paper <- function(...) scale_colour_manual(values = PAL, ...)

# ---- provenance: WHICH ANALYSIS EACH NUMBER AND FIGURE ORIGINALLY CAME FROM -----------------
# `source` records the FILE this Rmd read. That is not provenance: an aggregate CSV says nothing
# about which analysis computed it, under which statistic version, or whether it is still the
# pipeline's own output rather than a scratch prototype. ORIGIN records the ANALYSIS. Both go into
# paper_numbers and into the figure manifest, so a reviewer question ("where does 0.207 come from?")
# is answered by the artifact instead of by grepping the repo.
#
# Anything reading from .claude/scratch/ is flagged UNPROMOTED here, not silently treated as
# pipeline output -- a completed analysis whose only copy is in scratch is not reproducible.
# As of 2026-09-01 nothing is UNPROMOTED: the variance budget (12 keys) and the permuted-groups
# control (7 keys) -- the 19 that had no stage behind them -- became rules `psi_variance_decomposition`
# (step 03b) and `permuted_groups_control` (step 09p), and both are declared inputs of
# `rule paper_figures`. Keep the flagging mechanism: it is what made the gap countable.
ORIGIN <- list(
  atlas = "rMATS-turbo quantification (/nfs/data/IHEC/RNAseq/rmats/) -> atlas summary stage [leg R]",
  cohort = "01-gather-data.Rmd -> 05 create_aggregated_dt; aggregated_dt_filtered_{tf}.csv.gz (modelled cohort)",
  qc = "01-gather-data.Rmd QC pipeline, Policy A (all epirrs retained, qc_flag_count as covariate); QC scope is a SUPERSET of the modelled cohort",
  variance = "PSI variance decomposition, fit-free, leave-one-out group means: 03b-psi-variance-decomposition.R (rule psi_variance_decomposition) [leg A; DrEval-derived]",
  ml = "splicing_ml nested-CV pipeline (run_splicing_ml.py -> pipeline.py), grouped outer CV, per-fold inner hyperparameter search [leg B]",
  ablation = "splicing_ml --feature-groups ablation runs, seqnames-grouped CV [leg A]",
  tier1 = "Tier-1 ridge screen: 09s-ridge-screen.R per event -> 09s-aggregate.R FDR; SCREEN_STAT_VERSION 5, FEATURE_TABLE_VERSION 2 [leg C]",
  floor = "Detection-floor injection sweep: 09f-floor-sample.R -> 09s-ridge-screen.R under EPIATLAS_AS_SCREEN_INJECT -> 09f-floor-score.R (scored on p_pooled_z vs the real screen's 23.7 M null-z reference) [leg C]",
  permuted = "Permuted-groups negative control, paired per event: 09p-permute-groups.R (rule permuted_groups_control), supergroup LABELS permuted so fold count and sizes are preserved exactly [leg C]",
  protocol = "Paired-protocol effect on PSI: 06c-paired-protocol.R (rule paired_protocol), the 36 epirrs carrying BOTH an mRNA-Seq and a total-RNA-Seq uuid; effect estimated WITHIN sample [confound]",
  candidates = "Adjudication of the floor-free pooled-z candidates: screen_results.csv.gz (Tier-1) cross-referenced against floor_results.csv.gz (the injection sweep) -- no refits, reshape only [leg C]",
  design = "fixed by experimental design; not measured"
)
# Set per chunk. pn()/save_fig() ERROR if it is unset, so a new chunk cannot emit an
# unattributed number -- the failure mode this whole registry exists to prevent.
.ORIGIN_CUR <- NULL

# ---- figure writer -------------------------------------------------------------------------
# Vector PDF is the deliverable, written by ggsave. Rendering the DOCUMENT to PDF is deliberately
# avoided: this file's YAML is html_document, so the figure artifacts never depend on the LaTeX
# toolchain. That is not hypothetical caution -- `rule ml_analysis` sat broken for a week because
# 09-2's YAML said pdf_document (rmarkdown::render obeys the YAML, not the .html output_file it is
# handed) and pdflatex could not find bookmark.sty under TinyTeX 2021 against a 2026 remote. Fixed
# 2026-09-01 by correcting that header; the lesson stands regardless.
.FIG <- list()
save_fig <- function(
  plot,
  name,
  width,
  height,
  origin = .ORIGIN_CUR,
  source = NA_character_,
  shows = NA_character_
) {
  if (is.null(origin)) {
    stop(sprintf("save_fig('%s'): .ORIGIN_CUR unset -- attribute the figure", name))
  }
  f <- file.path(fig_dir, paste0(name, ".pdf"))
  ggsave(
    f,
    plot,
    width = width,
    height = height,
    units = "in",
    device = cairo_pdf
  )
  .FIG[[length(.FIG) + 1L]] <<- data.table::data.table(
    figure = name,
    file = f,
    width_in = width,
    height_in = height,
    shows = shows,
    origin = origin,
    source = paste(source, collapse = "; ")
  )
  message(sprintf("figure written: %s (%.1f x %.1f in)", f, width, height))
  invisible(f)
}

# ---- paper_numbers ------------------------------------------------------------------------
# EVERY number the manuscript states goes through pn(). `filter` is NOT optional: an unstated
# predicate is exactly the drift this file exists to kill. Live example -- |screen_R2| > 1 is
# 50.06% over all finite rows and 51.44% once note-carrying rows are dropped. Both correct.
.PN <- list()
pn <- function(
  key,
  value,
  unit,
  source,
  filter = NA_character_,
  n = NA_real_,
  stat_version = NA_integer_,
  note = NA_character_,
  origin = .ORIGIN_CUR
) {
  stopifnot(is.character(key), length(key) == 1L, nzchar(key))
  # No unattributed numbers. A key whose originating analysis is unknown cannot be defended in
  # review, so this is a hard error rather than an NA default.
  if (is.null(origin)) {
    stop(sprintf("pn('%s'): .ORIGIN_CUR unset -- attribute the number", key))
  }
  # setDT(list(...)), NOT data.table(...): `key` is a RESERVED ARGUMENT of data.table(), so
  # data.table(key = "cohort.uuids", ...) is read as "set that column as the table's key" and
  # dies with `some columns are not in the data.table: [cohort.uuids]`. Passing the list
  # positionally to setDT() makes `key` an ordinary column. (Same trap for `check.names` and
  # `stringsAsFactors`; the other column names here are safe.)
  .PN[[length(.PN) + 1L]] <<- data.table::setDT(list(
    key = key,
    value = as.numeric(value),
    unit = unit,
    n = as.numeric(n),
    filter = filter,
    source = source,
    origin = origin,
    stat_version = as.integer(stat_version),
    note = note
  ))
  invisible(value)
}

# PER OUTER FOLD, never collapsed. This is nested CV: every outer fold runs its own inner
# hyperparameter search, so the folds are DIFFERENT fitted models with different parameters. One
# mean AUROC asserts a single model that does not exist, hides the fold spread, and cannot meet
# the journal requirement to define error bars and give exact n. So read
# `metric_fold_data[[metric]]$points` (model x outer_fold x score) and summarise only when
# reporting -- always beside sd and the fold count. `metric_heatmap$z`/`$sd` are the fold mean and
# spread, both derivable from these points, so the heatmap is never read for numbers.
fold_metric_of <- function(payload_json, metric = "auroc") {
  if (is.na(payload_json) || !file.exists(payload_json)) {
    return(NULL)
  }
  d <- jsonlite::fromJSON(payload_json, simplifyVector = FALSE)[[1]]
  pts <- d$metric_fold_data[[metric]]$points
  if (is.null(pts) || !length(pts)) {
    return(NULL)
  }
  out <- data.table::rbindlist(lapply(pts, function(x) {
    data.table::data.table(
      model = x$model_name,
      outer_fold = as.integer(x$outer_fold),
      value = as.numeric(x$score)
    )
  }))
  out[, payload := payload_json][]
}

# Emit every fold AND the summary, so nothing downstream re-aggregates or guesses how.
pn_folds <- function(prefix, dt, source, filter, origin = .ORIGIN_CUR) {
  for (m in unique(dt$model)) {
    v <- dt[model == m][order(outer_fold)]
    for (i in seq_len(nrow(v))) {
      pn(
        sprintf("%s.%s.fold%d", prefix, m, v$outer_fold[i]),
        v$value[i],
        "auroc",
        source,
        filter = paste(filter, "- one outer fold; own tuned hyperparameters"),
        origin = origin
      )
    }
    pn(
      sprintf("%s.%s.mean", prefix, m),
      mean(v$value),
      "auroc",
      source,
      filter = paste(filter, "- mean over outer folds"),
      n = nrow(v),
      origin = origin
    )
    pn(
      sprintf("%s.%s.sd", prefix, m),
      stats::sd(v$value),
      "auroc",
      source,
      filter = paste(filter, "- sd over outer folds"),
      n = nrow(v),
      origin = origin
    )
  }
}

summarise_folds <- function(d, by) {
  d[,
    .(
      folds = .N,
      mean = mean(value),
      sd = stats::sd(value),
      min = min(value),
      max = max(value)
    ),
    by = by
  ]
}
find_payload <- function(root, et, split) {
  f <- Sys.glob(sprintf(
    "%s/%s_%s_both_%s/reports/subset_report_*.json",
    root,
    et,
    tf,
    split
  ))
  if (length(f)) f[1] else NA_character_
}

# ---- input availability ---------------------------------------------------------------------
# Gate every section on its inputs and print one manifest, rather than failing on the first
# missing file. What is still genuinely absent is the atlas summary (spec 6.2, blocked on the
# rMATS provenance audit) and the cohort-counts artifact (spec 6.3, self-bootstrapped below).
INPUTS <- list(
  screen = file.path(event_dir, "screen_results.csv.gz"),
  hits = file.path(event_dir, sprintf("tier1_hits_%s.txt", tf)),
  floor = file.path(event_dir, "floor", "floor_results.csv.gz"),
  session = sprintf("processed_data/session_09_1_ml_local_%s.rds", tf),
  variance = sprintf("processed_data/psi_variance_decomposition_%s.csv", tf),
  variance_alt = sprintf(
    ".claude/scratch/psi_variance_decomposition_%s.csv",
    tf
  ),
  # per-transcript_filter, like everything from 02-2 on: event `ID` is a row position in that
  # filter's event_gr, so one shared path reads back as silently wrong events under matching IDs.
  permuted = sprintf("processed_data/permuted_groups_control_%s.csv", tf),
  permuted_alt = ".claude/scratch/permuted_groups_control_result.csv",
  qc = "processed_data/qc_summary.csv",
  cohort = sprintf("processed_data/cohort_counts_%s.csv", tf),
  protocol = sprintf("processed_data/paired_protocol_dPSI_summary_%s.csv.gz", tf),
  protocol_conc = sprintf("processed_data/paired_protocol_concordance_%s.csv.gz", tf),
  protocol_pool = sprintf("processed_data/paired_protocol_pooled_summary_%s.csv.gz", tf),
  atlas = sprintf("processed_data/atlas_summary_%s.csv", tf)
)
have <- vapply(INPUTS, file.exists, logical(1))
# Both analyses were PROMOTED to pipeline stages on 2026-09-01 (rules `psi_variance_decomposition`
# and `permuted_groups_control`), and both are declared inputs of `rule paper_figures` -- so under
# Snakemake these files are always present and neither branch below fires. The scratch fallbacks are
# kept for STANDALONE runs of this Rmd only, and they warn loudly because a paper number sourced
# from scratch is not reproducible from the pipeline: the scratch copy can predate the stage, can
# have been produced at a different N_EVENTS, and has no declared provenance.
if (!have[["variance"]] && have[["variance_alt"]]) {
  INPUTS$variance <- INPUTS$variance_alt
  have[["variance"]] <- TRUE
  warning(
    "variance decomposition read from .claude/scratch/, NOT the pipeline output -- ",
    "run `snakemake processed_data/psi_variance_decomposition_", tf, ".csv`; ",
    "numbers from this path must not be quoted"
  )
}
if (!have[["permuted"]] && have[["permuted_alt"]]) {
  INPUTS$permuted <- INPUTS$permuted_alt
  have[["permuted"]] <- TRUE
  warning(
    "permuted-groups control read from .claude/scratch/, NOT the pipeline output -- ",
    "run `snakemake processed_data/permuted_groups_control_", tf, ".csv`; ",
    "the scratch copy was measured at n = 100 and carries no declared provenance"
  )
}
print(data.table(
  input = names(have),
  path = unlist(INPUTS[names(have)]),
  available = unname(have)
))
```

## 1. Atlas (leg R)

**Origin:** rMATS-turbo quantification over `/nfs/data/IHEC/RNAseq/rmats/{tf}/` (1,522 UUIDs, 745,545 events, 5 event classes), summarised once by the atlas-summary stage. **This file must never read that 136 GB tree directly.**

```{r atlas, eval = have[["atlas"]]}
.ORIGIN_CUR <- ORIGIN$atlas
# Requires the atlas summary artifact (spec 6.2): per-class event counts, UUID coverage,
# per-ontology PSI summary, library-strata counts, computed ONCE from the 136 GB rMATS tree.
# This file must never read that tree directly.
atlas <- fread(INPUTS$atlas)
print(atlas)
# `uuids` is repeated on every class row (it is a property of the run set, not of the class), so
# take one; `events` is per class, so sum.
stopifnot(uniqueN(atlas$uuids) == 1L)
# QUANTIFIED, measured from the data rather than taken from the command table's declaration. The two
# agree at 1,522 -- but only because 02-3b reads the count columns with colClasses="character".
# Without that, fread reads a two-sample value like "10,11" as the number 10.11 (comma as decimal
# separator), three N=2 runs appear to have lost a sample, and this key reads 1,519. The measurement
# is kept despite currently agreeing with the declaration: a real gap would silently mislabel every
# sample after it, and this is the headline count of the resource leg.
pn("atlas.uuids", atlas[["uuids"]][1], "count", INPUTS$atlas,
  filter = "RNA-seq uuids actually QUANTIFIED by rMATS (MATS.JC.txt field counts)",
  n = atlas[["runs"]][1])
if ("uuids_declared" %in% names(atlas)) {
  pn("atlas.uuids_declared", atlas[["uuids_declared"]][1], "count", INPUTS$atlas,
    filter = "uuids named in the rMATS command table's uuid_string -- a CLAIM, not a measurement",
    note = "currently equal to the quantified count; quote atlas.uuids, which is measured")
}
pn(
  "atlas.events",
  sum(atlas[["events"]]),
  "count",
  INPUTS$atlas,
  filter = "all 5 rMATS classes"
)
pn("atlas.runs", atlas[["runs"]][1], "count", INPUTS$atlas,
  filter = "rMATS invocations = library strata (paired/single x libtype x read length)")
# Per class, because the manuscript contrasts what was quantified against what was modelled, and
# that contrast is only legible per class: SE and MXE carry ~98% of the events while RI, the class
# where the epigenetic contribution sits, has just 2,459.
for (i in seq_len(nrow(atlas))) {
  pn(sprintf("atlas.events.%s", tolower(atlas$event_type[i])),
    atlas$events[i], "count", INPUTS$atlas,
    filter = sprintf("fromGTF.%s.txt, identical across all %d strata",
      atlas$event_type[i], atlas$n_runs_checked[i]))
}
```

```{r atlas-missing, eval = !have[["atlas"]], results='asis'}
cat(
  "> **Not available.** Blocked on the atlas summary artifact (spec 6.2), which is itself blocked on",
  "the provenance audit of `/nfs/data/IHEC/RNAseq/rmats/biotype_filtered/` (PATH-FORWARD R2):",
  "three RI files disagree and the smallest is live.\n"
)
```

## 2. Cohort counts

**Origin:** `01-gather-data.Rmd` (QC/metadata scope) and `05 create_aggregated_dt` -> `aggregated_dt_filtered_{tf}.csv.gz` (modelled scope). The two differ and conflating them is the recurring error: **modelled 379 epirrs / 415 UUIDs**, QC scope 405 / 441, a clean superset.

```{r cohort}
.ORIGIN_CUR <- ORIGIN$cohort
# Two cohorts, and conflating them is the recurring error: QC/metadata scope is 405 epirrs /
# 441 UUIDs, the MODELLED cohort is 379 / 415. Clean superset (26 QC rows never modelled, 0
# modelled UUIDs missing QC). The 26 are the AMED-CREST gap and ALL were mRNA-Seq.
# Read only. This used to derive the counts inline from aggregated_dt and cache them to a path that
# was NOT a declared output of rule paper_figures -- untracked by Snakemake, so a stale cache would
# have been read silently, on the numbers this project misquotes most often. It was also the only
# place this file touched aggregated_dt at all, against rule 2 in the header. Promoted 2026-09-01 to
# stage 05c (rule cohort_counts), now a declared input.
stopifnot(have[["cohort"]])
co_all <- fread(INPUTS$cohort)
co <- co_all[scope == "modelled"]
stopifnot(nrow(co) == 1L)
qcs <- if (have[["qc"]]) fread(INPUTS$qc) else NULL
print(co_all)

pn(
  "cohort.uuids",
  co$uuids[1],
  "count",
  INPUTS$cohort,
  filter = "modelled (aggregated_dt_filtered)"
)
pn("cohort.epirrs", co$epirrs[1], "count", INPUTS$cohort, filter = "modelled")
pn("cohort.mrna", co$mrna[1], "count", INPUTS$cohort, filter = "modelled")
pn(
  "cohort.total_rna",
  co$total_rna[1],
  "count",
  INPUTS$cohort,
  filter = "modelled"
)
pn(
  "cohort.both_protocols",
  co$both_protocols[1],
  "count",
  INPUTS$cohort,
  filter = "modelled"
)
pn(
  "cohort.total_rna_frac",
  co$total_rna[1] / co$uuids[1],
  "fraction",
  INPUTS$cohort,
  filter = "modelled",
  note = "54.7%: worse than the 214/227 near-balance older notes imply"
)
# THE RESOURCE CLAIM: the analysis used a small slice of what rMATS quantified. Derived, not
# retyped. Emitted HERE and not in section 1 because `co` is built in this chunk -- section 1 runs
# first, so an `exists("co")` guard there would silently never fire.
if (have[["atlas"]] && exists("atlas")) {
  pn(
    "atlas.modelled_uuid_frac",
    co$uuids[1] / atlas[["uuids"]][1],
    "fraction",
    "derived",
    filter = sprintf("modelled %d of %d atlas uuids", co$uuids[1], atlas[["uuids"]][1]),
    origin = ORIGIN$atlas,
    note = "the manuscript's '27% of the experiments'"
  )
}
if (!is.null(qcs)) {
  # kept ADJACENT to the modelled counts on purpose -- split apart, the 405-vs-379 drift recurs
  pn(
    "cohort.qc_epirrs",
    uniqueN(qcs$epirr_id_without_version),
    "count",
    INPUTS$qc,
    filter = "QC/metadata scope (superset)",
    origin = ORIGIN$qc
  )
}
```

## 3. PSI variance budget (leg A)

**Origin:** fit-free PSI variance decomposition, leave-one-out group means (DrEval-derived). Pipeline stage **03b** (`03b-psi-variance-decomposition.R`, rule `psi_variance_decomposition`), promoted out of `.claude/scratch/` on 2026-09-01. It sits before `create_aggregated_dt` on purpose: every input stops at stage 03 plus sample metadata, so the budget is established **before any epigenetic feature is considered** — which is the claim it exists to support.

```{r variance, eval = have[["variance"]], fig.width=6, fig.height=3}
.ORIGIN_CUR <- ORIGIN$variance
vb <- fread(INPUTS$variance)
vb[, c("event_type", "variability") := tstrsplit(stratum, " / ", fixed = TRUE)]
# LOO columns are the honest ones (in-sample group means over-fit); they differ from in-sample by
# <= 0.3 points here, but quote LOO.
vb[, within_event := 1 - r2_event_loo]
print(vb[, .(
  stratum,
  var_psi,
  between = r2_event_loo,
  within = within_event,
  ontology = r2_ontology_loo
)])

long <- melt(
  vb[, .(stratum, between = r2_event_loo, within = within_event)],
  id.vars = "stratum",
  variable.name = "component",
  value.name = "share"
)
p <- ggplot(long, aes(share, stratum, fill = component)) +
  geom_col(width = 0.7) +
  scale_fill_paper(name = NULL) +
  scale_x_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(
    x = "share of PSI variance",
    y = NULL,
    title = "Event identity dominates PSI variance"
  )
print(p)
save_fig(
  p,
  "fig2_variance_budget",
  5.5,
  2.6,
  shows = "between- vs within-event share of PSI variance, per Event Type x Variability stratum",
  source = INPUTS$variance
)

for (i in seq_len(nrow(vb))) {
  k <- gsub("[^A-Za-z]+", "_", tolower(vb$stratum[i]))
  pn(
    paste0("variance.between_event.", k),
    vb$r2_event_loo[i],
    "fraction",
    INPUTS$variance,
    filter = "leave-one-out group means",
    n = vb$n_rows[i]
  )
  pn(
    paste0("variance.within_event.", k),
    vb$within_event[i],
    "fraction",
    INPUTS$variance,
    filter = "leave-one-out group means",
    n = vb$n_rows[i]
  )
}
```

## 4. The artifacts (leg B)

**Origin:** `splicing_ml` nested-CV pipeline (`run_splicing_ml.py` -> `pipeline.py`), grouped outer CV with a per-fold inner hyperparameter search. Numbers are read from the small report payloads, never the multi-GB result pickles.

```{r artifacts}
.ORIGIN_CUR <- ORIGIN$ml
# Uses fold_metric_of()/find_payload() from section 0. Reads the small purpose-built report payloads,
# NOT the multi-GB splicing_ml_results pickles -- this file must stay fast enough to re-run on
# every figure tweak.
cfgs <- CJ(split = c("seqnames", "ontology"), et = c("RI", "SE"), unique = TRUE)
cfgs[,
  payload := vapply(
    seq_len(.N),
    function(i) find_payload("splicing_ml/output", et[i], split[i]),
    character(1)
  )
]
grid <- rbindlist(
  lapply(seq_len(nrow(cfgs)), function(i) {
    a <- fold_metric_of(cfgs$payload[i])
    if (is.null(a)) {
      return(NULL)
    }
    a[, `:=`(split = cfgs$split[i], et = cfgs$et[i])][]
  }),
  fill = TRUE
)
# The prior baseline must be exactly 0.500 IN EVERY FOLD -- the sanity floor that caught a
# transposed heatmap axis once.
bl <- grid[grepl("Baseline", model), value]
if (length(bl)) {
  stopifnot(all(abs(bl - 0.5) < 1e-9))
}
grid <- grid[!grepl("Baseline", model)]
print(summarise_folds(grid, c("split", "et", "model"))[order(split, et, model)])

if (nrow(grid)) {
  # One point per outer fold with a mean crossbar. A bar of the mean alone would assert a single
  # model where there are five, each with its own tuned hyperparameters.
  p <- ggplot(
    grid,
    aes(value, interaction(model, et, sep = " / "), colour = split)
  ) +
    stat_summary(
      fun = mean,
      geom = "crossbar",
      width = 0.55,
      linewidth = 0.25,
      position = position_dodge(width = 0.7)
    ) +
    geom_point(
      position = position_dodge(width = 0.7),
      size = 0.7,
      alpha = 0.85
    ) +
    geom_vline(xintercept = 0.5, linetype = 2, linewidth = 0.3) +
    scale_colour_paper(name = "CV grouping") +
    coord_cartesian(xlim = c(0.5, 1)) +
    labs(
      x = "test AUROC (one point per outer fold)",
      y = NULL,
      title = "Holding out cell types leaks event identity; holding out events does not"
    )
  print(p)
  save_fig(
    p,
    "fig3_artifacts",
    5.5,
    2.8,
    shows = "per-outer-fold test AUROC, seqnames- vs ontology-grouped CV (the event-identity leak)",
    source = cfgs$payload
  )
  for (i in seq_len(nrow(cfgs))) {
    d <- grid[split == cfgs$split[i] & et == cfgs$et[i]]
    if (!nrow(d)) {
      next
    }
    pn_folds(
      sprintf("cv.%s.%s.auroc", cfgs$split[i], tolower(cfgs$et[i])),
      d,
      cfgs$payload[i],
      "both variability tiers; classification"
    )
  }
  # Capacity gap PAIRED WITHIN FOLD: xgb and linear share the outer split, so the per-fold
  # difference is the paired contrast and its sd is interpretable.
  w <- dcast(grid, split + et + outer_fold ~ model, value.var = "value")
  if (all(c("linear", "xgb") %in% names(w))) {
    w[, gap := xgb - linear]
    gsum <- w[,
      .(folds = .N, mean = mean(gap), sd = stats::sd(gap)),
      by = .(split, et)
    ]
    print(gsum[order(split, et)])
    for (i in seq_len(nrow(gsum))) {
      pn(
        sprintf(
          "cv.%s.%s.xgb_minus_linear.mean",
          gsum$split[i],
          tolower(gsum$et[i])
        ),
        gsum$mean[i],
        "auroc",
        "derived",
        filter = "capacity gap = leak signature; paired within outer fold",
        n = gsum$folds[i]
      )
      pn(
        sprintf(
          "cv.%s.%s.xgb_minus_linear.sd",
          gsum$split[i],
          tolower(gsum$et[i])
        ),
        gsum$sd[i],
        "auroc",
        "derived",
        filter = "capacity gap; sd of the paired within-fold difference",
        n = gsum$folds[i]
      )
    }
  }
}
```

## 4b. The preprint's own split, and what survives it (leg B)

**Origin:** the same `splicing_ml` runs as section 4, reassembled as the argument the manuscript has to
make. No new computation.

The preprint reports two test sets separately and describes the first in its own words
(`sn-article.tex:411`): *"testing on **seen events but unknown cell/tissue types** leads to near-perfect
performances … emphasizing distinct mechanisms per event over cell-type specificity."* That design is
`ontology`-grouped CV. Held-out samples are new, but the **events are not** — an event's other samples
are in training, so a high-capacity model can fingerprint the event itself rather than learn anything
that transfers. The near-perfect number is the signature of that, and the conclusion drawn from it is
backwards.

This section walks the four steps, each a measured number, from the preprint's claim to what replaces it.

```{r preprint-correction, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6.5, fig.height=3.4}
.ORIGIN_CUR <- ORIGIN$ml
if (!nrow(grid)) {
  cat("splicing_ml payloads unavailable; section 4 did not load.\n")
} else {
  # STEP 1 + 2: the two splits side by side. `ontology` IS the preprint's "unknown cell/tissue types"
  # test set; `seqnames` holds out whole chromosomes, so the events are genuinely unseen.
  sm <- summarise_folds(grid, c("split", "et", "model"))[order(split, et, model)]
  print(sm)

  # STEP 3: THE LEAK SIGNATURE. Capacity gap = xgb - linear, PAIRED WITHIN OUTER FOLD -- both models
  # share the same deterministic split, so fold k is the same held-out data and the sd of the paired
  # difference is interpretable. A linear model cannot carve per-event islands out of feature space;
  # a boosted ensemble can. So a LARGE gap under ontology CV and NO gap under seqnames CV is the
  # fingerprint of event-identity memorisation, not of extra signal that only the tree can reach.
  w <- dcast(grid, split + et + outer_fold ~ model, value.var = "value")
  gap <- w[, .(folds = .N, mean = mean(xgb - linear), sd = stats::sd(xgb - linear)),
    by = .(split, et)][order(et, split)]
  print(gap)

  p <- ggplot(gap, aes(x = factor(split, levels = c("ontology", "seqnames")), y = mean,
      colour = `et`, group = et)) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.3, colour = PAL[7]) +
    geom_line(linewidth = 0.4) +
    geom_pointrange(aes(ymin = mean - sd, ymax = mean + sd), size = 0.3, linewidth = 0.4) +
    scale_colour_paper(name = NULL) +
    labs(
      x = NULL, y = "test AUROC, xgb − linear (paired within fold)",
      title = "The capacity gap is the leak, and it disappears when events are held out",
      subtitle = "ontology CV = the preprint's \"unknown cell/tissue types\"; seqnames CV = events genuinely unseen"
    ) +
    scale_x_discrete(labels = c(ontology = "seen events,\nunseen samples", seqnames = "unseen events"))
  print(p)
  save_fig(p, "fig3b_preprint_correction", 5.5, 3.0,
    shows = "capacity gap (xgb − linear, paired within outer fold) under the preprint's split vs event-held-out CV",
    source = cfgs$payload)
}
```

The steps above use only the two CV runs. **The full argument — including what the epigenome adds over
a sequence-only reference, and what the locus-matched null returns — is stated in section 7b**, which
has to come after the sections that produce those numbers.

## 5. Tier-1 screen (leg C)

**Origin:** `09s-ridge-screen.R` (one job per event, 34,146 events) -> `09s-aggregate.R` (BH/Storey FDR per `transcript_filter` x `Event Type` x `feature_set`). `SCREEN_STAT_VERSION` **5**, `FEATURE_TABLE_VERSION` **2**.

```{r tier1, eval = have[["screen"]], fig.width=6.5, fig.height=3.2}
.ORIGIN_CUR <- ORIGIN$tier1
scr <- fread(INPUTS$screen)
qthr <- 0.1
stopifnot(uniqueN(scr$stat_version) == 1L)
SV <- scr$stat_version[1]

fam <- scr[
  is.finite(q),
  .(
    m = .N,
    min_q = min(q),
    min_q_pz = min(q_pooled_z, na.rm = TRUE),
    pct_p05 = 100 * mean(p_emp <= 0.05),
    pct_ppz05 = 100 * mean(p_pooled_z <= 0.05, na.rm = TRUE)
  ),
  by = .(`Event Type`, feature_set)
][order(`Event Type`, feature_set)]
print(fam)

pn(
  "tier1.hits",
  sum(scr$q < qthr, na.rm = TRUE),
  "count",
  INPUTS$screen,
  filter = "q < 0.1, primary exceedance rule",
  stat_version = SV
)
pn(
  "tier1.hits_pooled_z",
  sum(scr$q_pooled_z < qthr, na.rm = TRUE),
  "count",
  INPUTS$screen,
  filter = "q_pooled_z < 0.1, floor-free sensitivity",
  stat_version = SV,
  note = "disagrees with primary since v5; 4 of 7 survive a same-n peer check"
)
pn(
  "tier1.min_q",
  min(fam$min_q),
  "q",
  INPUTS$screen,
  filter = "over all 6 families",
  stat_version = SV
)
pn(
  "tier1.min_q_pooled_z",
  min(fam$min_q_pz),
  "q",
  INPUTS$screen,
  filter = "over all 6 families",
  stat_version = SV
)
pn(
  "tier1.hits_file_bytes",
  file.info(INPUTS$hits)$size,
  "bytes",
  INPUTS$hits,
  filter = "0 bytes = no hits written by 09s-aggregate.R"
)

# p-uniformity: the calibration claim, stated per family because SE outnumbers RI ~18:1 and a
# pooled panel would be an SE panel with RI invisible.
ph <- scr[is.finite(p_emp)]
NBINS <- 20L
# the uniform expectation, per facet: m / nbins. Drawn so "flat" is judged against a line rather
# than by eye -- a p-histogram piled at 0 would mean an optimistic null and nothing concludable.
expect <- ph[, .(y = .N / NBINS), by = .(`Event Type`, feature_set)]
p <- ggplot(ph, aes(p_emp)) +
  geom_histogram(
    boundary = 0,
    bins = NBINS,
    fill = PAL[1],
    colour = "white",
    linewidth = 0.1
  ) +
  geom_hline(
    data = expect,
    aes(yintercept = y),
    linetype = 2,
    linewidth = 0.3,
    colour = PAL[6]
  ) +
  facet_grid(`Event Type` ~ feature_set, scales = "free_y") +
  labs(
    x = "empirical p (one-sided exceedance)",
    y = "events",
    title = sprintf("Calibrated null in every family (stat_version %d)", SV)
  )
print(p)
save_fig(
  p,
  "fig4_tier1_pvalues",
  6.5,
  3.0,
  shows = "empirical p histogram per (Event Type x feature_set) family vs the uniform expectation",
  source = INPUTS$screen
)

pn(
  "tier1.pct_p05_min",
  min(fam$pct_p05),
  "percent",
  INPUTS$screen,
  filter = "per family, expected 5",
  stat_version = SV
)
pn(
  "tier1.pct_p05_max",
  max(fam$pct_p05),
  "percent",
  INPUTS$screen,
  filter = "per family, expected 5",
  stat_version = SV
)

# Numeric health. Predicate stated TWICE over because both are quoted in the docs and they differ.
fin <- scr[is.finite(p_emp) & is.finite(screen_R2) & is.finite(effect)]
pn(
  "tier1.blowup_gt1",
  mean(abs(fin$screen_R2) > 1),
  "fraction",
  INPUTS$screen,
  filter = "|screen_R2| > 1 over all finite rows",
  stat_version = SV
)
pn(
  "tier1.blowup_gt1_notefree",
  fin[is.na(note) | note == "", mean(abs(screen_R2) > 1)],
  "fraction",
  INPUTS$screen,
  filter = "|screen_R2| > 1, note-carrying rows excluded",
  stat_version = SV
)
# >10 and >100 are quoted in the docs beside >1 and the thresholds get confused routinely
# (the n/lambda contrast belongs to >10 while the headline fraction is >1), so emit all three.
pn(
  "tier1.blowup_gt10",
  mean(abs(fin$screen_R2) > 10),
  "fraction",
  INPUTS$screen,
  filter = "|screen_R2| > 10 over all finite rows",
  stat_version = SV
)
pn(
  "tier1.blowup_gt100",
  mean(abs(fin$screen_R2) > 100),
  "fraction",
  INPUTS$screen,
  filter = "|screen_R2| > 100 over all finite rows",
  stat_version = SV
)
# n/lambda contrast MUST be faceted by feature space: the three spaces disagree in DIRECTION
# (local 57 vs 351, short 299 vs 347, long 357 vs 103 REVERSED). Pooling gives a meaningless 305/317.
bw <- fin[,
  .(
    pct_gt1 = 100 * mean(abs(screen_R2) > 1),
    n_blown = median(as.numeric(n_samples[abs(screen_R2) > 1])),
    n_rest = median(as.numeric(n_samples[abs(screen_R2) <= 1])),
    lam_blown = median(lambda[abs(screen_R2) > 1]),
    lam_rest = median(lambda[abs(screen_R2) <= 1])
  ),
  by = feature_set
]
print(bw)
for (i in seq_len(nrow(bw))) {
  fs <- bw$feature_set[i]
  pn(
    sprintf("tier1.blowup_pct.%s", fs),
    bw$pct_gt1[i],
    "percent",
    INPUTS$screen,
    filter = "|screen_R2| > 1, per feature space",
    stat_version = SV
  )
  # BOTH sides: a median for the blown group alone is not a contrast, and the three spaces
  # disagree in DIRECTION, so only the pair is interpretable.
  pn(
    sprintf("tier1.blowup_median_n.%s.blown", fs),
    bw$n_blown[i],
    "count",
    INPUTS$screen,
    filter = "|screen_R2| > 1, per feature space",
    stat_version = SV
  )
  pn(
    sprintf("tier1.blowup_median_n.%s.rest", fs),
    bw$n_rest[i],
    "count",
    INPUTS$screen,
    filter = "|screen_R2| <= 1, per feature space",
    stat_version = SV
  )
  pn(
    sprintf("tier1.blowup_median_lambda.%s.blown", fs),
    bw$lam_blown[i],
    "lambda",
    INPUTS$screen,
    filter = "|screen_R2| > 1, per feature space",
    stat_version = SV
  )
  pn(
    sprintf("tier1.blowup_median_lambda.%s.rest", fs),
    bw$lam_rest[i],
    "lambda",
    INPUTS$screen,
    filter = "|screen_R2| <= 1, per feature space",
    stat_version = SV
  )
}
```

```{r tier1-missing, eval = !have[["screen"]], results='asis'}
cat(
  "> **Not available**: `screen_results.csv.gz` missing. Run the screen and `screen_aggregate`.\n"
)
```

## 6. Statistical power of the Tier-1 screen (leg C)

**Origin:** detection-floor injection sweep — `09f-floor-sample.R` (596 events, 8 strata) -> `09s-ridge-screen.R` under `EPIATLAS_AS_SCREEN_INJECT` (**the production worker, not a re-implementation**) -> `09f-floor-score.R`, which scores on `p_pooled_z` against the real screen's 23,725,212 null-z values and self-validates against `09s-aggregate.R` to 5.6e-16 before scoring anything.

```{r floor, eval = have[["floor"]]}
.ORIGIN_CUR <- ORIGIN$floor
fl <- fread(INPUTS$floor)
stopifnot(all(
  c(
    "screen_R2",
    "detected",
    "p_pooled_z",
    "inject_rho",
    "blowup",
    "feature_set"
  ) %in%
    names(fl)
))

# CONSTRAINT 1: READ `detected`, NEVER recompute it, and never score on p_emp. The exceedance p is
# floored at 1/(R_used+1), so its minimum measures the ROTATION COUNT, not the data -- scoring
# detection on it would report a property of the sweep's own budget. 09f-floor-score.R decides
# detection on p_pooled_z against the real screen's 23.7 M null-z reference (leave-one-event-out
# standardised within Event Type x feature_set, self-validated against 09s-aggregate.R to 5.6e-16
# before it scores anything). Recomputing here would also break this file's rule 1.
#
# CONSTRAINT 2: the sweep injects into ONE feature space, but the worker screens all three, so
# two-thirds of these rows are spaces no signal was put into (they sit at R2 ~ -3.5 and would pile
# into the lowest bin). `local` only is deliberate: at rho=1 the injection fails to recover a signal
# that is BY CONSTRUCTION present in `long`'s own columns, because `long` runs at median lambda ~ 7
# against `local`'s ~ 8,050 -- three to four orders less regularisation. That is itself a finding:
# the screen's power lives in the event-overlapping features.
inj_fs <- if ("inject_fs" %in% names(fl)) {
  unique(stats::na.omit(fl$inject_fs))
} else {
  "local"
}
stopifnot(length(inj_fs) == 1L)
fl <- fl[feature_set == inj_fs]
message(sprintf(
  "scoring the injected space only: %s (%d rows)",
  inj_fs,
  nrow(fl)
))

# CONSTRAINT 3: rho <= 0.4 ONLY. rho is a VARIANCE FRACTION, not a correlation -- the worker builds
# sqrt(rho)*signal + sqrt(1-rho)*noise, so at rho=1 the noise coefficient is exactly 0 and the
# target is noiseless and 10-column-deterministic. Kernel ridge on the CONTROL features then fails
# erratically, inflating the null's SPREAD without shifting its mean, and since z =
# (R2 - null_mean)/null_sd, z FALLS as R2 rises. So rho=1 is DEGENERATE, not strong-signal, and the
# old "rho=1 must be detected" anchor was wrong (removed 2026-09-01). It is reported below as a
# diagnostic and excluded from every power and ceiling number.
RHO_MAX_VALID <- 0.4
```

```{r floor-anchors, eval = have[["floor"]]}
.ORIGIN_CUR <- ORIGIN$floor
# The one anchor that still holds: rho=0 must not be detected. This is the calibration claim that
# makes the screen's 0-hit result a real result rather than a broken test.
a0 <- fl[inject_rho == 0]
anch <- fl[,
  .(
    n = .N,
    med_R2 = median(screen_R2),
    med_null_sd = median(null_R2_sd, na.rm = TRUE),
    med_z = median(z_inj, na.rm = TRUE),
    det_pct = 100 * mean(detected)
  ),
  by = inject_rho
][order(inject_rho)]
print(anch)
stopifnot(nrow(a0) > 0L, mean(a0$detected) < 0.01)
pn(
  "power.null_detection_rate",
  mean(a0$detected),
  "fraction",
  INPUTS$floor,
  filter = sprintf("inject_rho = 0, %s space; MUST be ~0", inj_fs),
  n = nrow(a0),
  note = "calibrated null: the screen's 0 hits is a result, not a broken test"
)

# Quantify the degeneracy rather than merely asserting it: null sd blows up while R2 rises.
#
# MUST be stratified by Event Type AND restricted to numerically stable events. Pooled, this ratio
# reads 1.24 and the effect vanishes -- because blown-up events' median null sd FALLS over the same
# interval (RI 95.5 -> 57.9) and cancels the stable events' rise. Two populations moving in opposite
# directions, so a pooled median is meaningless. Stratified: RI x18.4, SE x26.5, while median R2
# rises only x10.0 / x19.5 -- and z = (R2 - null_mean)/null_sd, so z falls as R2 rises.
if (all(c(0.4, 1) %in% fl$inject_rho)) {
  infl <- fl[
    blowup == FALSE & inject_rho %in% c(0.4, 1),
    .(
      sd_ratio = median(null_R2_sd[inject_rho == 1], na.rm = TRUE) /
        median(null_R2_sd[inject_rho == 0.4], na.rm = TRUE),
      r2_ratio = median(screen_R2[inject_rho == 1]) /
        median(screen_R2[inject_rho == 0.4]),
      n1 = sum(inject_rho == 1)
    ),
    by = `Event Type`
  ]
  print(infl)
  for (i in seq_len(nrow(infl))) {
    etl <- tolower(infl$`Event Type`[i])
    pn(
      sprintf("power.rho1_null_sd_inflation.%s", etl),
      infl$sd_ratio[i],
      "ratio",
      INPUTS$floor,
      filter = "median null_R2_sd at rho=1 over rho=0.4; NUMERICALLY STABLE events only",
      n = infl$n1[i],
      note = "pooled over blowup this reads 1.24: blown-up events' null sd FALLS and cancels it"
    )
    pn(
      sprintf("power.rho1_r2_inflation.%s", etl),
      infl$r2_ratio[i],
      "ratio",
      INPUTS$floor,
      filter = "median screen_R2 at rho=1 over rho=0.4; stable events only",
      n = infl$n1[i],
      note = "R2 rises LESS than null sd does, which is why detection falls at rho=1"
    )
  }
  pn(
    "power.rho1_detection_rate",
    anch[inject_rho == 1]$det_pct / 100,
    "fraction",
    INPUTS$floor,
    filter = "inject_rho = 1 -- DEGENERATE, excluded from power; documents the artifact",
    n = anch[inject_rho == 1]$n
  )
}
```

```{r floor-power, eval = have[["floor"]], fig.width=6.5, fig.height=3.0}
.ORIGIN_CUR <- ORIGIN$floor
# `blowup` is JOINED by the scorer from the REAL screen's status (floor_events.tsv), never
# recomputed from the injected R2. Recomputing lets an event change stratum when a strong injection
# stabilises a fit that blows up on real PSI, so the per-rho rows stop being comparable -- measured,
# one stratum ran n = 184/177/181/176/183/184/180/228 where it must be constant.
fv <- fl[inject_rho <= RHO_MAX_VALID]
fv[,
  blow_lab := factor(
    ifelse(blowup, "blown up (|R2|>1 on real PSI)", "numerically stable"),
    levels = c("numerically stable", "blown up (|R2|>1 on real PSI)")
  )
]

# Stratum sizes MUST be constant across rho -- the check that catches a recomputed `blowup`.
szc <- fv[, .N, by = .(`Event Type`, blow_lab, inject_rho)][,
  .(n_distinct = uniqueN(N)),
  by = .(`Event Type`, blow_lab)
]
print(szc)
if (any(szc$n_distinct > 1L)) {
  warning(
    "stratum size varies across rho -- `blowup` looks recomputed, not joined"
  )
}

pw <- fv[,
  .(
    n = .N,
    med_R2 = median(screen_R2),
    det = sum(detected),
    rate = mean(detected)
  ),
  by = .(`Event Type`, blow_lab, inject_rho)
][order(`Event Type`, blow_lab, inject_rho)]
print(pw)

# THE HEADLINE: blown-up events have NO USABLE POWER at any effect size the design can represent,
# and they are ~50% of the real screen. Pooling them with the stable half averages that away.
#
# Say "no usable power", NOT "0% at every effect size": SE is a literal 0 of 150 in all 8 arms, but
# RI has exactly ONE detection over the whole grid (ID 64811 at rho=0.1, 1 of 146 = 0.68%), and it
# is variance collapse rather than recovered signal -- screen_R2 0.0118 against a null_R2_sd of
# 0.00168, so z = 6.9 off a trivial R2. It is also the SAME event that produces the
# min_detected_R2 = 0.0118 that 09f-floor-score.R prints for RI/Low, which is exactly why that
# statistic must never be quoted as a floor. Printed and annotated, not rounded away.
blow_det <- fv[blowup == TRUE & detected == TRUE]
if (nrow(blow_det)) {
  print(blow_det[, .(ID, `Event Type`, inject_rho, screen_R2, null_R2_sd, z_inj, p_pooled_z)])
}
for (etv in unique(pw$`Event Type`)) {
  for (bl in levels(pw$blow_lab)) {
    d <- pw[`Event Type` == etv & blow_lab == bl]
    if (!nrow(d)) {
      next
    }
    top <- d[which.max(rate)]
    k <- sprintf(
      "power.peak.%s.%s",
      tolower(etv),
      if (grepl("^blown", bl)) "blownup" else "stable"
    )
    pn(
      paste0(k, ".rate"),
      top$rate,
      "fraction",
      INPUTS$floor,
      filter = sprintf(
        "max detection over rho <= %.1f; %s space; %s; NOT pooled across strata",
        RHO_MAX_VALID,
        inj_fs,
        bl
      ),
      n = top$n
    )
    pn(
      paste0(k, ".at_median_r2"),
      top$med_R2,
      "r2",
      INPUTS$floor,
      filter = "median ACHIEVED out-of-fold partial R2 in the arm where power peaks",
      n = top$n,
      note = "quote achieved R2, never rho -- the rho->R2 map is nonlinear and event-dependent"
    )
  }
}

# THE SWEEP'S OWN PARAMETERS. Every power number above is conditional on these, and Nature's
# statistics section wants the exact n -- which for section 6 is the sweep size, not the screen's.
# They lived only in prose and comments until 2026-09-01, which meant the manuscript could quote a
# power figure with no machine-readable record of what it was measured over. All DERIVED from the
# artifact, so a re-run with a different grid updates them instead of contradicting them.
pn(
  "power.sweep_n_events",
  uniqueN(fl$ID),
  "count",
  INPUTS$floor,
  filter = sprintf("stratified sample screened under injection; %s space", inj_fs),
  note = "8 strata: Event Type x Variability x real-PSI blow-up status"
)
pn(
  "power.sweep_n_rho",
  uniqueN(fl$inject_rho),
  "count",
  INPUTS$floor,
  filter = sprintf(
    "rho grid points: %s",
    paste(sort(unique(fl$inject_rho)), collapse = ", ")
  )
)
pn(
  "power.rho_max_valid",
  RHO_MAX_VALID,
  "fraction",
  "by construction",
  filter = "largest rho whose response still carries noise; rho = 1 is noiseless and DEGENERATE",
  note = "rho is a VARIANCE fraction: sqrt(rho)*signal + sqrt(1-rho)*noise"
)
if ("R_used" %in% names(fl)) {
  pn(
    "power.sweep_rotations",
    stats::median(fl$R_used, na.rm = TRUE),
    "count",
    INPUTS$floor,
    filter = "matched control rotations per injected event; 200 for BOTH event types",
    note = "RI's production 818 buys exceedance depth the floor does not use -- it scores on p_pooled_z"
  )
}

# 50% power is never reached at any effect size the design can represent. State it as the bound.
pn(
  "power.max_overall",
  max(pw$rate),
  "fraction",
  INPUTS$floor,
  filter = sprintf(
    "best stratum x rho cell, rho <= %.1f, %s space",
    RHO_MAX_VALID,
    inj_fs
  ),
  note = "never reaches 50%: the 0-hit screen is a BOUND ON WHAT WAS TESTABLE, not a null finding"
)

p <- ggplot(pw, aes(med_R2, rate, colour = `Event Type`)) +
  geom_hline(yintercept = 0.5, linetype = 3, linewidth = 0.3, colour = PAL[7]) +
  geom_line(linewidth = 0.4) +
  geom_point(size = 0.9) +
  facet_wrap(~blow_lab) +
  scale_colour_paper(name = NULL) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  labs(
    x = "median achieved out-of-fold partial R² (injected)",
    y = "detected",
    title = "The screen never reaches 50% power, and none that is usable for blown-up events",
    caption = sprintf(
      "%s space; rho <= %.1f (rho=1 is degenerate); detection at q_pooled_z < 0.1",
      inj_fs,
      RHO_MAX_VALID
    )
  )
print(p)
save_fig(
  p,
  "fig5_power_curve",
  6.5,
  3.0,
  shows = "detection rate vs achieved injected R2, faceted by real-PSI numeric stability; rho <= 0.4",
  source = INPUTS$floor
)

# NOT WRITTEN, DELIBERATELY: min(screen_R2 | detected). 09f-floor-score.R prints it under an
# explicit "for reference only ... NOT a floor -- an outlier statistic ... Do not quote it" banner
# (it said "THE FLOOR" until 47aa538), and it must not be quoted -- detection is decided
# on z, so an event whose null variance has collapsed is detected at a trivial R2 while the median
# event at ten times that R2 is not. It is an outlier statistic over a heterogeneous set, and there
# is no floor here to report: the honest deliverable is the power curve above.
```

```{r floor-missing, eval = !have[["floor"]], results='asis'}
cat(
  "> **Not available.** `processed_data/event_models/",
  tf,
  "/floor/floor_results.csv.gz` is the declared input; run the `floor_*` rules.\n",
  sep = ""
)
```

## 6b. The floor-free pooled-z candidates (leg C)

**Origin:** `screen_results.csv.gz` cross-referenced against `floor_results.csv.gz`. No refits — the
adjudication is a reshape of two aggregates this file already reads.

The primary exceedance test gives **0 hits**. The floor-free sensitivity statistic `p_pooled_z` does
not: since `stat_version` 5 it flags a handful of events, and because the two tests disagree the
report cannot simply assert "0 hits by both". This section decides which of those flags survive, and
states the one comparison that looks decisive but is not.

```{r candidates, eval = have[["screen"]] && have[["floor"]]}
.ORIGIN_CUR <- ORIGIN$candidates
cand_ids <- unique(scr[is.finite(q_pooled_z) & q_pooled_z < qthr, ID])
cand <- scr[ID %in% cand_ids & is.finite(q_pooled_z) & q_pooled_z < qthr]

cat(sprintf("%d candidate events, %d candidate (event, space) rows\n\n",
  length(cand_ids), nrow(cand)))
print(cand[order(q_pooled_z), .(ID, `Event Type`, feature_set, n_samples,
  screen_R2 = round(screen_R2, 4), z = round(z_screen, 1),
  q_pz = signif(q_pooled_z, 3), null_sd = signif(null_R2_sd, 3),
  r2_confounds = round(r2_confounds, 2), r2_x_alone = round(r2_x_alone, 3))])

# The artifact/candidate boundary falls exactly on FEATURE SPACE, which is why all three are shown.
# The `local` candidates collapse to R2 -0.5 .. -510 in long/short -- the documented out-of-support
# extrapolation failure -- while the two `long`-only ones never qualify in `local` at all.
cat("\n--- all three spaces per candidate event: the split is by feature space ---\n")
print(dcast(
  scr[ID %in% cand_ids, .(ID, `Event Type`, feature_set,
    R2 = round(screen_R2, 3), q_pz = signif(q_pooled_z, 3))],
  ID + `Event Type` ~ feature_set, value.var = c("R2", "q_pz")
))
```

```{r candidates-adjudicate, eval = have[["screen"]] && have[["floor"]]}
.ORIGIN_CUR <- ORIGIN$candidates
# THE TEST: is a candidate's null_R2_sd anomalously tight -- i.e. is its z large because the null
# collapsed rather than because R2 is large? Reference is the injection sweep, whose events carry
# signal KNOWN to be present, so "what a healthy null looks like" is measured rather than assumed.
#
# CONDITION ON EVENT SIZE, NOT ON ACHIEVED R2. Size drives null_R2_sd independently of signal, so it
# is the fair matching variable. Conditioning on R2 instead INVERTS the answer and is invalid --
# see the trap documented in the next chunk.
fvc <- fl[inject_rho <= RHO_MAX_VALID & is.finite(screen_R2) & is.finite(null_R2_sd)]
adj <- rbindlist(lapply(seq_len(nrow(cand)), function(i) {
  nn <- cand$n_samples[i]
  peers <- fvc[abs(n_samples - nn) <= 0.25 * nn & screen_R2 > 0]
  data.table(
    ID = cand$ID[i], feature_set = cand$feature_set[i], n = nn,
    R2 = round(cand$screen_R2[i], 4), null_sd = signif(cand$null_R2_sd[i], 3),
    n_peers = nrow(peers),
    null_sd_pctile = if (nrow(peers) >= 20L) {
      round(100 * mean(peers$null_R2_sd < cand$null_R2_sd[i], na.rm = TRUE), 1)
    } else NA_real_
  )
}))
adj[, verdict := fifelse(is.na(null_sd_pctile), "unassessed",
  fifelse(null_sd_pctile < 1, "REFUTED - variance collapse",
    fifelse(null_sd_pctile < 20, "borderline - low null sd", "survives")))]
print(adj[order(-R2)])

for (i in seq_len(nrow(adj))) {
  pn(sprintf("candidates.null_sd_pctile.%d", adj$ID[i]), adj$null_sd_pctile[i], "percentile",
    INPUTS$floor,
    filter = sprintf("%s space; vs injected events within 25%% of n = %d, positive R2 only",
      adj$feature_set[i], adj$n[i]),
    n = adj$n_peers[i])
}
pn("candidates.n_flagged", length(cand_ids), "count", INPUTS$screen,
  filter = sprintf("distinct events with q_pooled_z < %g in any feature space", qthr))
pn("candidates.n_survives", sum(adj$verdict == "survives"), "count", INPUTS$floor,
  filter = "null_R2_sd at or above the 20th percentile of size-matched injected events")
pn("candidates.n_refuted", sum(grepl("^REFUTED", adj$verdict)), "count", INPUTS$floor,
  filter = "null_R2_sd below the 1st percentile of size-matched injected events",
  note = "both are `long`-only, with r2_confounds -117 and -7613: confound-projection instability")

# How often does the screen recover signal of the candidates' magnitude when it is KNOWN to be there?
# This is the number that says whether 4 survivors out of 32,318 SE events is a large hidden
# population or a handful. Restricted to numerically stable events, since blown-up ones have no
# usable power at any effect size (section 6).
rec <- fvc[blowup %in% FALSE & screen_R2 > 0.2]
if (nrow(rec) >= 20L) {
  pn("candidates.recovery_rate_at_comparable_r2", mean(rec$detected), "fraction", INPUTS$floor,
    filter = sprintf("injected, stable, achieved R2 > 0.2, %s space, rho <= %g",
      inj_fs, RHO_MAX_VALID),
    n = nrow(rec),
    note = "the screen recovers real signal of this magnitude ~1 time in 5")
  cat(sprintf(
    "\nrecovery of KNOWN signal at R2 > 0.2 (stable): %d of %d = %.1f%%\n",
    sum(rec$detected), nrow(rec), 100 * mean(rec$detected)))
}
```

```{r candidates-trap, eval = have[["screen"]] && have[["floor"]]}
.ORIGIN_CUR <- ORIGIN$candidates
# THE TRAP, printed rather than merely warned about, because it INVERTS the verdict and it is the
# obvious thing to try. Matching candidates to injected events on achieved R2 instead of on n makes
# the strongest candidate look like a ~560x outlier and "refutes" it.
#
# It is invalid: among REAL events, high R2 is usually PRODUCED BY the numerical instability that
# also inflates the null, so median null_R2_sd climbs monotonically with R2. Conditioning on R2
# therefore selects unstable events, and any well-behaved event is anomalous against them. Cause and
# effect are entangled; size is not.
trap <- scr[feature_set == "local" & `Event Type` == "SE" &
  is.finite(screen_R2) & is.finite(null_R2_sd)]
trap[, r2_bin := cut(screen_R2, c(-Inf, 0, 0.02, 0.05, 0.1, 0.2, 0.35, Inf))]
cat("REAL SE/local events: null_R2_sd rises with R2, so R2-matching selects unstable peers\n\n")
print(trap[, .(n = .N, med_null_sd = signif(median(null_R2_sd), 3)), by = r2_bin][order(r2_bin)])
cat("\nMatch on n, not on R2. See the chunk comment above.\n")
```

## 7. Ablation with sequence-only as the reference (leg A)

**Origin:** `splicing_ml --feature-groups` ablation runs, seqnames-grouped CV. The **all-features** arm is not an ablation output — it is the same seqnames config section 4 reads, reused as the reference, and is attributed to that analysis in the manifest.

```{r ablation, fig.width=5.5, fig.height=2.6}
.ORIGIN_CUR <- ORIGIN$ablation
acfg <- CJ(
  group = c("sequence", "histone+dnam"),
  et = c("RI", "SE"),
  unique = TRUE
)
acfg[,
  payload := vapply(
    seq_len(.N),
    function(i) {
      find_payload(
        file.path("splicing_ml/output_ablation", group[i]),
        et[i],
        "seqnames"
      )
    },
    character(1)
  )
]
# The all-features run is the THIRD group, not just an arithmetic term: the user asked for it on
# the figure, and it belongs there -- "sequence gets you 0.950 of 0.957" is only legible with all
# three bars visible. It comes from the same seqnames configs section 4 already read.
abl <- rbindlist(
  lapply(seq_len(nrow(acfg)), function(i) {
    a <- fold_metric_of(acfg$payload[i])
    if (is.null(a)) {
      return(NULL)
    }
    a[, `:=`(group = acfg$group[i], et = acfg$et[i])][]
  }),
  fill = TRUE
)
abl <- abl[!grepl("Baseline", model)]
# `full` = every feature group, seqnames CV. Same outer splits, so it stacks per fold.
full <- grid[split == "seqnames", .(et, model, outer_fold, value, payload)][,
  group := "all features"
]
abl <- rbind(abl, full, fill = TRUE)
abl[,
  group := factor(group, levels = c("histone+dnam", "sequence", "all features"))
]
print(summarise_folds(abl, c("group", "et", "model"))[order(group, et, model)])

# The prior baseline is 0.500 BY CONSTRUCTION -- a sanity floor, not the scientific reference.
# Sequence-only is the honest reference, and it turns "0.957" into "+0.007".
pn(
  "ablation.baseline_prior",
  0.5,
  "auroc",
  "by construction",
  filter = "DummyClassifier(strategy='prior'); identical in every fold"
)
for (g in levels(abl$group)) {
  for (etv in unique(abl$et)) {
    d <- abl[group == g & et == etv]
    if (!nrow(d)) {
      next
    }
    # The all-features arm is NOT an ablation run: it is the same seqnames config section 4
    # already read, reused here as the reference. Attribute it to that analysis, or the manifest
    # claims an ablation output that was never produced.
    pn_folds(
      sprintf("ablation.%s.%s.auroc", gsub("[^a-z]+", "_", g), tolower(etv)),
      d,
      d$payload[1],
      sprintf("%s; seqnames CV; both variability tiers", g),
      origin = if (g == "all features") ORIGIN$ml else ORIGIN$ablation
    )
  }
}

# Epigenetic gain PAIRED WITHIN OUTER FOLD, not as a difference of two means. The two runs share
# the same deterministic split (same seed, same group_col), so fold k is the same held-out
# chromosomes in both -- which is what makes the pairing valid. State that assumption.
gain <- merge(
  abl[group == "all features", .(et, model, outer_fold, full = value)],
  abl[group == "sequence", .(et, model, outer_fold, seq_only = value)],
  by = c("et", "model", "outer_fold")
)
gain[, epi_gain := full - seq_only]
gsum <- gain[,
  .(
    folds = .N,
    mean = mean(epi_gain),
    sd = stats::sd(epi_gain),
    min = min(epi_gain),
    max = max(epi_gain)
  ),
  by = .(et, model)
]
print(gsum[order(et, model)])
for (i in seq_len(nrow(gsum))) {
  pn(
    sprintf("ablation.epi_gain.%s.%s.mean", tolower(gsum$et[i]), gsum$model[i]),
    gsum$mean[i],
    "auroc",
    "derived",
    filter = "all features minus sequence-only; paired within outer fold",
    n = gsum$folds[i]
  )
  pn(
    sprintf("ablation.epi_gain.%s.%s.sd", tolower(gsum$et[i]), gsum$model[i]),
    gsum$sd[i],
    "auroc",
    "derived",
    filter = "sd of the paired within-fold gain",
    n = gsum$folds[i]
  )
}

if (nrow(abl)) {
  p <- ggplot(abl, aes(value, paste(et, model), colour = group)) +
    stat_summary(
      fun = mean,
      geom = "crossbar",
      width = 0.55,
      linewidth = 0.25,
      position = position_dodge(width = 0.75)
    ) +
    geom_point(
      position = position_dodge(width = 0.75),
      size = 0.7,
      alpha = 0.85
    ) +
    geom_vline(xintercept = 0.5, linetype = 2, linewidth = 0.3) +
    scale_colour_paper(name = NULL) +
    coord_cartesian(xlim = c(0.5, 1)) +
    labs(
      x = "test AUROC (one point per outer fold)",
      y = NULL,
      title = "Sequence carries it; marks add little"
    )
  print(p)
  save_fig(
    p,
    "fig6_ablation",
    5.5,
    2.8,
    shows = "per-outer-fold test AUROC by feature group (histone+dnam / sequence / all features)",
    origin = paste(ORIGIN$ablation, "+ all-features arm from", ORIGIN$ml),
    source = c(acfg$payload, unique(full$payload))
  )
}
```

## 7b. The argument end to end (legs B + C)

**Origin:** synthesis. Every figure below is read from `paper_numbers`, not recomputed.

**Placed here, not beside the figure in 4b, because it cites sections 5, 6 and 7** — the screen, the
power curve and the ablation. `pn()` keys exist only once their section has run, so stating this
argument next to its figure would cite keys that do not exist yet. It failed exactly that way on the
first attempt (`pnv()` errored on `ablation.baseline_prior`), which is what `pnv()` is for.

```{r preprint-correction-numbers, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
.ORIGIN_CUR <- ORIGIN$ml
# Every figure quoted in the prose below is read from paper_numbers, not retyped. `pnv()` errors on a
# missing key rather than rendering a blank, so a renamed key breaks the build instead of the claim.
.PNV <- rbindlist(.PN, fill = TRUE)
pnv <- function(k, digits = 3) {
  v <- .PNV[key == k, value]
  if (!length(v)) stop("prose cites a key that was never emitted: ", k)
  formatC(v[1], format = "f", digits = digits)
}
cat(sprintf("
**1. Under the preprint's own design, the models are near-perfect.** Held-out cell types, seen events:
xgb reaches AUROC **%s** (RI) and **%s** (SE). That reproduces the preprint's description.

**2. But the capacity gap gives it away.** Paired within fold, xgb exceeds linear by **%s ± %s** (RI)
and **%s ± %s** (SE). A linear model is bounded by one global hyperplane and cannot carve out per-event
islands; a boosted ensemble can. A gap that large, that consistently, is the signature of event-identity
memorisation rather than of signal only the tree can reach.

**3. Hold the events out and it collapses.** Under `seqnames` CV the gap is **%s ± %s** for RI —
*negative, and smaller than its own fold-to-fold spread, so not distinguishable from zero* — and
**%s ± %s** for SE. The near-perfect performance does not survive; xgb drops to **%s** (RI) and **%s**
(SE). **Generalisation to unseen events is therefore the better-founded result, the opposite of the
preprint's reading.**

**4. And what the epigenome adds, measured against sequence rather than against chance, is small.**
Sequence-only is the honest reference; the `DummyClassifier` prior is **%s** by construction and is a
sanity floor, not a baseline. Adding every epigenetic feature to sequence-only buys **%s ± %s** AUROC
for RI and **%s ± %s** for SE, paired within fold.

**5. The endpoint.** Asking the locus-specific question directly — can an event's PSI be predicted from
its own epigenetic state beyond what a matched but unrelated locus predicts — gives **%s hits** over
34,146 events at FDR 0.1 (minimum q **%s**), in a screen whose power never exceeds **%s%%** (section 6).
",
pnv("cv.ontology.ri.auroc.xgb.mean"), pnv("cv.ontology.se.auroc.xgb.mean"),
pnv("cv.ontology.ri.xgb_minus_linear.mean"), pnv("cv.ontology.ri.xgb_minus_linear.sd"),
pnv("cv.ontology.se.xgb_minus_linear.mean"), pnv("cv.ontology.se.xgb_minus_linear.sd"),
pnv("cv.seqnames.ri.xgb_minus_linear.mean"), pnv("cv.seqnames.ri.xgb_minus_linear.sd"),
pnv("cv.seqnames.se.xgb_minus_linear.mean"), pnv("cv.seqnames.se.xgb_minus_linear.sd"),
pnv("cv.seqnames.ri.auroc.xgb.mean"), pnv("cv.seqnames.se.auroc.xgb.mean"),
pnv("ablation.baseline_prior", 1),
pnv("ablation.epi_gain.ri.xgb.mean"), pnv("ablation.epi_gain.ri.xgb.sd"),
pnv("ablation.epi_gain.se.xgb.mean"), pnv("ablation.epi_gain.se.xgb.sd"),
pnv("tier1.hits", 0), pnv("tier1.min_q"),
formatC(100 * .PNV[key == "power.max_overall", value][1], format = "f", digits = 1)))
```

**What this must NOT say.** The exploited axis is **not** identified as cell-type identity, and the
manuscript must not name it that. The data show a sample-level component shared across loci; its cause
is open. Antagonistic marks correlate positively on the sample axis (H3K27ac vs H3K27me3, r ≈ 0.65)
while the spatial window axis is negative as biology predicts — at least as consistent with a global
signal-scaling component (ChIP efficiency, observed-vs-imputed provenance, depth, batch) as with
biology. Naming it would repeat the preprint's error in the opposite direction.

**Handle as an explicit self-correction.** The preprint is public and states the split itself; a
reviewer can find this. The correspondence is one of *design*, not a byte-identical replication — the
preprint used random forest at PSI ≤ ⅓ / ≥ ⅔ with a B/T-lymphocyte holdout, this uses the ablation
battery's models and thresholds on the same cohort.


## 8. Supplementary: permuted-groups control

**Origin:** permuted-groups negative control, paired per event. Pipeline stage **09p** (`09p-permute-groups.R`, rule `permuted_groups_control`), promoted out of `.claude/scratch/` on 2026-09-01. It sits beside `floor_sample` rather than at 03b because it samples its events *from* `screen_results.csv.gz` and reads the per-event feature tables.

```{r permuted, eval = have[["permuted"]]}
.ORIGIN_CUR <- ORIGIN$permuted
pg <- fread(INPUTS$permuted)
# dR2b = bounded-R2 gain from destroying group identity. `long`'s apparent gain is extrapolation
# relief, not group exploitation -- report it faceted by out-of-support severity (`oos`), never
# pooled, and never as a cell-type-identity claim (a permuted fold exposes batch, lab and protocol
# alongside ontology).
sm <- pg[,
  .(n = .N, median_dR2b = median(dR2b)),
  by = .(feature_set, event_type)
]
print(sm[order(feature_set, event_type)])
print(pg[,
  .(n = .N, median_dR2b = median(dR2b)),
  by = .(feature_set, oos)
][order(feature_set, oos)])
for (i in seq_len(nrow(sm))) {
  pn(
    sprintf(
      "permuted.median_dR2b.%s.%s",
      sm$feature_set[i],
      tolower(sm$event_type[i])
    ),
    sm$median_dR2b[i],
    "r2_bounded",
    INPUTS$permuted,
    filter = "paired per event, rotations off",
    n = sm$n[i]
  )
}
# Pooled over event types. Emitted ONLY because other notes quote a pooled figure; the faceted
# keys above are the ones to cite, because RI and SE differ materially (local 0.0090 vs 0.0014).
pl <- pg[, .(n = .N, median_dR2b = median(dR2b)), by = feature_set]
for (i in seq_len(nrow(pl))) {
  pn(
    sprintf("permuted.median_dR2b.%s.pooled", pl$feature_set[i]),
    pl$median_dR2b[i],
    "r2_bounded",
    INPUTS$permuted,
    filter = "pooled over event types - prefer the faceted keys",
    n = pl$n[i]
  )
}
# DERIVED, never hardcoded. This read "4 strata (2 spaces x 2 event types) of 50 rows each",
# which was true only at the n = 100 this happened to run at -- and `N_EVENTS` is now a config knob
# (config["permuted_groups_n_events"]), so any change would leave the stated predicate lying while
# the value silently updated. That is exactly the drift this file exists to kill.
.pg_strata <- pg[, .N, by = .(feature_set, event_type)]
pn(
  "permuted.n_events",
  uniqueN(pg$ID),
  "count",
  INPUTS$permuted,
  filter = sprintf(
    "distinct events; %d strata (%d feature spaces x %d event types), %s rows each",
    nrow(.pg_strata),
    uniqueN(pg$feature_set),
    uniqueN(pg$event_type),
    paste(sort(unique(.pg_strata$N)), collapse = "/")
  ),
  n = nrow(pg)
)
```

## 8b. The protocol confound (RI)

**Origin:** `06c-paired-protocol.R` (rule `paired_protocol`). Absent from the preprint entirely, and it
has to be stated: **total-RNA-Seq retains unspliced pre-mRNA, which inflates retained-intron PSI — and
RI is precisely where the epigenetic contribution is largest.** That coincidence is the reviewer's
first question, so the manuscript should raise it rather than be asked.

**Estimated WITHIN sample.** 36 epirrs carry both an mRNA-Seq and a total-RNA-Seq uuid, so their PSI
difference isolates protocol with cell type, donor, batch and every other per-sample factor held
fixed. A mixed-pool regression across all samples cannot do that: there, protocol is partly confounded
with whatever else differs between the two groups of samples.

```{r protocol, eval = have[["protocol"]]}
.ORIGIN_CUR <- ORIGIN$protocol
prot <- fread(INPUTS$protocol)
print(prot[, .(`Event Type`, n_obs, n_epirrs,
  mixed = signif(mixed_model_estimate, 3), mixed_se = signif(mixed_model_se, 2),
  mixed_p = signif(mixed_model_p_value, 3),
  paired = signif(paired_ttest_estimate, 3),
  ci = sprintf("[%.4f, %.4f]", paired_ttest_ci_low, paired_ttest_ci_high))])

for (i in seq_len(nrow(prot))) {
  et <- tolower(prot$`Event Type`[i])
  pn(sprintf("protocol.dpsi_paired.%s", et), prot$paired_ttest_estimate[i], "dpsi",
    INPUTS$protocol,
    filter = "total-RNA-Seq minus mRNA-Seq, paired t-test within epirr",
    n = prot$n_epirrs[i])
  pn(sprintf("protocol.dpsi_paired_ci_low.%s", et), prot$paired_ttest_ci_low[i], "dpsi",
    INPUTS$protocol, filter = "95% CI lower bound", n = prot$n_epirrs[i])
  pn(sprintf("protocol.dpsi_paired_ci_high.%s", et), prot$paired_ttest_ci_high[i], "dpsi",
    INPUTS$protocol, filter = "95% CI upper bound", n = prot$n_epirrs[i])
  # The mixed model uses every paired OBSERVATION rather than one mean per epirr, so its n is the
  # event x sample count; emitted beside the t-test because they answer the same question at
  # different levels and the manuscript should not silently pick one.
  pn(sprintf("protocol.dpsi_mixed.%s", et), prot$mixed_model_estimate[i], "dpsi",
    INPUTS$protocol,
    filter = "lme4 mixed model, random intercept per epirr; paired subset only",
    n = prot$n_obs[i])
  pn(sprintf("protocol.dpsi_mixed_p.%s", et), prot$mixed_model_p_value[i], "p",
    INPUTS$protocol, filter = "mixed model, two-sided", n = prot$n_obs[i])
}

if (have[["protocol_conc"]]) {
  conc <- fread(INPUTS$protocol_conc)
  print(conc[, .(`Event Type`, n, pearson = signif(pearson_r, 3),
    spearman = signif(spearman_r, 3))])
  for (i in seq_len(nrow(conc))) {
    pn(sprintf("protocol.concordance_r.%s", tolower(conc$`Event Type`[i])),
      conc$pearson_r[i], "correlation", INPUTS$protocol_conc,
      filter = "PSI of the same event in the same epirr, total-RNA-Seq vs mRNA-Seq",
      n = conc$n[i])
  }
}

# THE POOLED ESTIMATE, measured rather than quoted. Emitted so the manuscript compares two numbers
# from the same data instead of the paired estimate against a figure remembered from the preprint.
if (have[["protocol_pool"]]) {
  pool <- fread(INPUTS$protocol_pool)
  print(pool[, .(`Event Type`, n_events,
    median = signif(beta_pooled_median, 3),
    IQR = sprintf("[%.4f, %.4f]", beta_pooled_q25, beta_pooled_q75))])
  for (i in seq_len(nrow(pool))) {
    et <- tolower(pool$`Event Type`[i])
    pn(sprintf("protocol.dpsi_pooled_median.%s", et), pool$beta_pooled_median[i], "dpsi",
      INPUTS$protocol_pool,
      filter = "median per-event (total-RNA-Seq - mRNA-Seq) mean PSI over ALL samples, >= 5 per protocol",
      n = pool$n_events[i],
      note = "UPPER BOUND: absorbs cell-type composition, since protocol is not randomised across ontologies")
    # The ratio is the number the manuscript actually needs: it says how much of the naive protocol
    # coefficient survives once donor and cell type are held fixed.
    .pair <- prot[`Event Type` == pool$`Event Type`[i], paired_ttest_estimate]
    if (length(.pair) && is.finite(.pair[1]) && .pair[1] != 0) {
      pn(sprintf("protocol.pooled_over_paired.%s", et),
        pool$beta_pooled_median[i] / .pair[1], "ratio", "derived",
        filter = "pooled median / paired estimate; how much of the naive coefficient is confounding",
        n = pool$n_events[i])
    }
  }
}

# The cohort is not balanced, and the imbalance runs the wrong way for RI.
if (exists("co") && nrow(co)) {
  pn("protocol.total_rna_frac", co$total_rna[1] / co$uuids[1], "fraction",
    INPUTS$cohort, filter = "modelled uuids that are total-RNA-Seq",
    origin = ORIGIN$cohort)
}
```

```{r protocol-missing, eval = !have[["protocol"]], results='asis'}
cat("> **Not available**: run `snakemake processed_data/paired_protocol_dPSI_summary_",
  tf, ".csv.gz`.\n", sep = "")
```

**How to state it.** The within-sample effect is **small and positive for RI and near-zero for SE**,
which is the direction unspliced pre-mRNA predicts. Quote the paired estimate with its CI, and give
the mixed-model estimate beside it — they are the same question asked at different levels (one mean
per epirr vs every paired observation), and picking whichever is more convenient is the error to avoid.

**Do not quote a single "protocol effect" number as if the design were settled.** Both estimates are
computed above, from the same `aggregated_dt`, so they can be compared as measurements rather than one
measurement against a remembered figure. They answer different questions: the pooled per-event estimate
absorbs everything else that differs between the mRNA-Seq and total-RNA-Seq sample groups — cell-type
composition above all, since protocol is not randomised across ontologies — while the paired design
holds donor, cell type and batch fixed by construction. **The paired estimate is the causal quantity;
the pooled one is an upper bound that includes confounding**, and `protocol.pooled_over_paired.*` is
the fraction of the naive coefficient that does not survive controlling for sample identity.

## 9. Statistics appendix

**Origin:** experimental design, not measurement. Nature requires the replicate unit to be *defined*; it differs per analysis, so it is stated per analysis.

```{r stats-appendix}
.ORIGIN_CUR <- ORIGIN$design
# Nature requires a statistics section that DEFINES THE REPLICATES and gives exact n and exact
# non-significant P values. The replicate unit differs per analysis, so it is stated per analysis.
replicates <- data.table(
  analysis = c(
    "Atlas (leg R)",
    "Tier-1 screen",
    "Tier-1 fold structure",
    "splicing_ml CV",
    "Variance budget",
    "Permuted-groups control"
  ),
  replicate_unit = c(
    "RNA-seq experiment (UUID)",
    "splicing event",
    "ontology supergroup",
    "chromosome (seqnames) or ontology supergroup",
    "observed (event, sample) row",
    "splicing event, paired"
  ),
  test = c(
    "none (descriptive)",
    "one-sided exceedance p vs matched-locus null; BH/Storey FDR",
    "leave-one-ontology-supergroup-out",
    "grouped k-fold",
    "none (variance decomposition, LOO group means)",
    "paired difference"
  ),
  tails = c(NA, "one-sided", NA, NA, NA, "two-sided")
)

# The replicate UNIT and the TEST are design facts -- not measurable, correctly literal, and
# attributed to ORIGIN$design. The exact n is NOT a design fact, and leaving it out (or writing it by
# hand) is how a statistics section ends up claiming an n the pipeline never produced. So n is looked
# up from the keys already emitted above, and a missing one is an ERROR rather than a blank cell:
# Nature requires the exact n per analysis, and an appendix that silently omits one is worse than no
# appendix. Atlas is exempt -- leg R has no keys until the atlas summary artifact exists (section 1).
.PN_SO_FAR <- rbindlist(.PN, fill = TRUE)
.nkey <- function(k) {
  v <- .PN_SO_FAR[key == k, value]
  if (!length(v)) NA_real_ else v[1]
}
replicates[, n_key := c(
  "atlas.uuids",             # absent until the atlas summary exists
  "tier1.m_max",             # largest FDR family; set below
  NA_character_,             # fold structure: ~5 supergroups, a property of resolve_supergroup_folds
  "cohort.uuids",
  "variance.n_rows_total",   # set below
  "permuted.n_events"
)]
# two n's the sections above did not need as keys in their own right, emitted here so the appendix
# can cite them rather than restate them
if (exists("fam")) {
  pn("tier1.m_max", max(fam$m), "count", INPUTS$screen,
    filter = "largest (Event Type x feature_set) FDR family; events carrying a finite q",
    origin = ORIGIN$tier1)
}
if (exists("vb")) {
  pn("variance.n_rows_total", sum(vb[grepl("All", stratum), n_rows]), "count", INPUTS$variance,
    filter = "observed (event, sample) rows, both event types, All-variability strata",
    origin = ORIGIN$variance)
}
.PN_SO_FAR <- rbindlist(.PN, fill = TRUE)
replicates[, n := vapply(n_key, function(k) if (is.na(k)) NA_real_ else .nkey(k), numeric(1))]
print(replicates[, .(analysis, replicate_unit, n, test, tails)])

.missing <- replicates[!is.na(n_key) & is.na(n) & !grepl("^Atlas", analysis)]
if (nrow(.missing)) {
  stop("statistics appendix: no key backs the n for ",
    paste(.missing$analysis, collapse = ", "),
    " -- write the key or drop the row; do not hand-fill it")
}
```

## 10. Write `paper_numbers`

Emits `paper_numbers_{tf}.csv` (every quoted number, with `filter`, `source` **and** `origin`) and `figure_provenance_{tf}.csv` beside the PDFs, so the manuscript's images directory is self-describing.

```{r write-numbers}
# `filter` and `note` are prose and DO contain commas. That is fine and deliberate: fwrite quotes
# them and fread round-trips them losslessly (verified at 77 keys; 224 keys as of 2026-09-01).
# Mangling the text to suit `cut -d,` was considered and rejected -- it would degrade the one field
# whose job is to make a number's predicate legible. CONSUME THIS FILE WITH A REAL CSV PARSER.
PN <- rbindlist(.PN, fill = TRUE)
PN[, generated := format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")]
stopifnot(!anyDuplicated(PN$key))
# Every number must name the analysis that produced it. A missing `origin` means a pn() call
# slipped through with .ORIGIN_CUR unset, which pn() is supposed to make impossible -- so this is
# an assertion, not a warning.
stopifnot(!any(is.na(PN$origin) | !nzchar(PN$origin)))
out <- sprintf("processed_data/paper_numbers_%s.csv", tf)
fwrite(PN, out)
message(sprintf("paper_numbers: %d keys -> %s", nrow(PN), out))

# Figure manifest: same provenance contract for the figures. Written beside the PDFs so the
# manuscript's images directory is self-describing after sync_paper_figures copies it over.
FIGS <- rbindlist(.FIG, fill = TRUE)
if (nrow(FIGS)) {
  fout <- file.path(fig_dir, sprintf("figure_provenance_%s.csv", tf))
  fwrite(FIGS, fout)
  message(sprintf("figure manifest: %d figures -> %s", nrow(FIGS), fout))
  print(FIGS[, .(figure, shows)])
}

# How many numbers each analysis contributes -- the quickest read on whether an UNPROMOTED
# analysis is load-bearing for the manuscript.
print(PN[, .(keys = .N), by = origin][order(-keys)])
print(PN[, .(key, value = signif(value, 6), unit, origin)])
```

```{r session}
sessionInfo()
```
